### 2023-08-05 index.qmd
eda template doc

---
title: "Exploratory data analysis"
author: 
   - name: Marcus Beck, <mbeck@tbep.org>
institute: "Tampa Bay Estuary Program"
date: "`r Sys.Date()`"
format:
  html:
    theme: cosmo
    css: styles.css
    code-tools: 
      source: https://github.com/tbep-tech/temp-manu/index.qmd
code-fold: true
execute:
  echo: true
  warning: false
filters:
  - lightbox

lightbox: auto
---

```{r}
library(tidyverse)
library(lubridate)
library(EnvStats)
library(tbeptools)
```

### 2023-08-07 index.qmd
initial temp trend plots for all stations by year, month, strata in epc data

@@
       source: https://github.com/tbep-tech/temp-manu/index.qmd
-code-fold: true
 execute:
-  echo: true
+  echo: false
   warning: false
@@
 library(lubridate)
-library(EnvStats)
-library(tbeptools)
+library(here)
 ```
+
+```{r}
+knitr::include_graphics('figs/sktall.png')
+```
+
+# By month
+
+::: panel-tabset
+```{r}
+#| results: asis
+fls <- list.files(here('figs'), '_', full.names = F)
+mos <- fls %>% 
+  basename %>% 
+  gsub('^skt_|\\.png', '', .) %>% 
+  as.numeric() %>% 
+  sort() %>% 
+  month(label = T) %>% 
+  as.character()
+for(mo in seq_along(mos)){
+  cat('## ', mos[mo], '\n\n')
+
+  chnk <- paste0('![](figs/', fls[mo], ')\n')
+  cat(chnk)
+
+  cat('\n\n')
+
+}
+```
+:::

### 2023-08-07 R/figs.R

@@ -2,31 +2,46 @@ library(tidyverse)
 library(lubridate)
 library(EnvStats)
 library(tbeptools)
+library(ggmap)
+library(sf)
+library(patchwork)
+library(here)
 
-
-# kendall -------------------------------------------------------------------------------------
+# kendall all months --------------------------------------------------------------------------
 
 sktres <- epcdata %>% 
   select(bay_segment, station = epchc_station, SampleTime, lon = Longitude, lat = Latitude, yr, 
          mo, matches('Temp')) %>% 
   pivot_longer(matches('Temp'), names_to = 'var', values_to = 'val') %>% 
-  filter(yr < 2023) %>% 
   nest(.by = c('bay_segment', 'var', 'station', 'lon', 'lat')) %>% 
+  crossing(
+    ., 
+    tibble(
+      per = c('all', 'rec', 'cur'),
+      yrs = c('1974-2023', '2000-2016', '2016-2023')
+    )
+  ) %>% 
   mutate(
-    skt = purrr::pmap(list(station, var, data), function(station, var, data){
+    skt = purrr::pmap(list(station, var, yrs, data), function(station, var, yrs, data){
       
-      cat(station, var, '\n')
-
+      cat(station, var, yrs, '\n')
+      
+      yrsel <- strsplit(yrs, '-') %>% 
+        .[[1]] %>% 
+        as.numeric()
+      
+      # yr selection
       tomod <- data %>% 
         arrange(yr, mo) %>% 
-        na.omit()
+        na.omit() %>% 
+        filter(yr >= yrsel[1] & yr < yrsel[2])
       
-      # run sk test
-      ests_sk <- kendallSeasonalTrendTest(val ~ mo + yr, data = tomod)
+      # run tests
+      ests <- kendallSeasonalTrendTest(val ~ mo + yr, data = tomod)
       
       out <- tibble(
-        pval = ests_sk$p.value[2], 
-        slos = ests_sk$estimate[2],
+        pval = ests$p.value[2], 
+        slos = ests$estimate[2],
         n = nrow(tomod)
       )
       
@@ -37,72 +52,171 @@ sktres <- epcdata %>%
   select(-data) %>% 
   unnest(skt)
 
-# # basemap
-# dat_ext <- unname(st_bbox(tbseg))
[…]

### 2023-08-08 index.qmd
added satellite time period

@@
     code-tools: 
-      source: https://github.com/tbep-tech/temp-manu/index.qmd
+      source: https://github.com/tbep-tech/temp-manu/blob/main/index.qmd
 execute:

### 2023-08-08 R/figs.R

@@ -17,8 +17,8 @@ sktres <- epcdata %>%
   crossing(
     ., 
     tibble(
-      per = c('all', 'rec', 'cur'),
-      yrs = c('1974-2023', '2000-2016', '2016-2023')
+      per = c('all', 'rec', 'sat', 'cur'),
+      yrs = c('1974-2023', '2000-2016', '2007-2023', '2016-2023')
     )
   ) %>% 
   mutate(
@@ -85,8 +85,8 @@ toplo <- sktres %>%
       labels = c('Top', 'Mid', 'Bottom')
     ),
     yrs = factor(yrs, 
-      levels = c('1974-2023', '2000-2016', '2016-2023'), 
-      labels = c('1974-2022', '2000-2015', '2016-2022')
+      levels = c('1974-2023', '2007-2023', '2000-2016', '2016-2023'), 
+      labels = c('1974-2022', '2007-2022', '2000-2015', '2016-2022')
     )
   ) 
 
@@ -104,7 +104,7 @@ p <- ggmap(bsmap1_transparent) +
     caption = 'Outlines indicate p < 0.05'
   )
 
-png(here('figs/sktall.png'), height = 10, width = 9, family = 'serif', units = 'in', res = 300)
+png(here('figs/sktall.png'), height = 10, width = 12, family = 'serif', units = 'in', res = 300)
 p
 dev.off()
 
@@ -118,8 +118,8 @@ sktres <- epcdata %>%
   crossing(
     ., 
     tibble(
-      per = c('all', 'rec', 'cur'),
-      yrs = c('1974-2023', '2000-2016', '2016-2023')
+      per = c('all', 'rec', 'sat', 'cur'),
+      yrs = c('1974-2023', '2000-2016', '2007-2023', '2016-2023')
     )
   ) %>% 
   mutate(
@@ -192,6 +192,10 @@ sktres %>%
                  levels = c("Temp_Water_Top_degC", "Temp_Water_Mid_degC", "Temp_Water_Bottom_degC"), 
                  labels = c('Top', 'Mid', 'Bottom')
     ),
+    yrs = factor(yrs, 
+                 levels = c('1974-2023', '2007-2023', '2000-2016', '2016-2023'), 
+                 labels = c('1974-2022', '2007-2022', '2000-2015', '2016-2022')
+    ),
     mo = sprintf('%02d', mo), 
     sz = scales::rescale(abs(slos), c(3, 8))
   ) %>% 
@@ -215,7 +219,7 @@ sktres %>%
[…]

### 2023-08-09 index.qmd
added monthly summary plots by bay seg and text descriptions

@@
 
+# Complete record
+
+This plot shows the significant increasing or decreasing temperature trends for all long-term monitoring stations (EPCHC) in Tampa Bay, by period of record and sampling location in the water column.
+The periods of record are 1974-2022 (whole record), 2007-2022 ([NOAA record](https://coastwatch.noaa.gov/cw_html/cwViewer.html)), 2000-2015 (seagrass recovery), and 2016-2022 (seagrass decline).
+The latter two periods of record likely have low statistical power with limited sample size.
+Seasonal Mann-Kendall tests are used.
+
 ```{r}
@@
 
+These plots show significant increasing or decreasing temperature trends by month for long-term monitoring stations in Tampa Bay.
+The plots are organized by period of record and sampling location in the water column.
+Mann-Kendal tests are used.
+
 ::: panel-tabset
@@
 #| results: asis
-fls <- list.files(here('figs'), '_', full.names = F)
+fls <- list.files(here('figs'), 'skt_', full.names = F)
 mos <- fls %>% 
@@
 :::
+
+# Month summary by bay segment
+
+These plots show the percent of stations by month and bay segment that had significant increasing temperature trends for the specified period of record.
+Tabs are organized by top, mid, and bottom water column temperature measurements.
+The summaries are based on results for the monthly trend tests in the previous section.
+Decreasing trends are not shown.
+
+::: panel-tabset
+```{r}
+#| results: asis
+fls <- list.files(here('figs'), 'segsum_', full.names = F) %>% 
+  rev
+locs <- fls %>% 
+  basename %>% 
+  gsub('^segsum_|\\.png', '', .)
+for(loc in seq_along(locs)){
+  cat('## ', locs[loc], '\n\n')
+
+  chnk <- paste0('![](figs/', fls[loc], ')\n')
+  cat(chnk)
+
+  cat('\n\n')
+
+}
+```
+:::

### 2023-08-09 R/figs.R

@@ -224,3 +224,54 @@ sktres %>%
     dev.off()
     
   })
+
+# segment summary of monthly kendall by temp loc ----------------------------------------------
+
+sktres %>% 
+  summarise(
+    nsig = sum(pval < 0.05 & slos > 0),
+    cnt = n(),
+    nsigper = round(100 * nsig / cnt, 0),
+    .by = c('bay_segment', 'mo', 'yrs', 'var')
+  ) %>% 
+  mutate(
+    yrs = factor(yrs, 
+                 levels = c('1974-2023', '2007-2023', '2000-2016', '2016-2023'), 
+                 labels = c('1974-2022: whole record', '2007-2022: NOAA record', '2000-2015: seagrass recovery', '2016-2022: seagrass decline')
+    ),
+    mo = month(mo, label = T), 
+    bay_segment = factor(bay_segment, levels = c('LTB', 'MTB', 'HB', 'OTB')), 
+    nsigper = ifelse(nsigper == 0, '', nsigper), 
+    var = factor(var, 
+                 levels = c('Temp_Water_Top_degC', 'Temp_Water_Mid_degC', 'Temp_Water_Bottom_degC'),
+                 labels = c('Top', 'Mid', 'Bottom')
+    )
+  ) %>% 
+  group_nest(var) %>% 
+  deframe %>% 
+  iwalk(function(x, var){
+    
+    p <- ggplot(x, aes(x = mo, y = bay_segment, fill = nsig)) + 
+      geom_tile(color = 'darkgrey') +
+      geom_text(aes(label = nsigper)) +
+      scale_x_discrete(expand = c(0, 0)) + 
+      scale_y_discrete(expand = c(0, 0)) + 
+      scale_fill_distiller(palette = 'Reds', direction = 1) + 
+      facet_wrap(~ yrs, ncol = 1, scales = 'free_x') + 
+      theme(
+        strip.background = element_blank(), 
+        strip.text = element_text(hjust = 0, size = 12), 
+        legend.position = 'none'
+      ) + 
+      labs(
+        x = NULL, 
+        y = 'Bay segment'
+      )
+    
+    fl <- paste0('figs/segsum_', var, '.png')
+    png(here(fl), height = 8, width = 6, family = 'serif', units = 'in', res = 300)
+    print(p)
+    dev.off()
+    
+  })
+

### 2023-08-10 index.qmd
adding threshold plots

@@
 ```{r}
-knitr::include_graphics('figs/sktall.png')
+knitr::include_graphics(here('figs/sktall.png'))
 ```
@@
 :::
+
+# Threshold exceedances
+
+Temperature exceedences over time were estimated for major bay segments by using daily predictions from a Generalized Additive Model applied to each station.
+An example of a model is shown in the first plot.
+The date of the first and last instance each year for each of three temperature thresholds (20, 25, and 30 degrees C) were estimated.
+The total number of days in each year when the daily temperature was above the same thresholds were also estimated.
+The station level results were aggregated within each bay segment to produce the plots below.
+
+```{r}
+knitr::include_graphics(here('figs/gamex.png'))
+knitr::include_graphics(here('figs/threxceed.png'))
+knitr::include_graphics(here('figs/thrcount.png'))
+```

### 2023-08-10 R/figs.R

@@ -270,22 +270,137 @@ sktres %>%
       )
     
     fl <- paste0('figs/segsum_', var, '.png')
-    png(here(fl), height = 8, width = 6, family = 'serif', units = 'in', res = 300)
+    png(here(fl), height = 7, width = 6, family = 'serif', units = 'in', res = 300)
     print(p)
     dev.off()
     
   })
 
-# temp thresholds -----------------------------------------------------------------------------
 
+# station gam -------------------------------------------------------------------------------
+
+fl <- 'OTB_66_temptop'
+load(file = here(paste0('data/', fl, '.RData')))
+
+toplo <- get(fl)
+
+p <- show_prdseries(toplo, ylab = 'Surface tempeature (\u00B0 C') + 
+  labs(subtitle ='Station 66, OTB')
+
+png(here('figs/gamex.png'), height = 3, width = 6, family = 'serif', units = 'in', res = 300)
+print(p)
+dev.off()
+
+# temp thresholds doy -------------------------------------------------------------------------
+
+# model file with daily temp predictions for each station, bottom/surface temp
 load(file = here('data/moddat.RData'))
 
-modsum <- moddat %>% 
-  select(-mod) %>% 
+modsum <- moddat %>%
   mutate(
     cnts = purrr::map(prd, function(x){
       
-      tibble(thr = c(10:35)) %>%
+      tibble(thr = c(20, 25, 30)) %>%
+        group_nest(thr) %>%
+        mutate(data = list(x)) %>%
+        mutate(
+          data = purrr::pmap(list(data, thr), function(data, thr){
+            
+            data %>%
+              filter(yr < 2023) %>%
+              summarise(
+                cnt = sum(value > thr),
+                frt = date[which(value >= thr)[1]],
+                lst = date[rev(which(value >= thr))[1]],
+                .by = 'yr'
+              )
+            
+          })
+        ) %>% 
+        unnest('data')
+      
+    })
+  )
+
+toplo1 <- modsum %>% 
+  filter(param %in% 'temptop') %>% 
+  select(-AIC, -GCV, -R2, -prd) %>% 
[…]

### 2023-08-11 manu-draft.qmd
manu template added plus conceptual diagram

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "Title"
# bibliography: refs.bib
# csl: limnology-and-oceanography.csl
author: 
  - Marcus W Beck:
      email: mbeck@tbep.org
      institute: [TBEP]
      correspondence: true

institute:
  - TBEP: Tampa Bay Estuary Program, St. Petersburg, Florida 33701 USA 16802 USA

filters: 
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua

execute: 
  echo: false
  warning: false
  ft.keepnext: false
  ft.align: left
---

```{r, results = 'hide'}
library(knitr)
library(here)
```

## Abstract {.unnumbered}

## Introduction

## Methods

The conceptual stressor diagram @fig-concept.

## Results

## Discussion

## Acknowledgments {.unnumbered}

<!-- {{< pagebreak >}} -->

## Figures {.unnumbered}

![Conceptual stressor diagram](figs/concept.png){#fig-concept}
{{< pagebreak >}}

## Tables {.unnumbered}

<!-- {{< pagebreak >}} -->

## References {.unnumbered}

### 2023-08-12 index.qmd
add fim por to epc kendall

@@
 This plot shows the significant increasing or decreasing temperature trends for all long-term monitoring stations (EPCHC) in Tampa Bay, by period of record and sampling location in the water column.
-The periods of record are 1974-2022 (whole record), 2007-2022 ([NOAA record](https://coastwatch.noaa.gov/cw_html/cwViewer.html)), 2000-2015 (seagrass recovery), and 2016-2022 (seagrass decline).
+The periods of record are 1974-2022 (whole record), 2007-2022 ([NOAA record](https://coastwatch.noaa.gov/cw_html/cwViewer.html)), 1996-2020 (FIM record), 2000-2015 (seagrass recovery), and 2016-2022 (seagrass decline).
 The latter two periods of record likely have low statistical power with limited sample size.
@@
 #| results: asis
+#| out-width: 80%
 fls <- list.files(here('figs'), 'segsum_', full.names = F) %>% 

### 2023-08-12 R/dat_proc.R

@@ -105,8 +105,8 @@ phydat <- phyraw %>%
   select(date, Reference)
 
 hyddat <- hydraw %>% 
-  filter(Beg_end %in% 'B') %>% 
-  filter(Depth == max(Depth), .by = 'Reference') %>% 
+  filter(Beg_end %in% 'B') %>% # each location has a beginning and end log, take one 
+  filter(Depth == max(Depth), .by = 'Reference') %>% # some have depth profile, take bottom
   select(Reference, depth = Depth, temp = Temperature, sal = Salinity)
 
 fimdat <- phydat %>% 

### 2023-08-12 R/figs.R

@@ -12,14 +12,14 @@ library(wqtrends)
 
 sktres <- epcdata %>% 
   select(bay_segment, station = epchc_station, SampleTime, lon = Longitude, lat = Latitude, yr, 
-         mo, matches('Temp')) %>% 
+         mo, matches('Temp_Water_Top|Temp_Water_Bottom')) %>% 
   pivot_longer(matches('Temp'), names_to = 'var', values_to = 'val') %>% 
   nest(.by = c('bay_segment', 'var', 'station', 'lon', 'lat')) %>% 
   crossing(
     ., 
     tibble(
-      per = c('all', 'rec', 'sat', 'cur'),
-      yrs = c('1974-2023', '2000-2016', '2007-2023', '2016-2023')
+      per = c('all', 'fim', 'sat', 'rec', 'cur'),
+      yrs = c('1974-2023', '1996-2021', '2007-2023', '2000-2016', '2016-2023')
     )
   ) %>% 
   mutate(
@@ -82,12 +82,12 @@ toplo <- sktres %>%
     coefsgn = sign(slos),
     coefsgn = factor(coefsgn, levels = c('1', '-1'), labels = c('inc', 'dec')), 
     var = factor(var, 
-      levels = c("Temp_Water_Top_degC", "Temp_Water_Mid_degC", "Temp_Water_Bottom_degC"), 
-      labels = c('Top', 'Mid', 'Bottom')
+      levels = c("Temp_Water_Top_degC", "Temp_Water_Bottom_degC"), 
+      labels = c('Top', 'Bottom')
     ),
     yrs = factor(yrs, 
-      levels = c('1974-2023', '2007-2023', '2000-2016', '2016-2023'), 
-      labels = c('1974-2022', '2007-2022', '2000-2015', '2016-2022')
+      levels = c('1974-2023', '2007-2023', '1996-2021', '2000-2016', '2016-2023'), 
+      labels = c('1974-2022', '2007-2022', '1996-2020', '2000-2015', '2016-2022')
     )
   ) 
 
@@ -96,7 +96,7 @@ p <- ggmap(bsmap1_transparent) +
   facet_grid(var ~ yrs) +
   # scale_fill_gradient2(leglab, low = 'green', mid = 'grey',  high = 'tomato1', midpoint = 0) +
   scale_fill_gradientn(leglab, limits = c(-0.25, 0.25), colors = c('green', 'grey', 'tomato1')) +
[…]

### 2023-08-12 index.qmd
reproc kendall tests for both temp and sal, only whole period of record

@@
 
-This plot shows the significant increasing or decreasing temperature trends for all long-term monitoring stations (EPCHC) in Tampa Bay, by period of record and sampling location in the water column.
-The periods of record are 1974-2022 (whole record), 2007-2022 ([NOAA record](https://coastwatch.noaa.gov/cw_html/cwViewer.html)), 1996-2020 (FIM record), 2000-2015 (seagrass recovery), and 2016-2022 (seagrass decline).
-The latter two periods of record likely have low statistical power with limited sample size.
+This plot shows the significant increasing or decreasing salinity or temperature trends for all long-term monitoring stations (EPCHC) in Tampa Bay for the entire period of record (1974-2022) and sampling location in the water column.
 Seasonal Mann-Kendall tests are used.
@@
 ```{r}
+#| out-width: 70%
 knitr::include_graphics(here('figs/sktall.png'))
@@
 
-These plots show significant increasing or decreasing temperature trends by month for long-term monitoring stations in Tampa Bay.
-The plots are organized by period of record and sampling location in the water column.
-Mann-Kendal tests are used.
+These plots show significant increasing or decreasing temperature or salinity trends by month for long-term monitoring stations in Tampa Bay.
+The plots are organized by parameter and sampling location in the water column.
+Mann-Kendall tests are used.
 
@@
 
-  chnk <- paste0('![](figs/', fls[mo], ')\n')
+  chnk <- paste0('![](figs/', fls[mo], '){width=70%}\n')
   cat(chnk)
@@
 
-These plots show the percent of stations by month and bay segment that had significant increasing temperature trends for the specified period of record.
-Tabs are organized by top, mid, and bottom water column temperature measurements.
+These plots show the percent of stations by month and bay segment that had significant decreasing salinity trends or increasing temperature trends.
+Tabs are organized by top and bottom water column measurements.
 The summaries are based on results for the monthly trend tests in the previous section.
-Decreasing trends are not shown.
+Increasing trends for salinity and decreasing trends for temperature are not shown.
 

### 2023-08-12 R/figs.R

@@ -8,34 +8,25 @@ library(patchwork)
 library(here)
 library(wqtrends)
 
+yrsel <- c(1974, 2022)
+
 # kendall all months --------------------------------------------------------------------------
 
 sktres <- epcdata %>% 
   select(bay_segment, station = epchc_station, SampleTime, lon = Longitude, lat = Latitude, yr, 
-         mo, matches('Temp_Water_Top|Temp_Water_Bottom')) %>% 
-  pivot_longer(matches('Temp'), names_to = 'var', values_to = 'val') %>% 
+         mo, matches('Top|Bottom')) %>% 
+  filter(yr >= yrsel[1] & yr <= yrsel[2]) %>% 
+  pivot_longer(matches('Top|Bottom'), names_to = 'var', values_to = 'val') %>% 
   nest(.by = c('bay_segment', 'var', 'station', 'lon', 'lat')) %>% 
-  crossing(
-    ., 
-    tibble(
-      per = c('all', 'fim', 'sat', 'rec', 'cur'),
-      yrs = c('1974-2023', '1996-2021', '2007-2023', '2000-2016', '2016-2023')
-    )
-  ) %>% 
   mutate(
-    skt = purrr::pmap(list(station, var, yrs, data), function(station, var, yrs, data){
+    skt = purrr::pmap(list(station, var, data), function(station, var, data){
       
-      cat(station, var, yrs, '\n')
-      
-      yrsel <- strsplit(yrs, '-') %>% 
-        .[[1]] %>% 
-        as.numeric()
+      cat(station, var, '\n')
       
       # yr selection
       tomod <- data %>% 
         arrange(yr, mo) %>% 
-        na.omit() %>% 
-        filter(yr >= yrsel[1] & yr < yrsel[2])
+        na.omit()
       
       # run tests
       ests <- kendallSeasonalTrendTest(val ~ mo + yr, data = tomod)
@@ -64,7 +55,11 @@ bsmap1_transparent <- matrix(adjustcolor(bsmap1,
                              nrow = nrow(bsmap1))
 attributes(bsmap1_transparent) <- mapatt
 
-leglab <- "\u00B0C / year"
+leglab <- expression(paste(yr^{-1}))
+colrng <- range(sktres$slos) %>% 
+  abs %>% 
+  max
+colrng <- c(-1 * colrng, colrng)
 
[…]

### 2023-08-14 R/dat_proc.R
rename moddat and modsum as tempprd and tempsum

@@ -6,7 +6,7 @@ library(haven)
 library(here)
 library(sf)
 
-# GAMs for each station, save file ------------------------------------------------------------
+# temp GAMs for each station, save file -------------------------------------------------------
 
 # save model files for each station, separate for bottom/surface temp
 epcdata %>% 
@@ -53,12 +53,12 @@ epcdata %>%
     
   })
 
-# get daily predictions from GAM files --------------------------------------------------------
+# get daily temp predictions from GAM files ---------------------------------------------------
 
 fls <- list.files(here('data'), pattern = 'temp', full.names = T)
 obs <- gsub('\\.RData$', '', basename(fls))
 
-moddat <- tibble(obs = obs) %>% 
+tempprd <- tibble(obs = obs) %>% 
   mutate(fit = NA, prd = NA)
 for(i in seq_along(fls)){
   
@@ -69,19 +69,19 @@ for(i in seq_along(fls)){
   
   load(file = fl)
   toadd <- get(ob)
-  moddat[moddat$obs == ob, 'fit'][[1]] <- list(anlz_fit(toadd))
-  moddat[moddat$obs == ob, 'prd'][[1]] <- list(anlz_prdday(toadd))
+  tempprd[tempprd$obs == ob, 'fit'][[1]] <- list(anlz_fit(toadd))
+  tempprd[tempprd$obs == ob, 'prd'][[1]] <- list(anlz_prdday(toadd))
   
   rm(toadd)
   rm(list = ob)
   
 }
 
-moddat <- moddat %>% 
+tempprd <- tempprd %>% 
   separate(obs, c('bay_segment', 'station', 'param')) %>% 
   unnest('fit')
 
-save(moddat, file = here('data/moddat.RData'))
+save(tempprd, file = here('data/tempprd.RData'))
 
 # fim data ------------------------------------------------------------------------------------
 

### 2023-08-14 manu-draft.qmd
initial workflow for fim data import
adding documentation about thresholds

@@
 title: "Title"
-# bibliography: refs.bib
+bibliography: refs.bib
 # csl: limnology-and-oceanography.csl
@@
 
+Salinity tolerance: Ruppia \> Halodule\> Thalassia \> Syringodium
+
+Halodule:
+
+-   Temp: 20-30 (range, @lewis1985)
+
+-   Salinity: grows well in most salinity ranges [@lirman2003], like Syringodium but more tolerant [@lewis1985], 10-35 (range) or \< 10 (S. Scolaro)
+
+Thalassia:
+
+-   Temp: 20-30 (range, @lewis1985, @zieman75), 30-31 (upper, from SS)
+
+-   Salinity: 25 (lower, @lewis1985); 30, lowest at 5 (lower, @lirman2003), 33-38 (range, @phillips1960); 24-35 (range, @zieman1975 citing @phillips1960)
+
+Syringodium:
+
+-   Temp: 20-30 (range, @lewis1985), 29 (upper from SS)
+
+-   Salinity: 20 (lower, @lewis1985), 25 (lower, @lirman),
+
+@lewis1985 review of seagrass in Tampa Bay.
+@lirman2003 conducted exposure experiments to evaluate seagrass growth in response to a range of salinity conditions.
+@mcmillan1967 discusses growth of halodule, syringodium, thalassia, and ruppia in response to salinity increases (up to 75 psu), no info on lower limit.
+Cites @phillips1960 for a salinity range of Thalassia in Florida of 33 - 38 psu.
+@zieman1975 discusses seasonal variation of thalassia relative to temp and salinity
+
+-   Get FIM, Pinellas, Manatee data
+
 The conceptual stressor diagram @fig-concept.
@@
 
-![Conceptual stressor diagram](figs/concept.png){#fig-concept}
-{{< pagebreak >}}
+![Conceptual stressor diagram](figs/concept.png){#fig-concept} {{< pagebreak >}}
 

### 2023-08-14 manu-draft.qmd
rework threshold temp plots for diff ranges and fix polygon order

@@
 
+Focus on two time periods: full record and recent decline
+
+Focus on two physical parameters: temperature and salinity
+
+Temp background for TB: NOAA coastwatch trends
+
 ## Methods
@@
 
--   Salinity: 25 (lower, @lewis1985); 30, lowest at 5 (lower, @lirman2003), 33-38 (range, @phillips1960); 24-35 (range, @zieman1975 citing @phillips1960)
+-   Salinity: 25 (lower, @lewis1985); 30, lowest at 5 (lower, @lirman2003), 33-38 (range, @phillips1960); 24-35 (range, @zieman1975 citing @phillips1960); avg 28
 
@@
 
--   Salinity: 20 (lower, @lewis1985), 25 (lower, @lirman),
+-   Salinity: 20 (lower, @lewis1985), 25 (lower, @lirman2003); avg 22.5
 

### 2023-08-14 R/figs.R

@@ -284,7 +284,7 @@ tempsum <- tempprd %>%
   mutate(
     cnts = purrr::map(prd, function(x){
       
-      tibble(thr = c(20, 25, 30)) %>%
+      tibble(thr = c(29, 30, 31)) %>%
         group_nest(thr) %>%
         mutate(data = list(x)) %>%
         mutate(
@@ -345,17 +345,29 @@ toplo2 <- toplo1 %>%
   ) %>% 
   select(-data) %>% 
   unnest('reg') %>% 
-  arrange(param, thr, doy, yr)
+  mutate(
+    ind = case_when(
+      dts == 'frt' & yr == max(yr) ~ 1, 
+      dts == 'frt' & yr == min(yr) ~ 2,
+      dts == 'lst' & yr == min(yr) ~ 3,
+      dts == 'lst' & yr == max(yr) ~ 4
+    ),
+    .by = c(bay_segment, param, thr)
+  ) %>% 
+  arrange(bay_segment, param, thr, ind)
 
 p <- ggplot(toplo1, aes(x = yr, y = doy)) +
-  geom_polygon(data = toplo2, aes(x = yr, y = doy, fill = thr), alpha = 0.5) + 
+  geom_polygon(data = toplo2, aes(x = yr, y = doy, fill = thr), alpha = 0.5, rule = 'winding') + 
   geom_point(size = 0.5, aes(color = thr)) +
   geom_smooth(aes(group = dts, color = thr), method = 'lm', se = F, lwd = 1) +
   facet_grid(bay_segment ~ thr) +
   scale_colour_manual(values = c('coral', 'red2', 'darkred')) +
   scale_fill_manual(values = c('coral', 'red2', 'darkred')) +
   scale_x_continuous(expand = c(0,0)) +
-  scale_y_date(date_breaks = '3 months', date_labels = paste('%b', '1st')) +
+  scale_y_date(date_breaks = '2 months', date_labels = paste('%b', '1st')) +
+  coord_cartesian(
+    ylim = as.Date(c('2020-05-01', '2020-11-01'))
+  ) +
   theme_bw() + 
   theme(
     panel.grid.minor = element_blank(), 
@@ -384,7 +396,7 @@ tempsum <- tempprd %>%
   mutate(
     cnts = purrr::map(prd, function(x){
       
-      tibble(thr = c(20, 25, 30)) %>%
+      tibble(thr = c(29, 30, 31)) %>%
         group_nest(thr) %>%
         mutate(data = list(x)) %>%
         mutate(

### 2023-08-14 index.qmd
rename moddat and modsum as tempprd and tempsum
adding documentation about thresholds
working on salinity eval, added to index

@@
 
+::: panel-tabset
+
+## Temperature
+
 Temperature exceedences over time were estimated for major bay segments by using daily predictions from a Generalized Additive Model applied to each station.
 An example of a model is shown in the first plot.
-The date of the first and last instance each year for each of three temperature thresholds (20, 25, and 30 degrees C) were estimated.
+The date of the first and last instance each year for each of three temperature thresholds (29, 30, and 31 degrees C) were estimated.
 The total number of days in each year when the daily temperature was above the same thresholds were also estimated.
@@
 ```{r}
-knitr::include_graphics(here('figs/gamex.png'))
-knitr::include_graphics(here('figs/threxceed.png'))
-knitr::include_graphics(here('figs/thrcount.png'))
+knitr::include_graphics(here('figs/tempgamex.png'))
+knitr::include_graphics(here('figs/tempexceed.png'))
+knitr::include_graphics(here('figs/tempcount.png'))
 ```
+
+## Salinity
+
+Salinity exceedences below thresholds over time were estimated for major bay segments by using daily predictions from a Generalized Additive Model applied to each station.
+An example of a model is shown in the first plot.
+The date of the first and last instance each year for each of three salinity thresholds (20, 25, and 30 psu) were estimated.
+The total number of days in each year when the daily salinity was below the same thresholds were also estimated.
+The station level results were aggregated within each bay segment to produce the plots below.
+
+```{r}
+knitr::include_graphics(here('figs/saligamex.png'))
+knitr::include_graphics(here('figs/saliexceed.png'))
+knitr::include_graphics(here('figs/salicount.png'))
+```
+
+:::

### 2023-08-14 manu-draft.qmd
working on salinity eval, added to index

@@
 
--   Salinity: 25 (lower, @lewis1985); 30, lowest at 5 (lower, @lirman2003), 33-38 (range, @phillips1960); 24-35 (range, @zieman1975 citing @phillips1960); avg 28
+-   Salinity: 25 (lower, @lewis1985); 30, lowest at 5 (lower, @lirman2003), 33-38 (range, @phillips1960); 24-35 (range, @zieman1975 citing @phillips1960); 15 (SS)
 
@@
 
--   Salinity: 20 (lower, @lewis1985), 25 (lower, @lirman2003); avg 22.5
+-   Salinity: 20 (lower, @lewis1985), 25 (lower, @lirman2003); 20
 

### 2023-08-14 R/figs.R

@@ -1,3 +1,5 @@
+# setup ---------------------------------------------------------------------------------------
+
 library(tidyverse)
 library(lubridate)
 library(EnvStats)
@@ -9,6 +11,10 @@ library(here)
 library(wqtrends)
 
 yrsel <- c(1974, 2022)
+tempthr <- c(29, 30, 31)
+tempcol <- c('coral', 'red2', 'darkred')
+salithr <- c(20, 25, 30)
+salicol <- c('navyblue', 'dodgerblue2', 'slategray3')
 
 # kendall all months --------------------------------------------------------------------------
 
@@ -260,18 +266,17 @@ ktres %>%
     
   })
 
+# temp station gam --------------------------------------------------------------------------
 
-# station gam -------------------------------------------------------------------------------
-
-fl <- 'OTB_66_temptop'
+fl <- 'OTB_66_tempbot'
 load(file = here(paste0('data/', fl, '.RData')))
 
 toplo <- get(fl)
 
-p <- show_prdseries(toplo, ylab = 'Surface tempeature (\u00B0 C') + 
+p <- show_prdseries(toplo, ylab = 'Bottom temperature (\u00B0 C)') + 
   labs(subtitle ='Station 66, OTB')
 
-png(here('figs/gamex.png'), height = 3, width = 6, family = 'serif', units = 'in', res = 300)
+png(here('figs/tempgamex.png'), height = 3, width = 6, family = 'serif', units = 'in', res = 300)
 print(p)
 dev.off()
 
@@ -284,7 +289,7 @@ tempsum <- tempprd %>%
   mutate(
     cnts = purrr::map(prd, function(x){
       
-      tibble(thr = c(29, 30, 31)) %>%
+      tibble(thr = tempthr) %>%
         group_nest(thr) %>%
         mutate(data = list(x)) %>%
         mutate(
@@ -307,7 +312,7 @@ tempsum <- tempprd %>%
   )
 
 toplo1 <- tempsum %>% 
-  filter(param %in% 'temptop') %>% 
+  filter(param %in% 'tempbot') %>% 
   select(-AIC, -GCV, -R2, -prd) %>% 
   unnest('cnts') %>% 
   summarise(
@@ -357,12 +362,12 @@ toplo2 <- toplo1 %>%
   arrange(bay_segment, param, thr, ind)
 
 p <- ggplot(toplo1, aes(x = yr, y = doy)) +
[…]

### 2023-08-17 manu-draft.qmd
seagrass change fig added

@@
 
+Seagrass changes over time in @fig-seagrasschg.
+
+The conceptual stressor diagram @fig-concept.
+
 Salinity tolerance: Ruppia \> Halodule\> Thalassia \> Syringodium
@@
 
--   Temp: 20-30 (range, @lewis1985, @zieman75), 30-31 (upper, from SS)
+-   Temp: 20-30 (range, @lewis1985, @zieman1975), 30-31 (upper, from SS)
 
@@
 
-The conceptual stressor diagram @fig-concept.
-
 ## Results
@@
 ## Discussion
+
+Other areas showing seagrass loss - Florida Bay is different, less water flowing out of everglad and compounding SLR has elevated salinity and likely stress in other direction.
+Also, Biscayne Bay and IRL is a lot like OTB, poor flushing for example.
 
@@
 
-![Conceptual stressor diagram](figs/concept.png){#fig-concept} {{< pagebreak >}}
+![Seagrass changes over time in Tampa Bay for (a) areal coverage and (b) frequency occurrence of major species.
+Changes are shown for major bay segments.
+Note the different time scale between (a) and  (b).
+Coverage maps in (a) began in 1988 and seagrass transect monitoring in (b) began in 1998.
+Red lines in (a) show approximate capacity of seagrass coverage based on the baywide target of 40,000 acres.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
+](figs/seagrasschg.png){#fig-seagrasschg}
+
+{{< pagebreak >}}
+
+![Conceptual stressor diagram](figs/concept.png){#fig-concept}
+
+{{< pagebreak >}}
 

### 2023-08-17 R/figs.R

@@ -16,6 +16,98 @@ tempcol <- c('coral', 'red2', 'darkred')
 salithr <- c(20, 25, 30)
 salicol <- c('navyblue', 'dodgerblue2', 'slategray3')
 
+# seagrass loss -------------------------------------------------------------------------------
+
+load(file = url('https://github.com/tbep-tech/tbep-os-presentations/raw/master/data/sgsegest.RData'))
+
+sgsegest <- sgsegest %>% 
+  mutate(
+    segment = factor(segment, 
+                     levels = c("Old Tampa Bay", "Hillsborough Bay", "Middle Tampa Bay", "Lower Tampa Bay", 
+                                "Boca Ciega Bay", "Terra Ceia Bay", "Manatee River"),
+                     labels = c('OTB', 'HB', 'MTB', 'LTB', 'BCB', 'TCB', 'MR'))
+  )
+
+# segment coverage targets in 1k acres
+segtrgs <- tibble(
+  segment = factor(c(levels(sgsegest$segment), 'Total')), 
+  trgs = c(11.1, 1.751, 9.4, 7.4, 8.8, 1.1, 0.449, 40)
+)  
+
+savval <- c('Halodule', 'Syringodium', 'Thalassia')
+savleg <- expression(italic('H. wrightii'), italic('S. filiforme'), italic('T. testudinum'))
+savcol <- c('#ED90A4', '#CCA65A', '#7EBA68')
+names(savcol) <- savval
+
+toplo1 <- sgsegest %>%
+  filter(!segment %in% c('BCB', 'TCB', 'MR')) %>%
+  mutate(acres = acres / 1000) %>%
+  mutate(segment = forcats::fct_drop(segment))
+
+subsegtrgs <- segtrgs %>%
+  filter(segment %in% levels(toplo1$segment))
+
+p1 <- ggplot(toplo1, aes(x = factor(year), y = acres)) +
+  geom_bar(fill = '#00806E', stat = 'identity', colour = 'black', width = 0.6) +
+  geom_hline(data = subsegtrgs, aes(yintercept = trgs, color = 'Target')) +
+  scale_color_manual(values = 'red') +
+  facet_wrap(~segment, ncol = 4) +
+  theme_bw() + 
+  theme(panel.grid.minor =element_blank(),
+        panel.grid.major.x =element_blank(),
+        # plot.background = element_rect(fill = NA, color = NA),
+        axis.text.y = element_text(colour = 'black'),
[…]

### 2023-08-17 index.qmd
added time series plot of raw salinity and temp trends

@@
 ---
-title: "Exploratory data analysis"
+title: "Hot and fresh: Temperature, salinity, and seagrass change in a large Gulf coast estuary"
 author: 

### 2023-08-17 manu-draft.qmd
added time series plot of raw salinity and temp trends

@@
 
+Observed trends aggregated by year and bay segment across stations and months in @fig-saltempraw.
+
 The conceptual stressor diagram @fig-concept.
@@
 
+For @fig-saltempraw, salinity shows much higher inter-annual variability, but lower intra-year variation among stations.
+Trends are similar for bottom vs top, temperature shows stronger trend than salinity.
+
 ## Discussion
@@
 
+![Long-term salinity and temperature trends for major bay segments from 1975 to 2022.
+Points are colored by sampling location in the water column.
+Points show the average and 95% confidence interval across all stations and sampling months for each year in each bay segment.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
+](figs/saltempraw.png){#fig-saltempraw}
+
+{{< pagebreak >}}
+
 ![Conceptual stressor diagram](figs/concept.png){#fig-concept}

### 2023-08-17 R/figs.R

@@ -108,6 +108,61 @@ png(here('figs/seagrasschg.png'), height = 6, width = 8, family = 'serif', units
 print(p)
 dev.off()
 
+# raw temp and salinity changes ---------------------------------------------------------------
+
+toplo <- epcdata %>% 
+  select(bay_segment, epchc_station, SampleTime, yr, matches('Top|Bottom')) %>% 
+  filter(yr >= yrsel[1] & yr <= yrsel[2]) %>% 
+  pivot_longer(names_to = 'var', values_to = 'val', matches('Top|Bottom')) %>% 
+  mutate(
+    var = factor(var, 
+                 levels = c(c("Sal_Top_ppth", "Sal_Bottom_ppth", "Temp_Water_Top_degC", "Temp_Water_Bottom_degC"
+                 )), 
+                 labels = c("Sal_Top", "Sal_Bottom", "Temp_Top", "Temp_Bottom")
+                 ),
+    bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB'))
+  ) %>% 
+  separate(var, c('var', 'loc')) %>% 
+  mutate(
+    var = factor(var, levels = c('Sal', 'Temp'), labels = c('Salinity (psu)', 'Temperature (\u00B0 C)')),
+    loc = factor(loc, levels = c('Top', 'Bottom'))
+  ) %>% 
+  filter(!is.na(val)) %>% 
+  summarise(
+    avev = mean(val, na.rm = T),
+    lov = t.test(val, na.rm = T)$conf.int[1],
+    hiv = t.test(val, na.rm = T)$conf.int[2],
+    .by = c('bay_segment', 'yr', 'var', 'loc') 
+  )
+
+wd <- 0.5
+p <- ggplot(toplo, aes(x = yr, y = avev, group = loc, color = loc)) + 
+  geom_linerange(aes(ymin = lov, ymax = hiv), position = position_dodge2(width = wd), show.legend = F, alpha = 0.7) + 
+  geom_point(position = position_dodge2(width = wd)) +
+  # scale_x_continuous(breaks = seq(min(toplo$yr), max(toplo$yr), by = 3)) +
+  geom_smooth(method = 'lm', formula = y ~ x, se = F) +
+  facet_grid(var ~ bay_segment, scales = 'free_y', switch = 'y') +
+  scale_color_manual(values = c('steelblue1', 'steelblue4')) +
+  theme_bw(base_size = 14) + 
+  theme(
+    panel.grid.minor = element_blank(),
[…]

### 2023-08-17 index.qmd
moved all figs and R files for eda to separate folder

--- 2023-08-14 index.qmd
@@
 #| out-width: 70%
-knitr::include_graphics(here('figs/sktall.png'))
+knitr::include_graphics(here('figseda/sktall.png'))
 ```
@@
 
-  chnk <- paste0('![](figs/', fls[mo], '){width=70%}\n')
+  chnk <- paste0('![](figseda/', fls[mo], '){width=70%}\n')
   cat(chnk)
@@
 
-  chnk <- paste0('![](figs/', fls[loc], ')\n')
+  chnk <- paste0('![](figseda/', fls[loc], ')\n')
   cat(chnk)
@@
 ```{r}
-knitr::include_graphics(here('figs/tempgamex.png'))
-knitr::include_graphics(here('figs/tempexceed.png'))
-knitr::include_graphics(here('figs/tempcount.png'))
+knitr::include_graphics(here('figseda/tempgamex.png'))
+knitr::include_graphics(here('figseda/tempexceed.png'))
+knitr::include_graphics(here('figseda/tempcount.png'))
 ```
@@
 ```{r}
-knitr::include_graphics(here('figs/saligamex.png'))
-knitr::include_graphics(here('figs/saliexceed.png'))
-knitr::include_graphics(here('figs/salicount.png'))
+knitr::include_graphics(here('figseda/saligamex.png'))
+knitr::include_graphics(here('figseda/saliexceed.png'))
+knitr::include_graphics(here('figseda/salicount.png'))
 ```

### 2023-08-17 manu-draft.qmd
moved all figs and R files for eda to separate folder

@@
     number-sections: true
-title: "Title"
+title: "Hot and fresh: Temperature, salinity, and seagrass change in a large Gulf coast estuary"
 bibliography: refs.bib

### 2023-08-17 R/figseda.R

@@ -1,0 +1,670 @@
+# setup ---------------------------------------------------------------------------------------
+
+library(tidyverse)
+library(lubridate)
+library(EnvStats)
+library(tbeptools)
+library(ggmap)
+library(sf)
+library(patchwork)
+library(here)
+library(wqtrends)
+library(ggspatial)
+
+yrsel <- c(1974, 2022)
+tempthr <- c(29, 30, 31)
+tempcol <- c('coral', 'red2', 'darkred')
+salithr <- c(20, 25, 30)
+salicol <- c('navyblue', 'dodgerblue2', 'slategray3')
+
+# kendall all months --------------------------------------------------------------------------
+
+sktres <- epcdata %>% 
+  select(bay_segment, station = epchc_station, SampleTime, lon = Longitude, lat = Latitude, yr, 
+         mo, matches('Top|Bottom')) %>% 
+  filter(yr >= yrsel[1] & yr <= yrsel[2]) %>% 
+  pivot_longer(matches('Top|Bottom'), names_to = 'var', values_to = 'val') %>% 
+  nest(.by = c('bay_segment', 'var', 'station', 'lon', 'lat')) %>% 
+  mutate(
+    skt = purrr::pmap(list(station, var, data), function(station, var, data){
+      
+      cat(station, var, '\n')
+      
+      # yr selection
+      tomod <- data %>% 
+        arrange(yr, mo) %>% 
+        na.omit()
+      
+      # run tests
+      ests <- kendallSeasonalTrendTest(val ~ mo + yr, data = tomod)
+      
+      out <- tibble(
+        pval = ests$p.value[2][[1]], 
+        slos = ests$estimate[2][[1]],
+        n = nrow(tomod)
+      )
+      
+      return(out)
+      
+    })
+  ) %>% 
+  select(-data) %>% 
+  unnest(skt)
+
+# basemap
+dat_ext <- unname(st_bbox(st_buffer(tbseg, dist = 2000)))
+bsmap1 <- get_stamenmap(bbox = dat_ext, maptype = 'terrain-background', zoom = 11)
+
+# change opacity of basemap
+mapatt <- attributes(bsmap1)
+bsmap1_transparent <- matrix(adjustcolor(bsmap1,
+                                         alpha.f = 0.4),
+                             nrow = nrow(bsmap1))
[…]

### 2023-08-18 index.qmd
reproc index for correct fig path

@@
 #| results: asis
-fls <- list.files(here('figs'), 'skt_', full.names = F)
+fls <- list.files(here('figseda'), 'skt_', full.names = F)
 mos <- fls %>% 
@@
 #| out-width: 80%
-fls <- list.files(here('figs'), 'segsum_', full.names = F) %>% 
+fls <- list.files(here('figseda'), 'segsum_', full.names = F) %>% 
   rev

### 2023-08-18 manu-draft.qmd
finished kendall fig

--- 2023-08-17 manu-draft.qmd
@@
     number-sections: true
-title: "Title"
+title: "Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary"
 bibliography: refs.bib
@@
 
+Kendall test results are shown in @fig-kendall.
+
 ## Discussion
@@
 Red lines in (a) show approximate capacity of seagrass coverage based on the baywide target of 40,000 acres.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
-](figs/seagrasschg.png){#fig-seagrasschg}
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/seagrasschg.png){#fig-seagrasschg}
 
@@
 Points show the average and 95% confidence interval across all stations and sampling months for each year in each bay segment.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
-](figs/saltempraw.png){#fig-saltempraw}
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/saltempraw.png){#fig-saltempraw}
 
@@
 ![Conceptual stressor diagram](figs/concept.png){#fig-concept}
+
+{{< pagebreak >}}
+
+![Trends from 1974 to 2022 for salinity and temperature measurements at long-term monitoring stations in Tampa Bay.
+Results for (a) seasonal Kendall tests by station and monitoring location (top or bottom of the waer column) are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr-1).
+Summarized seasonal trends by month are shown for (b) top and (c) bottom measurements as the percent of stations in each bay segment with significant decreasing (salinity) or increasing (temperature) trends.
+Bay segment outlines are shown in (a); OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
+](figs/kendall.png){#fig-kendall}
 

### 2023-08-18 R/figs.R

@@ -240,7 +240,7 @@ ktres <- epcdata %>%
   select(-data) %>% 
   unnest(kt)
 
-# segment summaries by month
+# plots of segment summaries by month
 ktresplo <- ktres %>% 
   mutate(
     mo = month(mo, label = T), 
@@ -257,8 +257,8 @@ ktresplo <- ktres %>%
       data %>% 
         summarise(
           nsig = case_when(
-            varsimp == 'Sal' ~ -1 * sum(pval < 0.05 & slos < 0),
-            varsimp == 'Temp' ~ sum(pval < 0.05 & slos > 0)
+            varsimp == 'Salinity' ~ -1 * sum(pval < 0.05 & slos < 0),
+            varsimp == 'Temperature' ~ sum(pval < 0.05 & slos > 0)
           ),
           cnt = n(),
           nsigper = round(100 * nsig / cnt, 0),
@@ -272,10 +272,10 @@ ktresplo <- ktres %>%
   group_nest(loc) %>% 
   mutate(
     plo = purrr::pmap(list(loc, data), function(loc, data){
- 
+
       p <- ggplot(data, aes(x = mo, y = bay_segment, fill = nsigper)) + 
         geom_tile(color = 'darkgrey') +
-        geom_text(aes(label = perlab)) +
+        geom_text(aes(label = perlab), color = 'white', fontface = 'bold') +
         scale_x_discrete(expand = c(0, 0)) + 
         scale_y_discrete(expand = c(0, 0)) + 
         scale_fill_gradientn(leglab, limits = c(-100, 100), colors = c('blue', 'grey', 'tomato1')) +
@@ -283,7 +283,8 @@ ktresplo <- ktres %>%
         theme(
           strip.background = element_blank(), 
           strip.text = element_text(hjust = 0, size = 12), 
-          legend.position = 'none'
+          legend.position = 'none',
+          axis.text = element_text(size = 10)
         ) + 
         labs(
           x = NULL, 
@@ -312,16 +313,6 @@ colrng <- range(sktres$slos) %>%
   max
 colrng <- c(-1 * colrng, colrng)
 
-pthm <- theme_bw(base_family = 'serif', base_size = 14) +
-  theme(
-    legend.position = 'right',
-    # legend.box = 'vertical',
-    strip.background = element_blank(),
-    axis.title = element_blank(),
[…]

### 2023-08-21 manu-draft.qmd
added two summary tables for mixeff mods

@@
 library(here)
+library(flextable)
+
+load(file = here('tabs/mixtab.RData'))
+load(file = here('tabs/daytab.RData'))
 ```
@@
 
-<!-- {{< pagebreak >}} -->
+Summary of mixed-effects models evaluating increases in the number of days each year from 1974 to 2022 when salinity or temperature was below or above critical thresholds, respectively.
+The slope estimates in the "Both" column indicate trends in the number of days when both salinity and temperature were below or above critical thresholds.
+** p < 0.005, * p < 0.05, - no model.
+OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
+```{r}
+mixtab
+```
+
+{{< pagebreak >}}
+
+Number of days at the beginning and end of the period of record when salinity or temperature was below or above critical thresholds, respectively.
+Values are the estimated mean number of days (plus standard error) from 1974 to 2022.
+Columns for "Both" indicate the estimated number of days when both salinity and temperature were below or above critical thresholds.
+Note that not all models are significant (see (table 1)).
+** p < 0.005, * p < 0.05, - no model.
+OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
+```{r}
+daytab
+```
+
+{{< pagebreak >}}
 

### 2023-08-21 R/dat_proc.R

@@ -269,18 +269,25 @@ mixmods <- thrdat %>%
       if(inherits(mod, 'try-error'))
         return(NA)
       
+      effs <- estimate_means(mod, 'yr')
+      
       summod <- summary(mod)$coefficients
       
       out <- tibble(
         slo = summod['yr', 'Estimate'],
-        pvl = summod['yr', 'Pr(>|t|)']
+        pvl = summod['yr', 'Pr(>|t|)'], 
+        yrstr = effs$Mean[1],
+        yrstrse = effs$SE[1], 
+        yrend = effs$Mean[nrow(effs)],
+        yrendse = effs$SE[nrow(effs)]
       )
       
       return(out)
       
     })
   ) %>% 
-  unnest('mod')
+  unnest('mod') %>% 
+  ungroup()
 
 save(mixmods, file = here('data/mixmods.RData'), compress = 'xz')
 

### 2023-08-21 R/funcs.R

@@ -1,7 +1,7 @@
 # function for formatting p-values in tables
 p_ast <- function(x){
   
-  sig_cats <- c('**', '*', 'ns')
+  sig_cats <- c('**', '*', '')
   sig_vals <- c(-Inf, 0.005, 0.05, Inf)
   
   out <- cut(x, breaks = sig_vals, labels = sig_cats, right = FALSE)

### 2023-08-21 R/tabs.R

@@ -1,0 +1,87 @@
+library(tidyverse)
+library(here)
+library(flextable)
+
+source(here('R/funcs.R'))
+
+# mixef mod summary table ---------------------------------------------------------------------
+
+load(file = here('data/mixmods.RData'))
+
+totab <- mixmods %>% 
+  select(-yrstr, -yrstrse, -yrend, -yrendse) %>% 
+  mutate(
+    pvl = p_ast(pvl), 
+    slo = as.character(round(slo, 2)), 
+    salithr = gsub('^.*_', '', salithr),
+    tempthr = gsub('^.*_', '', tempthr), 
+    thrtyp = factor(thrtyp, levels = c('salicnt', 'tempcnt', 'bothcnt'), 
+                     labels = c('Salinity', 'Temperature', 'Both')),
+    bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB'))
+  ) %>% 
+  unite('slo', slo, pvl, sep = '') %>% 
+  mutate(
+    slo = ifelse(slo == 'NANA', '-', slo)
+  ) %>%
+  arrange(bay_segment, thrtyp) %>% 
+  pivot_wider(names_from = 'thrtyp', values_from = 'slo')
+
+mixtab <- totab %>% 
+  as_grouped_data(groups = 'bay_segment') %>% 
+  flextable() %>% 
+  add_header_row(values = c('', 'Thresholds', 'Thresholds', rep('Slopes', 3))) %>% 
+  merge_at(i = 1, j = c(2:3), part = 'header') %>% 
+  merge_at(i = 1, j = c(4:6), part = 'header') %>% 
+  set_header_labels(i = 2, values = c('Bay Segment', 'Salinity', 'Temperature', 'Salinity', 'Temperature', 'Both')) %>% 
+  padding(padding = 0, part = 'all') %>% 
+  font(part = 'all', fontname = 'Times New Roman') %>% 
+  fontsize(size = 9, part = 'body')
+
+save(mixtab, file = here('tabs/mixtab.RData'))
+
+# mixef summary start and end number of days --------------------------------------------------
+
+load(file = here('data/mixmods.RData'))
+
+totab <- mixmods %>% 
+  select(-slo, -pvl) %>% 
+  mutate(
+    yrstr = round(yrstr, 0), 
+    yrstrse = paste0('(', round(yrstrse, 1), ')'),
[…]

### 2023-08-23 manu-draft.qmd
adding mixeff model figure example

@@
 
+An example of mixed effects models for two selected thresholds is shown in @fig-mixeff.
+
 ## Discussion
@@
 
+![Example of mixed effects models for the estimated number of days by year that salinity (blue) or temperature (red) were below or above thresholds of 25 psu and 30 degrees C, respectively.
+The bottom row (black) shows the number of days when both salinity and temperature were below or above the thresholds.
+The models included station as a random effect for each bay segment, with grey lines indicating individual station trends and thicker lines indicating the overall model fit.
+Slopes for significant models are shown in the bottom right of each facet.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/mixeff.png){#fig-mixeff}
+
+{{< pagebreak >}}
+
 ## Tables {.unnumbered}

### 2023-08-23 R/dat_proc.R

@@ -6,6 +6,7 @@ library(haven)
 library(here)
 library(sf)
 library(lmerTest)
+library(modelbased)
 
 source(here('R/funcs.R'))
 

### 2023-08-24 manu-draft.qmd
added nickerson ref

@@
 ## Introduction
+
+Background: SST trends in Tampa Bay w/ EPC data, WFS, and deep GOM [@nickerson2023] showed that 1975 to 2022 trend for EPC was notable, but less so from 1998 to 2022.
+Also noted the trend was most pronounced in the winter, see Fig 3.
 
@@
 Summarized seasonal trends by month are shown for (b) top and (c) bottom measurements as the percent of stations in each bay segment with significant decreasing (salinity) or increasing (temperature) trends.
-Bay segment outlines are shown in (a); OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
-](figs/kendall.png){#fig-kendall}
+Bay segment outlines are shown in (a); OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/kendall.png){#fig-kendall}
 
@@
 The slope estimates in the "Both" column indicate trends in the number of days when both salinity and temperature were below or above critical thresholds.
-** p < 0.005, * p < 0.05, - no model.
+\*\* p \< 0.005, \* p \< 0.05, - no model.
 OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
+
 ```{r}
@@
 Note that not all models are significant (see (table 1)).
-** p < 0.005, * p < 0.05, - no model.
+\*\* p \< 0.005, \* p \< 0.05, - no model.
 OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
+
 ```{r}

### 2023-08-24 manu-draft.qmd
added refs about seagrass disease

@@
 
-Other areas showing seagrass loss - Florida Bay is different, less water flowing out of everglad and compounding SLR has elevated salinity and likely stress in other direction.
+Other areas showing seagrass loss - Florida Bay is different, less water flowing out of everglades and compounding SLR has elevated salinity and likely stress in other direction.
 Also, Biscayne Bay and IRL is a lot like OTB, poor flushing for example.
+
+Viral/disease effects on seagrass change, [@vanbogaert2018] for TB and [@duffin2021] for Florida Bay.
 

### 2023-08-24 manu-draft.qmd
added ref about heat stress and light
added Hensel et al to refs

@@
 Also noted the trend was most pronounced in the winter, see Fig 3.
+
+Heat stress combined with highlight accelerates decline of *E. acoroides*, used 36C [@zhang2023]
+
+Widgeongrass colonization in Chesapeake Bay as an opportunistic, heat-tolerant species that has replaced Eelgrass, although former is sensitive to nutrient pulses.
+Paper provides an example of implications on climate change and nutrient management on foundation species and system resiliency [@hensel2023].
 

### 2023-08-25 manu-draft.qmd
add pyro refs for discussion

@@
 
-## Acknowledgments {.unnumbered}
+OTB seagrass loss and role of pyrodinium, optimal temperatures from 28-31 [@lopez2021].
+For salinity, highest growth rates at 24 psu and above, with decline in growth rates at 20 psu [@usup1994], in Florida growth has occurred in psu 10-45, bloom concentrations observed above 15 psu [@phlips2006; @lopez2021].
+
+## Acknowledgments
 

### 2023-08-29 R/dat_proc.R
added nms fig but need to add caption, working gam but need to create plot and tables

@@ -1,3 +1,5 @@
+# setup ---------------------------------------------------------------------------------------
+
 library(wqtrends)
 library(tbeptools)
 library(tidyverse)
@@ -7,6 +9,7 @@ library(here)
 library(sf)
 library(lmerTest)
 library(modelbased)
+library(mgcv)
 
 source(here('R/funcs.R'))
 
@@ -329,7 +332,7 @@ thrdat <- cmbdat %>%
   unnest('cmbcnt') %>% 
   pivot_longer(names_to = 'thrtyp', values_to = 'cnt', matches('cnt')) %>% 
   summarise(
-    cnt = sum(cnt), 
+    cnt = runfunc(cnt), 
     .by = c(bay_segment, station, salithr, tempthr, thrtyp, yr)
   )
 
@@ -340,7 +343,7 @@ save(thrdat, file = here('data/thrdat.RData'), compress = 'xz')
 # thresholds to count by year, segment specific
 thrs <- targets %>% 
   filter(bay_segment %in% c('OTB', 'HB', 'MTB', 'LTB')) %>% 
-  select(bay_segment, chla_target)
+  select(bay_segment, thr = chla_thresh)
 
 # GAM models for chla by station with daily predictions for POR
 load(file = here('data/chlaprd.RData'))
@@ -349,15 +352,15 @@ load(file = here('data/chlaprd.RData'))
 chlthrdat <- chlaprd %>% 
   left_join(thrs, by = 'bay_segment') %>% 
   mutate(
-    cnts = purrr::pmap(list(chla_target, prd), function(chla_target, prd){
-      
+    cnts = purrr::pmap(list(thr, prd), function(thr, prd){
+
       out <- prd %>% 
         filter(yr < 2023) %>% 
         mutate(
-          chlathr = chla_target
+          cnt = value >= thr
         ) %>% 
         summarise(
-          cnt = sum(value >= chlathr), 
+          cnt = runfunc(cnt), 
           .by = c('yr')
         )
       
@@ -365,7 +368,7 @@ chlthrdat <- chlaprd %>%
       
     })
   ) %>% 
-  select(bay_segment, station, chla_thr = chla_target, cnts) %>% 
+  select(bay_segment, station, chla_thr = thr, cnts) %>% 
   mutate(
     chla_thr = paste0('chla_', chla_thr), 
     thrtyp = 'chlacnt'
[…]

### 2023-08-29 R/funcs.R

@@ -10,3 +10,14 @@ p_ast <- function(x){
   return(out)
   
 }
+
+# run length encoding for vector of TF, return longest run of T
+runfunc <- function(cnt){
+  x <- rle(cnt)
+  rn <- x$lengths[x$values]
+  if(length(rn) == 0)
+    out <- 0
+  else
+    out <- max(rn)
+  return(out)
+}

### 2023-08-29 R/modelscratch.R

@@ -1,170 +1,0 @@
-library(tidyverse)
-library(lmerTest)
-library(tbeptools)
-library(here)
-library(vegan)
-library(scales)
-
-load(file = here('data/thrdat.RData'))
-
-seglng <- c('Old Tampa Bay', 'Hillsborough Bay', 'Middle Tampa Bay', 'Lower Tampa Bay')
-segshr <- c('OTB', 'HB', 'MTB', 'LTB')
-
-##
-# seagrass fo
-transectocc <- anlz_transectocc(transect)
-transectavespp <- anlz_transectavespp(transectocc, by_seg = TRUE)
-
-fodat <- transectavespp %>% 
-  select(-nsites) %>% 
-  filter(bay_segment %in% segshr) %>% 
-  filter(Savspecies %in% c('Halodule', 'Syringodium', 'Thalassia')) %>% 
-  mutate(
-    Savspecies = droplevels(Savspecies)
-  ) %>% 
-  pivot_wider(names_from = 'Savspecies', values_from = 'foest', values_fill = 0)
-
-##
-# threshold count data
-
-# these had strong models
-salthr <- '25'
-tmpthr <- '30'
-
-cntdat <- thrdat %>% 
-  filter(salithr == paste0('sali_', salthr)) %>% 
-  filter(tempthr == paste0('temp_', tmpthr)) %>% 
-  summarise(
-    cnt = mean(cnt),
-    .by = c('thrtyp', 'bay_segment', 'yr')
-  ) %>% 
-  filter(yr >= min(fodat$yr) & yr <= max(fodat$yr)) %>% 
-  pivot_wider(names_from = thrtyp, values_from = cnt)
-
-##
-# load data
-
-load(file = url('https://github.com/tbep-tech/load-estimates/raw/main/data/totanndat.RData'))
-
-tndat <- totanndat %>% 
-  select(bay_segment, yr = year, tn_load) %>% 
-  filter(yr >= min(fodat$yr) & yr <= max(fodat$yr)) %>% 
-  filter(bay_segment %in% seglng) %>% 
-  mutate(
-    bay_segment = factor(bay_segment, levels = seglng, labels = segshr)
-  )
-
-##
-# chl data
-
-chldat <- epcdata %>% 
-  select(bay_segment, yr, chla) %>% 
-  summarise(
-    chla = mean(chla, na.rm = T),
[…]

### 2023-08-30 manu-draft.qmd
addl pyro content
added nms fig but need to add caption, working gam but need to create plot and tables
added gam results figure

@@
 
+NMS results in @fig-nms and GAM results in @fig-gamres.
+
 ## Discussion
@@
 For salinity, highest growth rates at 24 psu and above, with decline in growth rates at 20 psu [@usup1994], in Florida growth has occurred in psu 10-45, bloom concentrations observed above 15 psu [@phlips2006; @lopez2021].
+Optimal salinity range between 20-32 psu [@lopez2021].
 
@@
 
+![Ordination plots of Nonmetric Multidimensional Scaling results for annual estimates in each bay segment for total seagrass frequency occurrence (Freq.
+Occ.), counts of days above the temperature threshold (Temp.), counts of days below the salinity threshold (Sal.), and counts of days above or below for both, respectively (Sal.
++ Temp.).
+Points are sized by seagrass frequency occurrence.
+95% confidence ellipses are shown for observations in each major bay segment  OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/nms.png){#fig-nms}
+
+{{< pagebreak >}}
+
+![Predicted results from a Generalized Additive Model evaluating frequency occurrence of seagrass versus year (left), number of days in a year when temperature exceeded the threshold (top right), and number of days in a year when salinity was below the threshold (bottom right).
+Year was fit as a main effect, whereas the temperature and salinity counts were fit with interactions by year.
+The right plots show model predicted results for the threshold variables for selected year slices.
+2016 to 2022 are years with recent seagrass decline.
+Shaded areas are 95% confidence intervals.](figs/gamres.png){#fig-gamres}
+
+{{< pagebreak >}}
+
 ## Tables {.unnumbered}

### 2023-08-30 R/figs.R

@@ -17,6 +17,7 @@ library(ggord)
 library(scales)
 library(visreg)
 library(car)
+library(mgcv)
 
 yrsel <- c(1974, 2022)
 tempthr <- c(29, 30, 31)
@@ -543,10 +544,82 @@ tomod <- cmbdat %>%
 
 cmbmod <- gam(total ~ ti(salicnt) + ti(salicnt, yr) + ti(bothcnt) + ti(bothcnt, yr) + ti(yr) + ti(tempcnt) + ti(tempcnt, yr), data = tomod)
 
-visreg(cmbmod, 'bothcnt', by = 'yr', breaks = c(2000, 2005, 2012, 2016, 2020, 2022), scale = 'response', rug = 1)
-visreg(cmbmod, 'tempcnt', by = 'yr', breaks = c(2000, 2005, 2012, 2016, 2020, 2022), scale = 'response', rug = 1)
-visreg(cmbmod, 'salicnt', by = 'yr', breaks = c(2000, 2005, 2012, 2016, 2020, 2022), scale = 'response', rug = 1)
+# vifmod <- glm(total ~ salicnt + bothcnt + tempcnt + yr, data = tomod)
+# vif(vifmod)
 
-vifmod <- glm(total ~ salicnt + bothcnt + tempcnt + yr, data = tomod)
-vif(vifmod)
+yrbrks <- c(2000, 2010, 2016, 2020, 2022)
+# visreg(cmbmod, 'bothcnt', by = 'yr', breaks = yrbrks, scale = 'response', rug = 1)
+yrvis <- visreg(cmbmod, 'yr', scale = 'response', rug = 1, plot = F)
+yrfit <- yrvis$fit
+yrres <- yrvis$res
+tempvis <- visreg(cmbmod, 'tempcnt', by = 'yr', breaks = yrbrks, scale = 'response', rug = 1, plot = F)
+tempfit <- tempvis$fit
+tempres <- tempvis$res
+salivis <- visreg(cmbmod, 'salicnt', by = 'yr', breaks = yrbrks, scale = 'response', rug = 1, plot = F)
+salifit <- salivis$fit
+salires <- salivis$res
 
+alph <- 0.5
+ylb <- 'Freq. Occ.'
+lwd <- 0.8
+ptsz <- 1
+thm <- theme_minimal() + 
+  theme(
+    panel.grid.minor = element_blank(), 
+    strip.text = element_text(size = 11),
[…]

### 2023-08-30 manu-draft.qmd
add GAM summary table

@@
 load(file = here('tabs/daytab.RData'))
+load(file = here('tabs/gamtab.RData'))
 ```
@@
 ![Predicted results from a Generalized Additive Model evaluating frequency occurrence of seagrass versus year (left), number of days in a year when temperature exceeded the threshold (top right), and number of days in a year when salinity was below the threshold (bottom right).
-Year was fit as a main effect, whereas the temperature and salinity counts were fit with interactions by year.
+All variables were fit as main effects, including interactions with year.
 The right plots show model predicted results for the threshold variables for selected year slices.
@@
 
+Summary of a Generalized Additive Model evaluating frequency occurrence of seagrass versus year, number of days in a year when temperature exceeded the threshold, number of days in a year when salinity was below the threshold, and number of days in a year when both temperature and salinity were above or below thresholds.
+All variables were fit as main effects, including interactions with year.
+edf: estimate degrees of freedom, ref.df: reference degrees of freedom, F: F-statistic.
+* p < 0.05, ** p < 0.005.
+
+```{r}
+gamtab
+```
+
 ## References {.unnumbered}

### 2023-09-04 manu-draft.qmd
add total fo to fig 1
added spei/spi data and fig for meteo and wq trends

@@
 
-Observed trends aggregated by year and bay segment across stations and months in @fig-saltempraw.
+Observed trends aggregated by year and bay segment across stations and months in @fig-meteowqraw.
 
@@
 Changes are shown for major bay segments.
-Note the different time scale between (a) and (b).
-Coverage maps in (a) began in 1988 and seagrass transect monitoring in (b) began in 1998.
+Note the different time scale between (a) and (b); coverage maps in (a) began in 1988 and seagrass transect monitoring in (b) began in 1998.
 Red lines in (a) show approximate capacity of seagrass coverage based on the baywide target of 40,000 acres.
@@
 
-![Long-term salinity and temperature trends for major bay segments from 1975 to 2022.
-Points are colored by sampling location in the water column.
-Points show the average and 95% confidence interval across all stations and sampling months for each year in each bay segment.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/saltempraw.png){#fig-saltempraw}
+![Long-term precipitation, hydrologic load, Standard Precipitation Index (SPI), air temperature, salinity, and water temperature trends from 1975 to 2022.
+Points for salinity and water temperature are colored by sampling location in the water column and show the average (95% confidence interval) across all stations and sampling months for each year in each bay segment.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/meteowqraw.png){#fig-meteowqraw}
 

### 2023-09-04 R/dat_proc.R

@@ -10,8 +10,84 @@ library(sf)
 library(lmerTest)
 library(modelbased)
 library(mgcv)
+library(SPEI)
+library(rnoaa)
 
 source(here('R/funcs.R'))
+
+# SPEI, monthly rain, and monthly air temp ----------------------------------------------------
+
+noaa_key <- Sys.getenv('NOAA_KEY')
+
+yrs <- seq(1975, 2022)
+
+stameta <- ghcnd_stations()
+sta <- 'USW00012842'
+stalat <- stameta %>% 
+  filter(id %in% sta) %>% 
+  pull(latitude) %>% 
+  unique()
+
+res <- yrs %>%
+  tibble::enframe('name', 'year') %>%
+  dplyr::group_by(name) %>%
+  tidyr::nest() %>%
+  dplyr::mutate(
+    ests = purrr::map(data, function(x){
+      
+      yr <- x$year
+      cat(yr, '\n')
+      
+      start <- paste0(yr, "-01-01")
+      end <- paste0(yr, "-12-31")
+      
+      # download NOAA UWS rainfall and temperature station data
+      tia_dat <- meteo_pull_monitors(monitors = sta, 
+                                     date_min = start, date_max = end)
+      
+      out <- tia_dat
+      
+      return(out)
+      
+    })
+  )
+
+# tavg does not start until 1998, so reproduce from tmin/tmax
+resdat <- res %>% 
+  unnest('data') %>% 
+  unnest('ests') %>% 
+  ungroup() %>% 
+  select(date, prcp, tmax, tmin) %>% 
+  mutate(
+    yr = year(date),
+    mo = month(date),
+    tavg_c = (tmax + tmin) / 2,
+    tavg_c = tavg_c / 10,
+    tavg_c = na.approx(tavg_c),
+    precip_mm = prcp / 10, 
+    precip_in = precip_mm * 0.0393701
+  ) %>% 
+  select(date, yr, mo, tavg_c, precip_mm, precip_in)
+
+# monthly avg, sum
+speidat <- resdat %>% 
+  summarise(
+    precip_mm = sum(precip_mm),
+    precip_in = sum(precip_in), 
+    tavg_c = mean(tavg_c),
+    date = min(date),
+    .by = c('yr', 'mo')
+  ) %>% 
+  mutate(
+    PET = thornthwaite(tavg_c, stalat),
+    BAL = precip_mm - PET,
+    spei = as.numeric(spei(BAL, 12)$fitted), 
[…]

### 2023-09-04 R/figs.R

@@ -120,6 +120,89 @@ dev.off()
 
 # raw temp and salinity changes ---------------------------------------------------------------
 
+load(file = here('data/speidat.RData'))
+load(file = url("https://github.com/tbep-tech/load-estimates/raw/main/data/totanndat.RData"))
+
+yrsel <- range(speidat$yr)
+
+# hydro load, otb, mtb, ltb, hb
+hydrodat <- totanndat %>% 
+  select(year, bay_segment, hy_load) %>% 
+  filter(bay_segment %in% c('Hillsborough Bay', 'Old Tampa Bay', 'Lower Tampa Bay', 'Middle Tampa Bay')) %>% 
+  summarise(
+    hy_load = sum(hy_load) / 1e3, 
+    .by = 'year'
+  ) 
+
+toplo1 <- speidat %>% 
+  select(yr, precip_mm, tavg_c) %>% 
+  summarise(
+    precip_mm = sum(precip_mm), 
+    tavg_c = mean(tavg_c), 
+    .by = 'yr'
+  )
+
+thm <- theme_minimal() + 
+  theme(
+    strip.placement = 'outside', 
+    panel.grid.minor = element_blank(), 
+    axis.text.y = element_text(size = 12), 
+    legend.text = element_text(size= 12)
+  )
+
+p1 <- ggplot(toplo1, aes(x = yr, y = precip_mm / 1e3)) + 
+  geom_line() + 
+  geom_point() + 
+  geom_smooth(method = 'lm', se = F, formula = y~x, color = 'blue') +
+  thm +
+  theme(
+    axis.text.x = element_blank()
+  ) +
+  labs(
+    x = NULL, 
+    y = 'Precip. (m)'
+  )
+
+p2 <- ggplot(hydrodat, aes(x = year, y = hy_load)) +
+  geom_line() + 
+  geom_point() + 
+  geom_smooth(method = 'lm', se = F, formula = y~x, color = 'blue') +
+  coord_cartesian(xlim = yrsel) + 
+  thm +
+  theme(
+    axis.text.x = element_blank()
+  ) +
+  labs(
+    x = NULL, 
+    y = expression(paste('Hyd. load (', 10^3, ' t/yr)'))
+  )
+
+p3 <- ggplot(speidat, aes(x = date, y = spi, fill = spisign)) + 
+  geom_col(width = 35) +
+  scale_fill_manual(values = c('grey70', 'grey30'), guide = 'none') + 
+  thm + 
[…]

### 2023-09-05 manu-draft.qmd
minor fig aesthetic changes, added some latex equations for mixeff and gam

@@
 
-For @fig-saltempraw, salinity shows much higher inter-annual variability, but lower intra-year variation among stations.
+For @fig-meteowqraw, salinity shows much higher inter-annual variability, but lower intra-year variation among stations.
 Trends are similar for bottom vs top, temperature shows stronger trend than salinity.
@@
 
+$$
+Y_{ij} = \beta_0 + \beta_1X_{ij} + u_i + \epsilon_{ij}
+$$
+
 NMS results in @fig-nms and GAM results in @fig-gamres.
+
+GAM notation:
+
+$$
+Y_i = \beta_0 + f_1\left(\text{Sal}_i\right) + f_2\left(\text{Temp}_i\right) + f_3\left(\text{Both}_i\right) + f_4\left(\text{Year}_i, \text{Sal}_i\right) + f_5\left(\text{Year}_i, \text{Temp}_i\right) + f_6\left(\text{Year}_i, \text{Both}_i\right)  + \epsilon_i
+$$
 
@@
 edf: estimate degrees of freedom, ref.df: reference degrees of freedom, F: F-statistic.
-* p < 0.05, ** p < 0.005.
+\* p \< 0.05, \*\* p \< 0.005.
 

### 2023-09-05 R/figs.R

@@ -118,12 +118,10 @@ png(here('figs/seagrasschg.png'), height = 6, width = 8, family = 'serif', units
 print(p)
 dev.off()
 
-# raw temp and salinity changes ---------------------------------------------------------------
+# meteorological, salinity, water temp obs data -----------------------------------------------
 
 load(file = here('data/speidat.RData'))
 load(file = url("https://github.com/tbep-tech/load-estimates/raw/main/data/totanndat.RData"))
-
-yrsel <- range(speidat$yr)
 
 # hydro load, otb, mtb, ltb, hb
 hydrodat <- totanndat %>% 
@@ -167,7 +165,7 @@ p2 <- ggplot(hydrodat, aes(x = year, y = hy_load)) +
   geom_line() + 
   geom_point() + 
   geom_smooth(method = 'lm', se = F, formula = y~x, color = 'blue') +
-  coord_cartesian(xlim = yrsel) + 
+  coord_cartesian(xlim = range(speidat$yr)) + 
   thm +
   theme(
     axis.text.x = element_blank()
@@ -184,7 +182,7 @@ p3 <- ggplot(speidat, aes(x = date, y = spi, fill = spisign)) +
   theme(
     axis.text.x = element_blank()
   ) +
-  coord_cartesian(xlim = range(resmo$date)) +
+  coord_cartesian(xlim = range(speidat$date)) +
   labs(
     x = NULL, 
     y = 'SPI (z-values)'
@@ -234,7 +232,7 @@ toplo5 <- toplo %>%
   filter(var == 'Salinity (ppt)')
 p5 <- ggplot(toplo5, aes(x = yr, y = avev, group = loc, color = loc)) + 
   geom_linerange(aes(ymin = lov, ymax = hiv), position = position_dodge2(width = wd), show.legend = F, alpha = 0.7) + 
-  geom_point(position = position_dodge2(width = wd)) +
+  geom_point(position = position_dodge2(width = wd), size = 0.5) +
   # scale_x_continuous(breaks = seq(min(toplo$yr), max(toplo$yr), by = 3)) +
   geom_smooth(method = 'lm', formula = y ~ x, se = F) +
   facet_grid(~ bay_segment) +
@@ -254,7 +252,7 @@ toplo6 <- toplo %>%
   filter(var == 'Water temp. (\u00B0C)')
[…]

### 2023-09-08 manu-draft.qmd
redo nms with separate for both and temp/sal
kendall fig 1998 to pres, transect preceding load model, mixeff proc plot more flexible

@@
 
-![Trends from 1974 to 2022 for salinity and temperature measurements at long-term monitoring stations in Tampa Bay.
+![Trends from 1998 to 2022 for salinity and temperature measurements at long-term monitoring stations in Tampa Bay.
 Results for (a) seasonal Kendall tests by station and monitoring location (top or bottom of the waer column) are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr-1).
@@
 + Temp.).
+The left plot excludes counts of days with both temperature and salinity, whereas the right plot excludes the separate temperature and salinity metrics.
 Points are sized by seagrass frequency occurrence.

### 2023-09-15 manu-draft.qmd
more models of change over time

@@
 -   Get FIM, Pinellas, Manatee data
+
+### Models
+
+All models excluded Lower Tampa Bay because of minimal seagrass change over time. Also attempted to evaluate total and change as binary.
+
+GLMs of percent change each year by Sal, Temp, with year and bay segment interactions.
+Attempted different windows of time from approximate seagrass sample date, 30 days to one year.
+The best model was one that included an effect of Salinity and an interaction with bay segment.
+Hillsborough Bay showed a strong response of increasing likelihood of seagrass loss with number of days when salinity was below the threshold.
+
+Then GAMs were evaluated looking at total by Salinity, with interaction of year and bay segment.  Also created a model looking at percent change vs Temp and year/bay segment interaction.
+
+Show plots of both the glm and gams.
 

### 2023-09-15 R/dat_proc.R

@@ -465,7 +465,7 @@ save(chlthrdat, file = here('data/chlthrdat.RData'), compress = 'xz')
 transectocc <- anlz_transectocc(transect) 
 transectavespp <- transectocc %>% 
   anlz_transectavespp(by_seg = TRUE) %>% 
-  filter(bay_segment %in% c('HB', 'OTB', 'MTB')) %>% 
+  filter(bay_segment %in% c('HB', 'OTB', 'MTB', 'LTB')) %>% 
   filter(Savspecies %in% 'total') %>% 
   select(yr, bay_segment, foest)
 
@@ -476,7 +476,7 @@ trnptssub <- trnpts %>%
     Transect = TRAN_ID,
     bay_segment
   ) %>%
-  filter(bay_segment %in% c('OTB', 'HB', 'MTB'))
+  filter(bay_segment %in% c('OTB', 'HB', 'MTB', 'LTB'))
 trndts <- transectocc %>%
   ungroup() %>% 
   filter(Savspecies == 'total') %>%
@@ -580,7 +580,6 @@ thrdat <- cmbdat %>%
   unnest('cmbcnt') %>% 
   pivot_longer(names_to = 'thrtyp', values_to = 'cnt', matches('cnt')) %>% 
   filter(yr >= (min(year(trndts$date)) - 1)) %>% 
-  filter(!bay_segment == 'LTB') %>% 
   arrange(bay_segment, station, thrtyp, salithr, tempthr, date)
 
 ##

### 2023-09-18 manu-draft.qmd
added glm plot results and removed gam

@@
 
-NMS results in @fig-nms and GAM results in @fig-gamres.
-
-GAM notation:
-
-$$
-Y_i = \beta_0 + f_1\left(\text{Sal}_i\right) + f_2\left(\text{Temp}_i\right) + f_3\left(\text{Both}_i\right) + f_4\left(\text{Year}_i, \text{Sal}_i\right) + f_5\left(\text{Year}_i, \text{Temp}_i\right) + f_6\left(\text{Year}_i, \text{Both}_i\right)  + \epsilon_i
-$$
+NMS results in @fig-nms and seagrass decline results in @fig-sgmod
 
@@
 
-![Predicted results from a Generalized Additive Model evaluating frequency occurrence of seagrass versus year (left), number of days in a year when temperature exceeded the threshold (top right), and number of days in a year when salinity was below the threshold (bottom right).
-All variables were fit as main effects, including interactions with year.
-The right plots show model predicted results for the threshold variables for selected year slices.
-2016 to 2022 are years with recent seagrass decline.
-Shaded areas are 95% confidence intervals.](figs/gamres.png){#fig-gamres}
+![Predicted results from generalized linear models evaluating seagrass decline as (a) percent change between years and (b) probability of decline versus number of days when salinity was below the threshold.
+Models evaluated the response variable on the y-axes as a function of salinity days and interactions with year category (seagrass recovery prior to 2016, seagrass decline 2016 to present) and bay segment.
+Shaded areas are 95% confidence intervals.
+Colored lines in (b) as rug plots show instances of seagrass increase (blue) or decrease (red) by year.](figs/sgmod.png){#fig-sgmod}
 

### 2023-09-18 R/dat_proc.R

@@ -845,19 +845,179 @@ cmbdat <- fodat %>%
 
 save(cmbdat, file = here('data/cmbdat.RData'))
 
-# gam for cmbdat ------------------------------------------------------------------------------
 
-load(file = here('data/cmbdat.RData'))
+# seagrass decline models ---------------------------------------------------------------------
+
+load(file = here('data/thralltrndat.RData'))
+
+seglng <- c('Old Tampa Bay', 'Hillsborough Bay', 'Middle Tampa Bay', 'Lower Tampa Bay')
+segshr <- c('OTB', 'HB', 'MTB', 'LTB')
+
+##
+# seagrass fo
+transectocc <- anlz_transectocc(transect)
+transectave <- anlz_transectave(transectocc)
+
+fodat <- transectave %>% 
+  select(bay_segment, yr, total = foest) %>% 
+  filter(bay_segment %in% segshr) %>% 
+  mutate(
+    total = total / 100
+  ) %>% 
+  ungroup()
+
+trnptsshed <- trnpts %>% 
+  sf::st_set_geometry(NULL) %>% 
+  select(Transect = TRAN_ID, bay_segment) %>% 
+  unique()
+
+bbdat <- transectocc %>% 
+  ungroup() %>% 
+  filter(Savspecies %in% c('Halodule', 'Syringodium', 'Thalassia')) %>% 
+  filter(
+    bbest > 0, 
+    .by = c('Date', 'Transect')
+  ) %>% 
+  inner_join(trnptsshed, by = 'Transect') %>% 
+  filter(!bay_segment %in% c('BCB', 'TCB', 'MR')) %>% 
+  mutate(
+    yr = year(Date)
+  ) %>% 
+  summarise(
+    bbave = mean(bbest),
+    .by = c('bay_segment', 'yr')
+  )
+
+##
+# threshold count data
+
+# these had strong models
+salthr <- '25'
+tmpthr <- '30'
+
+# dys <- seq(30, 365, by = 30)
+# for(i in dys){
+#   
+#   thrtrndat <- thralltrndat %>% 
+#     filter(salithr == paste0('sali_', salthr)) %>% 
+#     filter(tempthr == paste0('temp_', tmpthr)) %>% 
+#     filter(dycnt <= i) %>% 
+#     summarise(
+#       cnt = runfunc(cnt),
[…]

### 2023-09-18 R/figs.R

@@ -19,6 +19,8 @@ library(visreg)
 library(car)
 library(mgcv)
 
+source(here('R/funcs.R'))
+
 yrsel <- c(1998, 2022)
 tempthr <- c(29, 30, 31)
 tempcol <- c('coral', 'red2', 'darkred')
@@ -588,92 +590,56 @@ png(here('figs/nms.png'), height = 4, width = 9, family = 'serif', units = 'in',
 print(p)
 dev.off()
 
-# gam results ---------------------------------------------------------------------------------
+# seagrass decline models ---------------------------------------------------------------------
 
-load(file = here('data/cmbmod.RData'))
+load(file = here('data/pchgmod.RData'))
+load(file = here('data/binomod.RData'))
 
-# load(file = here('data/cmbdat.RData'))
-# 
-# tomod <- cmbdat %>%
-#   filter(!bay_segment %in% 'LTB')
-# 
-# vifmod <- glm(total ~ Sal + Temp + Both + Year, data = tomod)
-# vif(vifmod)
+toplo1 <- getprd_fun(pchgmod)
+toplo2 <- getprd_fun(binomod)
 
-yrbrks <- c(2000, 2010, 2016, 2020, 2022)
-# visreg(cmbmod, 'Both', by = 'Year', breaks = yrbrks, scale = 'response', rug = 1)
-yrvis <- visreg(cmbmod, 'Year', scale = 'response', rug = 1, plot = F)
-yrfit <- yrvis$fit
-yrres <- yrvis$res
-tempvis <- visreg(cmbmod, 'Temp', by = 'Year', breaks = yrbrks, scale = 'response', rug = 1, plot = F)
-tempfit <- tempvis$fit
-tempres <- tempvis$res
-salivis <- visreg(cmbmod, 'Sal', by = 'Year', breaks = yrbrks, scale = 'response', rug = 1, plot = F)
-salifit <- salivis$fit
-salires <- salivis$res
-
-alph <- 0.5
-ylb <- 'Freq. Occ.'
-lwd <- 0.8
-ptsz <- 1
-thm <- theme_minimal() + 
+p1 <- ggplot(toplo1, aes(x = Sal)) + 
+  geom_ribbon(aes(ymin = visregLwr, ymax = visregUpr), fill = 'lightgrey', alpha = 0.5) + 
+  geom_line(aes(y = visregFit)) + 
+  geom_point(data = tomod, aes(x = Sal, y = pchg), color = 'dodgerblue2') +
[…]

### 2023-09-18 R/funcs.R

@@ -26,3 +26,31 @@ runfunc <- function(cnt){
     # }
   return(out)
 }
+
+# get model predictions for 
+getprd_fun <- function(modin){
+  
+  salfitotb <- visreg(modin, 'Sal', by = 'yrcat', cond = list(bay_segment = 'OTB'), scale = 'response', plot = F) %>% 
+    .$fit %>%  
+    mutate(
+      bay_segment = 'OTB'
+    )
+  salfithb <- visreg(modin, 'Sal', by = 'yrcat', cond = list(bay_segment = 'HB'), scale = 'response', plot = F) %>% 
+    .$fit %>% 
+    mutate(
+      bay_segment = 'HB'
+    ) 
+  salfitmtb <- visreg(modin, 'Sal', by = 'yrcat', cond = list(bay_segment = 'MTB'), scale = 'response', plot = F) %>% 
+    .$fit %>% 
+    mutate(
+      bay_segment = 'MTB'
+    )
+  
+  out <- bind_rows(salfitotb, salfithb, salfitmtb) %>% 
+    mutate(
+      bay_segment = factor(bay_segment, levels = segshr)
+    )
+  
+  return(out)
+  
+}

### 2023-09-18 R/tabs.R

@@ -90,34 +90,34 @@ save(daytab, file = here('tabs/daytab.RData'))
 
 # GAM summary ---------------------------------------------------------------------------------
 
-load(file = here('data/cmbmod.RData'))
-
-totab <- cmbmod %>% 
-  tidy() %>% 
-  mutate(
-    p.value = p_ast(p.value), 
-    term = gsub('^ti\\(|\\)$', '', term),
-    Effect = case_when(
-      grepl('\\,', term) ~ 'Interaction',
-      T ~ 'Main'
-    )
-  ) %>% 
-  mutate_if(is.numeric, round, 2) %>% 
-  unite('statistic', statistic, p.value, sep = '') %>% 
-  select(
-    Effect, 
-    Term = term, 
-    edf, 
-    ref.df, 
-    `F` = statistic
-  )
-
-gamtab <- totab %>% 
-  as_grouped_data(groups = 'Effect') %>% 
-  flextable() %>% 
-  padding(padding = 0, part = 'all') %>% 
-  font(part = 'all', fontname = 'Times New Roman') %>% 
-  align(j = 3:5, align = 'left', part = 'all')
-
-save(gamtab, file = here('tabs/gamtab.RData'))
+# load(file = here('data/cmbmod.RData'))
+# 
+# totab <- cmbmod %>% 
+#   tidy() %>% 
+#   mutate(
+#     p.value = p_ast(p.value), 
+#     term = gsub('^ti\\(|\\)$', '', term),
+#     Effect = case_when(
+#       grepl('\\,', term) ~ 'Interaction',
+#       T ~ 'Main'
+#     )
+#   ) %>% 
+#   mutate_if(is.numeric, round, 2) %>% 
+#   unite('statistic', statistic, p.value, sep = '') %>% 
+#   select(
+#     Effect, 
+#     Term = term, 
+#     edf, 
+#     ref.df, 
+#     `F` = statistic
+#   )
+# 
+# gamtab <- totab %>% 
+#   as_grouped_data(groups = 'Effect') %>% 
+#   flextable() %>% 
+#   padding(padding = 0, part = 'all') %>% 
+#   font(part = 'all', fontname = 'Times New Roman') %>% 
+#   align(j = 3:5, align = 'left', part = 'all')
+# 
+# save(gamtab, file = here('tabs/gamtab.RData'))
 

### 2023-09-19 manu-draft.qmd
redo all figs and tabs with temp before salinity

@@
 
-![Long-term precipitation, hydrologic load, Standard Precipitation Index (SPI), air temperature, salinity, and water temperature trends from 1975 to 2022.
+![Long-term air temperature, precipitation, hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1999 to 2022.
 Points for salinity and water temperature are colored by sampling location in the water column and show the average (95% confidence interval) across all stations and sampling months for each year in each bay segment.
@@
 
-![Trends from 1998 to 2022 for salinity and temperature measurements at long-term monitoring stations in Tampa Bay.
-Results for (a) seasonal Kendall tests by station and monitoring location (top or bottom of the waer column) are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr-1).
-Summarized seasonal trends by month are shown for (b) top and (c) bottom measurements as the percent of stations in each bay segment with significant decreasing (salinity) or increasing (temperature) trends.
+![Trends from 1998 to 2022 for temperature and salinity measurements at long-term monitoring stations in Tampa Bay.
+Results for (a) seasonal Kendall tests by station and monitoring location (top or bottom of the water column) are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr-1).
+Summarized seasonal trends by month are shown for (b) top and (c) bottom measurements as the percent of stations in each bay segment with significant increasing (temperature) or decreasing (salinity) trends.
 Bay segment outlines are shown in (a); OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/kendall.png){#fig-kendall}
@@
 
-![Example of mixed effects models for the estimated number of days by year that salinity (blue) or temperature (red) were below or above thresholds of 25 psu and 30 degrees C, respectively.
-The bottom row (black) shows the number of days when both salinity and temperature were below or above the thresholds.
+![Example of mixed effects models for the estimated number of days by year that temperature (red) or salinity (blue) were above or below thresholds of 30 degrees C or 25 psu, respectively.
+The bottom row (black) shows the number of days when both temperature and salinity were above or below the thresholds.
 The models included station as a random effect for each bay segment, with grey lines indicating individual station trends and thicker lines indicating the overall model fit.
@@
 ![Ordination plots of Nonmetric Multidimensional Scaling results for annual estimates in each bay segment for total seagrass frequency occurrence (Freq.
-Occ.), counts of days above the temperature threshold (Temp.), counts of days below the salinity threshold (Sal.), and counts of days above or below for both, respectively (Sal.
-+ Temp.).
-The left plot excludes counts of days with both temperature and salinity, whereas the right plot excludes the separate temperature and salinity metrics.
+Occ.), counts of days above the temperature threshold (Temp.), counts of days below the salinity threshold (Sal.), and counts of days above or below for both, respectively.
+The left plot excludes counts of days with both temperature and salinity above and below the threshods, whereas the right plot excludes the separate temperature and salinity metrics.
 Points are sized by seagrass frequency occurrence.
@@
 
-Summary of mixed-effects models evaluating increases in the number of days each year from 1974 to 2022 when salinity or temperature was below or above critical thresholds, respectively.
-The slope estimates in the "Both" column indicate trends in the number of days when both salinity and temperature were below or above critical thresholds.
+Summary of mixed-effects models evaluating increases in the number of days each year from 1974 to 2022 when temperature or salinity were above or below critical thresholds, respectively.
+The slope estimates in the "Both" column indicate trends in the number of days when both temperature and salinity were above or below critical thresholds.
 \*\* p \< 0.005, \* p \< 0.05, - no model.
@@
 
-Number of days at the beginning and end of the period of record when salinity or temperature was below or above critical thresholds, respectively.
+Number of days at the beginning and end of the period of record when temperature or salinity were above or below critical thresholds, respectively.
 Values are the estimated mean number of days (plus standard error) from 1974 to 2022.
-Columns for "Both" indicate the estimated number of days when both salinity and temperature were below or above critical thresholds.
-Note that not all models are significant (see (table 1)).
+Columns for "Both" indicate the estimated number of days when both temperature and salinity were above or below critical thresholds.
+Note that not all models are significant (see table 1).
 \*\* p \< 0.005, \* p \< 0.05, - no model.

### 2023-09-19 R/dat_proc.R

@@ -609,7 +609,7 @@ load(file = here("data/thrdat.RData"))
 
 # create mixed effects models of cnt ~ yr with station as random intercept
 mixmods <- thrdat %>% 
-  group_by(bay_segment, salithr, tempthr, thrtyp) %>% 
+  group_by(bay_segment, tempthr, salithr, thrtyp) %>% 
   nest() %>% 
   summarise(
     mod = purrr::map(data, function(x){
@@ -712,8 +712,8 @@ mixmodprds <- thrdat %>%
   mutate(
     bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB')), 
     thrtyp = factor(thrtyp, 
-                    levels = c('salicnt', 'tempcnt', 'bothcnt'), 
-                    labels = c(paste('Salinity <', salthr), paste('Temperature >', tmpthr), 'Both')
+                    levels = c('tempcnt', 'salicnt', 'bothcnt'), 
+                    labels = c(paste('Temperature >', tmpthr), paste('Salinity <', salthr), 'Both')
     )
   )
 

### 2023-09-19 R/figs.R

@@ -150,7 +150,20 @@ thm <- theme_minimal() +
     legend.text = element_text(size= 12)
   )
 
-p1 <- ggplot(toplo1, aes(x = yr, y = precip_mm / 1e3)) + 
+p1 <- ggplot(toplo1, aes(x = yr, y = tavg_c)) + 
+  geom_line() + 
+  geom_point() + 
+  geom_smooth(method = 'lm', se = F, formula = y~x, color = 'red2') +
+  thm +
+  theme(
+    axis.text.x = element_text(size = 12)
+  ) +
+  labs(
+    x = NULL, 
+    y = 'Air temp. (\u00B0 C)'
+  )
+
+p2 <- ggplot(toplo1, aes(x = yr, y = precip_mm / 1e3)) + 
   geom_line() + 
   geom_point() + 
   geom_smooth(method = 'lm', se = F, formula = y~x, color = 'blue') +
@@ -163,7 +176,7 @@ p1 <- ggplot(toplo1, aes(x = yr, y = precip_mm / 1e3)) +
     y = 'Precip. (m)'
   )
 
-p2 <- ggplot(hydrodat, aes(x = year, y = hy_load)) +
+p3 <- ggplot(hydrodat, aes(x = year, y = hy_load)) +
   geom_line() + 
   geom_point() + 
   geom_smooth(method = 'lm', se = F, formula = y~x, color = 'blue') +
@@ -177,7 +190,7 @@ p2 <- ggplot(hydrodat, aes(x = year, y = hy_load)) +
     y = expression(paste('Hyd. load (', 10^3, ' t/yr)'))
   )
 
-p3 <- ggplot(speidat, aes(x = date, y = spi, fill = spisign)) + 
+p4 <- ggplot(speidat, aes(x = date, y = spi, fill = spisign)) + 
   geom_col(width = 35) +
   scale_fill_manual(values = c('grey70', 'grey30'), guide = 'none') + 
   thm + 
@@ -188,19 +201,6 @@ p3 <- ggplot(speidat, aes(x = date, y = spi, fill = spisign)) +
   labs(
     x = NULL, 
     y = 'SPI (z-values)'
-  )
-
-p4 <- ggplot(toplo1, aes(x = yr, y = tavg_c)) + 
-  geom_line() + 
-  geom_point() + 
-  geom_smooth(method = 'lm', se = F, formula = y~x, color = 'red2') +
-  thm +
-  theme(
-    axis.text.x = element_text(size = 12)
-  ) +
-  labs(
-    x = NULL, 
[…]

### 2023-09-19 R/tabs.R

@@ -17,8 +17,8 @@ totab <- mixmods %>%
     slo = as.character(round(slo, 2)), 
     salithr = gsub('^.*_', '', salithr),
     tempthr = gsub('^.*_', '', tempthr), 
-    thrtyp = factor(thrtyp, levels = c('salicnt', 'tempcnt', 'bothcnt'), 
-                     labels = c('Salinity', 'Temperature', 'Both')),
+    thrtyp = factor(thrtyp, levels = c('tempcnt', 'salicnt', 'bothcnt'), 
+                     labels = c('Temperature', 'Salinity', 'Both')),
     bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB'))
   ) %>% 
   unite('slo', slo, pvl, sep = '') %>% 
@@ -34,7 +34,7 @@ mixtab <- totab %>%
   add_header_row(values = c('', 'Thresholds', 'Thresholds', rep('Slopes', 3))) %>% 
   merge_at(i = 1, j = c(2:3), part = 'header') %>% 
   merge_at(i = 1, j = c(4:6), part = 'header') %>% 
-  set_header_labels(i = 2, values = c('Bay Segment', 'Salinity', 'Temperature', 'Salinity', 'Temperature', 'Both')) %>% 
+  set_header_labels(i = 2, values = c('Bay Segment', 'Temperature', 'Salinity', 'Temperature', 'Salinity', 'Both')) %>% 
   padding(padding = 0, part = 'all') %>% 
   font(part = 'all', fontname = 'Times New Roman') %>% 
   fontsize(size = 9, part = 'body')
@@ -54,8 +54,8 @@ totab <- mixmods %>%
     yrendse = paste0('(', round(yrendse, 1), ')'),
     salithr = gsub('^.*_', '', salithr),
     tempthr = gsub('^.*_', '', tempthr), 
-    thrtyp = factor(thrtyp, levels = c('salicnt', 'tempcnt', 'bothcnt'), 
-                    labels = c('Salinity', 'Temperature', 'Both')),
+    thrtyp = factor(thrtyp, levels = c('tempcnt', 'salicnt', 'bothcnt'), 
+                    labels = c('Temperature', 'Salinity', 'Both')),
     bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB'))
   ) %>% 
   unite('yrstr', yrstr, yrstrse, sep = ' ') %>%
@@ -64,8 +64,8 @@ totab <- mixmods %>%
[…]

### 2023-09-20 manu-draft.qmd
created supplement file, all figs tables in main text have full year range where possible

@@
 
-```{r, results = 'hide'}
+```{r}
+#| results: hide
 library(knitr)
@@
 
-load(file = here('tabs/mixtab.RData'))
-load(file = here('tabs/daytab.RData'))
-load(file = here('tabs/gamtab.RData'))
+# load(file = here('tabs/mixtab.RData'))
+# load(file = here('tabs/daytab.RData'))
 ```
@@
 
-All models excluded Lower Tampa Bay because of minimal seagrass change over time. Also attempted to evaluate total and change as binary.
+All models excluded Lower Tampa Bay because of minimal seagrass change over time. Evaluated total fo, percent change between years, and
 
@@
 
-![Long-term air temperature, precipitation, hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1999 to 2022.
+![Long-term air temperature, precipitation, hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1975 to 2022.
 Points for salinity and water temperature are colored by sampling location in the water column and show the average (95% confidence interval) across all stations and sampling months for each year in each bay segment.
@@
 
-![Trends from 1998 to 2022 for temperature and salinity measurements at long-term monitoring stations in Tampa Bay.
+![Trends from 1975 to 2022 for temperature and salinity measurements at long-term monitoring stations in Tampa Bay.
 Results for (a) seasonal Kendall tests by station and monitoring location (top or bottom of the water column) are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr-1).
@@
 
-Summary of mixed-effects models evaluating increases in the number of days each year from 1974 to 2022 when temperature or salinity were above or below critical thresholds, respectively.
+<!-- Summary of mixed-effects models evaluating increases in the number of days each year from 1974 to 2022 when temperature or salinity were above or below critical thresholds, respectively.
 The slope estimates in the "Both" column indicate trends in the number of days when both temperature and salinity were above or below critical thresholds.
@@
 OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
+-->
 
-```{r}
-mixtab
-```
+<!-- ```{r} -->
+<!-- mixtab -->
+<!-- ``` -->
 
-{{< pagebreak >}}
+<!-- {{< pagebreak >}} -->
 
-Number of days at the beginning and end of the period of record when temperature or salinity were above or below critical thresholds, respectively.
-Values are the estimated mean number of days (plus standard error) from 1974 to 2022.
+<!-- Number of days at the beginning and end of the period of record when temperature or salinity were above or below critical thresholds, respectively.
+Values are the estimated mean number of days (plus standard error) from 1975 to 2022.
 Columns for "Both" indicate the estimated number of days when both temperature and salinity were above or below critical thresholds.
@@
 OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
+-->
 
-```{r}
-daytab
-```
+<!-- ```{r} -->
+<!-- daytab -->
+<!-- ``` -->
 
@@
 
-Summary of a Generalized Additive Model evaluating frequency occurrence of seagrass versus year, number of days in a year when temperature exceeded the threshold, number of days in a year when salinity was below the threshold, and number of days in a year when both temperature and salinity were above or below thresholds.
-All variables were fit as main effects, including interactions with year.
-edf: estimate degrees of freedom, ref.df: reference degrees of freedom, F: F-statistic.
-\* p \< 0.05, \*\* p \< 0.005.
-
-```{r}
-gamtab
-```
-
 ## References {.unnumbered}

### 2023-09-20 supplement.qmd
created supplement file, all figs tables in main text have full year range where possible

--- 2023-09-20 manu-draft.qmd
@@
     number-sections: true
-title: "Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary"
+title: "*Supplement* Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary"
 bibliography: refs.bib
@@
 
-# load(file = here('tabs/mixtab.RData'))
-# load(file = here('tabs/daytab.RData'))
+load(file = here('tabs/supp1mixtab.RData'))
+load(file = here('tabs/supp2mixtab.RData'))
+load(file = here('tabs/supp1daytab.RData'))
+load(file = here('tabs/supp2daytab.RData'))
 ```
-
-## Abstract {.unnumbered}
-
-## Introduction
-
-Background: SST trends in Tampa Bay w/ EPC data, WFS, and deep GOM [@nickerson2023] showed that 1975 to 2022 trend for EPC was notable, but less so from 1998 to 2022.
-Also noted the trend was most pronounced in the winter, see Fig 3.
-
-Heat stress combined with highlight accelerates decline of *E. acoroides*, used 36C [@zhang2023]
-
-Widgeongrass colonization in Chesapeake Bay as an opportunistic, heat-tolerant species that has replaced Eelgrass, although former is sensitive to nutrient pulses.
-Paper provides an example of implications on climate change and nutrient management on foundation species and system resiliency [@hensel2023].
-
-Focus on two time periods: full record and recent decline
-
-Focus on two physical parameters: temperature and salinity
-
-Temp background for TB: NOAA coastwatch trends
-
-## Methods
-
-Seagrass changes over time in @fig-seagrasschg.
-
-Observed trends aggregated by year and bay segment across stations and months in @fig-meteowqraw.
-
-The conceptual stressor diagram @fig-concept.
-
-Salinity tolerance: Ruppia \> Halodule\> Thalassia \> Syringodium
-
-Halodule:
-
--   Temp: 20-30 (range, @lewis1985)
-
--   Salinity: grows well in most salinity ranges [@lirman2003], like Syringodium but more tolerant [@lewis1985], 10-35 (range) or \< 10 (S. Scolaro)
-
-Thalassia:
-
--   Temp: 20-30 (range, @lewis1985, @zieman1975), 30-31 (upper, from SS)
-
--   Salinity: 25 (lower, @lewis1985); 30, lowest at 5 (lower, @lirman2003), 33-38 (range, @phillips1960); 24-35 (range, @zieman1975 citing @phillips1960); 15 (SS)
-
-Syringodium:
-
--   Temp: 20-30 (range, @lewis1985), 29 (upper from SS)
-
--   Salinity: 20 (lower, @lewis1985), 25 (lower, @lirman2003); 20
-
-@lewis1985 review of seagrass in Tampa Bay.
-@lirman2003 conducted exposure experiments to evaluate seagrass growth in response to a range of salinity conditions.
-@mcmillan1967 discusses growth of halodule, syringodium, thalassia, and ruppia in response to salinity increases (up to 75 psu), no info on lower limit.
-Cites @phillips1960 for a salinity range of Thalassia in Florida of 33 - 38 psu.
-@zieman1975 discusses seasonal variation of thalassia relative to temp and salinity
-
--   Get FIM, Pinellas, Manatee data
-
-### Models
-
-All models excluded Lower Tampa Bay because of minimal seagrass change over time. Evaluated total fo, percent change between years, and
-
-GLMs of percent change each year by Sal, Temp, with year and bay segment interactions.
-Attempted different windows of time from approximate seagrass sample date, 30 days to one year.
-The best model was one that included an effect of Salinity and an interaction with bay segment.
-Hillsborough Bay showed a strong response of increasing likelihood of seagrass loss with number of days when salinity was below the threshold.
-
-Then GAMs were evaluated looking at total by Salinity, with interaction of year and bay segment. Also created a model looking at percent change vs Temp and year/bay segment interaction.
-
-Show plots of both the glm and gams.
-
-## Results
-
-For @fig-meteowqraw, salinity shows much higher inter-annual variability, but lower intra-year variation among stations.
-Trends are similar for bottom vs top, temperature shows stronger trend than salinity.
-
-Kendall test results are shown in @fig-kendall.
-
-An example of mixed effects models for two selected thresholds is shown in @fig-mixeff.
-
-$$
-Y_{ij} = \beta_0 + \beta_1X_{ij} + u_i + \epsilon_{ij}
-$$
-
-NMS results in @fig-nms and seagrass decline results in @fig-sgmod
-
-## Discussion
-
-Other areas showing seagrass loss - Florida Bay is different, less water flowing out of everglades and compounding SLR has elevated salinity and likely stress in other direction.
-Also, Biscayne Bay and IRL is a lot like OTB, poor flushing for example.
-
-Viral/disease effects on seagrass change, [@vanbogaert2018] for TB and [@duffin2021] for Florida Bay.
-
-OTB seagrass loss and role of pyrodinium, optimal temperatures from 28-31 [@lopez2021].
-For salinity, highest growth rates at 24 psu and above, with decline in growth rates at 20 psu [@usup1994], in Florida growth has occurred in psu 10-45, bloom concentrations observed above 15 psu [@phlips2006; @lopez2021].
-Optimal salinity range between 20-32 psu [@lopez2021].
-
-## Acknowledgments
-
-<!-- {{< pagebreak >}} -->
 
@@
 
-![Seagrass changes over time in Tampa Bay for (a) areal coverage and (b) frequency occurrence of major species.
-Changes are shown for major bay segments.
-Note the different time scale between (a) and (b); coverage maps in (a) began in 1988 and seagrass transect monitoring in (b) began in 1998.
-Red lines in (a) show approximate capacity of seagrass coverage based on the baywide target of 40,000 acres.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/seagrasschg.png){#fig-seagrasschg}
+![Air temperature, precipitation, hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1998 to 2022.
+Points for salinity and water temperature are colored by sampling location in the water column and show the average (95% confidence interval) across all stations and sampling months for each year in each bay segment.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppmeteowqraw.png){#fig-suppmeteowqraw}
 
@@
 
-![Long-term air temperature, precipitation, hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1975 to 2022.
-Points for salinity and water temperature are colored by sampling location in the water column and show the average (95% confidence interval) across all stations and sampling months for each year in each bay segment.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/meteowqraw.png){#fig-meteowqraw}
-
+![Trends from 1998 to 2022 for temperature and salinity measurements at long-term monitoring stations in Tampa Bay.
+Results for (a) seasonal Kendall tests by station and monitoring location (top or bottom of the water column) are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr-1).
+Summarized seasonal trends by month are shown for (b) top and (c) bottom measurements as the percent of stations in each bay segment with significant increasing (temperature) or decreasing (salinity) trends.
+Bay segment outlines are shown in (a); OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppkendall.png){#fig-suppkendall}
 {{< pagebreak >}}
 
-![Conceptual stressor diagram](figs/concept.png){#fig-concept}
-
-{{< pagebreak >}}
-
-![Trends from 1975 to 2022 for temperature and salinity measurements at long-term monitoring stations in Tampa Bay.
-Results for (a) seasonal Kendall tests by station and monitoring location (top or bottom of the water column) are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr-1).
-Summarized seasonal trends by month are shown for (b) top and (c) bottom measurements as the percent of stations in each bay segment with significant increasing (temperature) or decreasing (salinity) trends.
-Bay segment outlines are shown in (a); OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/kendall.png){#fig-kendall}
-
-{{< pagebreak >}}
-
-![Example of mixed effects models for the estimated number of days by year that temperature (red) or salinity (blue) were above or below thresholds of 30 degrees C or 25 psu, respectively.
+![Example of mixed effects models for the estimated number of days by year that temperature (red) or salinity (blue) were above or below thresholds of 30 degrees C or 25 psu, respectively, from 1998 to 2022.
 The bottom row (black) shows the number of days when both temperature and salinity were above or below the thresholds.
@@
 Slopes for significant models are shown in the bottom right of each facet.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/mixeff.png){#fig-mixeff}
-
-{{< pagebreak >}}
-
-![Ordination plots of Nonmetric Multidimensional Scaling results for annual estimates in each bay segment for total seagrass frequency occurrence (Freq.
-Occ.), counts of days above the temperature threshold (Temp.), counts of days below the salinity threshold (Sal.), and counts of days above or below for both, respectively.
-The left plot excludes counts of days with both temperature and salinity above and below the threshods, whereas the right plot excludes the separate temperature and salinity metrics.
-Points are sized by seagrass frequency occurrence.
-95% confidence ellipses are shown for observations in each major bay segment OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/nms.png){#fig-nms}
-
-{{< pagebreak >}}
-
-![Predicted results from generalized linear models evaluating seagrass decline as (a) percent change between years and (b) probability of decline versus number of days when salinity was below the threshold.
-Models evaluated the response variable on the y-axes as a function of salinity days and interactions with year category (seagrass recovery prior to 2016, seagrass decline 2016 to present) and bay segment.
-Shaded areas are 95% confidence intervals.
-Colored lines in (b) as rug plots show instances of seagrass increase (blue) or decrease (red) by year.](figs/sgmod.png){#fig-sgmod}
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppmixeff.png){#fig-suppmixeff}
 
@@
 
-<!-- Summary of mixed-effects models evaluating increases in the number of days each year from 1974 to 2022 when temperature or salinity were above or below critical thresholds, respectively.
+Summary of mixed-effects models evaluating increases in the number of days each year from 1975 to 2022 when temperature or salinity were above or below critical thresholds, respectively.
 The slope estimates in the "Both" column indicate trends in the number of days when both temperature and salinity were above or below critical thresholds.
@@
 OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
--->
 
-<!-- ```{r} -->
-<!-- mixtab -->
-<!-- ``` -->
+```{r}
+supp1mixtab
+```
 
-<!-- {{< pagebreak >}} -->
+{{< pagebreak >}}
 
-<!-- Number of days at the beginning and end of the period of record when temperature or salinity were above or below critical thresholds, respectively.
+Summary of mixed-effects models evaluating increases in the number of days each year from 1998 to 2022 when temperature or salinity were above or below critical thresholds, respectively.
+The slope estimates in the "Both" column indicate trends in the number of days when both temperature and salinity were above or below critical thresholds.
+\*\* p \< 0.005, \* p \< 0.05, - no model.
+OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
+
+```{r}
+supp2mixtab
+```
+
+{{< pagebreak >}}
+
+Number of days at the beginning and end of the period of record when temperature or salinity were above or below critical thresholds, respectively.
 Values are the estimated mean number of days (plus standard error) from 1975 to 2022.
@@
 OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
--->
 
-<!-- ```{r} -->
-<!-- daytab -->
-<!-- ``` -->
+```{r}
+supp1daytab
+```
 
@@
 
-## References {.unnumbered}
+Number of days at the beginning and end of the period of record when temperature or salinity were above or below critical thresholds, respectively.
+Values are the estimated mean number of days (plus standard error) from 1998 to 2022.
+Columns for "Both" indicate the estimated number of days when both temperature and salinity were above or below critical thresholds.
+Note that not all models are significant (see table 1).
+\*\* p \< 0.005, \* p \< 0.05, - no model.
+OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
+
+```{r}
+supp2daytab
+```

### 2023-09-20 R/dat_proc.R

@@ -15,6 +15,7 @@ library(rnoaa)
 library(dataRetrieval)
 
 source(here('R/funcs.R'))
+yrsel <- c(1998, 2022)
 
 # SPEI, monthly rain, and monthly air temp ----------------------------------------------------
 
@@ -618,7 +619,7 @@ mixmods <- thrdat %>%
       if(inherits(mod, 'try-error'))
         return(NA)
       
-      effs <- estimate_means(mod, 'yr')
+      effs <- estimate_means(mod, at = 'yr = c(1976, 2022)')
       
       summod <- summary(mod)$coefficients
       
@@ -639,6 +640,46 @@ mixmods <- thrdat %>%
   ungroup()
 
 save(mixmods, file = here('data/mixmods.RData'), compress = 'xz')
+
+# supp mixef mods of threshold counts over time ----------------------------------------------------
+
+load(file = here("data/thrdat.RData"))
+
+# create mixed effects models of cnt ~ yr with station as random intercept
+suppmixmods <- thrdat %>% 
+  group_by(bay_segment, tempthr, salithr, thrtyp) %>% 
+  nest() %>% 
+  summarise(
+    mod = purrr::map(data, function(x){
+
+      tomod <- x %>% 
+        filter(yr >= yrsel[1] & yr <= yrsel[2])
+      
+      mod <- try(lmer(cnt ~ yr + (1|station), data = tomod, REML = F), silent = T)
+      if(inherits(mod, 'try-error'))
+        return(NA)
+      
+      effs <- estimate_means(mod, at = 'yr = c(1998, 2022)')
+    
+      summod <- summary(mod)$coefficients
+      
+      out <- tibble(
+        slo = summod['yr', 'Estimate'],
+        pvl = summod['yr', 'Pr(>|t|)'], 
+        yrstr = effs$Mean[1],
+        yrstrse = effs$SE[1], 
+        yrend = effs$Mean[nrow(effs)],
+        yrendse = effs$SE[nrow(effs)]
+      )
+      
+      return(out)
+      
+    })
+  ) %>% 
+  unnest('mod') %>% 
+  ungroup()
+
+save(suppmixmods, file = here('data/suppmixmods.RData'), compress = 'xz')
 
 # mixef mod predictions for select thresolds --------------------------------------------------
 
@@ -718,6 +759,86 @@ mixmodprds <- thrdat %>%
   )
 
 save(mixmodprds, file = here('data/mixmodprds.RData'))
+
[…]

### 2023-09-20 R/figs.R

@@ -156,7 +156,7 @@ p1 <- ggplot(toplo1, aes(x = yr, y = tavg_c)) +
   geom_smooth(method = 'lm', se = F, formula = y~x, color = 'red2') +
   thm +
   theme(
-    axis.text.x = element_text(size = 12)
+    axis.text.x = element_blank()
   ) +
   labs(
     x = NULL, 
@@ -193,9 +193,9 @@ p3 <- ggplot(hydrodat, aes(x = year, y = hy_load)) +
 p4 <- ggplot(speidat, aes(x = date, y = spi, fill = spisign)) + 
   geom_col(width = 35) +
   scale_fill_manual(values = c('grey70', 'grey30'), guide = 'none') + 
-  thm + 
+  thm  +
   theme(
-    axis.text.x = element_blank()
+    axis.text.x = element_text(size = 12)
   ) +
   coord_cartesian(xlim = range(speidat$date)) +
   labs(
@@ -205,7 +205,7 @@ p4 <- ggplot(speidat, aes(x = date, y = spi, fill = spisign)) +
 
 toplo <- epcdata %>% 
   select(bay_segment, epchc_station, SampleTime, yr, matches('Top|Bottom')) %>% 
-  filter(yr >= yrsel[1] & yr <= yrsel[2]) %>% 
+  filter(yr < 2023) %>% 
   pivot_longer(names_to = 'var', values_to = 'val', matches('Top|Bottom')) %>% 
   mutate(
     var = factor(var, 
@@ -276,7 +276,372 @@ png(here('figs/meteowqraw.png'), height = 8.5, width = 7, family = 'serif', unit
 print(p)
 dev.off()
 
+# supp meteorological, salinity, water temp obs data ------------------------------------------
+
+load(file = here('data/speidat.RData'))
+load(file = url("https://github.com/tbep-tech/load-estimates/raw/main/data/totanndat.RData"))
+
+# hydro load, otb, mtb, ltb, hb
+hydrodat <- totanndat %>% 
+  select(year, bay_segment, hy_load) %>% 
+  filter(bay_segment %in% c('Hillsborough Bay', 'Old Tampa Bay', 'Lower Tampa Bay', 'Middle Tampa Bay')) %>% 
+  summarise(
+    hy_load = sum(hy_load) / 1e3, 
+    .by = 'year'
+  ) %>% 
+  filter(year >= yrsel[1] & year <= yrsel[2])
+
[…]

### 2023-09-20 manu-draft.qmd
main text table for mixeff slopes and day predictions

@@
 
-# load(file = here('tabs/mixtab.RData'))
-# load(file = here('tabs/daytab.RData'))
+load(file = here('tabs/mixdaytab.RData'))
 ```
@@
 
-<!-- Summary of mixed-effects models evaluating increases in the number of days each year from 1974 to 2022 when temperature or salinity were above or below critical thresholds, respectively.
-The slope estimates in the "Both" column indicate trends in the number of days when both temperature and salinity were above or below critical thresholds.
+Summary of mixed-effects models evaluating increases in the number of days each year from 1975 to 2022 when temperature was above 30C, salinity was below 25ppt, or both temperature and salinity were above/below the thresholds.
+The start and end columns show the estimated number of days at the beginning and end of the period of record when temperature or salinity were above or below the thresholds.
+Values are the estimated mean number of days (plus standard error) from 1975 and 2022.
 \*\* p \< 0.005, \* p \< 0.05, - no model.
 OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
--->
 
-<!-- ```{r} -->
-<!-- mixtab -->
-<!-- ``` -->
-
-<!-- {{< pagebreak >}} -->
-
-<!-- Number of days at the beginning and end of the period of record when temperature or salinity were above or below critical thresholds, respectively.
-Values are the estimated mean number of days (plus standard error) from 1975 to 2022.
-Columns for "Both" indicate the estimated number of days when both temperature and salinity were above or below critical thresholds.
-Note that not all models are significant (see table 1).
-\*\* p \< 0.005, \* p \< 0.05, - no model.
-OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
--->
-
-<!-- ```{r} -->
-<!-- daytab -->
-<!-- ``` -->
+```{r}
+mixdaytab
+```
 

### 2023-09-20 R/tabs.R

@@ -6,6 +6,48 @@ library(broom)
 
 source(here('R/funcs.R'))
 
+# mixef mod summary table and start/end -------------------------------------------------------
+
+load(file = here('data/mixmods.RData'))
+
+salthr <- '25'
+tmpthr <- '30'
+
+totab <- mixmods %>% 
+  filter(salithr == paste0('sali_', salthr)) %>% 
+  filter(tempthr == paste0('temp_', tmpthr)) %>%  
+  select(-tempthr, -salithr) %>% 
+  mutate(
+    pvl = p_ast(pvl), 
+    slo = as.character(round(slo, 2)), 
+    yrstr = round(yrstr, 0), 
+    yrstrse = paste0('(', round(yrstrse, 1), ')'),
+    yrend = round(yrend, 0), 
+    yrendse = paste0('(', round(yrendse, 1), ')'),
+    thrtyp = factor(thrtyp, levels = c('tempcnt', 'salicnt', 'bothcnt'), 
+                    labels = c(paste('Temperature >', tmpthr), paste('Salinity <', salthr), 'Both')),
+    bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB'))
+  ) %>% 
+  unite('slo', slo, pvl, sep = '') %>% 
+  unite('yrstr', yrstr, yrstrse, sep = ' ') %>%
+  unite('yrend', yrend, yrendse, sep = ' ') %>% 
+  mutate(
+    slo = ifelse(slo == 'NANA', '-', slo),
+    yrstr = ifelse(grepl('NA', yrstr), '-', yrstr),
+    yrend = ifelse(grepl('NA', yrend), '-', yrend)
+  ) %>%
+  arrange(bay_segment, thrtyp)
+
+mixdaytab <- totab %>% 
+  as_grouped_data(groups = 'bay_segment') %>% 
+  flextable() %>% 
+  set_header_labels(i = 1, values = c('Bay Segment', 'Threshold', 'Slope', 'Start', 'End')) %>% 
+  padding(padding = 0, part = 'all') %>% 
+  width(j = 2, width = 1.5) %>% 
+  font(part = 'all', fontname = 'Times New Roman')
+
+save(mixdaytab, file = here('tabs/mixdaytab.RData'))
+
 # supp1 mixef mod summary table ----------------------------------------------------------------
 
 load(file = here('data/mixmods.RData'))

### 2023-09-20 manu-draft.qmd
hillsborough flow temp plot

@@
 
+![Relationship between water temperature and discharge in the Hillsborough River (USGS gage 02303000) over time.
+The relationship is shown for February (dry month) and August (wet month) grouped by historic (1975-1999) and contemporary (2000-2022) time periods.
+Regression lines with 95% confidence intervals for each temporal grouping are shown.](figs/flowtemp.png){#fig-flowtemp}
+
+{{< pagebreak >}}
+
 ## Tables {.unnumbered}

### 2023-09-20 R/figs.R

@@ -1270,18 +1270,36 @@ toplo <- gagedat %>%
   pivot_wider(names_from = 'var', values_from = 'val') %>%
   na.omit() %>%
   mutate(
-    yrgroup = ifelse(year < 2000, 'past', 'present'),
-    mo = month(date)
+    yrgroup = ifelse(year < 2000, '1975 - 1999', '2000 - 2022'),
+    mo = month(date, label = T, abbr = F)
   ) %>%
   # filter(flow_m3d < quantile(flow_m3d, 0.9)) %>%
-  filter(mo %in% c(2, 8)) #%>%
+  filter(mo %in% c('February', 'August')) #%>%
 # filter(temp_c > 1)
 
-ggplot(toplo, aes(x = flow_m3d, y = temp_c, group = yrgroup, color = yrgroup)) +
-  geom_point(size = 0.6) +
+p <- ggplot(toplo, aes(x = flow_m3d / 1e6, y = temp_c, group = yrgroup, fill = yrgroup, color = yrgroup)) +
+  geom_point(size = 0.6, show.legend = F) +
   scale_x_log10() +
   facet_grid(~mo) +
-  geom_smooth(se = T, method = 'lm', formula = y~ x)
+  theme_minimal() + 
+  scale_color_manual(values = c('tomato1', 'darkred')) +
+  scale_fill_manual(values = c('tomato1', 'darkred')) +
+  theme(
+    legend.position = 'top', 
+    panel.grid.minor = element_blank(), 
+    strip.text = element_text(size = 11)
+  ) +
+  geom_smooth(se = T, method = 'lm', formula = y ~ x) +
+  labs(
+    x = expression(paste('Hillsborough River discharge (', 10^6~m^3, ' / yr)')), 
+    y = 'Water temp. (\u00B0 C)',
+    color = 'Time period',
+    fill = 'Time period'
+  )
+
+png(here('figs/flowtemp.png'), height = 4, width = 6, family = 'serif', units = 'in', res = 600)
+print(p)
+dev.off()
 
 # # ports data threshold counts -----------------------------------------------------------------
 # 

### 2023-10-20 supplement.qmd
added supp fig for fim sal and temp trends

@@
 
+![Water temperature and salinity trends from 1996 to 2021 for stations from the Fisheries Independent Monitoring program.
+Points show the average (95% confidence interval) across all stations and sampling months for each year in each bay segment.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppfimtrnds.png){#fig-suppfimtrnds}
+
+{{< pagebreak >}}
+
 ![Trends from 1998 to 2022 for temperature and salinity measurements at long-term monitoring stations in Tampa Bay.
@@
 Bay segment outlines are shown in (a); OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppkendall.png){#fig-suppkendall}
+
 {{< pagebreak >}}

### 2023-10-20 R/figs.R

@@ -856,6 +856,59 @@ png(here('figs/suppkendall.png'), height = 7, width = 9.5, family = 'serif', uni
 print(p)
 dev.off()
 
+# supp fim temp, sal trends -------------------------------------------------------------------
+
+load(file = here('data/fimsgtempdat.RData'))
+
+toplo <- fimsgtempdat %>% 
+  select(date, temp, sal, bay_segment) %>% 
+  mutate(
+    yr = year(date), 
+    mo = month(date)
+  ) %>% 
+  pivot_longer(temp:sal) %>% 
+  summarise(
+    avev = mean(value, na.rm = T), 
+    lov = t.test(value)$conf.int[1], 
+    hiv = t.test(value)$conf.int[2], 
+    cnt = n(),
+    .by = c(bay_segment, yr, name)
+  ) %>% 
+  mutate(
+    bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB')),
+    name = factor(name, levels = c('temp', 'sal'), labels = c('Water temp. (\u00B0C)', 'Salinity (ppt)'))
+  )
+
+thm <- theme_minimal() + 
+  theme(
+    strip.placement = 'outside', 
+    panel.grid.minor = element_blank(), 
+    axis.text.y = element_text(size = 11), 
+    legend.text = element_text(size= 12), 
+    axis.text.x = element_text(size = 7)
+  )
+
+p <- ggplot(toplo, aes(x = yr, y = avev)) + 
+  geom_linerange(aes(ymin = lov, ymax = hiv), show.legend = F, alpha = 0.7, color = 'steelblue4') + 
+  geom_point(size = 0.75, color = 'steelblue4') +
+  # scale_x_continuous(breaks = seq(min(toplo$yr), max(toplo$yr), by = 3)) +
+  geom_smooth(method = 'lm', formula = y ~ x, se = F, color = 'steelblue4') +
+  facet_grid(name ~ bay_segment, switch = 'y', scales = 'free_y') +
+  thm +
+  theme(
+    strip.text = element_text(size = 12)
+  ) +
+  labs(
+    x = NULL, 
+    y = NULL,
+    color = NULL, 
+    shape = NULL
+  )
+
+png(here('figs/suppfimtrnds.png'), height = 3.5, width = 7, family = 'serif', units = 'in', res = 500)
+print(p)
+dev.off()
+
 # mixeff example plot -------------------------------------------------------------------------
 
 load(file = here('data/mixmodprds.RData'))

### 2023-10-23 manu-draft.qmd
adding study area text

@@
 
+-   Importance of seagrass as foundation species
+-   Tampa Bay context
+-   Climate change stressors
+-   Goals/objectives
+    -   Describe
+
 Background: SST trends in Tampa Bay w/ EPC data, WFS, and deep GOM [@nickerson2023] showed that 1975 to 2022 trend for EPC was notable, but less so from 1998 to 2022.
@@
 
+### Study area
+
+Tampa Bay is the largest open-water estuary in Florida covering 400 mi$^2$ (1,036 km$^2$) and one of the largest in the Gulf of Mexico.
+The watershed covers an 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee Rivers contributing a majority of freshwater inflow to the bay.
+Tampa Bay straddles the temperate and tropical boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer months.
+The watershed is heavily developed and includes over 3 million people with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023].
+The geology of the watershed is rich in phosphates and mining activities have greatly altered the landscape, with notable spills and releases of wastewater that have affected water quality and biological resources [@garrett2011; @beck2022].
+Important subtidal habitats include seagrasses, hard bottom, tidal flats, and oyster reefs, where a majority of management effort has focused on restoring and maintaining seagrass cover [@sherwood2017].
+Additional native habitats include intertidal wetlands (mangroves, salt marshes, salt barriers) and pine forests, oak hammocks, and freshwater wetlands present in upland habitats [@Robison20].
+Losses of native uplands and potentially restorable habitats to development in the watershed from 1990 to 2020 have been estimated at 295 mi$^2$ [763 km$^2$, @beck2023].
+
+The Bay is divided into distinct sub-segments defined by physical and natural boundaries to assist with water quality management activities [@lewis1985]: Old Tampa Bay (OTB) in the northwest; Hillsborough Bay (HB) in the northeast; Middle Tampa Bay (MTB); and Lower Tampa Bay (LTB) that connects to the Gulf of Mexico.
+Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality conditions primarily from nutrient inputs from wastewater and stormwater [@greening2014].
+Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia Rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small tributaries [@tbep0623].
+Notably, Old Tampa Bay has restricted circulation from multiple land bridges that traverse the bay [@sherwood2015; @luther2022].
+Recurring seasonal harmful algal blooms of the dinoflagellete *Pyrodinium bahamense* have contributed to exceedances of the chlorophyll-a regulatory standard in Old Tampa Bay [@lopez2023].
+By comparison, water quality conditions in Middle Tampa Bay and Lower Tampa Bay are generally better than the upper two bay segments primarily from more frequent water exchanges with the Gulf of Mexico and lower nutrient loading [@tbep0623].
+All bay segments are relatively shallow, with a baywide mean depth of approximately 3 m.
+Light penetration typically reaches bottom habitats under current conditions, although seagrasses were historically limited by high phytoplankton production that affected light environments prior to wastewater regulation, particularly in Old Tampa Bay and Hillsborough Bay [@greening2014; @johansson15].
+
+### Seagrass monitoring and change
+
+The long-term recovery of seagrass habitats in Tampa Bay is a nationally-recognized success story that demonstrates a successful management paradigm.
+From 1988 to 2016, seagrasses increased 79% (18,376) to a total of acres ()
+
 Seagrass changes over time in @fig-seagrasschg.
+
+### Water quality data
+
+EPC, FIM, Pinellas
+
+### Trend analysis
+
+### Links to seagrass
 
@@
 
-### Models
-
 All models excluded Lower Tampa Bay because of minimal seagrass change over time. Evaluated total fo, percent change between years, and
@@
 Viral/disease effects on seagrass change, [@vanbogaert2018] for TB and [@duffin2021] for Florida Bay.
+
+Alternative nutrient pathways and role of macroalgae, competition for resources, shading, etc.
 

### 2023-10-24 manu-draft.qmd
added text about sg trends

@@
 Tampa Bay is the largest open-water estuary in Florida covering 400 mi$^2$ (1,036 km$^2$) and one of the largest in the Gulf of Mexico.
-The watershed covers an 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee Rivers contributing a majority of freshwater inflow to the bay.
+The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee Rivers contributing a majority of freshwater inflow to the bay.
 Tampa Bay straddles the temperate and tropical boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer months.
@@
 Additional native habitats include intertidal wetlands (mangroves, salt marshes, salt barriers) and pine forests, oak hammocks, and freshwater wetlands present in upland habitats [@Robison20].
-Losses of native uplands and potentially restorable habitats to development in the watershed from 1990 to 2020 have been estimated at 295 mi$^2$ [763 km$^2$, @beck2023].
+Losses of native uplands and potentially restorable habitats to development in the watershed from 1990 to 2020 have been estimated at 188,429 acres [76,254 ha, @beck2023].
 
-The Bay is divided into distinct sub-segments defined by physical and natural boundaries to assist with water quality management activities [@lewis1985]: Old Tampa Bay (OTB) in the northwest; Hillsborough Bay (HB) in the northeast; Middle Tampa Bay (MTB); and Lower Tampa Bay (LTB) that connects to the Gulf of Mexico.
+Tampa Bay is divided into distinct sub-segments defined by physical and natural boundaries to assist with water quality management activities [@lewis1985]: Old Tampa Bay (OTB) in the northwest; Hillsborough Bay (HB) in the northeast; Middle Tampa Bay (MTB); and Lower Tampa Bay (LTB) that connects to the Gulf of Mexico.
 Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality conditions primarily from nutrient inputs from wastewater and stormwater [@greening2014].
@@
 
-The long-term recovery of seagrass habitats in Tampa Bay is a nationally-recognized success story that demonstrates a successful management paradigm.
-From 1988 to 2016, seagrasses increased 79% (18,376) to a total of acres ()
+The long-term recovery of seagrass habitats in Tampa Bay is a nationally-recognized success story that demonstrates application a successful management paradigm through the National Estuary Program [@greening2006; @greening2014; @sherwood2017].
+From 1988 to 2016, seagrasses increased 79% to a total areal cover of 41,655 acres (16,857 ha), surpassing the goal of restoring coverage to 95% of that which occurred in 1950 (Figure [-@fig-seagrasschg]a).
+Throughout this same period, nitrogen load estimates decreased by about 2/3 from their peak estimated in the mid-1970s as 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration projects in the watershed [@greening2014; @beck2019].
+These reductions in nutrient loadings resulted in largescale reductions in chlorophyll concentrations and light attenuation in the water column, creating favorable environments for seagrass growth.
+The most dramatic improvements in seagrass cover were observed in Old Tampa Bay where coverage increased by 122% or 4,465 acres (2,477 ha) to a total of 11,247 acres (4,511 ha) from 1988 estimates.
+Similar gains were observed in Middle Tampa Bay where seagrass cover increased by 86% or 4,465 acres (1,807 ha) to a total of 9,652 acres (3,906 ha) and Hillsborough Bay where cover increased from nearly zero acres to a total of 2,007 acres (810 ha).
+Seagrasses have generally been stable over time in Lower Tampa Bay.
 
-Seagrass changes over time in @fig-seagrasschg.
+From 2016 to present, dramatic losses of seagrasses have been observed in Tampa Bay, despite water quality conditions remaining relatively stable (Figure \[-@fig-seagrasschg\]a).
+Total cover in Tampa Bay has decreased by 28% from the 2016 peak by 11,518 acres (4,661 ha) to ta total of 30,137 acres (12,196 ha).
+Losses have been most pronounced in Old Tampa Bay (62%; 6,963acres/2,818 ha loss) and Hillsborough Bay (80%; 1,599acres/647 ha loss).
+The current estimate for Old Tampa Bay of 4,183 acres (1,693 ha) is the lowest ever recorded in that bay segment since mapping efforts began in the 1980s.
+Similarly, coverage in Middle Tampa Bay decreased by 20% (1,926 acres/779 ha loss).
+Coverage in Lower Tampa Bay has remained stable, with only a 2% loss that is likely within the mapping error for the coverage estimates.
+
+Two primary sources of data have been used to track seagrass change over time in Tampa Bay. The Southwest Florida Water Management District (SWFWMD)
 

### 2023-10-24 R/dat_proc.R

@@ -1450,7 +1450,8 @@ save(fimtempmod, file = here('data/fimtempmod.RData'))
 
 # all otb samples
 # four sites (polygons) E1:E4, 32 samples per year (eight per site)
-tmp <- read_excel(here('data-raw/pinellas_from_sd.xlsx'))
+tmp <- read_excel(here('data-raw/pinellas_from_sd.xlsx'), 
+                  col_types = 'text')
 
 pincotemp <- tmp %>% 
   select(
@@ -1467,7 +1468,7 @@ pincotemp <- tmp %>%
     SAV_Type
   ) %>%  
   mutate(
-    date = ymd(date), 
+    date = mdy(date), 
     hr = hour(hr), 
     SAV = case_when(
       SAV %in% c('n', 'N') ~ 0, 
@@ -1507,8 +1508,8 @@ sav <- pincotemp %>%
   select(-SAV, -SAV_Type) %>% 
   unique()
 
-
-# pincotemp <-  saltemp %>% 
-#   inner_join(sav, by = c('site', 'sample', 'date', ....))
+# rejoin temp, sal with sav ifno
+pincotemp <-  saltemp %>%
+  inner_join(sav, by = c('site', 'sample', 'date', 'hr', 'lat', 'lon'))
   
 # save(pincotemp, file = here('data/pincotemp.RData'))

### 2023-10-25 manu-draft.qmd
more methods

@@
 
-### Seagrass monitoring and change
+### Seagrass change in Tampa Bay
 
@@
 
-From 2016 to present, dramatic losses of seagrasses have been observed in Tampa Bay, despite water quality conditions remaining relatively stable (Figure \[-@fig-seagrasschg\]a).
+From 2016 to present, dramatic losses of seagrasses have been observed in Tampa Bay, despite water quality conditions remaining relatively stable (Figure [-@fig-seagrasschg]a).
 Total cover in Tampa Bay has decreased by 28% from the 2016 peak by 11,518 acres (4,661 ha) to ta total of 30,137 acres (12,196 ha).
@@
 
-Two primary sources of data have been used to track seagrass change over time in Tampa Bay. The Southwest Florida Water Management District (SWFWMD)
+### Seagrass data
+
+Two primary sources of data have been used to track seagrass change over time in Tampa Bay.
+The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s \[e.g., , @SWFWMD23\].
+These maps are created by photointerpretation of aerial images obtained at the end of the growing season, typically during November-December.
+The TBEP has used these maps to track progress towards achieving seagrass restoration goals as total cover in Tampa Bay.
+No species information is provided.
+A more detailed, but spatially-specific, data source is the Tampa Bay Interagency Seagrass Monitoring Program (Figure \[-@fig-seagrasschg\]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>).
+Annual transect surveys have been conducted since 1998 at 61 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017].
+This dataset provides species information, including abundance, cover, frequency occurrence, and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed.
+Although the areal maps provide the standard for assessment of restoration goals, the transect data allow for inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined.
+As such, the transect data were used below for comparison with temperature and salinity changes for the major bay segments.
+Other sources of seagrass data are described in the next section.
 
@@
 
-EPC, FIM, Pinellas
+Several datasets with distinct sample designs are available to assess long-term changes in water temperature and salinity in Tampa Bay.
+These datasets were evaluated individually to assess trends and relationships with seagrass change to provide a weight-of-evidence approach for potential causal relationships driving the recent decline.
+First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at 45 stations in the major bay segments since the late 1970s.
+These data provide the basis for regulatory assessments and compliance reporting for the nutrient TMDL in Tampa Bay.
+Water quality samples are collected at each station from surface water grabs (e.g., nutrients) or *in situ* measurements for physical parameters (e.g., salinity, temperature), where the latter includes measurements at the surface, mid-depth, and bottom.
+Time of sampling can vary, although most samples are collected from mid-morning to early afternoon.
+All surface and bottom salinity and temperature measurements for each of the 45 monitoring stations were evaluated herein.
+Trends were assessed for both surface and bottom samples, as described below, whereas only the bottom measurements were used for comparison to seagrass trends.
+The data were obtained using the tbeptools R package that imports the data directly from a stable web address provided by the EPC [@beck2021].
+
+The second dataset used to evaluate water quality trends was available from the Florida Fish and Wildlife Conservation Commission (FWC).
+The Fisheries Independent Monitoring (FIM) program administered by FWC provides monthly surveys of the entire nekton community in Tampa Bay, including species richness and abundance, using multiple survey gear types that target different habitats.
+A stratified sample design is used to target multiple habitats where unique sites are sampled each month.
+We used data from the 21.3 meter center-bag nearshore seine that specifically targets shallow habitats where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present).
+These data are collected by pulling the seine adjacent to the shore to sample approximately 140 m$^2$ of bay bottom [@schrandt2021].
+In addition to collecting fish and selected invertebrates, *in situ* physical measurements for water temperature and salinity are collected at the beginning and end of the seine haul, and typically at the surface and bottom.
+Only measurements at the beginning of the seine haul and from the bottom were used.
+Seagrass data are also provided for each site, with information on species and cover.
+Sites with greater than 50% cover of seagrass were identified as "seagrass" sites and those less than 50% or bare sediment were identified as "no seagrass" sites for comparison with temperature and salinity measurements.
+Sites exclusively with macroalgae were not included in the analysis.
+All FIM data from FWC staff by request.
+
+The third and final dataset evaluated was from the Pinellas County Department of Environmental Management (PCDEM).
+Surface waters in Pinellas county have been monitored since the 1990s, although a consistent stratified random sampling designed has been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES reporting.
+Data were obtained by recquest to PCDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (also available at \<https://wateratlas.usf.edu/\>).
+We focused primarily on OTB for the analysis of the PCDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments.
+Four distinct spatial zones in OTB are used to stratify the random selection of sample points that typically include 4 sample points per month in each zone.
+Water quality samples at each site are similar to those collected by EPC, where only bottom measurements for salinity and temperature were retained.
+Seagrass presence/absence is also recored at each site and all sites were defined as "seagrass" if only seagrass species were identified (any with macroalgae were excluded) and "no seagrass" if bare sediment was observed.
+
+All of the organizations that provided water quality datasets participate in the Southwest Florida Regional Ambient Monitoring Program.
+This *ad hoc* group has routinely met quarterly to ensure similar standards and methods are used for the collection and processing of surface water quality monitoring data.
+Split-samples between the organizations are also compared to assess precision between different laboratories.
+As such, the water quality measurements used herein are considered comparable, relative to the different sampling designs used by each program.
 
 ### Trend analysis
+
+The first goal of this analysis was to describe spatial and temporal trends in water temperature and salinity using the three water quality datasets described above.
+This assessment provided an indication of the extent of change in Tampa Bay as context for understanding potential relationships with seagrass change.
+An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions from long-term climate change.
+Meteorological data describing air temperature and precipitation were obtained from Tampa International Airport where daily measurements have been collected since 1939.
+The *rnoaa* R package [@chamberlain23] was used to retrieve these data starting from 1975 when EPC monitoring began in Tampa Bay.
+Annual averages for air temperature and cumulative annual precipitation were calculated.
+Additionally, the Standardized Precipitation Index (SPI) was estimated from the daily rainfall data to identify periods of time when rainfall significantly deviated from the long-term average (using the spi function from...).
 

### 2023-10-26 manu-draft.qmd
more methods

@@
 
+@lewis1985 review of seagrass in Tampa Bay.
+@lirman2003 conducted exposure experiments to evaluate seagrass growth in response to a range of salinity conditions.
+@mcmillan1967 discusses growth of halodule, syringodium, thalassia, and ruppia in response to salinity increases (up to 75 psu), no info on lower limit.
+Cites @phillips1960 for a salinity range of Thalassia in Florida of 33 - 38 psu.
+@zieman1975 discusses seasonal variation of thalassia relative to temp and salinity
+
 Focus on two time periods: full record and recent decline
@@
 Two primary sources of data have been used to track seagrass change over time in Tampa Bay.
-The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s \[e.g., , @SWFWMD23\].
+The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s \[e.g., @SWFWMD23\].
 These maps are created by photointerpretation of aerial images obtained at the end of the growing season, typically during November-December.
@@
 This *ad hoc* group has routinely met quarterly to ensure similar standards and methods are used for the collection and processing of surface water quality monitoring data.
-Split-samples between the organizations are also compared to assess precision between different laboratories.
+Split-samples evaluated by each organization are also compared to assess precision between different laboratories.
 As such, the water quality measurements used herein are considered comparable, relative to the different sampling designs used by each program.
@@
 
-The first goal of this analysis was to describe spatial and temporal trends in water temperature and salinity using the three water quality datasets described above.
+The first goal of the analysis was to describe spatial and temporal trends in water temperature and salinity using the three water quality datasets described above.
 This assessment provided an indication of the extent of change in Tampa Bay as context for understanding potential relationships with seagrass change.
-An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions from long-term climate change.
+An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions related to long-term (multi-decadal) climate change.
 Meteorological data describing air temperature and precipitation were obtained from Tampa International Airport where daily measurements have been collected since 1939.
@@
 Annual averages for air temperature and cumulative annual precipitation were calculated.
-Additionally, the Standardized Precipitation Index (SPI) was estimated from the daily rainfall data to identify periods of time when rainfall significantly deviated from the long-term average (using the spi function from...).
+Additionally, the Standardized Precipitation Index [SPI, @begueria2013] was estimated from the daily rainfall data to identify periods of time when rainfall significantly deviated from the long-term average [using the *spei* R package, @begueria23].
+Annual hydrologic loading data to Tampa Bay beginning in 1985 were also obtained for comparison to annual precipitation [@tbep0623].
+All climate and loading data were evaluated annually with simple linear regression trends to assess change over time.
+Water temperature and salinity trends using the EPC, FIM, and PDEM data were similarly evaluated by averaging the monthly data by year for each bay segment.
+
+Formal trend tests were used to assess station-level changes in water temperature and salinity in the EPC data.
+These analyses also provided a detailed spatial assessment of trends because the EPC data is the only dataset of the three where the same sites have been sampled over time.
+Seasonal Kendall trend tests were used to evaluate the monotonic change for temperature and salinity from 1975 to present at each water quality station [@hirsch1982; @millard13].
+The change per year was also evaluated for each parameter based on the slope estimates returned by each test.
+Kendall tests were also used to evaluate changes over time for each month to determine when the trends more most pronounced seasonally, e.g., all January estimates across years, all February estimates, etc.
+The percentage of stations in each bay segment with significant increasing temperature or decreasing salinity trends were evaluated for each month.
+All tests evaluated both surface and bottom measurements to assess potential differences by water depth.
 
@@
 
-Observed trends aggregated by year and bay segment across stations and months in @fig-meteowqraw.
+The second goal of the analysis was to evaluate if seagrass changes were linked to long-term changes in water temparerature and salinity, with particular attention on the time periods before and after 2016 (pre/post recovery).
+The conceptual model for evaluating these changes follows simple niche space theory where seagrass growth and reproduction is hypothesized to be greatest within optimal ranges for forcing factors that are present in the environment [@hutchinson1957; @vandermeer1972].
+In the simplest form, this can conceptualized as a bell curve with optimal conditions defined with a range of values (e.g., minimum and maximum temperatures where a species is typically observed), where reduced growth or mortality is observed outside of these ranges.
+Because both water temperature and salinity are evaluated herein, the same model can be conceptualized in two-dimensional space [@fig-concept].
+Seagrass growth can be limited when temperature is below or above the optimum range, when salinity is below or above the optimum, or when both temperature and salinity conditions are outside of the optimum range.
+Using this conceptual model, we can evaluate whether seagrasses are individually stressed by either temperature or salinity, or whether seagrasses are jointly stressed by both temperature and salinity.
+Based on the results of the trend tests, we hypothesized that seagrasses are likely stressed by both high temperature and low salinity.
+The statistical models described below were constructed specifically to assess these hypotheses following the conceptual model in @fig-concept.
+Of course, the optimal niche space can be defined in n-dimensions, but we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have generally not been limiting for growth in recent years.
 
-The conceptual stressor diagram @fig-concept.
+A fundamental challenge in describing niche space is identifying the boundaries within which the space occurs.
+In reality, the boundaries are typically not discrete and organisms can be found outside of these ranges, although at reduced abundances.
+Regardless, discrete limits are often used and substantial efforts have been made to identify these boundaries for seagrasses.
+In Tampa Bay, three dominant seagrass species are observed that include *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinium* (turtle grass) [@lewis1985].
+These species co-occur often in mixed beds, although differences in abundance are observed across salinity ranges.
+Shoal grass is tolerant of a wide range of salinity [@lirman2003], but is more abundant in oligohaline portions of Tampa Bay.
+Conversely, turtle grass is most often found in more euryhaline conditions near the mouth of Tampa Bay.
+Reported salinity ranges for each of these species varies depending on location, season, and other co-occurring factors like temperature [@phillips1960; @mcmillan1967; @zieman1975; @lewis1985], although most studies place lower limits of salinity in the range of 15-25 ppt.
+Optimal temperature ranges are similar between species, with reduced growth observed at temperatures above 30$^\circ$ C [@zieman1975;@lewis1985].
 
-Salinity tolerance: Ruppia \> Halodule\> Thalassia \> Syringodium
+Because of the uncertainty in defining thresholds for optimal temperature and salinity ranges, multiple thresholds were evaluated to both describe the potential for stress and how it may be related to changes in seagrass.
+Distinctions were not made between species, primarily due to lack of consensus between studies and the likely site-specific ranges that affect seagrass growth in Tampa Bay.
+First, we developed metrics of potential temperature and salinity stress by quantifying the maximum number of continuous days each year when temperature was above or salinity was below a given threshold.
+This approach assumed that stress could be observed based on duration of exposure (i.e., maximum number of continuous days each year) relative to a threshold that may are may not be outside of the optimum range for seagrasses.
+These metrics were quantified from the monthly long-term observation in the EPC data.
+To quantify daily counts each year, a continuous prediction of temperature or stress over time was estimated using Generalized Additive Models (GAMs) fit to the response variable and a single predictor for decimal year.
+The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends [@beck2022a].
+Model fit for each station was sufficient to calculate daily predictions for assessing the stressor metrics.
 
-Halodule:
-
--   Temp: 20-30 (range, @lewis1985)
-
--   Salinity: grows well in most salinity ranges [@lirman2003], like Syringodium but more tolerant [@lewis1985], 10-35 (range) or \< 10 (S. Scolaro)
-
-Thalassia:
-
--   Temp: 20-30 (range, @lewis1985, @zieman1975), 30-31 (upper, from SS)
-
--   Salinity: 25 (lower, @lewis1985); 30, lowest at 5 (lower, @lirman2003), 33-38 (range, @phillips1960); 24-35 (range, @zieman1975 citing @phillips1960); 15 (SS)
-
-Syringodium:
-
--   Temp: 20-30 (range, @lewis1985), 29 (upper from SS)
-
--   Salinity: 20 (lower, @lewis1985), 25 (lower, @lirman2003); 20
-
-@lewis1985 review of seagrass in Tampa Bay.
-@lirman2003 conducted exposure experiments to evaluate seagrass growth in response to a range of salinity conditions.
-@mcmillan1967 discusses growth of halodule, syringodium, thalassia, and ruppia in response to salinity increases (up to 75 psu), no info on lower limit.
-Cites @phillips1960 for a salinity range of Thalassia in Florida of 33 - 38 psu.
-@zieman1975 discusses seasonal variation of thalassia relative to temp and salinity
-
--   Get FIM, Pinellas, Manatee data
+Counts of the maximum number of days each year that temperature was above or salinity was below a thresold were obtained from the GAMs.
+This was done at each of 45 stations in the EPC data using temperature thresholds of 29, 30, and 31$^\circ$ C and salinity thresholds of 15, 20, and 25 ppt.
 
@@
 Viral/disease effects on seagrass change, [@vanbogaert2018] for TB and [@duffin2021] for Florida Bay.
+
+Role of additional stressors and why they weren't considered in the model - they're important but we can also demonstrate that light attenuation has been relatively stable over time, so may not be important to include in the models.
+
+Issues with thresholds, not definitive and those likely to stress seagrass are potentially not occurring at frequencies sufficient enough to model.
+Those used herein were chosen statistically.
+Point is that the temp and sal trends are in a direction such that thresholds that are more likely to stress seagrass are more likely to occur in the future.
+
+Case for continuous monitoring
+
+Are we just showing seagrasses are not present because of temperature limitations or if temperature indeed caused a reduction?
+
+Manamagement implication - we can't do much to control temp and salinity, but we can temper expectations for how to manage the bay.
+Adaptive capacity/resiliency is likely decreasing because of these "co-morbidities", so we have to temper expectations for other stressors that we can control.
+
+OTB is likely at a tipping point...
 

### 2023-10-26 manu-draft.qmd
add supplement gam fit stats

@@
 library(flextable)
+library(dplyr)
+library(tibble)
 
+load(file = here('data/gamfit.RData'))
 load(file = here('tabs/mixdaytab.RData'))
@@
 
+```{r}
+# r2 range for gam daily  models
+fitest <- gamfit %>% 
+  summarise(
+    minr2 = min(R2), 
+    maxr2 = max(R2), 
+    .by = parameter
+  ) %>% 
+  split(.$parameter)
+```
+
 The second goal of the analysis was to evaluate if seagrass changes were linked to long-term changes in water temparerature and salinity, with particular attention on the time periods before and after 2016 (pre/post recovery).
@@
 The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends [@beck2022a].
-Model fit for each station was sufficient to calculate daily predictions for assessing the stressor metrics.
+Model fit for each station was sufficient to calculate daily predictions for assessing the stressor metrics (see supplement, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models).
 

### 2023-10-26 supplement.qmd
add supplement gam fit stats

@@
 
+load(file = here('tabs/supptempfittab.RData'))
+load(file = here('tabs/suppsalifittab.RData'))
 load(file = here('tabs/supp1mixtab.RData'))
@@
 ## Tables {.unnumbered}
+
+```{r}
+#| tbl-cap: "Summary statistics for Generalized Additive Model fits to bottom water temperature data at all stations in Tampa Bay monitored by the Environmental Protection Commission of Hillsborough County.
+Models were fit on monthly data from 1975 to 2022 using decimal year as the single predictor.
+Fit statistics are Akaike Information Criterion (AIC), Generalized Cross-Validation (GCV) scores, and R-squared values of observed to predicted.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay."
+#| label: tbl-supptempfittab
+supptempfittab
+```
+
+{{< pagebreak >}}
+
+```{r}
+#| tbl-cap: "Summary statistics for Generalized Additive Model fits to bottom salinity data at all stations in Tampa Bay monitored by the Environmental Protection Commission of Hillsborough County.
+Models were fit on monthly data from 1975 to 2022 using decimal year as the single predictor.
+Fit statistics are Akaike Information Criterion (AIC), Generalized Cross-Validation (GCV) scores, and R-squared values of observed to predicted.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay."
+#| label: tbl-suppsalifittab
+suppsalifittab
+```
+
+{{< pagebreak >}}
 

### 2023-10-26 R/dat_proc.R

@@ -326,6 +326,45 @@ chlaprd <- chlaprd %>%
 
 save(chlaprd, file = here('data/chlaprd.RData'))
 
+
+# gam performance, temp/sal only --------------------------------------------------------------
+
+fls <- list.files(here('data'), 'salibot|tempbot')
+
+gamfitnest <- fls %>% 
+  tibble(fl = .) %>% 
+  group_nest(fl) %>% 
+  mutate(
+    data = purrr::pmap(list(fl, data), function(fl, data){
+      
+      cat(fl, '\n')
+      
+      obj <- gsub('\\.RData', '', fl)
+      load(file = paste0('data/', fl))
+      mod <- get(obj)
+      anlz_fit(mod)
+      
+    })
+  )
+
+gamfit <- gamfitnest %>% 
+  unnest('data') %>% 
+  separate(fl, into = c('bay_segment', 'epchc_station', 'parameter'), sep = '_', convert = T) %>% 
+  left_join(stations, by = c('bay_segment', 'epchc_station')) %>% 
+  mutate(
+    bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB')), 
+    parameter = factor(parameter, levels = c('tempbot.RData', 'salibot.RData'), 
+                       labels = c('temp', 'sal')
+    ), 
+    AIC = round(AIC, 0)
+  ) %>% 
+  mutate_at(vars(GCV, R2), round, 2) %>% 
+  mutate_at(vars(Latitude, Longitude), round, 3) %>% 
+  arrange(parameter, bay_segment, epchc_station) %>% 
+  select(parameter, bay_segment, epchc_station, Longitude, Latitude, everything()) 
+
+save(gamfit, file = here('data/gamfit.RData'))
+
 # count of days per year per station above/below thresholds -----------------------------------
 
 # thresholds to count by year, temp is above, salinity is below

### 2023-10-26 R/tabs.R

@@ -48,6 +48,40 @@ mixdaytab <- totab %>%
 
 save(mixdaytab, file = here('tabs/mixdaytab.RData'))
 
+# supp gam performance, temp/sal only ---------------------------------------------------------
+
+load(file = here('data/gamfit.RData'))
+
+fittab <- gamfit %>% 
+  mutate(
+    bay_segment = as.character(bay_segment),
+    bay_segment = ifelse(duplicated(bay_segment), '', bay_segment), 
+    .by = 'parameter'
+  ) %>% 
+  group_nest(parameter) %>% 
+  mutate(
+    data = purrr::map(data, function(x){
+      
+      x %>% 
+        knitr::kable(
+          col.names = c('Bay Segment', 'Station', 'Lon', 'Lat', 'AIC', 'GCV', 'R$^2$')
+        ) 
+      
+    })
+  )
+
+supptempfittab <- fittab %>% 
+  filter(parameter == 'temp') %>% 
+  pull('data') %>% 
+  .[[1]]
+suppsalifittab <- fittab %>% 
+  filter(parameter == 'sal') %>% 
+  pull('data') %>% 
+  .[[1]]
+
+save(supptempfittab, file = here('tabs/supptempfittab.RData'))
+save(suppsalifittab, file = here('tabs/suppsalifittab.RData'))
+
 # supp1 mixef mod summary table ----------------------------------------------------------------
 
 load(file = here('data/mixmods.RData'))

### 2023-10-27 manu-draft.qmd
add gam example figure to supp

@@
 Also noted the trend was most pronounced in the winter, see Fig 3.
+
+Good CC references for FL: https://www.mdpi.com/2071-1050/15/14/11364
 
@@
 Two primary sources of data have been used to track seagrass change over time in Tampa Bay.
-The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s \[e.g., @SWFWMD23\].
+The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s [e.g., @SWFWMD23].
 These maps are created by photointerpretation of aerial images obtained at the end of the growing season, typically during November-December.
@@
 No species information is provided.
-A more detailed, but spatially-specific, data source is the Tampa Bay Interagency Seagrass Monitoring Program (Figure \[-@fig-seagrasschg\]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>).
+A more detailed, but spatially-specific, data source is the Tampa Bay Interagency Seagrass Monitoring Program (Figure [-@fig-seagrasschg]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>).
 Annual transect surveys have been conducted since 1998 at 61 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017].
@@
 Reported salinity ranges for each of these species varies depending on location, season, and other co-occurring factors like temperature [@phillips1960; @mcmillan1967; @zieman1975; @lewis1985], although most studies place lower limits of salinity in the range of 15-25 ppt.
-Optimal temperature ranges are similar between species, with reduced growth observed at temperatures above 30$^\circ$ C [@zieman1975;@lewis1985].
+Optimal temperature ranges are similar between species, with reduced growth observed at temperatures above 30$^\circ$ C [@zieman1975; @lewis1985].
 

### 2023-10-27 supplement.qmd
add gam example figure to supp

@@
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppmeteowqraw.png){#fig-suppmeteowqraw}
+
+{{< pagebreak >}}
+
+![Examples of Generalized Additive Models of bottom water temperature and salinity for station 66 in Old Tampa Bay.
+Models were fit using a continuous smoothing spline for decimal year.
+Monthly observations are shown as points and daily predictions from the models as lines.](figs/suppgamex.png){#fig-supgamex}
 

### 2023-10-27 R/figs.R

@@ -443,6 +443,29 @@ png(here('figs/suppmeteowqraw.png'), height = 8.5, width = 7, family = 'serif', 
 print(p)
 dev.off()
 
+# supp gam temp, sali example -----------------------------------------------------------------
+
+fl <- 'OTB_66_tempbot'
+load(file = here(paste0('data/', fl, '.RData')))
+
+toplo <- get(fl)
+
+p1 <- show_prdseries(toplo, ylab = 'Bottom temperature (\u00B0C)') + 
+  labs(subtitle ='Station 66, OTB')
+
+fl <- 'OTB_66_salibot'
+load(file = here(paste0('data/', fl, '.RData')))
+
+toplo <- get(fl)
+
+p2 <- show_prdseries(toplo, ylab = 'Bottom salinity (ppt)')
+
+p <- p1 + p2 + plot_layout(ncol = 1)
+
+png(here('figs/suppgamex.png'), height = 5, width = 6, family = 'serif', units = 'in', res = 300)
+print(p)
+dev.off()
+
 # kendall -------------------------------------------------------------------------------------
 
 leglab <- expression(paste(yr^{-1}))

### 2023-10-27 manu-draft.qmd
methods draft done

@@
 This approach assumed that stress could be observed based on duration of exposure (i.e., maximum number of continuous days each year) relative to a threshold that may are may not be outside of the optimum range for seagrasses.
-These metrics were quantified from the monthly long-term observation in the EPC data.
-To quantify daily counts each year, a continuous prediction of temperature or stress over time was estimated using Generalized Additive Models (GAMs) fit to the response variable and a single predictor for decimal year.
+These metrics were quantified from the monthly long-term observations in the EPC data.
+To quantify daily counts each year, a continuous prediction of temperature and salinity over time at each of 45 stations was estimated using Generalized Additive Models (GAMs) fit to the response variable and a single predictor for decimal year [@wood17].
 The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends [@beck2022a].
@@
 
-Counts of the maximum number of days each year that temperature was above or salinity was below a thresold were obtained from the GAMs.
+Counts of the maximum number of days each year that temperature was above or salinity was below a thresold were obtained from the daily GAM predictions.
 This was done at each of 45 stations in the EPC data using temperature thresholds of 29, 30, and 31$^\circ$ C and salinity thresholds of 15, 20, and 25 ppt.
+The number of days when both temperature was above and salinity was below the thresholds was also estimated as a combined potential stress measure.
+Stressor metrics were further aggregated across stations in each bay segment using a mixed-effects regression model where the annual stressor counts for stations in a bay segment were fit against year (1975 to 2022) using a random intercept for station [@zuur2009a; @bates15].
+This produced an overall assessment of how the stressor metrics have changed over time by bay segment, including counts for temperature, salinity, and both relative to the thresholds.
 
-All models excluded Lower Tampa Bay because of minimal seagrass change over time. Evaluated total fo, percent change between years, and
+For comparison to seagrass, the annual metrics were also referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year as above.
+Transect surveys are typically done in the fall at the end of the growing season.
+The average date for each year across the subset of transects in each bay segment was estimated and the stressor metrics were referenced accordingly.
+For example, if the average transect date for a bay segment was September 15th in 2009 and Oct 1st in 2010, the 2010 stressor counts were based on all days between the two average dates.
+This was repeated for all years from 1998 to 2022 when transect data were available.
+Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment.
+The stressor metrics were compared to various response measures at the bay segments-scale from the transect data, including frequency occurrence each year (all species), percent change in frequency occurrence between years, and increase/decrease in frequency occurrence between years as a binary (1/0) variable.
+Generalized linear regression models were used to evaluate each of the seagrass response variables relative to the stressor metrics.
+Interaction factors were also included for bay segment and time period (recovery pre-2016, decline post-2016).
+The most parsimonious models were identified by forward and backward variable selection and comparison with AIC values [@sakamoto1986; @venables2002b].
+All models excluded Lower Tampa Bay because of minimal seagrass change over time.
 
-GLMs of percent change each year by Sal, Temp, with year and bay segment interactions.
-Attempted different windows of time from approximate seagrass sample date, 30 days to one year.
-The best model was one that included an effect of Salinity and an interaction with bay segment.
-Hillsborough Bay showed a strong response of increasing likelihood of seagrass loss with number of days when salinity was below the threshold.
-
-Then GAMs were evaluated looking at total by Salinity, with interaction of year and bay segment. Also created a model looking at percent change vs Temp and year/bay segment interaction.
-
-Show plots of both the glm and gams.
+Additional models were constructed for the FIM and PDEM datasets to evaluate seagrass change relative to temperature and salinity.
+These models did not include stressor metrics, but rather used direct measurements of salinity and temperature as independent variables.
+Models of temperature or salinity to predict daily measurements are not possible with the FIM and PDEM monitoring designs because random locations were used for each sample date.
+However, the sample sizes for the FIM and PDEM models were much larger than those for the EPC data, where the latter was limited by the annual timestep of the transect surveys.
+This allowed for more flexible models that are less limited by degrees of freedom.
+GAMs were used to evaluate seagrass percent cover for the FIM data using continuous temperature and salinity as independent variables.
+Interactions with factor variables for bay segment and time period as above were also included.
+The GAMs also evaluated interactions between salinity and temperature to determine if both factors had an additive effect on seagrass change.
+The GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated (no bay segment interaction) and the seagrass response variable was presence/absence (i.e., a binomial distribution was used).
 

### 2023-10-27 manu-draft.qmd
add linear trend tables

@@
 load(file = here('data/gamfit.RData'))
+load(file = here('tabs/temptrndtab.RData'))
+load(file = here('tabs/saltrndtab.RData'))
 load(file = here('tabs/mixdaytab.RData'))
@@
 Water temperature and salinity trends using the EPC, FIM, and PDEM data were similarly evaluated by averaging the monthly data by year for each bay segment.
+Linear trends for these data were evaluated based on averages of all stations and months within each year and bay segment, which allowed for comparable statistical power between datasets with different sample designs.
 
@@
 
-### Links to seagrass
+### Quantifying potential stress
 
@@
 
-The second goal of the analysis was to evaluate if seagrass changes were linked to long-term changes in water temparerature and salinity, with particular attention on the time periods before and after 2016 (pre/post recovery).
-The conceptual model for evaluating these changes follows simple niche space theory where seagrass growth and reproduction is hypothesized to be greatest within optimal ranges for forcing factors that are present in the environment [@hutchinson1957; @vandermeer1972].
-In the simplest form, this can conceptualized as a bell curve with optimal conditions defined with a range of values (e.g., minimum and maximum temperatures where a species is typically observed), where reduced growth or mortality is observed outside of these ranges.
-Because both water temperature and salinity are evaluated herein, the same model can be conceptualized in two-dimensional space [@fig-concept].
+The second goal of the analysis was to evaluate if seagrass changes were linked to long-term changes in water temparerature and salinity, with particular attention on differences between bay segments and the time periods before and after 2016 (pre/post recovery).
+The conceptual model for evaluating these changes describes the niche space where seagrass growth and reproduction is hypothesized to be greatest within optimal ranges for forcing factors that are present in the environment [@hutchinson1957; @vandermeer1972].
+In the simplest form, this can conceptualized as a bell curve with optimal conditions defined with a range of values for a single parameter (e.g., minimum and maximum temperatures where a species is typically observed), where reduced growth or mortality is observed outside of these ranges.
+Because both water temperature and salinity are evaluated, the same model can be conceptualized in two-dimensional space [@fig-concept].
 Seagrass growth can be limited when temperature is below or above the optimum range, when salinity is below or above the optimum, or when both temperature and salinity conditions are outside of the optimum range.
-Using this conceptual model, we can evaluate whether seagrasses are individually stressed by either temperature or salinity, or whether seagrasses are jointly stressed by both temperature and salinity.
-Based on the results of the trend tests, we hypothesized that seagrasses are likely stressed by both high temperature and low salinity.
-The statistical models described below were constructed specifically to assess these hypotheses following the conceptual model in @fig-concept.
-Of course, the optimal niche space can be defined in n-dimensions, but we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have generally not been limiting for growth in recent years.
+Based on the results of the trend tests, we hypothesized that seagrasses are likely stressed by both high temperature and low salinity (bottom right @fig-concept).
+Although the optimal niche space can be defined in multiple dimensions for many parameters, we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have generally not been limiting for growth in recent years.
 
-A fundamental challenge in describing niche space is identifying the boundaries within which the space occurs.
-In reality, the boundaries are typically not discrete and organisms can be found outside of these ranges, although at reduced abundances.
-Regardless, discrete limits are often used and substantial efforts have been made to identify these boundaries for seagrasses.
-In Tampa Bay, three dominant seagrass species are observed that include *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinium* (turtle grass) [@lewis1985].
+A fundamental challenge in describing niche space is identifying the boundaries for optimal conditions.
+In Tampa Bay, the three dominant seagrass species are *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinium* (turtle grass) [@lewis1985].
 These species co-occur often in mixed beds, although differences in abundance are observed across salinity ranges.
 Shoal grass is tolerant of a wide range of salinity [@lirman2003], but is more abundant in oligohaline portions of Tampa Bay.
-Conversely, turtle grass is most often found in more euryhaline conditions near the mouth of Tampa Bay.
+Conversely, turtle grass less tolerant of low salinity and is most often found in more euryhaline conditions near the mouth of Tampa Bay.
 Reported salinity ranges for each of these species varies depending on location, season, and other co-occurring factors like temperature [@phillips1960; @mcmillan1967; @zieman1975; @lewis1985], although most studies place lower limits of salinity in the range of 15-25 ppt.
@@
 This produced an overall assessment of how the stressor metrics have changed over time by bay segment, including counts for temperature, salinity, and both relative to the thresholds.
+
+### Links to seagrass
 
@@
 
+### Temperature and salinity trends
+
 For @fig-meteowqraw, salinity shows much higher inter-annual variability, but lower intra-year variation among stations.
@@
 Kendall test results are shown in @fig-kendall.
+
+Linear trends by dataset
+
+### Stressor metrics
 
@@
 
-NMS results in @fig-nms and seagrass decline results in @fig-sgmod
+NMS results in @fig-nms
+
+### Seagrass response
+
+Seagrass decline results in @fig-sgmod, @fig-sggammod
 
@@
 Points are sized by seagrass frequency occurrence.
-95% confidence ellipses are shown for observations in each major bay segment OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/nms.png){#fig-nms}
+95% confidence ellipses are shown for observations in each major bay segment.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/nms.png){#fig-nms}
 
@@
 ![Predicted results from generalized linear models evaluating seagrass decline as (a) percent change between years and (b) probability of decline versus number of days when salinity was below the threshold.
-Models evaluated the response variable on the y-axes as a function of salinity days and interactions with year category (seagrass recovery prior to 2016, seagrass decline 2016 to present) and bay segment.
+Models evaluated the response variable on the y-axes as a function of salinity days and interactions with time period (seagrass recovery prior to 2016, seagrass decline 2016 to present) and bay segment.
 Shaded areas are 95% confidence intervals.
-Colored lines in (b) as rug plots show instances of seagrass increase (blue) or decrease (red) by year.](figs/sgmod.png){#fig-sgmod}
+Colored lines in (b) as rug plots show instances of seagrass increase (blue) or decrease (red) by year.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/sgmod.png){#fig-sgmod}
 
@@
 
-![Relationship between water temperature and discharge in the Hillsborough River (USGS gage 02303000) over time.
-The relationship is shown for February (dry month) and August (wet month) grouped by historic (1975-1999) and contemporary (2000-2022) time periods.
-Regression lines with 95% confidence intervals for each temporal grouping are shown.](figs/flowtemp.png){#fig-flowtemp}
+![Predicted results from generalized additive models evaluating seagrass percent cover compared to (a) salinity and (b) water temperature.
+Models include interactions with time period (seagrass recovery prior to 2016, seagrass decline 2016 to present) and bay segment.
+Shaded areas are 95% confidence intervals.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/sggammod.png){#fig-sggammod}
 
@@
 ## Tables {.unnumbered}
+
+```{r}
+#| label: tbl-temptrndtab
+#| tbl-cap: "Bottom temperature trends ($^\\circ$C) by bay segments and datasets.
+Start year describes the range of the trend test to the present year (2022).
+The starting value is the estimated temperature at the start year and the end value is the estimated temperature at 2022.
+Total change is the difference between the two.
+Datasets evaluated were from the Environmental Protection Commission of Hillsborough County (EPC), Fisheries Independent Monitoring (FIM), and Pinellas County Department of Environmental Management (PDEM).
+Note that PDEM includes data only for the western portion of Old Tampa Bay from 2004 to present and EPC is the only dataset beginning prior to 1996.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
+** p < 0.005, * p < 0.05"
+temptrndtab
+```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: tbl-saltrndtab
+#| tbl-cap: "Bottom salinity trends (ppt) by bay segments and datasets.
+Start year describes the range of the trend test to the present year (2022).
+The starting value is the estimated salinity at the start year and the end value is the estimated temperature at 2022.
+Total change is the difference between the two.
+Datasets evaluated were from the Environmental Protection Commission of Hillsborough County (EPC), Fisheries Independent Monitoring (FIM), and Pinellas County Department of Environmental Management (PDEM).
+Note that PDEM includes data only for the western portion of Old Tampa Bay from 2004 to present and EPC is the only dataset beginning prior to 1996.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
+** p < 0.005, * p < 0.05"
+saltrndtab
+```
+
+{{< pagebreak >}}
 

### 2023-10-27 supplement.qmd
add linear trend tables

@@
 
-![Examples of Generalized Additive Models of bottom water temperature and salinity for station 66 in Old Tampa Bay.
-Models were fit using a continuous smoothing spline for decimal year.
-Monthly observations are shown as points and daily predictions from the models as lines.](figs/suppgamex.png){#fig-supgamex}
+![Water temperature and salinity trends from 1996 to 2021 for stations from the Fisheries Independent Monitoring program.
+Points show the average (95% confidence interval) across all stations and sampling months for each year in each bay segment.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppfimtrnds.png){#fig-suppfimtrnds}
 
@@
 
-![Water temperature and salinity trends from 1996 to 2021 for stations from the Fisheries Independent Monitoring program.
-Points show the average (95% confidence interval) across all stations and sampling months for each year in each bay segment.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppfimtrnds.png){#fig-suppfimtrnds}
+![Water temperature and salinity trends from 2004 to 2022 for stations from the Pinellas County Department of Environmental Management in Old Tampa Bay.
+Points show the average (95% confidence interval) across all stations and sampling months for each year in each bay segment.](figs/supppincotrnds.png){#fig-supppincotrnds}
 
@@
 Bay segment outlines are shown in (a); OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppkendall.png){#fig-suppkendall}
+
+{{< pagebreak >}}
+
+![Relationship between water temperature and discharge in the Hillsborough River (USGS gage 02303000) over time.
+The relationship is shown for February (dry month) and August (wet month) grouped by historic (1975-1999) and contemporary (2000-2022) time periods.
+Regression lines with 95% confidence intervals for each temporal grouping are shown.](figs/suppflowtemp.png){#fig-suppflowtemp}
+
+{{< pagebreak >}}
+
+![Examples of Generalized Additive Models of bottom water temperature and salinity for station 66 in Old Tampa Bay.
+Models were fit using a continuous smoothing spline for decimal year.
+Monthly observations are shown as points and daily predictions from the models as lines.](figs/suppgamex.png){#fig-supgamex}
 

### 2023-10-27 R/dat_proc.R

@@ -1519,10 +1519,12 @@ pincotemp <- tmp %>%
       T ~ NaN
     ), 
     level = factor(level, levels = c('Surface', 'Middle', 'Bottom')), 
-    levelint = as.numeric(level)
+    levelint = as.numeric(level), 
+    lon = ifelse(sign(as.numeric(lon)) == 1, -1 * as.numeric(lon), as.numeric(lon))
   ) %>% 
   filter(SAV %in% c(0, 1)) %>% 
-  mutate_at(vars(lat, lon, temp, sal), as.numeric)
+  mutate_at(vars(lat, lon, temp, sal), as.numeric) %>% 
+  filter(lat < 35) # one outlier at 72
 
 # get salinity, temperature by lowest level (not always bottom)
 saltemp <- pincotemp %>% 
@@ -1554,6 +1556,11 @@ sav <- pincotemp %>%
 
 # rejoin temp, sal with sav ifno
 pincotemp <-  saltemp %>%
-  inner_join(sav, by = c('site', 'sample', 'date', 'hr', 'lat', 'lon'))
+  inner_join(sav, by = c('site', 'sample', 'date', 'hr', 'lat', 'lon')) %>% 
+  mutate(
+    yr = year(date), 
+    mo = month(date)
+  ) %>% 
+  filter(yr > 2003) # only a few
   
 save(pincotemp, file = here('data/pincotemp.RData'))

### 2023-10-27 R/figs.R

@@ -17,6 +17,7 @@ library(scales)
 library(visreg)
 library(car)
 library(mgcv)
+library(mixmeta)
 
 source(here('R/funcs.R'))
 
@@ -226,10 +227,31 @@ toplo <- epcdata %>%
   filter(!is.na(val)) %>% 
   summarise(
     avev = mean(val, na.rm = T),
+    se = sd(val, na.rm = T)^2,
     lov = t.test(val, na.rm = T)$conf.int[1],
     hiv = t.test(val, na.rm = T)$conf.int[2],
     .by = c('bay_segment', 'yr', 'var', 'loc') 
   )
+
+# # get mixed-effect meta-analysis linear preds
+# mixmets <- toplo %>% 
+#   group_nest(bay_segment, var, loc) %>% 
+#   mutate(
+#     data = purrr::map(data, function(x){
+#       
+#       mod <- mixmeta(avev ~ yr, S = se, random = ~1|yr, data = x, method = 'reml')
+#       
+#       out <- x %>% 
+#         mutate(
+#           prd = predict(mod)
+#         ) %>% 
+#         select(yr, prd)
+#       
+#       return(out)
+#       
+#     }) 
+#   ) %>% 
+#   unnest(data)
 
 wd <- 0.5
 
@@ -241,7 +263,8 @@ p5 <- ggplot(toplo5, aes(x = yr, y = avev, group = loc, color = loc)) +
   geom_linerange(aes(ymin = lov, ymax = hiv), position = position_dodge2(width = wd), show.legend = F, alpha = 0.7) + 
   geom_point(position = position_dodge2(width = wd), size = 0.5) +
   # scale_x_continuous(breaks = seq(min(toplo$yr), max(toplo$yr), by = 3)) +
-  geom_smooth(method = 'lm', formula = y ~ x, se = F) +
+  geom_smooth(method = 'lm', se = F, formula = y ~ x) +
+  # geom_line(data = mixmets %>% filter(var == 'Water temp. (\u00B0C)'), aes(y = prd)) +
   facet_grid(~ bay_segment) +
   scale_color_manual(values = c( 'steelblue4', 'steelblue1')) +
   thm +
@@ -261,7 +284,8 @@ p6 <- ggplot(toplo6, aes(x = yr, y = avev, group = loc, color = loc)) +
   geom_linerange(aes(ymin = lov, ymax = hiv), position = position_dodge2(width = wd), show.legend = F, alpha = 0.7) + 
[…]

### 2023-10-29 manu-draft.qmd
started results text

@@
 
+source(here('R/funcs.R'))
+
+load(file = here('data/speidat.RData'))
 load(file = here('data/gamfit.RData'))
@@
 In Tampa Bay, the three dominant seagrass species are *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinium* (turtle grass) [@lewis1985].
-These species co-occur often in mixed beds, although differences in abundance are observed across salinity ranges.
-Shoal grass is tolerant of a wide range of salinity [@lirman2003], but is more abundant in oligohaline portions of Tampa Bay.
-Conversely, turtle grass less tolerant of low salinity and is most often found in more euryhaline conditions near the mouth of Tampa Bay.
+These species co-occur often in mixed beds throughout the bay, although some differences in abundance are observed across salinity ranges.
+Shoal grass is tolerant of a wide range of salinity [@lirman2003], but is more abundant in oligo/mesohaline portions of Tampa Bay.
+Conversely, turtle grass is less tolerant of low salinity and is most often found in more euryhaline conditions near the mouth of Tampa Bay.
 Reported salinity ranges for each of these species varies depending on location, season, and other co-occurring factors like temperature [@phillips1960; @mcmillan1967; @zieman1975; @lewis1985], although most studies place lower limits of salinity in the range of 15-25 ppt.
@@
 
-For @fig-meteowqraw, salinity shows much higher inter-annual variability, but lower intra-year variation among stations.
-Trends are similar for bottom vs top, temperature shows stronger trend than salinity.
+```{r}
+data(speidat)
+airtemp <- speidat %>% 
+  select(yr, tavg_c) %>% 
+  summarise(temp = mean(tavg_c), .by = yr)
+airtempmod <- lm(temp ~ yr, airtemp)
+airtempslo <- round(coefficients(summary(airtempmod))[2, 1], 2)
+airtemppvl <- p_txt(coefficients(summary(airtempmod))[2, 4])
+airtemprsq <- round(summary(airtempmod)$adj.r.squared, 2)
+airtempprd <- predict(airtempmod, se = T)
+airtempstr <- paste0(round(airtempprd$fit[1], 1), ' (+/-', round(airtempprd$se.fit[1], 2), ' st. err.)')
+airtempend <- paste0(round(airtempprd$fit[nrow(airtemp)], 1), ' (+/-', round(airtempprd$se.fit[nrow(airtemp)], 2), ' st. err.)')
+airtempdif <- round(airtempprd$fit[nrow(airtemp)]- airtempprd$fit[1], 1)
 
-Kendall test results are shown in @fig-kendall.
+prp <- speidat %>% 
+  select(yr, precip_mm) %>% 
+  summarise(prptot = sum(precip_mm), .by = yr)
+prpmod <- lm(prptot ~ yr, prp)
+prpslo <- round(coefficients(summary(prpmod))[2, 1], 2)
+prppvl <- p_txt(coefficients(summary(prpmod))[2, 4])
+prprsq <- round(summary(prpmod)$adj.r.squared, 2)
+prpprd <- predict(prpmod, se = T)
+prpstr <- paste0(round(prpprd$fit[1], 1), ' (+/-', round(prpprd$se.fit[1], 2), ' st. err.)')
+prpend <- paste0(round(prpprd$fit[nrow(prp)], 1), ' (+/-', round(prpprd$se.fit[nrow(prp)], 2), ' st. err.)')
+prpdif <- round(prpprd$fit[nrow(prp)]- prpprd$fit[1], 1)
 
-Linear trends by dataset
+# load(file = url("https://github.com/tbep-tech/load-estimates/raw/main/data/totanndat.RData"))
+# 
+# # hydro load, otb, mtb, ltb, hb
+# hydrodat <- totanndat %>% 
+#   select(year, bay_segment, hy_load) %>% 
+#   filter(bay_segment %in% c('Hillsborough Bay', 'Old Tampa Bay', 'Lower Tampa Bay', 'Middle Tampa Bay')) %>% 
+#   summarise(
+#     hy_load = sum(hy_load) / 1e3, 
+#     .by = 'year'
+#   ) 
+```
+
+Long-term meteorological data from Tampa International Airport showed significantly increasing trends for air temperature an precipitation [@fig-meteowqraw].
+Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`).
+Mean annual air temperature in 1975 was `r airtempstr`, whereas current mean annual air temperature in 2022 was `r airtempend`, showing an overall increase in the period of record of `r airtempdif` $^\circ$.
+Total precipitation has increased by `r prpslo` mm per year (`r prppvl`, $R^2$ `r prprsq`).
+Total precipitation in 1975 was `r prpstr`, whereas current total precipitation in 2022 was `r prpend`, showing an overall increase in the period of record of `r prpdif` mm.
+Trends in total annual hydrologic load from 1985 to present followed total annual precipitation, although the linear model was not significant likely due to a shorter period of record.
+The SPI showed notable anomalies in precipitation, with pronounced rainy periods in 1980, the late 1990s, 2005, and the last ten years.
+
+Increasing water temperature and decreasing salinity generally followed the meteorological trends for all three *in situ* datasets (EPC, FIM, and PDEM, @fig-meteowqraw, @tbl-temptrndtab, @tbl-saltrndtab, figures in supplement).
+The strongest trends were observed for the EPC dataset which had the longest record from 1975 to 2022.
+The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths in Tampa Bay ([@fig-meteowqraw]).
+Trends in water temperature were similar across bay segments with significant increases varying from 0.3 to 0.4 $^\circ$C per year, equating to 1.4 (OTB) to 2.0 (HB) $^\circ$C total increase from 1975 to present (@tbl-temptrndtab).
+Salinity trends were similar between bay segments, although overall salinity was higher for bay segments closer to the Gulf of Mexico.
+Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, totaling an overall decrease of 2.7 and 2.0 ppt from 1975 to present.
+As expected, intra-annual variability of water temperature was much higher than salinity following the seasonal variation in temperature ([@fig-meteowqraw]).
+This variation may also be related to differences among the monitoring stations in each segment.
+Statistically significant increasing temperature trends were also observed for the FIM datasets from 1996 to present in OTB and from 2004 to present in HB (@tbl-temptrndtab).
+No significant temperature or salinity trends were observed in the other bay segments for the FIM and PDEM datasets from 1996 to present or 2004 present (see figures in supplement).
+However, the EPC dataset showed significant increasing temperature across all time periods and bay segments.
+Salinity trends from the EPC dataset were not significant from 1996 to present, but were significantly decreasing from 2004 to present in all bay segments (@tbl-saltrndtab).
+
+The EPC dataset was also used to provide detailed information on station-level trends from 1975 to present (@fig-kendall).
+All stations in both the top and bottom of the water column had significantly increasing temperature and decreasing salinity from 1975 to present, excluding one station in HB that did not have significantly decreasing bottom salinity (Figure [-@fig-kendall]a).
+Seasonally, most stations had more significant trends in the summer, early fall period for both temperature and salinity (Figures [-@fig-kendall]b,c), although some variation was observed between bay segments.
+Temperature trends were more often observed to occur in the summer, early fall for the upper bay segments (OTB, HB), whereas the lower bay segments (MTB, LTB) had more significant trends during the spring.
+Seasonal trends in salinity did not vary as much between segments, although OTB and LTB had more significantly decreasing trends than the other bay segments.
+Again, differences between top or bottom trends did not differ seasonally.
 

### 2023-10-29 R/funcs.R

@@ -11,6 +11,20 @@ p_ast <- function(x){
   
 }
 
+# function for formatting p-values in text
+p_txt <- function(x){
+  
+  sig_cats <- c('p < 0.005', 'p < 0.05', 'ns')
+  sig_vals <- c(-Inf, 0.005, 0.05, Inf)
+  
+  out <- cut(x, breaks = sig_vals, labels = sig_cats, right = FALSE)
+  out <- as.character(out)
+  
+  return(out)
+  
+}
+
+
 # run length encoding for vector of TF, return longest run of T
 runfunc <- function(cnt){
   lngth <- length(na.omit(cnt))

### 2023-10-30 manu-draft.qmd
more results text

@@
 
-Long-term meteorological data from Tampa International Airport showed significantly increasing trends for air temperature an precipitation [@fig-meteowqraw].
+Long-term meteorological data from Tampa International Airport showed significantly increasing trends for air temperature and precipitation [@fig-meteowqraw].
 Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`).
@@
 The strongest trends were observed for the EPC dataset which had the longest record from 1975 to 2022.
-The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths in Tampa Bay ([@fig-meteowqraw]).
+The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths in Tampa Bay (@fig-meteowqraw).
 Trends in water temperature were similar across bay segments with significant increases varying from 0.3 to 0.4 $^\circ$C per year, equating to 1.4 (OTB) to 2.0 (HB) $^\circ$C total increase from 1975 to present (@tbl-temptrndtab).
 Salinity trends were similar between bay segments, although overall salinity was higher for bay segments closer to the Gulf of Mexico.
-Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, totaling an overall decrease of 2.7 and 2.0 ppt from 1975 to present.
-As expected, intra-annual variability of water temperature was much higher than salinity following the seasonal variation in temperature ([@fig-meteowqraw]).
+Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, totaling an overall decrease of 2.7 and 2.0 ppt from 1975 to present.
+As expected, intra-annual variability of water temperature was much higher than salinity following the larger seasonal variation in temperature ([@fig-meteowqraw]).
 This variation may also be related to differences among the monitoring stations in each segment.
-Statistically significant increasing temperature trends were also observed for the FIM datasets from 1996 to present in OTB and from 2004 to present in HB (@tbl-temptrndtab).
+Significant increasing temperature trends were also observed for the FIM dataset from 1996 to present in OTB and from 2004 to present in HB (@tbl-temptrndtab).
 No significant temperature or salinity trends were observed in the other bay segments for the FIM and PDEM datasets from 1996 to present or 2004 present (see figures in supplement).
@@
 
-An example of mixed effects models for two selected thresholds is shown in @fig-mixeff.
+Linear mixed-effects models showed varying trends in the potential stressor metrics for the number of days when temperature was above, salinity was below, or both occurred relative to different thresholds.
+All of the temperature models for each of the three thresholds showed significantly increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$ (see supplement).
+For all bay segments, the estimated slopes for the number of days when temperature was above 30 $^\circ$C were similar among bay segments, varying from 0.9 days per year in LTB to 1.1 days per year in MTB.
+Likewise, the average number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied from 8 (OTB, HB) to 21 (LTB) days in 1975 to 55 (HB) to 63 (LTB) days in 2022.
 
-$$
-Y_{ij} = \beta_0 + \beta_1X_{ij} + u_i + \epsilon_{ij}
-$$
+The salinity models were less similar between bay segments compared to the temperature models (see supplement).
+None of the bay segments had a significantly increasing number of days when salinity was below 15 ppt.
+For MTB, only the number of days when salinity was below 25 ppt was significantly increasing at a rate of 0.67 days per year.
+Both HB and OTB had significantly increasing number of days per year when salinity was below 20 and 25 ppt, although the slopes varied such that the rates in OTB were nearly double those in HB.
+The number of days per year when salinity was below 20 ppt in OTB increased by 1 day per year, whereas the number of days in HB increased by 0.5 days per year from 1975 to present.
+Likewise, the average number of days when salinity was 20 ppt at the beginning and end of the period of record varied from 11 (HB) to 23 (OTB) days in 1975 to 25 (HB) to 68 (LTB) days in 2022.
+
+Times when both temperature was above a threshold and salinity was below a threshold also varied by bay segment.
+None of the models were significant for LTB, four were significant for MTB, seven were significant for HB, and four were significant for OTB.
+The number of days when temperature was above 29 $^\circ$C and salinity was below 25 ppt had the largest slopes of 1.4, 1.2, and 0.7 days per year for OTB, HB, and MTB, respectively.
+Likewise, the average number of days for both varied from the beginning to the end of the period of record from 6, 12, and 1 day(s) in 1975 to 72, 68, and 24 in 2022 for OTB, HB, and MTB respectively.
+
+@fig-mixeff provides visual examples of the mixed-effects models for the estimated number of days over time for each bay segment when temperature was above 30$^\circ$, salinity was below 25 ppt, and when both occurred.
+These thresholds were chosen moving forward because they are slightly more conservative than those with the largest slopes (i.e., higher temperature threshold), while still showing significant increases over time (excluding LTB).
+Temperature trends were similar among segments, whereas the number of days when salinity was below the threshold varied by proximity to the Gulf of Mexico (Table 3).
+The number of days when both temperature was above and salinity was below the threshold generally followed the trends for the number of days when salinity was below the threshold, although the slopes varied.
+These thresholds were used for comparison to seagrass changes described below, based primarily on the statistical strength of the trends and the variance of counts among bay segments.
+That is, more restrictive thresholds did not provide sufficient counts of days per year to develop models and the chosen thresholds were based primarily on statistical considerations rather than ecological considerations, as discussed below.
+
+### Seagrass response
 
 NMS results in @fig-nms
-
-### Seagrass response
 

### 2023-11-03 manu-draft.qmd
minor results edits

@@
 Mean annual air temperature in 1975 was `r airtempstr`, whereas current mean annual air temperature in 2022 was `r airtempend`, showing an overall increase in the period of record of `r airtempdif` $^\circ$.
-Total precipitation has increased by `r prpslo` mm per year (`r prppvl`, $R^2$ `r prprsq`).
+Similarly, total precipitation has increased by `r prpslo` mm per year (`r prppvl`, $R^2$ `r prprsq`).
 Total precipitation in 1975 was `r prpstr`, whereas current total precipitation in 2022 was `r prpend`, showing an overall increase in the period of record of `r prpdif` mm.
@@
 Increasing water temperature and decreasing salinity generally followed the meteorological trends for all three *in situ* datasets (EPC, FIM, and PDEM, @fig-meteowqraw, @tbl-temptrndtab, @tbl-saltrndtab, figures in supplement).
+Note that for tables [-@tbl-temptrndtab] and [-@tbl-saltrntab], comparable time periods were evaluated between the datasets when possible given the different sample sizes, and therefore power, to detect trends.
 The strongest trends were observed for the EPC dataset which had the longest record from 1975 to 2022.
@@
 Trends in water temperature were similar across bay segments with significant increases varying from 0.3 to 0.4 $^\circ$C per year, equating to 1.4 (OTB) to 2.0 (HB) $^\circ$C total increase from 1975 to present (@tbl-temptrndtab).
-Salinity trends were similar between bay segments, although overall salinity was higher for bay segments closer to the Gulf of Mexico.
+Salinity trends were similar between bay segments, although overall salinity was of course higher for bay segments closer to the Gulf of Mexico.
 Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, totaling an overall decrease of 2.7 and 2.0 ppt from 1975 to present.
 As expected, intra-annual variability of water temperature was much higher than salinity following the larger seasonal variation in temperature ([@fig-meteowqraw]).
-This variation may also be related to differences among the monitoring stations in each segment.
+This variation may also be related to differences among the monitoring stations in each segment as the averages in @fig-meteowqraw combine monthly and spatial variation.
 Significant increasing temperature trends were also observed for the FIM dataset from 1996 to present in OTB and from 2004 to present in HB (@tbl-temptrndtab).

### 2023-11-05 manu-draft.qmd
table 3 as knitr

@@
 library(here)
-library(flextable)
 library(dplyr)
@@
 prpdif <- round(prpprd$fit[nrow(prp)]- prpprd$fit[1], 1)
-
-# load(file = url("https://github.com/tbep-tech/load-estimates/raw/main/data/totanndat.RData"))
-# 
-# # hydro load, otb, mtb, ltb, hb
-# hydrodat <- totanndat %>% 
-#   select(year, bay_segment, hy_load) %>% 
-#   filter(bay_segment %in% c('Hillsborough Bay', 'Old Tampa Bay', 'Lower Tampa Bay', 'Middle Tampa Bay')) %>% 
-#   summarise(
-#     hy_load = sum(hy_load) / 1e3, 
-#     .by = 'year'
-#   ) 
 ```
@@
 Increasing water temperature and decreasing salinity generally followed the meteorological trends for all three *in situ* datasets (EPC, FIM, and PDEM, @fig-meteowqraw, @tbl-temptrndtab, @tbl-saltrndtab, figures in supplement).
-Note that for tables [-@tbl-temptrndtab] and [-@tbl-saltrntab], comparable time periods were evaluated between the datasets when possible given the different sample sizes, and therefore power, to detect trends.
+Note that for tables [-@tbl-temptrndtab] and [-@tbl-saltrndtab], comparable time periods were evaluated between the datasets when possible given the different sample sizes, and therefore power, to detect trends.
 The strongest trends were observed for the EPC dataset which had the longest record from 1975 to 2022.
-The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths in Tampa Bay (@fig-meteowqraw).
-Trends in water temperature were similar across bay segments with significant increases varying from 0.3 to 0.4 $^\circ$C per year, equating to 1.4 (OTB) to 2.0 (HB) $^\circ$C total increase from 1975 to present (@tbl-temptrndtab).
+The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths and minimal stratification in Tampa Bay (@fig-meteowqraw).
+Trends in water temperature were similar across bay segments with significant increases varying from 0.3 to 0.4 $^\circ$C per year, with a total increase of 1.4 (OTB) to 2.0 (HB) $^\circ$C from 1975 to present (@tbl-temptrndtab).
 Salinity trends were similar between bay segments, although overall salinity was of course higher for bay segments closer to the Gulf of Mexico.
-Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, totaling an overall decrease of 2.7 and 2.0 ppt from 1975 to present.
-As expected, intra-annual variability of water temperature was much higher than salinity following the larger seasonal variation in temperature ([@fig-meteowqraw]).
-This variation may also be related to differences among the monitoring stations in each segment as the averages in @fig-meteowqraw combine monthly and spatial variation.
+Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, with a total decrease of 2.7 (OTB) and 2.0 (LTB) ppt from 1975 to present.
+As expected, intra-annual variability of water temperature was much higher than salinity following the larger seasonal variation in temperature (@fig-meteowqraw).
+This variation may also be related to differences among the monitoring stations in each segment because the averages combine monthly and spatial variation.
 Significant increasing temperature trends were also observed for the FIM dataset from 1996 to present in OTB and from 2004 to present in HB (@tbl-temptrndtab).
-No significant temperature or salinity trends were observed in the other bay segments for the FIM and PDEM datasets from 1996 to present or 2004 present (see figures in supplement).
+No significant temperature or salinity trends were observed in the other bay segments for the FIM and PDEM datasets from 1996 to present nor 2004 to present (see figures in supplement).
 However, the EPC dataset showed significant increasing temperature across all time periods and bay segments.
@@
 The EPC dataset was also used to provide detailed information on station-level trends from 1975 to present (@fig-kendall).
-All stations in both the top and bottom of the water column had significantly increasing temperature and decreasing salinity from 1975 to present, excluding one station in HB that did not have significantly decreasing bottom salinity (Figure [-@fig-kendall]a).
-Seasonally, most stations had more significant trends in the summer, early fall period for both temperature and salinity (Figures [-@fig-kendall]b,c), although some variation was observed between bay segments.
-Temperature trends were more often observed to occur in the summer, early fall for the upper bay segments (OTB, HB), whereas the lower bay segments (MTB, LTB) had more significant trends during the spring.
-Seasonal trends in salinity did not vary as much between segments, although OTB and LTB had more significantly decreasing trends than the other bay segments.
+All stations had significantly increasing temperature and decreasing salinity from 1975 to present in both the top and bottom of the water column, excluding one station in HB that did not have significantly decreasing bottom salinity (Figure [-@fig-kendall]a).
+Seasonally, most stations had more significant trends in the summer, early fall period for both temperature and salinity (Figure [-@fig-kendall]b, [-@fig-kendall]c), although some variation was observed between bay segments.
+Temperature trends were more often observed in the summer, early fall for the upper bay segments (OTB, HB), whereas the lower bay segments (MTB, LTB) had more significant trends during the spring.
+Seasonal trends in salinity did not vary as much between segments, although OTB and LTB had more stations with significantly decreasing trends than the other bay segments.
 Again, differences between top or bottom trends did not differ seasonally.
@@
 Linear mixed-effects models showed varying trends in the potential stressor metrics for the number of days when temperature was above, salinity was below, or both occurred relative to different thresholds.
-All of the temperature models for each of the three thresholds showed significantly increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$ (see supplement).
-For all bay segments, the estimated slopes for the number of days when temperature was above 30 $^\circ$C were similar among bay segments, varying from 0.9 days per year in LTB to 1.1 days per year in MTB.
+All of the temperature models for each of the three thresholds showed significantly increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (see supplement).
+The estimated slopes for the number of days when temperature was above 30 $^\circ$C were similar among bay segments, varying from 0.9 days per year in LTB to 1.1 days per year in MTB.
 Likewise, the average number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied from 8 (OTB, HB) to 21 (LTB) days in 1975 to 55 (HB) to 63 (LTB) days in 2022.
 
-The salinity models were less similar between bay segments compared to the temperature models (see supplement).
-None of the bay segments had a significantly increasing number of days when salinity was below 15 ppt.
+The salinity models were less similar between bay segments compared to the temperature models, primarily because of the natural variation in salinity along the bay's longitudinal axis (see supplement).
+None of the bay segments had significantly increasing number of days when salinity was below 15 ppt.
 For MTB, only the number of days when salinity was below 25 ppt was significantly increasing at a rate of 0.67 days per year.
@@
 The number of days when temperature was above 29 $^\circ$C and salinity was below 25 ppt had the largest slopes of 1.4, 1.2, and 0.7 days per year for OTB, HB, and MTB, respectively.
-Likewise, the average number of days for both varied from the beginning to the end of the period of record from 6, 12, and 1 day(s) in 1975 to 72, 68, and 24 in 2022 for OTB, HB, and MTB respectively.
+Likewise, the average number of days when both temperature was above 29 $^\circ$C and salinity was below 25 ppt from the beginning to the end of the period of record varied from 6, 12, and 1 day(s) in 1975 to 72, 68, and 24 days in 2022 for OTB, HB, and MTB, respectively.
 
 @fig-mixeff provides visual examples of the mixed-effects models for the estimated number of days over time for each bay segment when temperature was above 30$^\circ$, salinity was below 25 ppt, and when both occurred.
-These thresholds were chosen moving forward because they are slightly more conservative than those with the largest slopes (i.e., higher temperature threshold), while still showing significant increases over time (excluding LTB).
-Temperature trends were similar among segments, whereas the number of days when salinity was below the threshold varied by proximity to the Gulf of Mexico (Table 3).
+Temperature trends were similar among segments, whereas the number of days when salinity was below the threshold varied by proximity to the Gulf of Mexico (@tbl-mixdaytab).
 The number of days when both temperature was above and salinity was below the threshold generally followed the trends for the number of days when salinity was below the threshold, although the slopes varied.
 These thresholds were used for comparison to seagrass changes described below, based primarily on the statistical strength of the trends and the variance of counts among bay segments.
-That is, more restrictive thresholds did not provide sufficient counts of days per year to develop models and the chosen thresholds were based primarily on statistical considerations rather than ecological considerations, as discussed below.
+That is, more restrictive thresholds did not provide sufficient counts of days per year to develop models and the chosen thresholds were based primarily on statistical considerations.
+The thresholds likely have minimal ecological significance, but rather provide a useful metric for tracking physical changes in the water column that are likely correlated with seagrass change.
 
@@
 
-Summary of mixed-effects models evaluating increases in the number of days each year from 1975 to 2022 when temperature was above 30C, salinity was below 25ppt, or both temperature and salinity were above/below the thresholds.
+```{r}
+#| label: tbl-mixdaytab
+#| tbl-cap: "Summary of mixed-effects models evaluating increases in the number of days each year from 1975 to 2022 when temperature was above 30 $^\\circ$C, salinity was below 25 ppt, or both temperature and salinity were above/below the thresholds.
 The start and end columns show the estimated number of days at the beginning and end of the period of record when temperature or salinity were above or below the thresholds.
 Values are the estimated mean number of days (plus standard error) from 1975 and 2022.
-\*\* p \< 0.005, \* p \< 0.05, - no model.
-OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
-
-```{r}
+** p < 0.005, * p < 0.05.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay."
 mixdaytab

### 2023-11-05 R/tabs.R

@@ -1,6 +1,5 @@
 library(tidyverse)
 library(here)
-library(flextable)
 library(mgcv)
 library(broom)
 
@@ -190,12 +189,10 @@ totab <- mixmods %>%
   arrange(bay_segment, thrtyp)
 
 mixdaytab <- totab %>% 
-  as_grouped_data(groups = 'bay_segment') %>% 
-  flextable() %>% 
-  set_header_labels(i = 1, values = c('Bay Segment', 'Threshold', 'Slope', 'Start', 'End')) %>% 
-  padding(padding = 0, part = 'all') %>% 
-  width(j = 2, width = 1.5) %>% 
-  font(part = 'all', fontname = 'Times New Roman')
+  mutate(
+    bay_segment = ifelse(duplicated(bay_segment), '', as.character(bay_segment))
+  ) %>%
+  knitr::kable(col.names = c('Bay Segment', 'Threshold', 'Slope', 'Start', 'End'))
 
 save(mixdaytab, file = here('tabs/mixdaytab.RData'))
 

### 2023-11-05 manu-draft.qmd
pinco mod scratch

@@
 These models did not include stressor metrics, but rather used direct measurements of salinity and temperature as independent variables.
-Models of temperature or salinity to predict daily measurements are not possible with the FIM and PDEM monitoring designs because random locations were used for each sample date.
+Models of temperature or salinity to predict daily measurements were not possible with the FIM and PDEM monitoring designs because random locations were used for each sample date.
 However, the sample sizes for the FIM and PDEM models were much larger than those for the EPC data, where the latter was limited by the annual timestep of the transect surveys.

### 2023-11-05 R/funcs.R

@@ -69,7 +69,7 @@ getprd_fun <- function(modin, depvar = 'Sal'){
   
 }
 
-# get model predictions gam
+# get model predictions gam, fim
 getprd_fun2 <- function(mod, indvar = c('sal', 'temp')){
   
   indvar <- match.arg(indvar)
@@ -92,6 +92,32 @@ getprd_fun2 <- function(mod, indvar = c('sal', 'temp')){
     ) %>% 
     unnest(data)
 
+  return(out)
+  
+}
+
+
+# get model predictions gam, pinco
+getprd_fun3 <- function(mod, indvar = c('sal', 'temp')){
+  
+  out <- unique(mod$model) %>% 
+    group_by(yrcat) %>% 
+    nest() %>% 
+    mutate(
+      data = purrr::pmap(list(yrcat), function(yrcat){
+        
+        condls <- list(yrcat = yrcat)
+        prd <- visreg(mod, xvar = indvar, cond = condls, plot = F, scale = 'response')
+        
+        out <- prd$fit %>% 
+          select(-yrcat)
+        
+        return(out)
+        
+      })
+    ) %>% 
+    unnest(data)
+  
   return(out)
   
 }

### 2023-11-05 R/pincomod_scratch.R

@@ -1,0 +1,93 @@
+data(pincotemp)
+
+tomod <- pincotemp %>%
+  mutate(
+    yrcat = case_when(
+      yr <= 2015 ~ 'Recovery (pre - 2016)',
+      T ~ 'Decline (2016 - present)'
+    ), 
+    yrcat = factor(yrcat, levels = c('Recovery (pre - 2016)', 'Decline (2016 - present)'))
+  ) #%>% 
+  # filter(yr > 2008) %>%
+  # filter(
+  # month(date) >= 7 & month(date) <= 11
+  # ) #%>%
+  # filter(site == 'E1') %>%
+  # filter(hr %in% c(9:12))
+
+# models
+pincosalmod <- gam(allsg ~ yrcat + s(sal, by = yrcat), data = tomod, family = binomial(link = 'logit'),
+                 method = 'REML')
+
+pincotempmod <- gam(allsg ~ yrcat + s(temp, by = yrcat), data = tomod, family = binomial(link = 'logit'),
+                  method = 'REML')
+
+toplo1 <- getprd_fun3(pincosalmod, 'sal')
+toplo2 <- getprd_fun3(pincotempmod, 'temp')
+
+thm <- theme_bw() + 
+  theme(
+    panel.grid.minor = element_blank(), 
+    strip.background = element_blank(),
+    strip.text = element_text(size = 12), 
+    axis.text = element_text(size = 11), 
+    axis.title = element_text(size = 12)
+  )
+
+# separate plots by bay segment
+p1 <- ggplot(toplo1, aes(x = sal, y = visregFit)) + 
+  geom_rug(data = tomod[tomod$allsg == 0,], aes(x = sal, y = allsg), sides = 'b', linewidth = 1, color = 'red') +
+  geom_rug(data = tomod[tomod$allsg == 1,], aes(x = sal, y = allsg), sides = 't', linewidth = 1, color = 'blue') +
+  geom_ribbon(aes(ymin = visregLwr, ymax = visregUpr), fill = 'dodgerblue2', alpha = 0.2) +
+  geom_line(color = 'dodgerblue2', size = 1) +
+  coord_cartesian(
+    ylim = c(0, max(toplo1$visregUpr))#,
+    # xlim = c(10, 30)
+  ) +
+  facet_grid(~ yrcat, scales = 'free') +
+  thm + 
+  labs(
+    x = 'Salinity (psu)', 
[…]

### 2023-11-06 manu-draft.qmd
update all figs, tables, results with 2022 fim data

@@
 This variation may also be related to differences among the monitoring stations in each segment because the averages combine monthly and spatial variation.
-Significant increasing temperature trends were also observed for the FIM dataset from 1996 to present in OTB and from 2004 to present in HB (@tbl-temptrndtab).
-No significant temperature or salinity trends were observed in the other bay segments for the FIM and PDEM datasets from 1996 to present nor 2004 to present (see figures in supplement).
+Significant increasing temperature trends were also observed for the FIM dataset from 1996 to present in OTB and from 2004 to present in OTB and HB (@tbl-temptrndtab).
+No significant temperature or salinity trends were observed in the other bay segments for the FIM and PDEM datasets from 1996 to present nor from 2004 to present (see figures in supplement).
 However, the EPC dataset showed significant increasing temperature across all time periods and bay segments.

### 2023-11-07 manu-draft.qmd
redo transect stressor metric models and figure

@@
 Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment.
-The stressor metrics were compared to various response measures at the bay segments-scale from the transect data, including frequency occurrence each year (all species), percent change in frequency occurrence between years, and increase/decrease in frequency occurrence between years as a binary (1/0) variable.
-Generalized linear regression models were used to evaluate each of the seagrass response variables relative to the stressor metrics.
-Interaction factors were also included for bay segment and time period (recovery pre-2016, decline post-2016).
+The stressor metrics were compared to frequency occurrence (all species) each year by bay segment(all species).
+Generalized linear regression models were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre-2016, decline post-2016).
+Two models wre evaluated, one with the temperature and salinity metrics and another with the both metrics given that the latter was a combined metric of the former.
+Interactions were included among all the predictor variables to test for changes in the relationships by bay segment, time period, and different levels of the stressor metrics.
 The most parsimonious models were identified by forward and backward variable selection and comparison with AIC values [@sakamoto1986; @venables2002b].
@@
 Additional models were constructed for the FIM and PDEM datasets to evaluate seagrass change relative to temperature and salinity.
-These models did not include stressor metrics, but rather used direct measurements of salinity and temperature as independent variables.
-Models of temperature or salinity to predict daily measurements were not possible with the FIM and PDEM monitoring designs because random locations were used for each sample date.
+These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs.
 However, the sample sizes for the FIM and PDEM models were much larger than those for the EPC data, where the latter was limited by the annual timestep of the transect surveys.
@@
 Interactions with factor variables for bay segment and time period as above were also included.
-The GAMs also evaluated interactions between salinity and temperature to determine if both factors had an additive effect on seagrass change.
+The GAMs also evaluated interactions between salinity and temperature to determine if both factors had a combined effect on seagrass change.
 The GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated (no bay segment interaction) and the seagrass response variable was presence/absence (i.e., a binomial distribution was used).
@@
 
-![Ordination plots of Nonmetric Multidimensional Scaling results for annual estimates in each bay segment for total seagrass frequency occurrence (Freq.
-Occ.), counts of days above the temperature threshold (Temp.), counts of days below the salinity threshold (Sal.), and counts of days above or below for both, respectively.
-The left plot excludes counts of days with both temperature and salinity above and below the threshods, whereas the right plot excludes the separate temperature and salinity metrics.
-Points are sized by seagrass frequency occurrence.
-95% confidence ellipses are shown for observations in each major bay segment.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/nms.png){#fig-nms}
-
-{{< pagebreak >}}
-
-![Predicted results from generalized linear models evaluating seagrass decline as (a) percent change between years and (b) probability of decline versus number of days when salinity was below the threshold.
-Models evaluated the response variable on the y-axes as a function of salinity days and interactions with time period (seagrass recovery prior to 2016, seagrass decline 2016 to present) and bay segment.
+![Predicted results from generalized linear models evaluating seagrass frequency occurrence by time periods versus (a) the number of days when salinity was below the threshold and (b) the number of days when temperature was above the threshold.
+Time time periods were seagrass recovery prior to 2016 and seagrass decline 2016 to present.
 Shaded areas are 95% confidence intervals.
-Colored lines in (b) as rug plots show instances of seagrass increase (blue) or decrease (red) by year.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/sgmod.png){#fig-sgmod}
+Points are estimates of seagrass frequency and stressor metrics by bay segment.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/tempsalmod.png){#fig-sgmod}
 

### 2023-11-07 R/dat_proc.R

@@ -1153,32 +1153,21 @@ cmbdat <- fodat %>%
     Temp = tempcnt, 
     Both = bothcnt
   ) %>% 
-  arrange(bay_segment, Year) %>% 
-  mutate(
-    chng = c(NA, sign(diff(total))), 
-    chng = ifelse(chng == -1, 1, 0), 
-    .by = 'bay_segment'
-  ) %>% 
-  filter(!is.na(chng))
+  arrange(bay_segment, Year)
 
 tomod <- cmbdat %>% 
   mutate(
-    pchg = (total - lag(total)) / lag(total), 
-    chg = c(NA, sign(diff(total))), 
-    chg = ifelse(chg == -1, 1, 0),
-    .by = bay_segment, 
     yrcat = cut(Year, breaks = c(-Inf, 2016, Inf), labels = c('Recovery (pre - 2016)', 'Decline (2016 - present)'), right = F)
   ) %>% 
-  filter(!is.na(pchg)) %>% 
   filter(bay_segment != 'LTB')
 
-pchgmod <- glm(pchg ~ bay_segment*Sal*yrcat + bay_segment*Temp*yrcat, data = tomod) %>% 
-  step(trace = 0)
-binomod <- glm(chg ~ bay_segment*Sal*yrcat + bay_segment*Temp*yrcat, data = tomod, family = binomial('logit')) %>% 
-  step(trace = 0) 
+tempsalmod <- lm(total ~ bay_segment*yrcat*Temp*Sal, data = tomod) %>% 
+  step()
+bothmod <- mod <- lm(total ~ bay_segment*yrcat*Both, data = tomod) %>% 
+  step()
 
-save(pchgmod, file = here('data/pchgmod.RData'))
-save(binomod, file = here('data/binomod.RData'))
+save(tempsalmod, file = here('data/tempsalmod.RData'))
+save(bothmod, file = here('data/bothmod.RData'))
 
 # USGS gage data ------------------------------------------------------------------------------
 

### 2023-11-07 R/figs.R

@@ -1214,21 +1214,30 @@ dev.off()
 
 # seagrass decline models ---------------------------------------------------------------------
 
-load(file = here('data/pchgmod.RData'))
-load(file = here('data/binomod.RData'))
+load(file = here('data/tempsalmod.RData'))
+load(file = here('data/bothmod.RData'))
 
-toplo1 <- getprd_fun(pchgmod)
-toplo2 <- getprd_fun(binomod)
+toplo1a <- visreg(tempsalmod, xvar = 'Sal', by = 'yrcat', scale = 'response', plot = F, cond = list(bay_segment = 'OTB'))$fit
+toplo1b <- visreg(tempsalmod, xvar = 'Sal', by = 'yrcat', scale = 'response', plot = F, cond = list(bay_segment = 'HB'))$fit
+toplo1c <- visreg(tempsalmod, xvar = 'Sal', by = 'yrcat', scale = 'response', plot = F, cond = list(bay_segment = 'MTB'))$fit
+toplo1 <- bind_rows(toplo1a, toplo1b, toplo1c)
+toplo2a <- visreg(tempsalmod, xvar = 'Temp', by = 'yrcat', scale = 'response', plot = F, cond = list(bay_segment = 'OTB'))$fit
+toplo2b <- visreg(tempsalmod, xvar = 'Temp', by = 'yrcat', scale = 'response', plot = F, cond = list(bay_segment = 'HB'))$fit
+toplo2c <- visreg(tempsalmod, xvar = 'Temp', by = 'yrcat', scale = 'response', plot = F, cond = list(bay_segment = 'MTB'))$fit
+toplo2 <- bind_rows(toplo2a, toplo2b, toplo2c)
+
+tomod <- tempsalmod$model
 
 p1 <- ggplot(toplo1, aes(x = Sal)) + 
   geom_ribbon(aes(ymin = visregLwr, ymax = visregUpr), fill = 'lightgrey', alpha = 0.5) + 
   geom_line(aes(y = visregFit)) + 
-  geom_point(data = tomod, aes(x = Sal, y = pchg), color = 'dodgerblue2') +
-  facet_grid(bay_segment ~ yrcat) + 
+  geom_point(data = tomod, aes(x = Sal, y = total), color = 'dodgerblue2') +
+  facet_grid(bay_segment ~ yrcat, scales = 'free_y') + 
+  coord_cartesian(ylim = c(0, 1)) +
   labs(
-    y = '% change in seagrass', 
+    y = 'Frequency Occurrence', 
[…]

### 2023-11-08 R/figs.R
remove nms fig

@@ -1152,66 +1152,6 @@ png(here('figs/suppmixeff.png'), height = 6, width = 7, family = 'serif', units 
 print(p)
 dev.off()
 
-# nms -----------------------------------------------------------------------------------------
-
-load(file = here('data/cmbdat.RData'))
-
-cmbdat <- cmbdat %>% 
-  filter(!bay_segment %in% 'LTB')
-
-toord1 <- cmbdat %>% 
-  dplyr::select(-Year, -Chla, -bbave, -Both, -bay_segment) %>% 
-  mutate_all(rescale, to = c(0, 1))
-
-orddat1 <- metaMDS(toord1, k = 2, trymax = 100)
-
-toord2 <- cmbdat %>% 
-  dplyr::select(-Year, -Chla, -bbave, -Sal, -Temp, -bay_segment) %>% 
-  mutate_all(rescale, to = c(0, 1))
-
-orddat2 <- metaMDS(toord2, k = 2, trymax = 100)
-
-grps <- factor(cmbdat$bay_segment)
-
-vec_ext <- 1.75
-coord_fix <- F
-size <- toord1$total
-repel <- F
-force <- T
-arrow <- 0.2
-txt <- 3
-alpha <- 0.8
-ext <- 1.25
-exp <- 0.1
-parse <- F
-ellipse <- T
-cols <- c("#3B9AB2", "#EBCC2A", "#F21A00")# c("#1F78B4", "#33A02C", "#E31A1C")
-vec_lab <- list(
-  'total'= 'Freq Occ',
-  'Sal' = 'Sal', 
-  'Temp' = 'Temp', 
-  'Both' = 'Both'
-)
-grp_title <- 'Bay segment'
-sizelab <- 'Freq Occ'
-
-p1 <- ggord(orddat1, axes = c('1', '2'), cols = cols, ellipse = ellipse, grp_in = grps,
-            parse = parse, vec_ext = vec_ext, coord_fix = coord_fix, size = size, 
-            repel = repel, arrow = arrow, txt = txt, alpha = alpha, ext = ext, 
-            exp = exp, grp_title = grp_title, sizelab = sizelab, vec_lab = vec_lab, 
-            force = force)
-p2 <- ggord(orddat2, axes = c('1', '2'), cols = cols, ellipse = ellipse, grp_in = grps,
-            parse = parse, vec_ext = vec_ext, coord_fix = coord_fix, size = size, 
-            repel = repel, arrow = arrow, txt = txt, alpha = alpha, ext = ext, 
[…]

### 2023-11-08 manu-draft.qmd
remove nms fig
better GLMs for seagrass change and final fig, removed scratch scripts, removed earlier model figs

@@
 load(file = here('tabs/mixdaytab.RData'))
+load(file = here('data/sgmods.RData'))
 ```
@@
 The stressor metrics were compared to frequency occurrence (all species) each year by bay segment(all species).
-Generalized linear regression models were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre-2016, decline post-2016).
+Generalized linear regression models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre-2016, decline post-2016).
 Two models wre evaluated, one with the temperature and salinity metrics and another with the both metrics given that the latter was a combined metric of the former.
-Interactions were included among all the predictor variables to test for changes in the relationships by bay segment, time period, and different levels of the stressor metrics.
+Interactions were included between temperature, salinity, and time period or the both metric and time period depending on the model.
+Bay segment was included as a main effect without interactions.
 The most parsimonious models were identified by forward and backward variable selection and comparison with AIC values [@sakamoto1986; @venables2002b].
@@
 These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs.
-However, the sample sizes for the FIM and PDEM models were much larger than those for the EPC data, where the latter was limited by the annual timestep of the transect surveys.
-This allowed for more flexible models that are less limited by degrees of freedom.
-GAMs were used to evaluate seagrass percent cover for the FIM data using continuous temperature and salinity as independent variables.
-Interactions with factor variables for bay segment and time period as above were also included.
-The GAMs also evaluated interactions between salinity and temperature to determine if both factors had a combined effect on seagrass change.
-The GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated (no bay segment interaction) and the seagrass response variable was presence/absence (i.e., a binomial distribution was used).
+GLMs were used for the FIM data to evaluate seagrass percent cover using continuous temperature and salinity as independent variables.
+Interactions with time period and a main effect for bay segment were also included as above.
+GLMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated (no bay segment interaction) and the seagrass response variable was presence/absence (i.e., a binomial distribution was used).
+Input data for models for the FIM and PDEM datasets were further subset to include only months from July to November to reduce potential seasonal effects.
 
@@
 
-NMS results in @fig-nms
+GLMs for the "Both" model did not include a year category and the interaction of temp/sal in the first model was negligible (fig holds temp/sal at mean? for respective plots).
 
-Seagrass decline results in @fig-sgmod, @fig-sggammod
+Seagrass decline results in @fig-tempsalmod, @fig-sggammod
 
@@
 
-![Predicted results from generalized linear models evaluating seagrass frequency occurrence by time periods versus (a) the number of days when salinity was below the threshold and (b) the number of days when temperature was above the threshold.
-Time time periods were seagrass recovery prior to 2016 and seagrass decline 2016 to present.
+![Predicted results from generalized linear models evaluating seagrass changes versus temperature or salinity predictors for the (a) EPC and (b) FIM datasets.
+The predictors for the (a) were the number of days when temperature was above the threshold and the number of days when temperature was below the threshold.
+The predictors for (b) were measured temperature and salinity at the bottom of the water column.
+Time periods were seagrass recovery prior to 2016 and seagrass decline after 2016.
 Shaded areas are 95% confidence intervals.
-Points are estimates of seagrass frequency and stressor metrics by bay segment.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/tempsalmod.png){#fig-sgmod}
-
-{{< pagebreak >}}
-
-![Predicted results from generalized additive models evaluating seagrass percent cover compared to (a) salinity and (b) water temperature.
-Models include interactions with time period (seagrass recovery prior to 2016, seagrass decline 2016 to present) and bay segment.
-Shaded areas are 95% confidence intervals.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/sggammod.png){#fig-sggammod}
+Points are observed data that include aggregated effects of bay segment included in the models.](figs/sgmods.png){#fig-sgmod}
 

### 2023-11-08 R/dat_proc.R

@@ -1280,7 +1280,7 @@ habdat <- habraw %>%
     ) 
   ) %>%
   select(Reference, BottomVeg, BottomVegRatio, bvegcat) %>% 
-  filter(!is.na(bvegcat) | bvegcat == 'mc') %>% 
+  filter(!(is.na(bvegcat) | bvegcat == 'mc')) %>% 
   left_join(phydat, by = 'Reference') %>% 
   select(-geometry) %>% 
   mutate(
@@ -1294,7 +1294,8 @@ habdat <- habraw %>%
     ),
     sgcov = sum(vegcov, na.rm = T),
     .by = Reference
-  )
+  ) %>% 
+  filter(sgcov <= 100) # some above 100, probably an error
 
 # combine seagrass p/a with fimtempdat
 fimsgtempdat <- fimtempdat %>% 
@@ -1305,7 +1306,9 @@ fimsgtempdat <- fimtempdat %>%
     lat = st_coordinates(.)[, 2], 
     yr = year(date)
   ) %>% 
-  st_set_geometry(NULL)
+  st_set_geometry(NULL) %>% 
+  filter(!is.na(temp)) %>% 
+  filter(!is.na(sal))
 
 save(fimsgtempdat, file = here('data/fimsgtempdat.RData'), compress = 'xz')
 
@@ -1386,8 +1389,10 @@ pincotemp <-  saltemp %>%
     mo = month(date), 
     bay_segment = 'OTB'
   ) %>% 
-  filter(yr > 2003) # only a few
-  
+  filter(yr > 2003) %>% # only a few
+  filter(!is.na(temp)) %>%
+  filter(!is.na(sal))
+
 save(pincotemp, file = here('data/pincotemp.RData'))
 
 # seagrass declines models, epc, fim, and pinco -----------------------------------------------
@@ -1405,23 +1410,28 @@ modprep <- function(dat){
         yr <= 2016 ~ 'pre', 
         yr > 2016 ~ 'post'
       ), 
-      yrcat = factor(yrcat, levels = c('pre', 'post'))
+      yrcat = factor(yrcat, levels = c('pre', 'post'), 
+                     labels = c('Recovery (pre - 2016)', 'Decline (post - 2016)'))
     )
 
 }
 
 epctomod <- modprep(epccmbdat)
-fimtomod <- modprep(fimsgtempdat)
-pincotomod <- modprep(pincotemp)
[…]

### 2023-11-08 R/figs.R

@@ -1152,118 +1152,52 @@ png(here('figs/suppmixeff.png'), height = 6, width = 7, family = 'serif', units 
 print(p)
 dev.off()
 
-# seagrass decline models ---------------------------------------------------------------------
+# seagrass change plots -----------------------------------------------------------------------
 
-load(file = here('data/tempsalmod.RData'))
-load(file = here('data/bothmod.RData'))
+load(file = here('data/sgmods.RData'))
 
-toplo1a <- visreg(tempsalmod, xvar = 'Sal', by = 'yrcat', scale = 'response', plot = F, cond = list(bay_segment = 'OTB'))$fit
-toplo1b <- visreg(tempsalmod, xvar = 'Sal', by = 'yrcat', scale = 'response', plot = F, cond = list(bay_segment = 'HB'))$fit
-toplo1c <- visreg(tempsalmod, xvar = 'Sal', by = 'yrcat', scale = 'response', plot = F, cond = list(bay_segment = 'MTB'))$fit
-toplo1 <- bind_rows(toplo1a, toplo1b, toplo1c)
-toplo2a <- visreg(tempsalmod, xvar = 'Temp', by = 'yrcat', scale = 'response', plot = F, cond = list(bay_segment = 'OTB'))$fit
-toplo2b <- visreg(tempsalmod, xvar = 'Temp', by = 'yrcat', scale = 'response', plot = F, cond = list(bay_segment = 'HB'))$fit
-toplo2c <- visreg(tempsalmod, xvar = 'Temp', by = 'yrcat', scale = 'response', plot = F, cond = list(bay_segment = 'MTB'))$fit
-toplo2 <- bind_rows(toplo2a, toplo2b, toplo2c)
+epcmod1 <- sgmods$epcmod1
+fimmod <- sgmods$fimmod
+pincomod <- sgmods$pincomod
 
-tomod <- tempsalmod$model
+##
+# epc model plot
 
-p1 <- ggplot(toplo1, aes(x = Sal)) + 
-  geom_ribbon(aes(ymin = visregLwr, ymax = visregUpr), fill = 'lightgrey', alpha = 0.5) + 
-  geom_line(aes(y = visregFit)) + 
-  geom_point(data = tomod, aes(x = Sal, y = total), color = 'dodgerblue2') +
-  facet_grid(bay_segment ~ yrcat, scales = 'free_y') + 
[…]

### 2023-11-08 R/funcs.R

@@ -41,83 +41,99 @@ runfunc <- function(cnt){
   return(out)
 }
 
-# get model predictions glm
-getprd_fun <- function(modin, depvar = 'Sal'){
+# prep epc, fim, pinco data prior to modelign 
+modprep <- function(dat){
   
-  fitotb <- visreg(modin, depvar, by = 'yrcat', cond = list(bay_segment = 'OTB'), scale = 'response', plot = F) %>% 
-    .$fit %>%  
+  dat %>% 
+    filter(!bay_segment %in% c('LTB')) %>% 
     mutate(
-      bay_segment = 'OTB'
+      yrcat = case_when(
+        yr <= 2016 ~ 'pre', 
+        yr > 2016 ~ 'post'
+      ), 
+      yrcat = factor(yrcat, levels = c('pre', 'post'), 
+                     labels = c('Recovery (pre - 2016)', 'Decline (2016 - present)'))
     )
-  fithb <- visreg(modin, depvar, by = 'yrcat', cond = list(bay_segment = 'HB'), scale = 'response', plot = F) %>% 
-    .$fit %>% 
-    mutate(
-      bay_segment = 'HB'
-    ) 
-  fitmtb <- visreg(modin, depvar, by = 'yrcat', cond = list(bay_segment = 'MTB'), scale = 'response', plot = F) %>% 
-    .$fit %>% 
-    mutate(
-      bay_segment = 'MTB'
-    )
-  
-  out <- bind_rows(fitotb, fithb, fitmtb) %>% 
-    mutate(
-      bay_segment = factor(bay_segment, levels = segshr)
-    )
-  
-  return(out)
   
 }
 
-# get model predictions gam, fim
-getprd_fun2 <- function(mod, indvar = c('sal', 'temp')){
-  
-  indvar <- match.arg(indvar)
-  
-  out <- unique(mod$model[, c('bay_segment', 'yrcat')]) %>% 
-    group_by(bay_segment, yrcat) %>% 
-    nest() %>% 
-    mutate(
-      data = purrr::pmap(list(bay_segment, yrcat), function(bay_segment, yrcat){
-        
-        condls <- list(bay_segment = bay_segment, yrcat = yrcat, fctcrs = paste(yrcat, bay_segment, sep = ':'))
-        prd <- visreg(mod, xvar = indvar, cond = condls, plot = F, scale = 'response')
-        
-        out <- prd$fit %>% 
[…]

### 2023-11-08 R/pincomod_scratch.R

@@ -1,93 +1,0 @@
-data(pincotemp)
-
-tomod <- pincotemp %>%
-  mutate(
-    yrcat = case_when(
-      yr <= 2015 ~ 'Recovery (pre - 2016)',
-      T ~ 'Decline (2016 - present)'
-    ), 
-    yrcat = factor(yrcat, levels = c('Recovery (pre - 2016)', 'Decline (2016 - present)'))
-  ) #%>% 
-  # filter(yr > 2008) %>%
-  # filter(
-  # month(date) >= 7 & month(date) <= 11
-  # ) #%>%
-  # filter(site == 'E1') %>%
-  # filter(hr %in% c(9:12))
-
-# models
-pincosalmod <- gam(allsg ~ yrcat + s(sal, by = yrcat), data = tomod, family = binomial(link = 'logit'),
-                 method = 'REML')
-
-pincotempmod <- gam(allsg ~ yrcat + s(temp, by = yrcat), data = tomod, family = binomial(link = 'logit'),
-                  method = 'REML')
-
-toplo1 <- getprd_fun3(pincosalmod, 'sal')
-toplo2 <- getprd_fun3(pincotempmod, 'temp')
-
-thm <- theme_bw() + 
-  theme(
-    panel.grid.minor = element_blank(), 
-    strip.background = element_blank(),
-    strip.text = element_text(size = 12), 
-    axis.text = element_text(size = 11), 
-    axis.title = element_text(size = 12)
-  )
-
-# separate plots by bay segment
-p1 <- ggplot(toplo1, aes(x = sal, y = visregFit)) + 
-  geom_rug(data = tomod[tomod$allsg == 0,], aes(x = sal, y = allsg), sides = 'b', linewidth = 1, color = 'red') +
-  geom_rug(data = tomod[tomod$allsg == 1,], aes(x = sal, y = allsg), sides = 't', linewidth = 1, color = 'blue') +
-  geom_ribbon(aes(ymin = visregLwr, ymax = visregUpr), fill = 'dodgerblue2', alpha = 0.2) +
-  geom_line(color = 'dodgerblue2', size = 1) +
-  coord_cartesian(
-    ylim = c(0, max(toplo1$visregUpr))#,
-    # xlim = c(10, 30)
-  ) +
-  facet_grid(~ yrcat, scales = 'free') +
-  thm + 
-  labs(
-    x = 'Salinity (psu)', 
[…]

### 2023-11-08 manu-draft.qmd
results draft done, but need to add model tables

@@
 
-GLMs for the "Both" model did not include a year category and the interaction of temp/sal in the first model was negligible (fig holds temp/sal at mean? for respective plots).
+```{r}
+epcmodsum1 <- modtxt_fun(sgmods$epcmod1)
+epcmodsum2 <- modtxt_fun(sgmods$epcmod2)
+fimmodsum <- modtxt_fun(sgmods$fimmod)
+```
 
-Seagrass decline results in @fig-tempsalmod, @fig-sggammod
+Linear models to assess the potential effects of temperature and salinity on seagrass changes provided modest evidence that seagrass loss after 2016 was driven by climate stressors.
+The most parsimonious model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence (`r epcmodsum1`).
+An interaction of the metrics with time period was also observed, with notable differences by predictor (Figure [-@fig-sgmod]a).
+For the period prior to 2016, frequency occurrence of seagrass increased as the number of days where temperature was above 30\$\^\\circ\$ increased, whereas no relationship was observed after 2016.
+However, the interaction between time period and temperature was not significant, suggesting no difference between time periods.
+The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (p \< 0.01), Figure [-@fig-sgmod]a).
+The interaction between the temperature and salinity metric was included in the most parsimonious model (p \< 0.1), although the shape of the relationship of temperature with time period or salinity with time period did not noticeably change using different condition values for salinity or temperature, respectively.
+The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (`r epcmodsum2`), although the interaction term between the both metric and time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016.
+
+The most parsimonious model for the FIM dataset also showed potential temperature and salinity associations with seagrass change after 2016, although explanatory power was lower than the EPC models likely due to the larger spread of data (`r fimmodsum`).
+Increases in temperature and decreases in salinity were both associated with reductions in percent cover after 2016, whereas no relationships were observed prior to 2016 (p \< 0.05 and p \< 0.1 for the interactions between temperature, salinity with time period, respectively, Figure [-@fig-sgmod]b).
+The interaction between salinity and temperature was not included as a significant term in the final model.
+Lastly, the most parsimonious model for the PDEM dataset only included time period as a marginally significant variable and the overall model fit was not significant.
 
 ## Discussion
+
+Takehome results: notable temp, salinity changes, some evidence of association with seagrass change.
+Of the limited evidence, effects of high temp, low salinity may just be additive, no support for interactions.
+This was shown by the FIM models and lack of significance for the "both" EPC model.
+Also, note that the PDEM data may have had low power, limited spatially (only west side of bay segment) and even shorter time period (supp fig, sal trends support this).
 

### 2023-11-08 R/figs.R

@@ -1165,9 +1165,9 @@ pincomod <- sgmods$pincomod
 
 epcplo <- modplo_fun(
   mod = epcmod1, 
-  xlab1 = 'days temperature > 30 \u00B0C', 
+  xlab1 = '# days temperature > 30 \u00B0C', 
   ylab1 = 'Frequency occurrence', 
-  xlab2 = 'days salinity < 25 ppt',
+  xlab2 = '# days salinity < 25 ppt',
   title1 = '(a) EPC', 
   ismetric = T
 )

### 2023-11-08 manu-draft.qmd
added summary of glm table

@@
 load(file = here('tabs/mixdaytab.RData'))
+load(file = here('tabs/modtab.RData'))
 load(file = here('data/sgmods.RData'))
@@
 The most parsimonious model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence (`r epcmodsum1`).
-An interaction of the metrics with time period was also observed, with notable differences by predictor (Figure [-@fig-sgmod]a).
+An interaction of the metrics with time period was also observed, with notable differences by predictor (Figure [-@fig-sgmod]a, @tbl-modtab).
 For the period prior to 2016, frequency occurrence of seagrass increased as the number of days where temperature was above 30\$\^\\circ\$ increased, whereas no relationship was observed after 2016.
 However, the interaction between time period and temperature was not significant, suggesting no difference between time periods.
-The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (p \< 0.01), Figure [-@fig-sgmod]a).
+The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (p \< 0.01), Figure [-@fig-sgmod]a, @tbl-modtab).
 The interaction between the temperature and salinity metric was included in the most parsimonious model (p \< 0.1), although the shape of the relationship of temperature with time period or salinity with time period did not noticeably change using different condition values for salinity or temperature, respectively.
-The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (`r epcmodsum2`), although the interaction term between the both metric and time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016.
+The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (`r epcmodsum2`), although the interaction term between the both metric and time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab).
 
 The most parsimonious model for the FIM dataset also showed potential temperature and salinity associations with seagrass change after 2016, although explanatory power was lower than the EPC models likely due to the larger spread of data (`r fimmodsum`).
-Increases in temperature and decreases in salinity were both associated with reductions in percent cover after 2016, whereas no relationships were observed prior to 2016 (p \< 0.05 and p \< 0.1 for the interactions between temperature, salinity with time period, respectively, Figure [-@fig-sgmod]b).
+Increases in temperature and decreases in salinity were both associated with reductions in percent cover after 2016, whereas no relationships were observed prior to 2016 (p \< 0.05 and p \< 0.1 for the interactions between temperature, salinity with time period, respectively, Figure [-@fig-sgmod]b, @tbl-modtab).
 The interaction between salinity and temperature was not included as a significant term in the final model.
-Lastly, the most parsimonious model for the PDEM dataset only included time period as a marginally significant variable and the overall model fit was not significant.
+Lastly, the most parsimonious model for the PDEM dataset only included time period as a marginally significant variable and the overall model fit was not significant (@tbl-modtab).
 
@@
 
+```{r}
+#| label: tbl-modtab
+#| tbl-cap: 'Summary of the most parsimonious Generalized Linear Models (GLMs) developed to evaluate the effects of temperature and salinity variables on seagrass change between time periods.
+Models were developed for different datasets described in the model column.
+The significant intercept term for each model is not shown and described factor conditions not explicitly shown by the remaining terms (eg., OTB and pre-2016).
+The model term estimates include the standard error in parentheses.
+Note that the "Temp" and "Sal" terms for the EPC models describe metrics that measure the number of days when temperature was above or salinity was below a threshold value, whereas the terms in all other models describe observed temperature or salinity (see text for description).
+** p < 0.005, * p < 0.05, $^.$ p < 0.1.
+EPC: Environmental Protection Commission, FIM: Fisheries Independent Monitoring, PDEM: Pinellas County Department of Environmental Management.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.'
+modtab
+```
+
+{{< pagebreak >}}
+
 ## References {.unnumbered}

### 2023-11-08 R/funcs.R

@@ -11,6 +11,19 @@ p_ast <- function(x){
   
 }
 
+# function for formatting p-values in tables
+p_ast2 <- function(x){
+  
+  sig_cats <- c('**', '*', '$^.$', '')
+  sig_vals <- c(-Inf, 0.005, 0.05, 0.1, Inf)
+  
+  out <- cut(x, breaks = sig_vals, labels = sig_cats, right = FALSE)
+  out <- as.character(out)
+  
+  return(out)
+  
+}
+
 # function for formatting p-values in text
 p_txt <- function(x){
   
@@ -69,7 +82,8 @@ modprep <- function(dat){
         yr > 2016 ~ 'post'
       ), 
       yrcat = factor(yrcat, levels = c('pre', 'post'), 
-                     labels = c('Recovery (pre - 2016)', 'Decline (2016 - present)'))
+                     labels = c('Recovery (pre - 2016)', 'Decline (2016 - present)')), 
+      bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB'))
     )
   
 }

### 2023-11-08 R/tabs.R

@@ -230,6 +230,42 @@ suppsalifittab <- fittab %>%
 save(supptempfittab, file = here('tabs/supptempfittab.RData'))
 save(suppsalifittab, file = here('tabs/suppsalifittab.RData'))
 
+
+# glm performance -----------------------------------------------------------------------------
+
+load(file = here("data/sgmods.RData"))
+
+# epcmod1
+modtab <- sgmods %>% 
+  lapply(tidy) %>% 
+  enframe() %>% 
+  unnest(value) %>%
+  filter(term != '(Intercept)') %>% 
+  mutate(p.value = p_ast2(p.value)) %>% 
+  mutate_if(is.numeric, formatC, digits = 2) %>% 
+  mutate(
+    std.error = paste0('(', std.error, ')'), 
+    name = case_when(
+      name == 'epcmod1' ~ 'EPC 1', 
+      name == 'epcmod2' ~ 'EPC 2',
+      name == 'fimmod' ~ 'FIM',
+      name == 'pincomod' ~ 'PDEM'
+    ), 
+    term = gsub('temp', 'Temp', term), 
+    term = gsub('sal', 'Sal', term),
+    term = gsub('both', 'Both', term),
+    term = gsub('yrcatDecline \\(post - 2016\\)', 'Time_post', term),
+    term = gsub('yrcatDecline \\(post - 2016\\)', 'Time_post', term), 
+    term = gsub('bay_segment', 'Baysegment_', term),
+    name = ifelse(duplicated(name), '', name)
+  ) %>% 
+  unite('estimate', estimate, std.error, sep = ' ') %>% 
+  unite('statistic', statistic, p.value, sep = '') %>%
+  select(Model = name, Term = term, Estimate = estimate, `t-statistic` = statistic) %>% 
+  knitr::kable()
+
+save(modtab, file = here('tabs/modtab.RData'))
+
 # supp1 mixef mod summary table ----------------------------------------------------------------
 
 load(file = here('data/mixmods.RData'))

### 2023-11-28 manu-draft.qmd
proofing methods, results, starting discussion

@@
 
-Tampa Bay is the largest open-water estuary in Florida covering 400 mi$^2$ (1,036 km$^2$) and one of the largest in the Gulf of Mexico.
+Tampa Bay is the largest open-water estuary in Florida covering 400 mi$^2$ (1,036 km$^2$) and the second largest in the Gulf of Mexico.
 The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee Rivers contributing a majority of freshwater inflow to the bay.
@@
 The geology of the watershed is rich in phosphates and mining activities have greatly altered the landscape, with notable spills and releases of wastewater that have affected water quality and biological resources [@garrett2011; @beck2022].
-Important subtidal habitats include seagrasses, hard bottom, tidal flats, and oyster reefs, where a majority of management effort has focused on restoring and maintaining seagrass cover [@sherwood2017].
-Additional native habitats include intertidal wetlands (mangroves, salt marshes, salt barriers) and pine forests, oak hammocks, and freshwater wetlands present in upland habitats [@Robison20].
+Important subtidal habitats include seagrasses, hard bottom, tidal flats, and oyster reefs, where a majority of management efforts have historically focused on restoring and maintaining seagrass cover [@sherwood2017].
+Additional native habitats include intertidal wetlands (mangroves, salt marshes, salt barrens) and pine forests, oak hammocks, and freshwater wetlands present in upland habitats [@Robison20].
 Losses of native uplands and potentially restorable habitats to development in the watershed from 1990 to 2020 have been estimated at 188,429 acres [76,254 ha, @beck2023].
@@
 Tampa Bay is divided into distinct sub-segments defined by physical and natural boundaries to assist with water quality management activities [@lewis1985]: Old Tampa Bay (OTB) in the northwest; Hillsborough Bay (HB) in the northeast; Middle Tampa Bay (MTB); and Lower Tampa Bay (LTB) that connects to the Gulf of Mexico.
-Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality conditions primarily from nutrient inputs from wastewater and stormwater [@greening2014].
+Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality primarily from nutrient inputs from wastewater and stormwater [@greening2014].
 Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia Rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small tributaries [@tbep0623].
@@
 
-The long-term recovery of seagrass habitats in Tampa Bay is a nationally-recognized success story that demonstrates application a successful management paradigm through the National Estuary Program [@greening2006; @greening2014; @sherwood2017].
-From 1988 to 2016, seagrasses increased 79% to a total areal cover of 41,655 acres (16,857 ha), surpassing the goal of restoring coverage to 95% of that which occurred in 1950 (Figure [-@fig-seagrasschg]a).
-Throughout this same period, nitrogen load estimates decreased by about 2/3 from their peak estimated in the mid-1970s as 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration projects in the watershed [@greening2014; @beck2019].
-These reductions in nutrient loadings resulted in largescale reductions in chlorophyll concentrations and light attenuation in the water column, creating favorable environments for seagrass growth.
-The most dramatic improvements in seagrass cover were observed in Old Tampa Bay where coverage increased by 122% or 4,465 acres (2,477 ha) to a total of 11,247 acres (4,511 ha) from 1988 estimates.
+The long-term recovery of seagrass habitats in Tampa Bay is a nationally-recognized success story that demonstrates application of a successful management paradigm through the National Estuary Program [@greening2006; @greening2014; @sherwood2017].
+From 1988 to 2016, seagrasses increased 79% to 41,655 acres (16,857 ha), surpassing the regional management goal of restoring coverage to 95% of a benchmark 1950s coverage (Figure [-@fig-seagrasschg]a).
+Throughout this same period, nitrogen load estimates decreased by about 2/3 from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration projects in the watershed [@greening2014; @beck2019].
+These reductions in nutrient loadings resulted in large scale reductions in chlorophyll concentrations and light attenuation in the water column, creating favorable environments for seagrass growth.
+The most dramatic improvements in seagrass cover were observed in Old Tampa Bay where coverage increased by 122% or 4,465 acres (2,477 ha) from 1988 estimates to a total of 11,247 acres (4,511 ha).
 Similar gains were observed in Middle Tampa Bay where seagrass cover increased by 86% or 4,465 acres (1,807 ha) to a total of 9,652 acres (3,906 ha) and Hillsborough Bay where cover increased from nearly zero acres to a total of 2,007 acres (810 ha).
@@
 
-From 2016 to present, dramatic losses of seagrasses have been observed in Tampa Bay, despite water quality conditions remaining relatively stable (Figure [-@fig-seagrasschg]a).
-Total cover in Tampa Bay has decreased by 28% from the 2016 peak by 11,518 acres (4,661 ha) to ta total of 30,137 acres (12,196 ha).
-Losses have been most pronounced in Old Tampa Bay (62%; 6,963acres/2,818 ha loss) and Hillsborough Bay (80%; 1,599acres/647 ha loss).
+From 2016 to present, dramatic losses of seagrasses have been observed in Tampa Bay, despite nutrient loading and light attenuation remaining relatively stable (Figure [-@fig-seagrasschg]a).
+Total cover in Tampa Bay has decreased by 28% from the 2016 peak by 11,518 acres (4,661 ha) to a total of 30,137 acres (12,196 ha).
+Losses have been most pronounced in Old Tampa Bay (62%; 6,963 acres/2,818 ha loss) and Hillsborough Bay (80%; 1,599 acres/647 ha loss).
 The current estimate for Old Tampa Bay of 4,183 acres (1,693 ha) is the lowest ever recorded in that bay segment since mapping efforts began in the 1980s.
-Similarly, coverage in Middle Tampa Bay decreased by 20% (1,926 acres/779 ha loss).
-Coverage in Lower Tampa Bay has remained stable, with only a 2% loss that is likely within the mapping error for the coverage estimates.
+Coverage in Middle Tampa Bay decreased by 20% (1,926 acres/779 ha loss), whereas coverage in Lower Tampa Bay has remained stable, with only a 2% loss that is likely within the range of mapping error for the coverage estimates.
 
@@
 As such, the transect data were used below for comparison with temperature and salinity changes for the major bay segments.
-Other sources of seagrass data are described in the next section.
+Additional sources of seagrass data are described in the next section.
 
@@
 These data provide the basis for regulatory assessments and compliance reporting for the nutrient TMDL in Tampa Bay.
-Water quality samples are collected at each station from surface water grabs (e.g., nutrients) or *in situ* measurements for physical parameters (e.g., salinity, temperature), where the latter includes measurements at the surface, mid-depth, and bottom.
+Water quality samples are collected at each station from surface water grabs (e.g., nutrients) or *in situ* measurements for physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom.
 Time of sampling can vary, although most samples are collected from mid-morning to early afternoon.
@@
 A stratified sample design is used to target multiple habitats where unique sites are sampled each month.
-We used data from the 21.3 meter center-bag nearshore seine that specifically targets shallow habitats where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present).
+We used data from the 21.3 meter center-bag nearshore seine that targets shallow habitats where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present).
 These data are collected by pulling the seine adjacent to the shore to sample approximately 140 m$^2$ of bay bottom [@schrandt2021].
@@
 Seagrass data are also provided for each site, with information on species and cover.
-Sites with greater than 50% cover of seagrass were identified as "seagrass" sites and those less than 50% or bare sediment were identified as "no seagrass" sites for comparison with temperature and salinity measurements.
+Total percent cover for all species at a site was used for comparison with temperature and salinity measurements.
 Sites exclusively with macroalgae were not included in the analysis.
-All FIM data from FWC staff by request.
+All FIM data were provided from FWC staff by request.
 
@@
 Surface waters in Pinellas county have been monitored since the 1990s, although a consistent stratified random sampling designed has been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES reporting.
-Data were obtained by recquest to PCDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (also available at \<https://wateratlas.usf.edu/\>).
+Data were obtained by request to PCDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (also available at \<https://wateratlas.usf.edu/\>).
 We focused primarily on OTB for the analysis of the PCDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments.
@@
 Water quality samples at each site are similar to those collected by EPC, where only bottom measurements for salinity and temperature were retained.
-Seagrass presence/absence is also recored at each site and all sites were defined as "seagrass" if only seagrass species were identified (any with macroalgae were excluded) and "no seagrass" if bare sediment was observed.
+Seagrass presence/absence is also recorded at each site and all sites were defined as "seagrass" if only seagrass species were identified (any with macroalgae were excluded) and "no seagrass" if bare sediment was observed.
 
@@
 An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions related to long-term (multi-decadal) climate change.
-Meteorological data describing air temperature and precipitation were obtained from Tampa International Airport where daily measurements have been collected since 1939.
+For comparison to water quality conditions, meteorological data describing air temperature and precipitation were obtained from Tampa International Airport where daily measurements have been collected since 1939.
 The *rnoaa* R package [@chamberlain23] was used to retrieve these data starting from 1975 when EPC monitoring began in Tampa Bay.
@@
 All climate and loading data were evaluated annually with simple linear regression trends to assess change over time.
-Water temperature and salinity trends using the EPC, FIM, and PDEM data were similarly evaluated by averaging the monthly data by year for each bay segment.
+Water temperature and salinity trends using the EPC, FIM, and PDEM data were similarly evaluated by averaging the monthly data each year for each bay segment.
 Linear trends for these data were evaluated based on averages of all stations and months within each year and bay segment, which allowed for comparable statistical power between datasets with different sample designs.
@@
 The change per year was also evaluated for each parameter based on the slope estimates returned by each test.
-Kendall tests were also used to evaluate changes over time for each month to determine when the trends more most pronounced seasonally, e.g., all January estimates across years, all February estimates, etc.
+Kendall tests were also used to evaluate changes over time for each month to determine when the trends were most pronounced seasonally, e.g., all January estimates across years, all February estimates, etc.
 The percentage of stations in each bay segment with significant increasing temperature or decreasing salinity trends were evaluated for each month.
@@
 
-The second goal of the analysis was to evaluate if seagrass changes were linked to long-term changes in water temparerature and salinity, with particular attention on differences between bay segments and the time periods before and after 2016 (pre/post recovery).
+The second goal of the analysis was to evaluate if seagrass changes were linked to long-term changes in water temperature and salinity, with particular attention on differences between bay segments and the time periods before and after 2016 (pre/post seagrass recovery).
 The conceptual model for evaluating these changes describes the niche space where seagrass growth and reproduction is hypothesized to be greatest within optimal ranges for forcing factors that are present in the environment [@hutchinson1957; @vandermeer1972].
-In the simplest form, this can conceptualized as a bell curve with optimal conditions defined with a range of values for a single parameter (e.g., minimum and maximum temperatures where a species is typically observed), where reduced growth or mortality is observed outside of these ranges.
-Because both water temperature and salinity are evaluated, the same model can be conceptualized in two-dimensional space [@fig-concept].
-Seagrass growth can be limited when temperature is below or above the optimum range, when salinity is below or above the optimum, or when both temperature and salinity conditions are outside of the optimum range.
+In the simplest form, this can conceptualized as a bell curve with optimal conditions defined within a range of values for a single parameter (e.g., minimum and maximum temperatures where a species is typically observed), where reduced growth or mortality is observed outside of these ranges.
+Because both water temperature and salinity were evaluated, the same model can be conceptualized in two-dimensional space [@fig-concept].
+Seagrass growth can be limited when temperature is below or above the optimum range, when salinity is below or above the optimum range, or when both temperature and salinity conditions are outside of the optimum range.
 Based on the results of the trend tests, we hypothesized that seagrasses are likely stressed by both high temperature and low salinity (bottom right @fig-concept).
@@
 
-A fundamental challenge in describing niche space is identifying the boundaries for optimal conditions.
+A fundamental challenge describing niche space is identifying the boundaries for optimal conditions.
 In Tampa Bay, the three dominant seagrass species are *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinium* (turtle grass) [@lewis1985].
@@
 Shoal grass is tolerant of a wide range of salinity [@lirman2003], but is more abundant in oligo/mesohaline portions of Tampa Bay.
-Conversely, turtle grass is less tolerant of low salinity and is most often found in more euryhaline conditions near the mouth of Tampa Bay.
+Conversely, turtle grass is less tolerant of low salinity and is more abundant in more euryhaline conditions near the mouth of Tampa Bay.
 Reported salinity ranges for each of these species varies depending on location, season, and other co-occurring factors like temperature [@phillips1960; @mcmillan1967; @zieman1975; @lewis1985], although most studies place lower limits of salinity in the range of 15-25 ppt.
@@
 Because of the uncertainty in defining thresholds for optimal temperature and salinity ranges, multiple thresholds were evaluated to both describe the potential for stress and how it may be related to changes in seagrass.
-Distinctions were not made between species, primarily due to lack of consensus between studies and the likely site-specific ranges that affect seagrass growth in Tampa Bay.
+Distinctions were not made between species, primarily due to lack of consensus between studies and likely site-specific ranges that affect seagrass growth in Tampa Bay.
 First, we developed metrics of potential temperature and salinity stress by quantifying the maximum number of continuous days each year when temperature was above or salinity was below a given threshold.
@@
 These metrics were quantified from the monthly long-term observations in the EPC data.
-To quantify daily counts each year, a continuous prediction of temperature and salinity over time at each of 45 stations was estimated using Generalized Additive Models (GAMs) fit to the response variable and a single predictor for decimal year [@wood17].
+To quantify daily counts each year, a continuous prediction of temperature and salinity over time at each of 45 stations was estimated using Generalized Additive Models (GAMs) fit to  temperature or salinity and a single predictor for decimal year [@wood17].
 The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends [@beck2022a].
@@
 
-Counts of the maximum number of days each year that temperature was above or salinity was below a thresold were obtained from the daily GAM predictions.
-This was done at each of 45 stations in the EPC data using temperature thresholds of 29, 30, and 31$^\circ$ C and salinity thresholds of 15, 20, and 25 ppt.
+Counts of the maximum continuous number of days each year that temperature was above or salinity was below a threshold were obtained from the daily GAM predictions.
+This was done at each of the 45 stations in the EPC data using temperature thresholds of 29, 30, and 31$^\circ$ C and salinity thresholds of 15, 20, and 25 ppt.
 The number of days when both temperature was above and salinity was below the thresholds was also estimated as a combined potential stress measure.
@@
 Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment.
-The stressor metrics were compared to frequency occurrence (all species) each year by bay segment(all species).
-Generalized linear regression models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre-2016, decline post-2016).
-Two models wre evaluated, one with the temperature and salinity metrics and another with the both metrics given that the latter was a combined metric of the former.
+The stressor metrics were compared to frequency occurrence (all species) each year by bay segment.
+Generalized Linear regression Models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre-2016, decline post-2016).
+Two models were evaluated, one with the temperature and salinity metrics and another with the both metric given that the latter was a combined metric of the former.
 Interactions were included between temperature, salinity, and time period or the both metric and time period depending on the model.
 Bay segment was included as a main effect without interactions.
-The most parsimonious models were identified by forward and backward variable selection and comparison with AIC values [@sakamoto1986; @venables2002b].
+The most parsimonious models were identified by forward and backward selection of the independent variables and comparison with AIC values [@sakamoto1986; @venables2002b].
 All models excluded Lower Tampa Bay because of minimal seagrass change over time.
@@
 Interactions with time period and a main effect for bay segment were also included as above.
-GLMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated (no bay segment interaction) and the seagrass response variable was presence/absence (i.e., a binomial distribution was used).
-Input data for models for the FIM and PDEM datasets were further subset to include only months from July to November to reduce potential seasonal effects.
+GLMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated (no bay segment interaction) and the seagrass response variable was presence/absence (i.e., a binomial distribution was used) to model the probability of observing seagrass.
+Input data for models for the FIM and PDEM datasets were further subset to include only months from July to November to describe seagrasses during the growing season and to reduce potential seasonal effects.
 
@@
 Note that for tables [-@tbl-temptrndtab] and [-@tbl-saltrndtab], comparable time periods were evaluated between the datasets when possible given the different sample sizes, and therefore power, to detect trends.
-The strongest trends were observed for the EPC dataset which had the longest record from 1975 to 2022.
+The strongest trends were observed for the EPC dataset, which had the longest record from 1975 to 2022.
 The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths and minimal stratification in Tampa Bay (@fig-meteowqraw).
@@
 An interaction of the metrics with time period was also observed, with notable differences by predictor (Figure [-@fig-sgmod]a, @tbl-modtab).
-For the period prior to 2016, frequency occurrence of seagrass increased as the number of days where temperature was above 30\$\^\\circ\$ increased, whereas no relationship was observed after 2016.
+For the period prior to 2016, frequency occurrence of seagrass increased as the number of days where temperature was above 30$^\circ$ increased, whereas no relationship was observed after 2016.
 However, the interaction between time period and temperature was not significant, suggesting no difference between time periods.
-The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (p \< 0.01), Figure [-@fig-sgmod]a, @tbl-modtab).
-The interaction between the temperature and salinity metric was included in the most parsimonious model (p \< 0.1), although the shape of the relationship of temperature with time period or salinity with time period did not noticeably change using different condition values for salinity or temperature, respectively.
-The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (`r epcmodsum2`), although the interaction term between the both metric and time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab).
+The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (p \< 0.01, Figure [-@fig-sgmod]a, @tbl-modtab).
+The interaction between the temperature and salinity metric was included in the most parsimonious model (marginally significant, p \< 0.1), although the shape of the relationship of temperature with time period or salinity with time period did not noticeably change using different conditional values for salinity or temperature, respectively.
+The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (`r epcmodsum2`), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab).
 
-The most parsimonious model for the FIM dataset also showed potential temperature and salinity associations with seagrass change after 2016, although explanatory power was lower than the EPC models likely due to the larger spread of data (`r fimmodsum`).
+The most parsimonious model for the FIM dataset also showed potential temperature and salinity associations with seagrass change after 2016, although explanatory power was lower than the EPC models likely due to larger spread of the data (`r fimmodsum`).
 Increases in temperature and decreases in salinity were both associated with reductions in percent cover after 2016, whereas no relationships were observed prior to 2016 (p \< 0.05 and p \< 0.1 for the interactions between temperature, salinity with time period, respectively, Figure [-@fig-sgmod]b, @tbl-modtab).
@@
 
-Takehome results: notable temp, salinity changes, some evidence of association with seagrass change.
+Global increases in temperature and precipitation related to climate change have contributed to widespread alteration of the structure and functioning of natural environments.
+At a local scale, these changes have been demonstrated herein for Tampa Bay where the evaluation of long-term trends in water temperature and salinity showed clear changes mirrored by long-term changes in air temperature and precipitation.
+Over the previous fifty years, the Bay has demonstrated a consistent trend towards hotter and fresher conditions, where water temperature has approximately increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year.
+These changes were demonstrated in three long-term datasets with different sampling methods and periods of record.
+Understandably, the trends were most consistently observed in the EPC dataset with the longest period of record covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season locations farther from the mouth of the Bay.
+These long-term changes manifested into consistent trends in the stressor metrics, where the continuous number of days each year when temperature was above, salinity was below, or both occurred showed consistent increases across the period of record.
+Notably, bottom and surface trends for water temperature and salinity in the EPC dataset were similar, suggesting that any potential stress on biotic resources related to these factors are likely occurring throughout the water column.
+
+Similar regional, long-term changes in coastal waters have been observed by others.
+@nickerson2023 evaluated sea surface trends at a larger spatial scale for Tampa Bay, the West Florida Continental Shelf, and the adjacent Gulf of Mexico.
+Temperature trends were similar for Tampa Bay (the EPC dataset was also used), although the authors note that the power of detecting the trends decreased for shorter time periods, as independently verified with the FIM and PDEM datasets herein.
+@nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, which somewhat conflicts with our results showing distinct increases in the summer, particularly for the OTB and HB segments.
+However, our assessment evaluated trends at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to the larger spatial area.
+These differences are understandable given the analysis goals and methods for data synthesis.
+Our results showing notable increases in temperature and decreases in salinity for OTB and HB for the summer months  is likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.
+These uppper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to shallower depths.
+Regardless, @nickerson2023 state a Gulf-wide increase in water temperature of 0.1 - 0.5 $^\circ$C per decade, which is comparable to our estimates.
+
+Our relatively simple modeling approach provides some evidence that climate-related stressors can explain the recent loss in seagrass in Tampa Bay.
+The models did not provide a consistent, nor statistically powerful, explanation that increasing temperature and decreasing salinity were key drives.
+However, a weight-of-evidence approach evaluating all models together does suggest a pattern that demonstrates the value of considering multiple datasets and models in explaining noisy ecological patterns.
+
+The seagrass loss in Tampa Bay since 2016 is a notable phenomenon that is not limited to our study area.
+Losses have been observed throughout southwest Florida during this time period, including...
+These regional losses suggest that large-scale stressors are driving these changes, both in Tampa Bay and elsewhere, leading us to our initial hypothesis that climate-related stressors could explain the change in Tampa Bay.
+and
+
+clearTakehome results: notable temp, salinity changes, some evidence of association with seagrass change.
 Of the limited evidence, effects of high temp, low salinity may just be additive, no support for interactions.
@@
 
-![Conceptual stressor diagram](figs/concept.png){#fig-concept}
+![Conceptual stressor diagram demonstrating a two-dimensional niche space for temperature and salinity.](figs/concept.png){#fig-concept}
 

### 2023-11-28 R/tabs.R

@@ -2,6 +2,7 @@ library(tidyverse)
 library(here)
 library(mgcv)
 library(broom)
+library(tbeptools)
 
 source(here('R/funcs.R'))
 

### 2023-11-29 manu-draft.qmd
more discussion

@@
 These changes were demonstrated in three long-term datasets with different sampling methods and periods of record.
-Understandably, the trends were most consistently observed in the EPC dataset with the longest period of record covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season locations farther from the mouth of the Bay.
+Understandably, the trends were most consistently observed in the EPC dataset with the longest period of record covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season at locations farther from the mouth of the Bay.
 These long-term changes manifested into consistent trends in the stressor metrics, where the continuous number of days each year when temperature was above, salinity was below, or both occurred showed consistent increases across the period of record.
@@
 These differences are understandable given the analysis goals and methods for data synthesis.
-Our results showing notable increases in temperature and decreases in salinity for OTB and HB for the summer months  is likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.
-These uppper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to shallower depths.
+Our results showing increases in temperature and decreases in salinity for OTB and HB for the summer months are likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.
+These upper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to more rapid warming with shallower depths.
 Regardless, @nickerson2023 state a Gulf-wide increase in water temperature of 0.1 - 0.5 $^\circ$C per decade, which is comparable to our estimates.
 
-Our relatively simple modeling approach provides some evidence that climate-related stressors can explain the recent loss in seagrass in Tampa Bay.
-The models did not provide a consistent, nor statistically powerful, explanation that increasing temperature and decreasing salinity were key drives.
+Our relatively simple modeling approach provided some evidence that climate-related stressors can explain the recent loss in seagrass in Tampa Bay.
+The models did not provide a consistent, nor statistically powerful, explanation that increasing temperature and decreasing salinity were key drivers.
 However, a weight-of-evidence approach evaluating all models together does suggest a pattern that demonstrates the value of considering multiple datasets and models in explaining noisy ecological patterns.
+Model results for the EPC and FIM datasets both suggested that increasing temperature and decreasing salinity were associated with potential seagrass loss post-2016, described primarily using separate interaction terms of temperature or salinity with time period.
+For the EPC model, the interaction term was not significant for temperature, whereas the interaction was marginally significant for the salinity metric such that a negative association was observed post-2016 as compared to pre-2016 where frequency occurrence of seagrass declined with the increasing number of days when salinity was below 25 ppt.
+Likewise, the FIM models had a significant interaction term for the association of the temperature with time period and a marginally significant interaction term for the association of salinity with time period, such that percent cover of seagrasses showed a decline with increasing temperature and decreasing salinity post-2016.
+An important distinction between the EPC and FIM models is that the former evaluated the number of days above/below thresholds each year to quantify increases in annual stress associated with temperature and salinity, whereas the latter evaluated observed temperature and salinity values at the time of seagrass sampling.
+As such, both models attempted to describe the role of these stressors on potential seagrass change, but use different independent variables given the different sampling designs of each monitoring programs.
+These differences highlight challenges describing noisy relationships in long-term ecological datasets, while also demonstrating the utility of our weight-of-evidence approach to describe these relationships.
+
+An additional caveat of our models was the use of "thresholds" to define potential stressor metrics for temperature and salinity on an annual time scale.
+Our choice to use 30$^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed, as shown by the mixed-effects linear models for each bay segment.
+That is, sufficient change and variation in the independendent variables for the models of seagrass change were needed to statistically describe potential relationships.
+These thresholds are not to be considered ecologically significant since most of the reported values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity.
+Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values (i.e., higher temperature, lower salinity), conditions in Tampa Bay in recent years are generally within the ecological niche for seagrasses.
+This does not suggest that these factors are unimportant, both currently and in the future.
+Our models suggested that temperature and salinity are potentially associated with seagrass loss and the consistent long-term trend in temperature and salinity suggest that conditions are very likely to become more stressful in the futre.
+Moreover, the GAMs used to model continuous temperature and salinity at each EPC monitoring station likely did not fully describe the current extent of stress given that extreme low or high values were often not captured by the daily predictions (see supplement).
+Regardless of the above limitations, the role of climate stressors on seagrass change is likely to continue in the future as suggested by our results.
+
+Additional limitations of our model may relate to an incomplete description of factors influencing seagrass growth.
+This relates to the inclusion of additional drivers as well as issues related to correlation/causation.
+For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, nitrogen in particular, in affecting light environments for seagrass growth.
+As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences could additionally be explained by temperature or salinity.
+Our use of time period (pre-/post-2016) as a categorical variable partially addressed this issue.
+Modeling seagrass change as related to temperature or salinity for the whole record without time period as a predictor would have shown a spurious correlation of both with seagrass given the long-term recovery of Tampa Bay related to nutrient reduction, increased water clarity, and increased seagrass coverage.
+Thus, time period was necessary to control for these confounding relationships.
+Additionally, light attenuation  has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay (OTB in particular where the most losses have been seen).
+This suggests that other factors are influencing seagrass growth and the inclusion of light attenuation in the models was likely not critical for describing the roles of temperature and salinity.
+A final consideration for our models relates to how seagrasses may influence their environment, particularly for the EPC and PDEM datasets where temperature and salinity were measured at the same locations as seagrass.
+For example, temperature may simply be higher in locations where seagrasses are present, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses.
+This explanation cannot be ruled out with the existing datasets, although the trend analyses showing the consistent long-term changes and our weight-of-evidence modeling approach suggest that climate-related stressors driving seagrass loss is a more likely scenario.
 
 The seagrass loss in Tampa Bay since 2016 is a notable phenomenon that is not limited to our study area.
-Losses have been observed throughout southwest Florida during this time period, including...
-These regional losses suggest that large-scale stressors are driving these changes, both in Tampa Bay and elsewhere, leading us to our initial hypothesis that climate-related stressors could explain the change in Tampa Bay.
-and
-
-clearTakehome results: notable temp, salinity changes, some evidence of association with seagrass change.
-Of the limited evidence, effects of high temp, low salinity may just be additive, no support for interactions.
-This was shown by the FIM models and lack of significance for the "both" EPC model.
-Also, note that the PDEM data may have had low power, limited spatially (only west side of bay segment) and even shorter time period (supp fig, sal trends support this).
+Losses have been observed throughout southwest Florida during this time period, including Sarasota Bay directly south of Tampa Bay and Charlotte Harbor further south.
+These regional losses suggest that large-scale stressors are driving these changes, leading us to our initial hypothesis that climate-related stressors could explain the change in Tampa Bay.
+Based on our results, the losses elsewhere may potentially be explained by temperature and salinity and are worth exploring in other southwest Florida coastal regions and estuaries where long-term datasets exist.
+Additional factors that could explain these changes are also likely co-occurring with climate stress, some of which are unique to Tampa Bay and others that are likely more pervasive.
+For Tampa Bay, annual summer/fall blooms of the toxic dinoflagellate *Pyrodinium bahamense* have occurred in OTB since 2008 and the specific relationships of these blooms with seagrass change is unclear, athough the expectation is that seagrass growth may be limited by the degradation of the light environment with algal growth.
+These blooms are exacerbated by the hydrologic conditions in OTB that contribute to relatively longer residence times, whereas the effect of warming temperature and decreasing salinity in OTB will further complicate the understanding of how these blooms manifest and persist each year.
+In Tampa Bay and elsewhere, the persistence of macroalgal species and how these may inluence seagrass growth and recovery h as also been a concern...
 
@@
 
-Role of additional stressors and why they weren't considered in the model - they're important but we can also demonstrate that light attenuation has been relatively stable over time, so may not be important to include in the models.
-
-Issues with thresholds, not definitive and those likely to stress seagrass are potentially not occurring at frequencies sufficient enough to model.
-Those used herein were chosen statistically.
-Point is that the temp and sal trends are in a direction such that thresholds that are more likely to stress seagrass are more likely to occur in the future.
-
-Case for continuous monitoring
-
-Are we just showing seagrasses are not present because of temperature limitations or if temperature indeed caused a reduction?
-
-Manamagement implication - we can't do much to control temp and salinity, but we can temper expectations for how to manage the bay.
-Adaptive capacity/resiliency is likely decreasing because of these "co-morbidities", so we have to temper expectations for other stressors that we can control.
-
-OTB is likely at a tipping point...
-
 Alternative nutrient pathways and role of macroalgae, competition for resources, shading, etc.
@@
 Optimal salinity range between 20-32 psu [@lopez2021].
+
+### Conclusions
+
+Case for continuous monitoring
+
+OTB is likely at a tipping point...
+
+Manamagement implication - we can't do much to control temp and salinity, but we can temper expectations for how to manage the bay.
+Adaptive capacity/resiliency is likely decreasing because of these "co-morbidities", so we have to temper expectations for other stressors that we can control.
 

### 2023-11-30 manu-draft.qmd
more discussion

@@
 Also noted the trend was most pronounced in the winter, see Fig 3.
+
+Preprint about warming SW Fl estuaries: <https://www.researchsquare.com/article/rs-3528678/v1>
 
@@
 Our choice to use 30$^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed, as shown by the mixed-effects linear models for each bay segment.
-That is, sufficient change and variation in the independendent variables for the models of seagrass change were needed to statistically describe potential relationships.
+That is, sufficient change and variation in the independent variables for the models of seagrass change were needed to statistically describe potential relationships.
 These thresholds are not to be considered ecologically significant since most of the reported values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity.
@@
 
-The seagrass loss in Tampa Bay since 2016 is a notable phenomenon that is not limited to our study area.
+The seagrass loss in Tampa Bay since 2016 is a notable phenomenon that is not limited to our study area [@lizcano-sandoval2022].
 Losses have been observed throughout southwest Florida during this time period, including Sarasota Bay directly south of Tampa Bay and Charlotte Harbor further south.
@@
 Additional factors that could explain these changes are also likely co-occurring with climate stress, some of which are unique to Tampa Bay and others that are likely more pervasive.
-For Tampa Bay, annual summer/fall blooms of the toxic dinoflagellate *Pyrodinium bahamense* have occurred in OTB since 2008 and the specific relationships of these blooms with seagrass change is unclear, athough the expectation is that seagrass growth may be limited by the degradation of the light environment with algal growth.
-These blooms are exacerbated by the hydrologic conditions in OTB that contribute to relatively longer residence times, whereas the effect of warming temperature and decreasing salinity in OTB will further complicate the understanding of how these blooms manifest and persist each year.
-In Tampa Bay and elsewhere, the persistence of macroalgal species and how these may inluence seagrass growth and recovery h as also been a concern...
+For Tampa Bay, annual summer/fall blooms of the toxic dinoflagellate *Pyrodinium bahamense* have occurred in OTB since 2008 [@lopez2023] and the specific relationships of these blooms with seagrass change is unclear, although the expectation is that seagrass growth may be limited by the degradation of the light environment with algal growth.
+These blooms are exacerbated by the hydrologic conditions in OTB that contribute to relatively longer residence times, whereas the effect of warming temperature and decreasing salinity in OTB will further complicate the understanding of how these blooms manifest and persist each year [@usup1994; @phlips2006; @lopez2021].
 
-Other areas showing seagrass loss - Florida Bay is different, less water flowing out of everglades and compounding SLR has elevated salinity and likely stress in other direction.
-Also, Biscayne Bay and IRL is a lot like OTB, poor flushing for example.
+Additional biotic factors could be influencing regional patterns in seagrass growth.
+In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern.
+Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present.
+Competitive differences between seagrasses and macroalgae are poorly understood in these systems, in addition to insufficient data to clearly describe growth patterns and nutrient cycling related to macroalgae.
+Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth \[@beck2022; @scolaro2023; @tomasko2023\], whereas the role of chronic nutrient loading and changing climatic conditions in promoting macroalgae growth and interactios with seagrasses is less understood, particularly in recent years.
+Finally, additional research has focused on how diseases and pathogens can influence seagrass growth patterns in Florida [@robblee1991; @duffin2021; @vanbogaert2018].
+For example, the parasitic slime mold *Labryinthula* spp.
+that causes seagrass wasting disease has been known to infect *Thalassia testudinium* in Tampa Bay [@blakesley2001], although it is unclear if these infections have had large-scale, population-level effects.
+Existing esearch has primarily focused on describing spatial patterns, past die-off events, or immunology of these pathogens [@robblee1991; @duffin2021].
 
-Viral/disease effects on seagrass change, [@vanbogaert2018] for TB and [@duffin2021] for Florida Bay.
-
-Alternative nutrient pathways and role of macroalgae, competition for resources, shading, etc.
-
-OTB seagrass loss and role of pyrodinium, optimal temperatures from 28-31 [@lopez2021].
-For salinity, highest growth rates at 24 psu and above, with decline in growth rates at 20 psu [@usup1994], in Florida growth has occurred in psu 10-45, bloom concentrations observed above 15 psu [@phlips2006; @lopez2021].
-Optimal salinity range between 20-32 psu [@lopez2021].
+Lastly, our result showing that salinity has decreased in Tampa Bay is generally contrary to expectations for coastal systems consistent with sea-level rise projections [@alarcon2024].
+Salinity changes with sea-level rise have been documented to contribute to numerous alterations of subtidal and nearshore habitats.
+In southwest Florida, the most common example is migration of tidal wetlands, primary mangroves, upland as porewater salinity and water levels have gradually changed over the past few decades.
+Alteration of salinity regimes for surface and groundwater resources are also well documented.
+In Florida Bay, for example, widespread decline of *Thalassia testudinium* has been partially attributed to elevated salinity levels beyond the optimal range for the species [@hall2016].
+Although drought has been implicated in the hypersaline conditions observed in Florida Bay, sea level rise is expected to further alter salinity dynamics in the region.
+@dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Greater Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters.
+In Tampa Bay, the long-term trends of decreasing salinity suggest that the hydrologic load into the system has had a greater influence on salinity regimes than the effects of sea-level rise.
+This hypothesis is supported by our assessment of precipitation patterns over time, where the long-term increase is inversely correlated with the decrease in salinity.
 
 ### Conclusions
+
+This study provided a detailed assessment of long-term water temperature and salinity changes in Tampa Bay supported by datasets from three long-term monitoring programs of different length and sampling design.
+An evaluation of each dataset showed a clear pattern of increasing temperature and decreasing salinity mirrored by long-term changes in air temperature and precipitation, suggesting that TampaBay has become hotter and fresher with the trends likely continuing in the future.
+Simple regression
 

### 2023-11-30 supplement.qmd
more discussion

@@
 
-![Relationship between water temperature and discharge in the Hillsborough River (USGS gage 02303000) over time.
+<!-- ![Relationship between water temperature and discharge in the Hillsborough River (USGS gage 02303000) over time.
 The relationship is shown for February (dry month) and August (wet month) grouped by historic (1975-1999) and contemporary (2000-2022) time periods.
-Regression lines with 95% confidence intervals for each temporal grouping are shown.](figs/suppflowtemp.png){#fig-suppflowtemp}
+Regression lines with 95% confidence intervals for each temporal grouping are shown.](figs/suppflowtemp.png){#fig-suppflowtemp} -->
 
-{{< pagebreak >}}
+<!-- {{< pagebreak >}} -->
 

### 2023-12-01 manu-draft.qmd
finish discussion text but need to add refs

@@
 Understandably, the trends were most consistently observed in the EPC dataset with the longest period of record covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season at locations farther from the mouth of the Bay.
-These long-term changes manifested into consistent trends in the stressor metrics, where the continuous number of days each year when temperature was above, salinity was below, or both occurred showed consistent increases across the period of record.
+These long-term changes manifested into consistent trends in the stressor metrics, where the continuous number of days each year when temperature was above, salinity was below, or both crossed thresholds showed consistent increases across the period of record.
 Notably, bottom and surface trends for water temperature and salinity in the EPC dataset were similar, suggesting that any potential stress on biotic resources related to these factors are likely occurring throughout the water column.
@@
 For Tampa Bay, annual summer/fall blooms of the toxic dinoflagellate *Pyrodinium bahamense* have occurred in OTB since 2008 [@lopez2023] and the specific relationships of these blooms with seagrass change is unclear, although the expectation is that seagrass growth may be limited by the degradation of the light environment with algal growth.
-These blooms are exacerbated by the hydrologic conditions in OTB that contribute to relatively longer residence times, whereas the effect of warming temperature and decreasing salinity in OTB will further complicate the understanding of how these blooms manifest and persist each year [@usup1994; @phlips2006; @lopez2021].
+These blooms are exacerbated by the hydrologic conditions in OTB that contribute to relatively longer residence times.
+The effect of warming temperature and decreasing salinity in OTB will further complicate the understanding of how these blooms manifest and persist each year [@usup1994; @phlips2006; @lopez2021].
 
@@
 Competitive differences between seagrasses and macroalgae are poorly understood in these systems, in addition to insufficient data to clearly describe growth patterns and nutrient cycling related to macroalgae.
-Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth \[@beck2022; @scolaro2023; @tomasko2023\], whereas the role of chronic nutrient loading and changing climatic conditions in promoting macroalgae growth and interactios with seagrasses is less understood, particularly in recent years.
+Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth \[@beck2022; @scolaro2023; @tomasko2023\], whereas the role of chronic nutrient loading and changing climatic conditions in promoting growth and interactions with seagrasses is less understood, particularly in recent years.
 Finally, additional research has focused on how diseases and pathogens can influence seagrass growth patterns in Florida [@robblee1991; @duffin2021; @vanbogaert2018].
@@
 that causes seagrass wasting disease has been known to infect *Thalassia testudinium* in Tampa Bay [@blakesley2001], although it is unclear if these infections have had large-scale, population-level effects.
-Existing esearch has primarily focused on describing spatial patterns, past die-off events, or immunology of these pathogens [@robblee1991; @duffin2021].
+Existing research has primarily focused on describing spatial patterns, past die-off events, or immunology of these pathogens [@robblee1991; @duffin2021].
 
@@
 Salinity changes with sea-level rise have been documented to contribute to numerous alterations of subtidal and nearshore habitats.
-In southwest Florida, the most common example is migration of tidal wetlands, primary mangroves, upland as porewater salinity and water levels have gradually changed over the past few decades.
+In southwest Florida, the most common ecological example is migration of tidal wetlands, primary mangroves, upland as porewater salinity and water levels have gradually changed over the past few decades.
 Alteration of salinity regimes for surface and groundwater resources are also well documented.
@@
 This study provided a detailed assessment of long-term water temperature and salinity changes in Tampa Bay supported by datasets from three long-term monitoring programs of different length and sampling design.
-An evaluation of each dataset showed a clear pattern of increasing temperature and decreasing salinity mirrored by long-term changes in air temperature and precipitation, suggesting that TampaBay has become hotter and fresher with the trends likely continuing in the future.
-Simple regression
+An evaluation of each dataset showed a clear pattern of increasing temperature and decreasing salinity mirrored by long-term changes in air temperature and precipitation, suggesting that Tampa Bay has become hotter and fresher with the trends likely continuing in the future.
+Simple regression models provided weak, but partially-supporting evidence that these changes can be linked to recent seagrass losses since 2016.
+Our models suggested that rising temperatures and decreasing salinities are additively affecting seagrasses, rather than multiplicatively as evidenced by lack of significant interactions of these stressors.
+Future analyses may show more significant associations between physical habitat conditions and seagrass change as the trends are very likely to continue to push seagrass further outside of their tolerance ranges.
+These analyses should be supported by additional data, particularly high-resolution continuous monitoring data that provide a more precise assessment of stress at multiple time-scales.
 
-Case for continuous monitoring
-
-OTB is likely at a tipping point...
-
-Manamagement implication - we can't do much to control temp and salinity, but we can temper expectations for how to manage the bay.
-Adaptive capacity/resiliency is likely decreasing because of these "co-morbidities", so we have to temper expectations for other stressors that we can control.
+Area managers should consider how these climate-related stressors may alter the effectiveness of intervention activities aimed at supporting ecological resources in Tampa Bay.
+Management actions that have historically been effective may have degraded capacity to support ecosystem status due to changing resilience of ecological resources with climate change.
+For example, nitrogen load reductions have been effective at restoring Tampa Bay, primarily due to wastewater load allocations considered to be within the nutrient assimilative capacity of Tampa Bay.
+As Tampa Bay becomes hotter and fresher, similar load allocations may no longer be effectively assimilated.
+Strategies can be pursued to reduce allowable load inputs from regulated entities and to create additional resilience through more aggressive management strategies (e.g., additional stormwater controls, hydrological modification).
+These considerations are especially critical for upper parts of Tampa Bay where a majority of seagrass loss has occurred and where trends have been most pronounced.
+Reversal of recent trends may be more likely to occur if aggressive actions and controls are pursued sooner rather than later, given the challenges of restoring degraded systems due to hysteresis.
 

### 2023-12-01 manu-draft.qmd
adding refs to discussion

@@
 Background: SST trends in Tampa Bay w/ EPC data, WFS, and deep GOM [@nickerson2023] showed that 1975 to 2022 trend for EPC was notable, but less so from 1998 to 2022.
-Also noted the trend was most pronounced in the winter, see Fig 3.
-
-Preprint about warming SW Fl estuaries: <https://www.researchsquare.com/article/rs-3528678/v1>
+Also noted the trend was most pronounced in the winter, see Fig 3
 
 Good CC references for FL: https://www.mdpi.com/2071-1050/15/14/11364
+
+el nino effects, see Greening and Janicki 2006
 
@@
 
-Global increases in temperature and precipitation related to climate change have contributed to widespread alteration of the structure and functioning of natural environments.
+Global increases in temperature and precipitation related to climate change have contributed to widespread alteration of the structure and functioning of natural environments [@osland2015; @oliver2018].
 At a local scale, these changes have been demonstrated herein for Tampa Bay where the evaluation of long-term trends in water temperature and salinity showed clear changes mirrored by long-term changes in air temperature and precipitation.
@@
 
-Similar regional, long-term changes in coastal waters have been observed by others.
+Similar regional, long-term changes in coastal waters and estuaries have been observed by others [@nickerson2023; @shiir].
 @nickerson2023 evaluated sea surface trends at a larger spatial scale for Tampa Bay, the West Florida Continental Shelf, and the adjacent Gulf of Mexico.
-Temperature trends were similar for Tampa Bay (the EPC dataset was also used), although the authors note that the power of detecting the trends decreased for shorter time periods, as independently verified with the FIM and PDEM datasets herein.
+Temperature trends were similar for Tampa Bay (the EPC dataset was also used), although the power to detect trends decreased for shorter time periods, as independently verified with the FIM and PDEM datasets herein.
 @nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, which somewhat conflicts with our results showing distinct increases in the summer, particularly for the OTB and HB segments.
@@
 These upper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to more rapid warming with shallower depths.
-Regardless, @nickerson2023 state a Gulf-wide increase in water temperature of 0.1 - 0.5 $^\circ$C per decade, which is comparable to our estimates.
+Additionally, @shiir provided a recent assessment of a 2023 heatwave in south Florida, supported by a 20-year trend assessment that suggested estuaries were warming at nearly double the rate of the Gulf of Mexico.
+The upper limit of our warming estimates for Tampa Bay is comparable.
 
@@
 The models did not provide a consistent, nor statistically powerful, explanation that increasing temperature and decreasing salinity were key drivers.
-However, a weight-of-evidence approach evaluating all models together does suggest a pattern that demonstrates the value of considering multiple datasets and models in explaining noisy ecological patterns.
+However, evaluating all models together does suggest a pattern that demonstrates the value of considering multiple datasets and models to explain noisy ecological patterns.
 Model results for the EPC and FIM datasets both suggested that increasing temperature and decreasing salinity were associated with potential seagrass loss post-2016, described primarily using separate interaction terms of temperature or salinity with time period.
@@
 That is, sufficient change and variation in the independent variables for the models of seagrass change were needed to statistically describe potential relationships.
-These thresholds are not to be considered ecologically significant since most of the reported values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity.
+These thresholds are likely not ecologically significant since most of the reported values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity [@phillips1960; @mcmillan1967; @zieman1975; @lirman2003].
 Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values (i.e., higher temperature, lower salinity), conditions in Tampa Bay in recent years are generally within the ecological niche for seagrasses.
@@
 This relates to the inclusion of additional drivers as well as issues related to correlation/causation.
-For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, nitrogen in particular, in affecting light environments for seagrass growth.
-As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences could additionally be explained by temperature or salinity.
+For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, nitrogen in particular, in affecting light environments for seagrass growth \[[@greening2014a; @sherwood2017a].
+As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences after accounting for light attenuation could additionally be explained by temperature or salinity.
 Our use of time period (pre-/post-2016) as a categorical variable partially addressed this issue.
-Modeling seagrass change as related to temperature or salinity for the whole record without time period as a predictor would have shown a spurious correlation of both with seagrass given the long-term recovery of Tampa Bay related to nutrient reduction, increased water clarity, and increased seagrass coverage.
+Modeling seagrass change as related to temperature or salinity for the entire record would have shown a spurious correlation of both with seagrass given the long-term recovery of seagrass.
 Thus, time period was necessary to control for these confounding relationships.
-Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay (OTB in particular where the most losses have been seen).
+Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay (OTB in particular where the most losses have been seen) \[@tbep0696; @Beck20a\].
 This suggests that other factors are influencing seagrass growth and the inclusion of light attenuation in the models was likely not critical for describing the roles of temperature and salinity.
@@
 For example, temperature may simply be higher in locations where seagrasses are present, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses.
-This explanation cannot be ruled out with the existing datasets, although the trend analyses showing the consistent long-term changes and our weight-of-evidence modeling approach suggest that climate-related stressors driving seagrass loss is a more likely scenario.
+This explanation cannot be ruled out with the existing datasets, although the trend analyses showing the consistent long-term changes and our models suggest that climate-related stressors driving seagrass loss is a more likely scenario.
 
@@
 Additional factors that could explain these changes are also likely co-occurring with climate stress, some of which are unique to Tampa Bay and others that are likely more pervasive.
-For Tampa Bay, annual summer/fall blooms of the toxic dinoflagellate *Pyrodinium bahamense* have occurred in OTB since 2008 [@lopez2023] and the specific relationships of these blooms with seagrass change is unclear, although the expectation is that seagrass growth may be limited by the degradation of the light environment with algal growth.
-These blooms are exacerbated by the hydrologic conditions in OTB that contribute to relatively longer residence times.
-The effect of warming temperature and decreasing salinity in OTB will further complicate the understanding of how these blooms manifest and persist each year [@usup1994; @phlips2006; @lopez2021].
+For Tampa Bay, annual summer/fall blooms of the toxic dinoflagellate *Pyrodinium bahamense* have occurred in OTB since 2008 [@usup1994; @lopez2023] and the specific relationships of these blooms with seagrass change is unclear, although the expectation is that seagrass growth may be limited by the degradation of the light environment with algal growth.
+These blooms are exacerbated by the hydrologic conditions in OTB that contribute to relatively longer residence times [@phlips2006; @lopez2021].
+The effect of warming temperature and decreasing salinity in OTB will further complicate the understanding of how these blooms manifest and persist each year [e.g., @stelling2023].
 
 Additional biotic factors could be influencing regional patterns in seagrass growth.
-In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern.
-Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present.
-Competitive differences between seagrasses and macroalgae are poorly understood in these systems, in addition to insufficient data to clearly describe growth patterns and nutrient cycling related to macroalgae.
+In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern [@jei2022; @scolaro2023].
+Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present in Tampa Bay \[@Beck20b\].
+Competitive differences between seagrasses and macroalgae are poorly understood in these systems \[but see @bell1997\], in addition to insufficient data in Tampa Bay to clearly describe growth patterns and nutrient cycling related to macroalgae.
 Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth \[@beck2022; @scolaro2023; @tomasko2023\], whereas the role of chronic nutrient loading and changing climatic conditions in promoting growth and interactions with seagrasses is less understood, particularly in recent years.
@@
 Lastly, our result showing that salinity has decreased in Tampa Bay is generally contrary to expectations for coastal systems consistent with sea-level rise projections [@alarcon2024].
-Salinity changes with sea-level rise have been documented to contribute to numerous alterations of subtidal and nearshore habitats.
-In southwest Florida, the most common ecological example is migration of tidal wetlands, primary mangroves, upland as porewater salinity and water levels have gradually changed over the past few decades.
+Salinity changes with sea-level rise have been documented to contribute to numerous alterations of subtidal and nearshore habitats [@brinson1995; @white2017].
+In southwest Florida, the most common ecological example is migration of tidal wetlands, primary mangroves, upland as porewater salinity and water levels have gradually changed over the past few decades [@borchert2018].
 Alteration of salinity regimes for surface and groundwater resources are also well documented.
@@
 
-Area managers should consider how these climate-related stressors may alter the effectiveness of intervention activities aimed at supporting ecological resources in Tampa Bay.
-Management actions that have historically been effective may have degraded capacity to support ecosystem status due to changing resilience of ecological resources with climate change.
-For example, nitrogen load reductions have been effective at restoring Tampa Bay, primarily due to wastewater load allocations considered to be within the nutrient assimilative capacity of Tampa Bay.
+Area managers should consider how these climate-related stressors may alter the effectiveness of intervention activities aimed at protecting ecological resources in Tampa Bay.
+Management actions that have historically been effective may have inadequate capacity to maintain ecosystem status due to changing resilience of ecological resources with climate change.
+For example, nitrogen load reductions have been effective at restoring Tampa Bay, primarily due to wastewater load allocations considered to be within the nutrient assimilative capacity of Tampa Bay [@greening2006; @greening2014].
 As Tampa Bay becomes hotter and fresher, similar load allocations may no longer be effectively assimilated.
-Strategies can be pursued to reduce allowable load inputs from regulated entities and to create additional resilience through more aggressive management strategies (e.g., additional stormwater controls, hydrological modification).
-These considerations are especially critical for upper parts of Tampa Bay where a majority of seagrass loss has occurred and where trends have been most pronounced.
+Strategies can be pursued to reduce allowable load inputs from regulated entities and to create additional resilience through more aggressive management strategies (e.g., additional stormwater controls, hydrological modifications).
+These considerations are especially critical for upper parts of Tampa Bay where a majority of seagrass loss has occurred and where temperature and salinity trends have been most pronounced.
 Reversal of recent trends may be more likely to occur if aggressive actions and controls are pursued sooner rather than later, given the challenges of restoring degraded systems due to hysteresis.
@@
 
-<!-- {{< pagebreak >}} -->
+This work benefited from constructive discussions with the Technical Advisory Committee of the Tampa Bay Estuary Program (TBEP).
+TBEP funding for this work stems from EPA Section 320 Grant Funds, and the TBEP’s local government partners (Hillsborough, Manatee, Pasco, and Pinellas Counties; the Cities of Clearwater, St.
+Petersburg, and Tampa; Tampa Bay Water; and the Southwest Florida Water Management District) through contributions to the operating budget.
+We are indebted to the hundreds of field and administrative staff for collecting and maintaining the valuable long-term monitoring data used in this study.
+
+{{< pagebreak >}}
 

### 2023-12-04 manu-draft.qmd
esco csl

@@
 bibliography: refs.bib
-# csl: limnology-and-oceanography.csl
+csl: estuaries-and-coasts.csl
 author:
@@
 This does not suggest that these factors are unimportant, both currently and in the future.
-Our models suggested that temperature and salinity are potentially associated with seagrass loss and the consistent long-term trend in temperature and salinity suggest that conditions are very likely to become more stressful in the futre.
+Our models suggested that temperature and salinity are potentially associated with seagrass loss and the consistent long-term trend in temperature and salinity suggest that conditions are very likely to become more stressful in the future.
 Moreover, the GAMs used to model continuous temperature and salinity at each EPC monitoring station likely did not fully describe the current extent of stress given that extreme low or high values were often not captured by the daily predictions (see supplement).
@@
 This relates to the inclusion of additional drivers as well as issues related to correlation/causation.
-For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, nitrogen in particular, in affecting light environments for seagrass growth \[[@greening2014a; @sherwood2017a].
+For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, nitrogen in particular, in affecting light environments for seagrass growth [@greening2014a; @sherwood2017a].
 As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences after accounting for light attenuation could additionally be explained by temperature or salinity.
@@
 Thus, time period was necessary to control for these confounding relationships.
-Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay (OTB in particular where the most losses have been seen) \[@tbep0696; @Beck20a\].
+Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay (OTB in particular where the most losses have been seen) [@tbep0696; @Beck20a].
 This suggests that other factors are influencing seagrass growth and the inclusion of light attenuation in the models was likely not critical for describing the roles of temperature and salinity.
@@
 In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern [@jei2022; @scolaro2023].
-Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present in Tampa Bay \[@Beck20b\].
-Competitive differences between seagrasses and macroalgae are poorly understood in these systems \[but see @bell1997\], in addition to insufficient data in Tampa Bay to clearly describe growth patterns and nutrient cycling related to macroalgae.
-Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth \[@beck2022; @scolaro2023; @tomasko2023\], whereas the role of chronic nutrient loading and changing climatic conditions in promoting growth and interactions with seagrasses is less understood, particularly in recent years.
+Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present in Tampa Bay [@Beck20b].
+Competitive differences between seagrasses and macroalgae are poorly understood in these systems [but see @bell1997], in addition to insufficient data in Tampa Bay to clearly describe growth patterns and nutrient cycling related to macroalgae.
+Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth [@beck2022; @scolaro2023; @tomasko2023], whereas the role of chronic nutrient loading and changing climatic conditions in promoting growth and interactions with seagrasses is less understood, particularly in recent years.
 Finally, additional research has focused on how diseases and pathogens can influence seagrass growth patterns in Florida [@robblee1991; @duffin2021; @vanbogaert2018].

### 2023-12-04 manu-draft.qmd
editing references and author list

@@
       correspondence: true
+  - Kerry Flaherty-Walia:
+      email: kfwalia@tbep.org
+      institute: [TBEP]
+      correspondence: false
+  - Sheila Scolaro:
+      email: sscolaro@tbep.org
+      institute: [TBEP]
+      correspondence: false
+  - Maya C Burke:
+      email: mburke@tbep.org
+      institute: [TBEP]
+      correspondence: false
+  - Edward T Sherwood:
+      email: esherwood@tbep.org
+      institute: [TBEP]
+      correspondence: false
 
 institute:
-  - TBEP: Tampa Bay Estuary Program, St. Petersburg, Florida 33701 USA 16802 USA
+  - TBEP: Tampa Bay Estuary Program, 263 13th Ave S., Suite 350, St. Petersburg, Florida 33701 USA
 
@@
 The watershed is heavily developed and includes over 3 million people with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023].
-The geology of the watershed is rich in phosphates and mining activities have greatly altered the landscape, with notable spills and releases of wastewater that have affected water quality and biological resources [@garrett2011; @beck2022].
+The geology of the watershed is rich in phosphates and mining activities have greatly altered the landscape, with notable spills and releases of wastewater that have affected water quality and biological resources [@garrett2011; @beck2022a].
 Important subtidal habitats include seagrasses, hard bottom, tidal flats, and oyster reefs, where a majority of management efforts have historically focused on restoring and maintaining seagrass cover [@sherwood2017].
@@
 Surface waters in Pinellas county have been monitored since the 1990s, although a consistent stratified random sampling designed has been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES reporting.
-Data were obtained by request to PCDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (also available at \<https://wateratlas.usf.edu/\>).
+Data were obtained by request to PCDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (also available at <https://wateratlas.usf.edu/>).
 We focused primarily on OTB for the analysis of the PCDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments.
@@
 To quantify daily counts each year, a continuous prediction of temperature and salinity over time at each of 45 stations was estimated using Generalized Additive Models (GAMs) fit to temperature or salinity and a single predictor for decimal year [@wood17].
-The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends [@beck2022a].
+The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends.
 Model fit for each station was sufficient to calculate daily predictions for assessing the stressor metrics (see supplement, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models).
@@
 Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`).
-Mean annual air temperature in 1975 was `r airtempstr`, whereas current mean annual air temperature in 2022 was `r airtempend`, showing an overall increase in the period of record of `r airtempdif` $^\circ$.
+Mean annual air temperature in 1975 was `r airtempstr`, whereas current mean annual air temperature in 2022 was `r airtempend`, showing an overall increase in the period of record of `r airtempdif` $^\circ$C.
 Similarly, total precipitation has increased by `r prpslo` mm per year (`r prppvl`, $R^2$ `r prprsq`).
@@
 This relates to the inclusion of additional drivers as well as issues related to correlation/causation.
-For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, nitrogen in particular, in affecting light environments for seagrass growth [@greening2014a; @sherwood2017a].
+For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, nitrogen in particular, in affecting light environments for seagrass growth [@greening2014; @sherwood2017].
 As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences after accounting for light attenuation could additionally be explained by temperature or salinity.
@@
 Competitive differences between seagrasses and macroalgae are poorly understood in these systems [but see @bell1997], in addition to insufficient data in Tampa Bay to clearly describe growth patterns and nutrient cycling related to macroalgae.
-Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth [@beck2022; @scolaro2023; @tomasko2023], whereas the role of chronic nutrient loading and changing climatic conditions in promoting growth and interactions with seagrasses is less understood, particularly in recent years.
+Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth [@beck2022a; @scolaro2023; @tomasko2023], whereas the role of chronic nutrient loading and changing climatic conditions in promoting growth and interactions with seagrasses is less understood, particularly in recent years.
 Finally, additional research has focused on how diseases and pathogens can influence seagrass growth patterns in Florida [@robblee1991; @duffin2021; @vanbogaert2018].
@@
 This work benefited from constructive discussions with the Technical Advisory Committee of the Tampa Bay Estuary Program (TBEP).
-TBEP funding for this work stems from EPA Section 320 Grant Funds, and the TBEP’s local government partners (Hillsborough, Manatee, Pasco, and Pinellas Counties; the Cities of Clearwater, St.
+TBEP funding for this work stems from EPA Section 320 Grant Funds, and the TBEP's local government partners (Hillsborough, Manatee, Pasco, and Pinellas Counties; the Cities of Clearwater, St.
 Petersburg, and Tampa; Tampa Bay Water; and the Southwest Florida Water Management District) through contributions to the operating budget.

### 2023-12-05 manu-draft.qmd
added map figure

@@
 
-Tampa Bay is divided into distinct sub-segments defined by physical and natural boundaries to assist with water quality management activities [@lewis1985]: Old Tampa Bay (OTB) in the northwest; Hillsborough Bay (HB) in the northeast; Middle Tampa Bay (MTB); and Lower Tampa Bay (LTB) that connects to the Gulf of Mexico.
+Tampa Bay is divided into distinct sub-segments defined by physical and natural boundaries to assist with water quality management activities [@lewis1985]: Old Tampa Bay (OTB) in the northwest; Hillsborough Bay (HB) in the northeast; Middle Tampa Bay (MTB); and Lower Tampa Bay (LTB) that connects to the Gulf of Mexico (Figure [-@fig-map]a).
 Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality primarily from nutrient inputs from wastewater and stormwater [@greening2014].
@@
 These datasets were evaluated individually to assess trends and relationships with seagrass change to provide a weight-of-evidence approach for potential causal relationships driving the recent decline.
-First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at 45 stations in the major bay segments since the late 1970s.
+First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at 45 stations in the major bay segments since the late 1970s (Figure [-@fig-map]b).
 These data provide the basis for regulatory assessments and compliance reporting for the nutrient TMDL in Tampa Bay.
@@
 A stratified sample design is used to target multiple habitats where unique sites are sampled each month.
-We used data from the 21.3 meter center-bag nearshore seine that targets shallow habitats where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present).
+We used data from the 21.3 meter center-bag nearshore seine that targets shallow habitats where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present, Figure [-@fig-map]c).
 These data are collected by pulling the seine adjacent to the shore to sample approximately 140 m$^2$ of bay bottom [@schrandt2021].
@@
 Data were obtained by request to PCDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (also available at <https://wateratlas.usf.edu/>).
-We focused primarily on OTB for the analysis of the PCDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments.
+We focused primarily on OTB for the analysis of the PCDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments (Figure [-@fig-map]d).
 Four distinct spatial zones in OTB are used to stratify the random selection of sample points that typically include 4 sample points per month in each zone.
@@
 
+![Map of Tampa Bay and the three datasets used for trend analysis, including (a) bay segments, 2022 seagrass coverage (green), and transect starting points (red), (b) Environmental Protection Commissions (EPC) long-term monitoring sites, (c) Fisheries Independent Monitoring (FIM) random sampling for near-shore seine hauls, and (d) OTB portion of Pinellas County Department of Environmental Management (PDEM) random sampling.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/map.png){#fig-map}
+
+{{< pagebreak >}}
+
 ![Seagrass changes over time in Tampa Bay for (a) areal coverage and (b) frequency occurrence of major species.

### 2023-12-05 R/figs.R

@@ -18,6 +18,7 @@ library(visreg)
 library(car)
 library(mgcv)
 library(mixmeta)
+library(maps)
 
 source(here('R/funcs.R'))
 
@@ -30,6 +31,132 @@ tempthr <- c(29, 30, 31)
 tempcol <- c('coral', 'red2', 'darkred')
 salithr <- c(20, 25, 30)
 salicol <- c('navyblue', 'dodgerblue2', 'slategray3')
+
+# map -----------------------------------------------------------------------------------------
+
+fl <- paste0(tempdir(), '/sgdat2022.RData')
+download.file('https://github.com/tbep-tech/hmpu-workflow/raw/master/data/sgdat2022.RData', destfile = fl)
+load(file = fl)
+
+data(pincotemp)
+data(fimsgtempdat)
+data(epcdata)
+
+sgdat <- sgdat2022 %>% 
+  filter(FLUCCSCODE %in% c(9113, 9116)) %>% 
+  st_simplify(5, preserveTopology = F)
+
+flpoly <- map_data('state', 'florida') %>% 
+  st_as_sf(coords = c('long', 'lat'), crs = 4326) %>% 
+  summarise(geometry = st_combine(geometry)) %>%
+  st_cast("POLYGON")
+
+bbox <- st_bbox(tbseg)
+
+minset <- ggplot() + 
+  geom_sf(data = flpoly, fill = 'grey', color = NA) + 
+  geom_sf(data = st_as_sfc(bbox), fill = NA, color = 'black', linewidth = 1) + 
+  theme_void() +
+  theme( 
+    panel.background = element_rect(fill = '#FFFFFF', colour = 'white'), 
+    panel.border = element_rect(colour = 'black', fill = 'transparent')
+  ) 
+
+segcent <- tbseg %>% 
+  st_centroid() 
+
+thm <- theme(
+  panel.grid = element_blank(), 
+  axis.title = element_blank(), 
+  axis.text.y = element_text(size = 6),
+  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 6), 
+  axis.ticks = element_blank()
+)
+
+m1 <- ggplot() + 
+  ggspatial::annotation_map_tile(zoom = 11, type = 'cartolight', cachedir = system.file("rosm.cache", package = "ggspatial")) +
[…]

### 2023-12-06 manu-draft.qmd
working on intro

@@
 
--   Importance of seagrass as foundation species
--   Tampa Bay context
--   Climate change stressors
--   Goals/objectives
-    -   Describe
+The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale.
+Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services[@fourqurean2012; @orth2020; @orth2023].
+Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ in areal cover since the 1980s [@waycott2009].
+These losses are comparable, if not more, than other critical coastal environments, which represents an existential threat to the natural and economic services that these habitats provide.
+Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities, whereas natural disturbances such as storm events and disease have also been implicated.
+Management actions and interventions aimed at mitigating loss and ultimately supporting restoration require an adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline.
 
-Background: SST trends in Tampa Bay w/ EPC data, WFS, and deep GOM [@nickerson2023] showed that 1975 to 2022 trend for EPC was notable, but less so from 1998 to 2022.
-Also noted the trend was most pronounced in the winter, see Fig 3
+The sustained coverage or restoration of seagrasses by means other than planting requires management actions that create suitable conditions supportive for growth.
+A long-standing approach adopted by numerous management entities has been the control of nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses.
+There are limited examples of successful recovery of seagrass through control of nutrient inputs, with notable exceptions including Tampa Bay\...
+examples in Orth 2023.
+Lack of recovery of seagrasses through nutrient controls alone suggests insufficient reductions of nutrient inputs to promote restoration, such as for ecosystems where hysteresis may substantially limit the reverse of a trajectory deline.
+For coastal systems More likely, the role of alternative stressors may also be inadequately considered and addressed in management paradigms.
 
-Good CC references for FL: https://www.mdpi.com/2071-1050/15/14/11364
+Climate change has complicated the understanding of ecosystem response to conventational stressors, presenting new challenges and expectations for how ecological resources will respond to management actions.
+rising temperature
 
@@
 
+This study focuses on Tampa Bay...
+
+Background: SST trends in Tampa Bay w/ EPC data, WFS, and deep GOM [@nickerson2023] showed that 1975 to 2022 trend for EPC was notable, but less so from 1998 to 2022.
+Also noted the trend was most pronounced in the winter, see Fig 3
+
+Good CC references for FL: https://www.mdpi.com/2071-1050/15/14/11364
+
 @lewis1985 review of seagrass in Tampa Bay.
@@
 
-Focus on two time periods: full record and recent decline
+Temp background for TB: NOAA coastwatch trends
 
-Focus on two physical parameters: temperature and salinity
-
-Temp background for TB: NOAA coastwatch trends
+This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay Estuary over the last fifty years.
+Three datasets of varying sample design and temporal coverage were used to assess the primary hypothesis that conditions in Tampa Bay are trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges.
+This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring data describing cover of individual species, and synoptic seagrass data collected with the water quality data.
+Specific focus was on the period from prior to and after 2016, when seagrasses recovered baywide and then declined by 28% despite water quality conditions remaining relatively stable, suggesting that additional stressors beyond light requirements may explain the recent decline.
+The long-term recovery of Tampa Bay as primarily shown through the recovery of seagrasses to 2016 is a nationally recognized success story.
+This analysis provides a more recent assessment of current conditions by examining climate-related stressors and seagrass loss.
+Our primary goal was to relate recent seagrass changes to climate stressors to provide a motivating example for the management community in the Tampa Bay region that a shifting ecological baseline presents new challenges beyond the known success story.
+The results are also provided as a cautionary example of the subtle but increasing role of climate-related stressors on the resiliency of estuarine systems, prompting a reassessment of existing management paradigms to accommodate current and anticipated future conditions.
 

### 2023-12-06 R/figs.R

@@ -55,7 +55,7 @@ bbox <- st_bbox(tbseg)
 
 minset <- ggplot() + 
   geom_sf(data = flpoly, fill = 'grey', color = NA) + 
-  geom_sf(data = st_as_sfc(bbox), fill = NA, color = 'black', linewidth = 1) + 
+  geom_sf(data = st_as_sfc(bbox), fill = NA, color = 'black', linewidth = 0.5) + 
   theme_void() +
   theme( 
     panel.background = element_rect(fill = '#FFFFFF', colour = 'white'), 

### 2023-12-07 manu-draft.qmd
into draft almost done but needs work

@@
 Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities, whereas natural disturbances such as storm events and disease have also been implicated.
-Management actions and interventions aimed at mitigating loss and ultimately supporting restoration require an adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline.
+Management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].
 
-The sustained coverage or restoration of seagrasses by means other than planting requires management actions that create suitable conditions supportive for growth.
+The sustained coverage or restoration of seagrasses in stressed environments requires management actions that create suitable conditions supportive for growth.
 A long-standing approach adopted by numerous management entities has been the control of nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses.
-There are limited examples of successful recovery of seagrass through control of nutrient inputs, with notable exceptions including Tampa Bay\...
-examples in Orth 2023.
-Lack of recovery of seagrasses through nutrient controls alone suggests insufficient reductions of nutrient inputs to promote restoration, such as for ecosystems where hysteresis may substantially limit the reverse of a trajectory deline.
-For coastal systems More likely, the role of alternative stressors may also be inadequately considered and addressed in management paradigms.
+There are limited examples of successful recovery of seagrass through control of nutrient inputs alone, primarily because of the difficulty in identifying and regulating both point and diffuse non-point sources.
+Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014]and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in areal coverage through sustained and long-term reductions in nutrients.
+Cooperation among management, regulatory, and private sectors were critical aspects of both these examples for successful implementation of nutrient controls.
+In other cases, the reversal of seagrass losses through nutrient reductions may not be possible because of system hysteresis where the path to recovery is not the same as the path to decline.
+Costly alternative management actions in addition to nutrient reductions are needed in these situations.
+Management actions are much more effective in estuarine sytems that are stressed but not past critical thresholds that define an alternative state.
 
-Climate change has complicated the understanding of ecosystem response to conventational stressors, presenting new challenges and expectations for how ecological resources will respond to management actions.
-rising temperature
+Climate change has complicated the understanding of ecosystem response to conventional stressors, presenting new challenges and expectations for how ecological resources will respond to management actions.
+In addition to sea-level rise, the most anticipated effects of climate change in coastal environments are increasing temperature and alteration of precipitation patterns.
+These changes will profoundly alter physical habitats, creating suboptimal or uninhabitable conditions for many species.
+@lefcheck2017 evaluated interactions between water clarity and rising temperatures with over 30 years of data in Chesapeake Bay.
+The effects of decreased water clarity for seagrasses in shallow habitats were exacerbated by acute warming related to climate change, demonstrating a concerning synergy of multiple stressors.
+Species shifts are also expected to occur as changing physical conditions decrease the competitive advantages of historically abundant species.
+Again in Chesapeake Bay, the abundance of widgeongrass (*Ruppia maritima*) has been mediated by nutrient reductions, while also outcompeting and replacing formerly dominant eelgrass (*Zostera marina*) given higher tolerance to increased temperatures [@hensel2023].
+Changing frequency and severity of precipicitation patterns may further alter the distribution and abundance of seagrasses by pushing species outside of optimal salinity tolerance ranges.
+These changes may produce hypersaline conditions under periods of prolonged drought or more freshwater conditions with increased storm events.
+Seagrasses may respond differently under individual or multiple stressors [@lefcheck2017; @zhang2023] , suggesting place-based assessments of response to climate change provide valuable insights for managers responsible for maintaining ecological resources.
 
-el nino effects, see Greening and Janicki 2006
-
-Heat stress combined with highlight accelerates decline of *E. acoroides*, used 36C [@zhang2023]
-
-Widgeongrass colonization in Chesapeake Bay as an opportunistic, heat-tolerant species that has replaced Eelgrass, although former is sensitive to nutrient pulses.
-Paper provides an example of implications on climate change and nutrient management on foundation species and system resiliency [@hensel2023].
-
-This study focuses on Tampa Bay...
-
-Background: SST trends in Tampa Bay w/ EPC data, WFS, and deep GOM [@nickerson2023] showed that 1975 to 2022 trend for EPC was notable, but less so from 1998 to 2022.
-Also noted the trend was most pronounced in the winter, see Fig 3
-
-Good CC references for FL: https://www.mdpi.com/2071-1050/15/14/11364
-
-@lewis1985 review of seagrass in Tampa Bay.
-@lirman2003 conducted exposure experiments to evaluate seagrass growth in response to a range of salinity conditions.
-@mcmillan1967 discusses growth of halodule, syringodium, thalassia, and ruppia in response to salinity increases (up to 75 psu), no info on lower limit.
-Cites @phillips1960 for a salinity range of Thalassia in Florida of 33 - 38 psu.
-@zieman1975 discusses seasonal variation of thalassia relative to temp and salinity
-
-Temp background for TB: NOAA coastwatch trends
+In Tampa Bay, Florida, seagrasses are a primary indicator of bay health that have been the focus of management efforts over the last thirty years.
+Through successful reduction of nitrogen loads by 2/3 from , seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres.
+Seagrass distribution and abundance has been dynamic throughout this period, overall responding positively to increases in water clarity with nutrient load reductions, while also demonstrating more short-term variability in response to regional climate events [@greening2006].
+Since 2016, seagrass areal cover has decreased by nearly 1/3 despite relatively stable water quality with adequate light penetration that has been historically shown to support growth.
+Factors that have influenced this recent decline are unknown and the effects of climate change on physical conditions have been implicated as potential drivers.
+Recent work has demonstrated a broad long-term trend of increasing water temperature in Tampa Bay [@nickerson2023], whereas changing salinity has not been well-described, nor have potential links to seagrass change.
+Tampa Bay is rich with historical data that can be used to evaluate long-term trends and to develop models that can explain relationships of seagrass change with water quality conditions.
+This information can fill a critical knowledge gap that can inform regional management activities, while also demonstrating the confounding effects of climate change with ecosystem response to conventional stressors.
 

### 2023-12-08 manu-draft.qmd
text edits to intro, adding cites

@@
 
+{{< pagebreak >}}
+
 ## Introduction
@@
 Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services[@fourqurean2012; @orth2020; @orth2023].
-Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ in areal cover since the 1980s [@waycott2009].
-These losses are comparable, if not more, than other critical coastal environments, which represents an existential threat to the natural and economic services that these habitats provide.
-Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities, whereas natural disturbances such as storm events and disease have also been implicated.
+Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021].
+These losses are comparable, if not more, than other critical coastal environments [@duarte2008a], which represents an existential threat to the natural and economic services that these habitats provide.
+Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @orth2006; @burkholder2007], whereas natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020].
 Management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].
 
-The sustained coverage or restoration of seagrasses in stressed environments requires management actions that create suitable conditions supportive for growth.
-A long-standing approach adopted by numerous management entities has been the control of nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses.
+The sustained coverage or restoration of seagrasses in stressed environments requires management actions that create conditions supportive for growth.
+A long-standing approach adopted by numerous management entities has been the control of nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @greening2014] .
 There are limited examples of successful recovery of seagrass through control of nutrient inputs alone, primarily because of the difficulty in identifying and regulating both point and diffuse non-point sources.
 Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014]and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in areal coverage through sustained and long-term reductions in nutrients.
-Cooperation among management, regulatory, and private sectors were critical aspects of both these examples for successful implementation of nutrient controls.
-In other cases, the reversal of seagrass losses through nutrient reductions may not be possible because of system hysteresis where the path to recovery is not the same as the path to decline.
-Costly alternative management actions in addition to nutrient reductions are needed in these situations.
-Management actions are much more effective in estuarine sytems that are stressed but not past critical thresholds that define an alternative state.
+Cooperation among management, regulatory, and private sectors were critical aspects of both these examples for successful implementation of nutrient controls [@tango2016; @sherwood2016].
+In other cases, the reversal of seagrass losses through nutrient reductions may not be possible because of system hysteresis, where the path to recovery is not the same as the path to decline [@maxwell2016].
+Costly alternative management actions, in addition to nutrient reductions, are needed in these situations.
+Management actions are much more effective in estuaries that are stressed but not past critical thresholds that define an alternative state.
 
@@
 In Tampa Bay, Florida, seagrasses are a primary indicator of bay health that have been the focus of management efforts over the last thirty years.
-Through successful reduction of nitrogen loads by 2/3 from , seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres.
+Through successful reduction of nitrogen loads by 2/3 pre-regulation, seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres (16,857 ha).
 Seagrass distribution and abundance has been dynamic throughout this period, overall responding positively to increases in water clarity with nutrient load reductions, while also demonstrating more short-term variability in response to regional climate events [@greening2006].
@@
 Factors that have influenced this recent decline are unknown and the effects of climate change on physical conditions have been implicated as potential drivers.
-Recent work has demonstrated a broad long-term trend of increasing water temperature in Tampa Bay [@nickerson2023], whereas changing salinity has not been well-described, nor have potential links to seagrass change.
+Following global trends, recent work has demonstrated a broad long-term trend of increasing water temperature in Tampa Bay [@nickerson2023].
+Changing salinity has not been well-described, nor have potential links to seagrass change.
 Tampa Bay is rich with historical data that can be used to evaluate long-term trends and to develop models that can explain relationships of seagrass change with water quality conditions.
@@
 Three datasets of varying sample design and temporal coverage were used to assess the primary hypothesis that conditions in Tampa Bay are trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges.
-This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring data describing cover of individual species, and synoptic seagrass data collected with the water quality data.
-Specific focus was on the period from prior to and after 2016, when seagrasses recovered baywide and then declined by 28% despite water quality conditions remaining relatively stable, suggesting that additional stressors beyond light requirements may explain the recent decline.
-The long-term recovery of Tampa Bay as primarily shown through the recovery of seagrasses to 2016 is a nationally recognized success story.
-This analysis provides a more recent assessment of current conditions by examining climate-related stressors and seagrass loss.
-Our primary goal was to relate recent seagrass changes to climate stressors to provide a motivating example for the management community in the Tampa Bay region that a shifting ecological baseline presents new challenges beyond the known success story.
+This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring data describing cover of individual species, and synoptic seagrass data collected with the water quality surveys.
+Specific focus was on the period from prior to and after 2016, when seagrasses recovered baywide and then declined despite water quality conditions remaining relatively stable.
+Our primary goal was to relate recent seagrass changes to climate stressors to provide a motivating example for the management community in the Tampa Bay region that a shifting ecological baseline presents new challenges beyond the known success story of long-term bay recovery.
 The results are also provided as a cautionary example of the subtle but increasing role of climate-related stressors on the resiliency of estuarine systems, prompting a reassessment of existing management paradigms to accommodate current and anticipated future conditions.

### 2023-12-10 manu-draft.qmd
intro draft done complete with cites

@@
 
-The sustained coverage or restoration of seagrasses in stressed environments requires management actions that create conditions supportive for growth.
+The sustained coverage or restoration of seagrasses in stressed environments requires management actions that create supportive conditions for growth.
 A long-standing approach adopted by numerous management entities has been the control of nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @greening2014] .
@@
 Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014]and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in areal coverage through sustained and long-term reductions in nutrients.
-Cooperation among management, regulatory, and private sectors were critical aspects of both these examples for successful implementation of nutrient controls [@tango2016; @sherwood2016].
-In other cases, the reversal of seagrass losses through nutrient reductions may not be possible because of system hysteresis, where the path to recovery is not the same as the path to decline [@maxwell2016].
+Cooperation among management, regulatory, public, and private sectors were critical aspects of both examples [@tango2016; @sherwood2016].
+In other cases, the reversal of seagrass losses through nutrient reductions alone may not be possible because of system hysteresis, where the path to recovery is not the same as the path to decline [@maxwell2016].
 Costly alternative management actions, in addition to nutrient reductions, are needed in these situations.
@@
 
-Climate change has complicated the understanding of ecosystem response to conventional stressors, presenting new challenges and expectations for how ecological resources will respond to management actions.
+Climate change has complicated the understanding of ecosystem response to conventional stressors, presenting new challenges and expectations for how ecological resources will respond to management actions [@statham2012a; @sherwood2013].
 In addition to sea-level rise, the most anticipated effects of climate change in coastal environments are increasing temperature and alteration of precipitation patterns.
-These changes will profoundly alter physical habitats, creating suboptimal or uninhabitable conditions for many species.
-@lefcheck2017 evaluated interactions between water clarity and rising temperatures with over 30 years of data in Chesapeake Bay.
+These changes will profoundly alter physical habitats, creating suboptimal or uninhabitable conditions for many species [@madeira2012; @lefcheck2017; @hensel2023].
+@lefcheck2017 evaluated interactive effects of water clarity and rising temperatures on seagrasses with over 30 years of data in Chesapeake Bay.
 The effects of decreased water clarity for seagrasses in shallow habitats were exacerbated by acute warming related to climate change, demonstrating a concerning synergy of multiple stressors.
@@
 Again in Chesapeake Bay, the abundance of widgeongrass (*Ruppia maritima*) has been mediated by nutrient reductions, while also outcompeting and replacing formerly dominant eelgrass (*Zostera marina*) given higher tolerance to increased temperatures [@hensel2023].
-Changing frequency and severity of precipicitation patterns may further alter the distribution and abundance of seagrasses by pushing species outside of optimal salinity tolerance ranges.
+Changing frequency and severity of precipitation patterns may further alter the distribution and abundance of seagrasses by pushing species outside of optimal salinity tolerance ranges [@rasheed2011; @webster2021].
 These changes may produce hypersaline conditions under periods of prolonged drought or more freshwater conditions with increased storm events.
-Seagrasses may respond differently under individual or multiple stressors [@lefcheck2017; @zhang2023] , suggesting place-based assessments of response to climate change provide valuable insights for managers responsible for maintaining ecological resources.
+Seagrasses may respond differently under individual or multiple stressors [@lefcheck2017; @zhang2023], suggesting place-based assessments of response to climate change provide valuable insights for managers responsible for maintaining ecological resources.
 
-In Tampa Bay, Florida, seagrasses are a primary indicator of bay health that have been the focus of management efforts over the last thirty years.
-Through successful reduction of nitrogen loads by 2/3 pre-regulation, seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres (16,857 ha).
+In Tampa Bay, Florida, seagrasses are a primary indicator of bay health that have been the focus of management efforts over the last thirty years [@sherwood2017].
+Through successful reduction of nitrogen loads by 2/3 pre-regulation, seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres (16,857 ha) [@greening2006; @greening2014].
 Seagrass distribution and abundance has been dynamic throughout this period, overall responding positively to increases in water clarity with nutrient load reductions, while also demonstrating more short-term variability in response to regional climate events [@greening2006].
@@
 Factors that have influenced this recent decline are unknown and the effects of climate change on physical conditions have been implicated as potential drivers.
-Following global trends, recent work has demonstrated a broad long-term trend of increasing water temperature in Tampa Bay [@nickerson2023].
-Changing salinity has not been well-described, nor have potential links to seagrass change.
+Following global trends, recent work has demonstrated a broad long-term trend of increasing water temperature in Tampa Bay, although at a relatively coarse scale [@nickerson2023].
+Changing salinity has not been well-described, nor have potential links of changing salinity and temperature with recent seagrass change.
 Tampa Bay is rich with historical data that can be used to evaluate long-term trends and to develop models that can explain relationships of seagrass change with water quality conditions.
@@
 Three datasets of varying sample design and temporal coverage were used to assess the primary hypothesis that conditions in Tampa Bay are trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges.
-This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring data describing cover of individual species, and synoptic seagrass data collected with the water quality surveys.
-Specific focus was on the period from prior to and after 2016, when seagrasses recovered baywide and then declined despite water quality conditions remaining relatively stable.
+This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring describing cover of individual species, and synoptic seagrass data collected with the water quality surveys.
+Specific focus was on the periods prior to and after 2016, when seagrasses recovered baywide and then declined despite water quality conditions remaining relatively stable.
 Our primary goal was to relate recent seagrass changes to climate stressors to provide a motivating example for the management community in the Tampa Bay region that a shifting ecological baseline presents new challenges beyond the known success story of long-term bay recovery.
-The results are also provided as a cautionary example of the subtle but increasing role of climate-related stressors on the resiliency of estuarine systems, prompting a reassessment of existing management paradigms to accommodate current and anticipated future conditions.
+The results are also provided as a cautionary example of the subtle but increasing role of climate-related stressors on the resiliency of estuarine systems at the tropical-temperate boundary, prompting a reassessment of existing management paradigms to accommodate current and anticipated future conditions.
 
@@
 The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee Rivers contributing a majority of freshwater inflow to the bay.
-Tampa Bay straddles the temperate and tropical boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer months.
+Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer months.
 The watershed is heavily developed and includes over 3 million people with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023].

### 2023-12-10 manu-draft.qmd
abstract draft added

@@
 ## Abstract {.unnumbered}
+
+Climate change is an existential threat to environmental resources that sustain human health and the environment.
+In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth.
+Changing physical conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be associated climate stress.
+Three datasets of varying sample design and coverage were evaluated to characterize physical environments in Tampa Bay and the potential relationships with seagrass change.
+Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 ppt per year.
+These changes varied spatially and seasonally, with the most dramatic changes being observed in the upper bay segments during the summer months.
+Simple linear models provided limited evidence that the recent seagrass decline is associated with hotter and fresher conditions.
+Trends in warming and increased precipitation in the region will continue, further creating suboptimal conditions for seagrasses in Tampa Bay.
+These results provide a motivating example for resource managers to consider the likelihood of reduced resilience of estuarine resources due to shifting ecological baselines from climate change.
+Conventional management approaches that focus on limiting nutrient loads should be continued, but the effectiveness may be limited and additional actions will be needed to maintain ecosystem health in the future.
 
@@
 
-The third and final dataset evaluated was from the Pinellas County Department of Environmental Management (PCDEM).
-Surface waters in Pinellas county have been monitored since the 1990s, although a consistent stratified random sampling designed has been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES reporting.
-Data were obtained by request to PCDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (also available at <https://wateratlas.usf.edu/>).
-We focused primarily on OTB for the analysis of the PCDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments (Figure [-@fig-map]d).
+The third and final dataset evaluated was from the Pinellas County Department of Environmental Management (PDEM).
+Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES reporting.
+Data were obtained by request to PDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (also available at <https://wateratlas.usf.edu/>).
+We focused primarily on OTB for the analysis of the PDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments (Figure [-@fig-map]d).
 Four distinct spatial zones in OTB are used to stratify the random selection of sample points that typically include 4 sample points per month in each zone.
@@
 Petersburg, and Tampa; Tampa Bay Water; and the Southwest Florida Water Management District) through contributions to the operating budget.
-We are indebted to the hundreds of field and administrative staff for collecting and maintaining the valuable long-term monitoring data used in this study.
+We are indebted to the hundreds of field and technical staff for collecting and maintaining the valuable long-term monitoring data used in this study.
 
@@
 Summarized seasonal trends by month are shown for (b) top and (c) bottom measurements as the percent of stations in each bay segment with significant increasing (temperature) or decreasing (salinity) trends.
-Bay segment outlines are shown in (a); OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/kendall.png){#fig-kendall}
+Bay segment outlines are shown in (a); OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/kendall.png){#fig-kendall}
 

### 2023-12-10 manu-draft.qmd
fixing reference file

@@
 A fundamental challenge describing niche space is identifying the boundaries for optimal conditions.
-In Tampa Bay, the three dominant seagrass species are *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinium* (turtle grass) [@lewis1985].
+In Tampa Bay, the three dominant seagrass species are *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinum* (turtle grass) [@lewis1985].
 These species co-occur often in mixed beds throughout the bay, although some differences in abundance are observed across salinity ranges.

### 2023-12-11 manu-draft.qmd
minor text edit

@@
 In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth.
-Changing physical conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be associated climate stress.
+Changing physical conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be linked to climate stress.
 Three datasets of varying sample design and coverage were evaluated to characterize physical environments in Tampa Bay and the potential relationships with seagrass change.
 Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 ppt per year.
-These changes varied spatially and seasonally, with the most dramatic changes being observed in the upper bay segments during the summer months.
+These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay segments during the summer months.
 Simple linear models provided limited evidence that the recent seagrass decline is associated with hotter and fresher conditions.
@@
 Lastly, our result showing that salinity has decreased in Tampa Bay is generally contrary to expectations for coastal systems consistent with sea-level rise projections [@alarcon2024].
-Salinity changes with sea-level rise have been documented to contribute to numerous alterations of subtidal and nearshore habitats [@brinson1995; @white2017].
-In southwest Florida, the most common ecological example is migration of tidal wetlands, primary mangroves, upland as porewater salinity and water levels have gradually changed over the past few decades [@borchert2018].
+Salinity changes with sea-level rise have caused numerous alterations of subtidal and nearshore habitats [@brinson1995; @white2017].
+In southwest Florida, the most common ecological example is migration of tidal wetlands, primarily mangroves, upland as porewater salinity and water levels have gradually changed over the past few decades [@borchert2018].
 Alteration of salinity regimes for surface and groundwater resources are also well documented.

### 2023-12-13 manu-draft.qmd
proofing manu draft

@@
 The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale.
-Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services[@fourqurean2012; @orth2020; @orth2023].
+Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023].
 Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021].
@@
 The sustained coverage or restoration of seagrasses in stressed environments requires management actions that create supportive conditions for growth.
-A long-standing approach adopted by numerous management entities has been the control of nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @greening2014] .
+A long-standing approach adopted by numerous management entities has been the control of nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @greening2014].
 There are limited examples of successful recovery of seagrass through control of nutrient inputs alone, primarily because of the difficulty in identifying and regulating both point and diffuse non-point sources.
-Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014]and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in areal coverage through sustained and long-term reductions in nutrients.
+Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014] and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in seagrass areal coverage through sustained and long-term reductions in nutrients.
 Cooperation among management, regulatory, public, and private sectors were critical aspects of both examples [@tango2016; @sherwood2016].
@@
 In Tampa Bay, Florida, seagrasses are a primary indicator of bay health that have been the focus of management efforts over the last thirty years [@sherwood2017].
-Through successful reduction of nitrogen loads by 2/3 pre-regulation, seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres (16,857 ha) [@greening2006; @greening2014].
+Through successful reduction of nitrogen loads by 2/3 pre-regulation, seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres [16,857 ha, @greening2006; @greening2014].
 Seagrass distribution and abundance has been dynamic throughout this period, overall responding positively to increases in water clarity with nutrient load reductions, while also demonstrating more short-term variability in response to regional climate events [@greening2006].
-Since 2016, seagrass areal cover has decreased by nearly 1/3 despite relatively stable water quality with adequate light penetration that has been historically shown to support growth.
+Since 2016, seagrass areal cover has decreased by nearly 1/3 despite relatively stable water quality with adequate light penetration that has been historically shown to support growth [@tbep0696; @Beck20a].
 Factors that have influenced this recent decline are unknown and the effects of climate change on physical conditions have been implicated as potential drivers.
@@
 
-This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay Estuary over the last fifty years.
+This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay estuary over the last fifty years.
 Three datasets of varying sample design and temporal coverage were used to assess the primary hypothesis that conditions in Tampa Bay are trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges.
@@
 The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee Rivers contributing a majority of freshwater inflow to the bay.
-Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer months.
+Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer.
 The watershed is heavily developed and includes over 3 million people with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023].
@@
 Two primary sources of data have been used to track seagrass change over time in Tampa Bay.
-The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s [e.g., @SWFWMD23].
+The Southwest Florida Water Management District has estimated areal coverage of seagrasses approximately biennially since the late 1980s (Figure [-@fig-seagrasschg]a, available at <https://data-swfwmd.opendata.arcgis.com/>).
 These maps are created by photointerpretation of aerial images obtained at the end of the growing season, typically during November-December.
-The TBEP has used these maps to track progress towards achieving seagrass restoration goals as total cover in Tampa Bay.
 No species information is provided.
 A more detailed, but spatially-specific, data source is the Tampa Bay Interagency Seagrass Monitoring Program (Figure [-@fig-seagrasschg]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>).
-Annual transect surveys have been conducted since 1998 at 61 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017].
-This dataset provides species information, including abundance, cover, frequency occurrence, and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed.
-Although the areal maps provide the standard for assessment of restoration goals, the transect data allow for inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined.
+Annual transect surveys have been conducted since 1998 at 62 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017].
+This dataset provides species information on abundance, cover, frequency occurrence, and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed.
+Although the areal maps provide the standard for assessment of restoration goals, the transect data allow inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined.
 As such, the transect data were used below for comparison with temperature and salinity changes for the major bay segments.
@@
 First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at 45 stations in the major bay segments since the late 1970s (Figure [-@fig-map]b).
-These data provide the basis for regulatory assessments and compliance reporting for the nutrient TMDL in Tampa Bay.
-Water quality samples are collected at each station from surface water grabs (e.g., nutrients) or *in situ* measurements for physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom.
+These data are used for regulatory assessments for the nutrient TMDL in Tampa Bay.
+Water quality samples are collected at each station from surface water grabs (e.g., nutrients) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom.
 Time of sampling can vary, although most samples are collected from mid-morning to early afternoon.
@@
 The Fisheries Independent Monitoring (FIM) program administered by FWC provides monthly surveys of the entire nekton community in Tampa Bay, including species richness and abundance, using multiple survey gear types that target different habitats.
-A stratified sample design is used to target multiple habitats where unique sites are sampled each month.
+A stratified sample design is used to target multiple habitats at different sites sampled each month.
 We used data from the 21.3 meter center-bag nearshore seine that targets shallow habitats where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present, Figure [-@fig-map]c).
@@
 Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES reporting.
-Data were obtained by request to PDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (also available at <https://wateratlas.usf.edu/>).
-We focused primarily on OTB for the analysis of the PDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments (Figure [-@fig-map]d).
+Data were obtained by request to PDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (Figure [-@fig-map]d, also available at <https://wateratlas.usf.edu/>).
+We focused primarily on OTB for the analysis of the PDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments.
 Four distinct spatial zones in OTB are used to stratify the random selection of sample points that typically include 4 sample points per month in each zone.
-Water quality samples at each site are similar to those collected by EPC, where only bottom measurements for salinity and temperature were retained.
+Water quality samples at each site are similar to those collected by EPC, where only bottom measurements for salinity and temperature were retained for analysis.
 Seagrass presence/absence is also recorded at each site and all sites were defined as "seagrass" if only seagrass species were identified (any with macroalgae were excluded) and "no seagrass" if bare sediment was observed.
@@
 The change per year was also evaluated for each parameter based on the slope estimates returned by each test.
-Kendall tests were also used to evaluate changes over time for each month to determine when the trends were most pronounced seasonally, e.g., all January estimates across years, all February estimates, etc.
-The percentage of stations in each bay segment with significant increasing temperature or decreasing salinity trends were evaluated for each month.
+Kendall tests were also used to evaluate changes over time for each month across years to determine when the trends were most pronounced seasonally, e.g., all January estimates across years, all February estimates, etc.
+The percentage of stations in each bay segment with significantly increasing temperature or decreasing salinity trends were evaluated for each month.
 All tests evaluated both surface and bottom measurements to assess potential differences by water depth.
@@
 In the simplest form, this can conceptualized as a bell curve with optimal conditions defined within a range of values for a single parameter (e.g., minimum and maximum temperatures where a species is typically observed), where reduced growth or mortality is observed outside of these ranges.
-Because both water temperature and salinity were evaluated, the same model can be conceptualized in two-dimensional space [@fig-concept].
+Because both water temperature and salinity were evaluated, the same model can be conceptualized in two-dimensional space ([@fig-concept]).
 Seagrass growth can be limited when temperature is below or above the optimum range, when salinity is below or above the optimum range, or when both temperature and salinity conditions are outside of the optimum range.
-Based on the results of the trend tests, we hypothesized that seagrasses are likely stressed by both high temperature and low salinity (bottom right @fig-concept).
+Based on the results of the trend tests, we hypothesized that seagrasses are likely stressed by both high temperature and low salinity (bottom right, @fig-concept).
 Although the optimal niche space can be defined in multiple dimensions for many parameters, we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have generally not been limiting for growth in recent years.
@@
 
-Because of the uncertainty in defining thresholds for optimal temperature and salinity ranges, multiple thresholds were evaluated to both describe the potential for stress and how it may be related to changes in seagrass.
+Because of the uncertainty in defining thresholds for optimal temperature and salinity ranges, multiple thresholds were evaluated to describe the potential for stress and how it may be related to changes in seagrass.
 Distinctions were not made between species, primarily due to lack of consensus between studies and likely site-specific ranges that affect seagrass growth in Tampa Bay.
@@
 These metrics were quantified from the monthly long-term observations in the EPC data.
-To quantify daily counts each year, a continuous prediction of temperature and salinity over time at each of 45 stations was estimated using Generalized Additive Models (GAMs) fit to temperature or salinity and a single predictor for decimal year [@wood17].
+To quantify daily counts each year, a continuous prediction of temperature and salinity over time at each of 45 stations was estimated using Generalized Additive Models (GAMs) fit to temperature or salinity with a single predictor for decimal year [@wood17].
 The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends.
-Model fit for each station was sufficient to calculate daily predictions for assessing the stressor metrics (see supplement, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models).
+Model fit for each station was sufficient to calculate daily predictions to assess potential stressor metrics (see supplement, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models).
 
@@
 
-For comparison to seagrass, the annual metrics were also referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year as above.
+For comparison to seagrass, the annual metrics were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends.
 Transect surveys are typically done in the fall at the end of the growing season.
-The average date for each year across the subset of transects in each bay segment was estimated and the stressor metrics were referenced accordingly.
+The average sampling date for each year across the subset of transects in each bay segment was estimated and the stressor metrics were referenced accordingly.
 For example, if the average transect date for a bay segment was September 15th in 2009 and Oct 1st in 2010, the 2010 stressor counts were based on all days between the two average dates.
@@
 Generalized Linear regression Models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre-2016, decline post-2016).
-Two models were evaluated, one with the temperature and salinity metrics and another with the both metric given that the latter was a combined metric of the former.
-Interactions were included between temperature, salinity, and time period or the both metric and time period depending on the model.
+Two models were evaluated, one with the temperature and salinity metrics together and another with the both metric given that the latter was a combined metric of the former.
+Interactions were included between temperature, salinity, and time period or the both metric and time period, depending on the model.
 Bay segment was included as a main effect without interactions.
@@
 GLMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated (no bay segment interaction) and the seagrass response variable was presence/absence (i.e., a binomial distribution was used) to model the probability of observing seagrass.
-Input data for models for the FIM and PDEM datasets were further subset to include only months from July to November to describe seagrasses during the growing season and to reduce potential seasonal effects.
+Model input data were further subset to include only months from July to November to describe seagrasses during the growing season and to reduce potential seasonal effects.
 
@@
 
-Long-term meteorological data from Tampa International Airport showed significantly increasing trends for air temperature and precipitation [@fig-meteowqraw].
+Long-term meteorological data from Tampa International Airport showed significantly increasing trends for air temperature and precipitation ([@fig-meteowqraw]).
 Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`).
-Mean annual air temperature in 1975 was `r airtempstr`, whereas current mean annual air temperature in 2022 was `r airtempend`, showing an overall increase in the period of record of `r airtempdif` $^\circ$C.
+Mean annual air temperature in 1975 was `r airtempstr` $^\circ$C, whereas current mean annual air temperature in 2022 was `r airtempend` $^\circ$C, showing an overall increase in the period of record of `r airtempdif` $^\circ$C.
 Similarly, total precipitation has increased by `r prpslo` mm per year (`r prppvl`, $R^2$ `r prprsq`).
-Total precipitation in 1975 was `r prpstr`, whereas current total precipitation in 2022 was `r prpend`, showing an overall increase in the period of record of `r prpdif` mm.
+Total precipitation in 1975 was `r prpstr` mm, whereas current total precipitation in 2022 was `r prpend` mm, showing an overall increase in the period of record of `r prpdif` mm.
 Trends in total annual hydrologic load from 1985 to present followed total annual precipitation, although the linear model was not significant likely due to a shorter period of record.
-The SPI showed notable anomalies in precipitation, with pronounced rainy periods in 1980, the late 1990s, 2005, and the last ten years.
+The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and the last ten years.
 
@@
 The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths and minimal stratification in Tampa Bay (@fig-meteowqraw).
-Trends in water temperature were similar across bay segments with significant increases varying from 0.3 to 0.4 $^\circ$C per year, with a total increase of 1.4 (OTB) to 2.0 (HB) $^\circ$C from 1975 to present (@tbl-temptrndtab).
+Trends in water temperature were similar across bay segments with significant increases varying from 0.03 to 0.04 $^\circ$C per year, with a total increase of 1.4 (OTB) to 2.0 (HB) $^\circ$C from 1975 to present (@tbl-temptrndtab).
 Salinity trends were similar between bay segments, although overall salinity was of course higher for bay segments closer to the Gulf of Mexico.
@@
 All stations had significantly increasing temperature and decreasing salinity from 1975 to present in both the top and bottom of the water column, excluding one station in HB that did not have significantly decreasing bottom salinity (Figure [-@fig-kendall]a).
-Seasonally, most stations had more significant trends in the summer, early fall period for both temperature and salinity (Figure [-@fig-kendall]b, [-@fig-kendall]c), although some variation was observed between bay segments.
-Temperature trends were more often observed in the summer, early fall for the upper bay segments (OTB, HB), whereas the lower bay segments (MTB, LTB) had more significant trends during the spring.
+Seasonally, most stations had more significant trends in the summer, early fall period for both temperature and salinity (Figure [-@fig-kendall]b, c), although some variation was observed between bay segments.
+Temperature trends were more often observed in the summer, early fall for the upper bay segments (OTB, HB), whereas the lower bay segments (MTB, LTB) had more seasonally homogeneous trends.
 Seasonal trends in salinity did not vary as much between segments, although OTB and LTB had more stations with significantly decreasing trends than the other bay segments.
-Again, differences between top or bottom trends did not differ seasonally.
+Again, differences between top or bottom salinity trends were not observed.
 
@@
 Linear mixed-effects models showed varying trends in the potential stressor metrics for the number of days when temperature was above, salinity was below, or both occurred relative to different thresholds.
-All of the temperature models for each of the three thresholds showed significantly increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (see supplement).
+All of the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed significantly increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (see supplement).
 The estimated slopes for the number of days when temperature was above 30 $^\circ$C were similar among bay segments, varying from 0.9 days per year in LTB to 1.1 days per year in MTB.
@@
 The salinity models were less similar between bay segments compared to the temperature models, primarily because of the natural variation in salinity along the bay's longitudinal axis (see supplement).
-None of the bay segments had significantly increasing number of days when salinity was below 15 ppt.
-For MTB, only the number of days when salinity was below 25 ppt was significantly increasing at a rate of 0.67 days per year.
+None of the bay segments had significantly increasing number of days per year when salinity was below 15 ppt.
+For MTB, only the number of days when salinity was below 25 ppt was significantly increasing at a rate of 0.7 days per year.
 Both HB and OTB had significantly increasing number of days per year when salinity was below 20 and 25 ppt, although the slopes varied such that the rates in OTB were nearly double those in HB.
-The number of days per year when salinity was below 20 ppt in OTB increased by 1 day per year, whereas the number of days in HB increased by 0.5 days per year from 1975 to present.
-Likewise, the average number of days when salinity was 20 ppt at the beginning and end of the period of record varied from 11 (HB) to 23 (OTB) days in 1975 to 25 (HB) to 68 (LTB) days in 2022.
+The number of days per year when salinity was below 25 ppt in OTB increased by 2 days per year, whereas the number of days in HB increased by 0.9 days per year from 1975 to present.
+Likewise, the average number of days per year when salinity was continuously below 25 ppt at the beginning and end of the period of record was 128 (OTB) and 131 (HB) days in 1975 and 171 (HB) and 219 (OTB) days in 2022.
 
-Times when both temperature was above a threshold and salinity was below a threshold also varied by bay segment.
-None of the models were significant for LTB, four were significant for MTB, seven were significant for HB, and four were significant for OTB.
+The number of days when both temperature was above a threshold and salinity was below a threshold also varied by bay segment.
+Four models were significant for OTB, seven for HB, four for MTB, and none for LTB.
 The number of days when temperature was above 29 $^\circ$C and salinity was below 25 ppt had the largest slopes of 1.4, 1.2, and 0.7 days per year for OTB, HB, and MTB, respectively.
-Likewise, the average number of days when both temperature was above 29 $^\circ$C and salinity was below 25 ppt from the beginning to the end of the period of record varied from 6, 12, and 1 day(s) in 1975 to 72, 68, and 24 days in 2022 for OTB, HB, and MTB, respectively.
+Likewise, the average number of days when both temperature was above 29 $^\circ$C and salinity was below 25 ppt from the beginning to the end of the period of record varied from 6, 12, and 1 day(s) in 1975 to 72, 68, and 34 days in 2022 for OTB, HB, and MTB, respectively.
 
@@
 Temperature trends were similar among segments, whereas the number of days when salinity was below the threshold varied by proximity to the Gulf of Mexico (@tbl-mixdaytab).
-The number of days when both temperature was above and salinity was below the threshold generally followed the trends for the number of days when salinity was below the threshold, although the slopes varied.
-These thresholds were used for comparison to seagrass changes described below, based primarily on the statistical strength of the trends and the variance of counts among bay segments.
+The number of days when both temperature was above and salinity was below the threshold generally followed the trends for the number of days when salinity was below the threshold, with the slopes varying by bay segment.
+These thresholds were used for comparison to seagrass changes described below, based primarily on the statistical strength of the trends and the variance of counts across stations within each bay segment (points in @fig-mixeff).
 That is, more restrictive thresholds did not provide sufficient counts of days per year to develop models and the chosen thresholds were based primarily on statistical considerations.
@@
 However, the interaction between time period and temperature was not significant, suggesting no difference between time periods.
-The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (p \< 0.01, Figure [-@fig-sgmod]a, @tbl-modtab).
+The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure [-@fig-sgmod]a, @tbl-modtab).
 The interaction between the temperature and salinity metric was included in the most parsimonious model (marginally significant, p \< 0.1), although the shape of the relationship of temperature with time period or salinity with time period did not noticeably change using different conditional values for salinity or temperature, respectively.
@@
 At a local scale, these changes have been demonstrated herein for Tampa Bay where the evaluation of long-term trends in water temperature and salinity showed clear changes mirrored by long-term changes in air temperature and precipitation.
-Over the previous fifty years, the Bay has demonstrated a consistent trend towards hotter and fresher conditions, where water temperature has approximately increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year.
+Over the previous fifty years, the Bay has consistently trended towards hotter and fresher conditions, where water temperature has approximately increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year.
 These changes were demonstrated in three long-term datasets with different sampling methods and periods of record.
@@
 @nickerson2023 evaluated sea surface trends at a larger spatial scale for Tampa Bay, the West Florida Continental Shelf, and the adjacent Gulf of Mexico.
-Temperature trends were similar for Tampa Bay (the EPC dataset was also used), although the power to detect trends decreased for shorter time periods, as independently verified with the FIM and PDEM datasets herein.
+Temperature trends were similar for Tampa Bay (the EPC dataset was also used), although the power to detect trends decreased for shorter time periods, as verified herein with the FIM and PDEM datasets.
 @nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, which somewhat conflicts with our results showing distinct increases in the summer, particularly for the OTB and HB segments.
@@
 Additionally, @shiir provided a recent assessment of a 2023 heatwave in south Florida, supported by a 20-year trend assessment that suggested estuaries were warming at nearly double the rate of the Gulf of Mexico.
-The upper limit of our warming estimates for Tampa Bay is comparable.
+The upper limit of our warming estimate for Tampa Bay is comparable.
 
@@
 An additional caveat of our models was the use of "thresholds" to define potential stressor metrics for temperature and salinity on an annual time scale.
-Our choice to use 30$^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed, as shown by the mixed-effects linear models for each bay segment.
+Our choice to use 30 $^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed, as shown by the mixed-effects linear models for each bay segment.
 That is, sufficient change and variation in the independent variables for the models of seagrass change were needed to statistically describe potential relationships.
@@
 Thus, time period was necessary to control for these confounding relationships.
-Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay (OTB in particular where the most losses have been seen) [@tbep0696; @Beck20a].
+Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay [OTB in particular where the most losses have been seen, @tbep0696; @Beck20a].
 This suggests that other factors are influencing seagrass growth and the inclusion of light attenuation in the models was likely not critical for describing the roles of temperature and salinity.
@@
 For example, temperature may simply be higher in locations where seagrasses are present, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses.
-This explanation cannot be ruled out with the existing datasets, although the trend analyses showing the consistent long-term changes and our models suggest that climate-related stressors driving seagrass loss is a more likely scenario.
+This explanation cannot be ruled out with the existing datasets, although the trend analyses and models suggest that climate-related stressors driving seagrass loss is a more likely scenario.
 
@@
 Salinity changes with sea-level rise have caused numerous alterations of subtidal and nearshore habitats [@brinson1995; @white2017].
-In southwest Florida, the most common ecological example is migration of tidal wetlands, primarily mangroves, upland as porewater salinity and water levels have gradually changed over the past few decades [@borchert2018].
+In southwest Florida, the most common ecological example is migration of mangroves upland as porewater salinity and water levels have gradually increased over the past few decades [@borchert2018].
 Alteration of salinity regimes for surface and groundwater resources are also well documented.
@@
 Although drought has been implicated in the hypersaline conditions observed in Florida Bay, sea level rise is expected to further alter salinity dynamics in the region.
-@dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Greater Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters.
+@dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters.
 In Tampa Bay, the long-term trends of decreasing salinity suggest that the hydrologic load into the system has had a greater influence on salinity regimes than the effects of sea-level rise.
@@
 Our models suggested that rising temperatures and decreasing salinities are additively affecting seagrasses, rather than multiplicatively as evidenced by lack of significant interactions of these stressors.
-Future analyses may show more significant associations between physical habitat conditions and seagrass change as the trends are very likely to continue to push seagrass further outside of their tolerance ranges.
+Future analyses may show more significant associations between physical habitat conditions and seagrass change as the trends are very likely to continue to push seagrasses further outside of their tolerance ranges.
 These analyses should be supported by additional data, particularly high-resolution continuous monitoring data that provide a more precise assessment of stress at multiple time-scales.
@@
 
-This work benefited from constructive discussions with the Technical Advisory Committee of the Tampa Bay Estuary Program (TBEP).
+This work benefited from constructive discussions with the Technical Advisory Committee of the Tampa Bay Estuary Program (TBEP), Mike Wessel, Ray Pribble, and Dave Tomasko.
 TBEP funding for this work stems from EPA Section 320 Grant Funds, and the TBEP's local government partners (Hillsborough, Manatee, Pasco, and Pinellas Counties; the Cities of Clearwater, St.
@@
 
+![Conceptual stressor diagram demonstrating a two-dimensional niche space for temperature and salinity.](figs/concept.png){#fig-concept}
+
+{{< pagebreak >}}
+
 ![Long-term air temperature, precipitation, hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1975 to 2022.
@@
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/meteowqraw.png){#fig-meteowqraw}
-
-{{< pagebreak >}}
-
-![Conceptual stressor diagram demonstrating a two-dimensional niche space for temperature and salinity.](figs/concept.png){#fig-concept}
 

### 2023-12-14 manu-draft.qmd
proof fig and table captions

@@
 
-![Map of Tampa Bay and the three datasets used for trend analysis, including (a) bay segments, 2022 seagrass coverage (green), and transect starting points (red), (b) Environmental Protection Commissions (EPC) long-term monitoring sites, (c) Fisheries Independent Monitoring (FIM) random sampling for near-shore seine hauls, and (d) OTB portion of Pinellas County Department of Environmental Management (PDEM) random sampling.
+![Map of Tampa Bay and the three datasets used for trend analysis, including (a) bay segments, 2022 seagrass coverage (green), and transect starting points (red), (b) Environmental Protection Commission (EPC) long-term monitoring sites, (c) Fisheries Independent Monitoring (FIM) random sampling for near-shore seine hauls, and (d) OTB portion of Pinellas County Department of Environmental Management (PDEM) random sampling.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/map.png){#fig-map}
@@
 
-![Seagrass changes over time in Tampa Bay for (a) areal coverage and (b) frequency occurrence of major species.
+![Seagrass changes over time in Tampa Bay for (a) areal coverage 1988 - 2022 from mapping and (b) frequency occurrence of major species 1998 - 2022 from annual transect monitoring.
 Changes are shown for major bay segments.
-Note the different time scale between (a) and (b); coverage maps in (a) began in 1988 and seagrass transect monitoring in (b) began in 1998.
 Red lines in (a) show approximate capacity of seagrass coverage based on the baywide target of 40,000 acres.
@@
 
-![Conceptual stressor diagram demonstrating a two-dimensional niche space for temperature and salinity.](figs/concept.png){#fig-concept}
+![Conceptual stressor diagram demonstrating a two-dimensional niche space for temperature and salinity. Tampa Bay is trending towards the bottom right. ](figs/concept.png){#fig-concept}
 
@@
 ![Long-term air temperature, precipitation, hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1975 to 2022.
-Points for salinity and water temperature are colored by sampling location in the water column and show the average (95% confidence interval) across all stations and sampling months for each year in each bay segment.
+Points for water temperature and salinity are colored by sampling location in the water column and show the average (95% confidence interval) across all stations in each bay segment and sampling months for each year.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/meteowqraw.png){#fig-meteowqraw}
@@
 ![Trends from 1975 to 2022 for temperature and salinity measurements at long-term monitoring stations in Tampa Bay.
-Results for (a) seasonal Kendall tests by station and monitoring location (top or bottom of the water column) are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr-1).
+Results for seasonal Kendall tests by station and monitoring location (top or bottom of the water column) are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr$^{-1}$).
 Summarized seasonal trends by month are shown for (b) top and (c) bottom measurements as the percent of stations in each bay segment with significant increasing (temperature) or decreasing (salinity) trends.
@@
 
-![Example of mixed effects models for the estimated number of days by year that temperature (red) or salinity (blue) were above or below thresholds of 30 degrees C or 25 psu, respectively.
-The bottom row (black) shows the number of days when both temperature and salinity were above or below the thresholds.
-The models included station as a random effect for each bay segment, with grey lines indicating individual station trends and thicker lines indicating the overall model fit.
+![Example of mixed effects models for the estimated number of days per year that temperature (red) or salinity (blue) were above or below thresholds of 30 $^\circ$C or 25 psu, respectively.
+The bottom row (black) shows the number of days when both temperature was above and salinity was below the thresholds.
+The models included station as a random effect for each bay segment, with grey lines indicating individual station trends, grey points as actual number of days, and thicker lines indicating the overall model fit.
 Slopes for significant models are shown in the bottom right of each facet.
@@
 ![Predicted results from generalized linear models evaluating seagrass changes versus temperature or salinity predictors for the (a) EPC and (b) FIM datasets.
-The predictors for the (a) were the number of days when temperature was above the threshold and the number of days when temperature was below the threshold.
+The predictors for (a) were the number of days when temperature was above the threshold and the number of days when salinity was below the threshold.
 The predictors for (b) were measured temperature and salinity at the bottom of the water column.
@@
 Start year describes the range of the trend test to the present year (2022).
-The starting value is the estimated salinity at the start year and the end value is the estimated temperature at 2022.
+The starting value is the estimated salinity at the start year and the end value is the estimated salinity at 2022.
 Total change is the difference between the two.
@@
 Models were developed for different datasets described in the model column.
-The significant intercept term for each model is not shown and described factor conditions not explicitly shown by the remaining terms (eg., OTB and pre-2016).
+The significant intercept term for each model is not shown and described factor conditions not explicitly shown by the remaining terms (e.g., OTB and pre-2016).
 The model term estimates include the standard error in parentheses.

### 2023-12-14 supplement.qmd
proof fig and table captions

@@
 ![Air temperature, precipitation, hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1998 to 2022.
-Points for salinity and water temperature are colored by sampling location in the water column and show the average (95% confidence interval) across all stations and sampling months for each year in each bay segment.
+Points for water temperature and salinity are colored by sampling location in the water column and show the average (95% confidence interval) across all stations in each bay segment and sampling months for each year.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppmeteowqraw.png){#fig-suppmeteowqraw}
@@
 
-![Water temperature and salinity trends from 1996 to 2021 for stations from the Fisheries Independent Monitoring program.
-Points show the average (95% confidence interval) across all stations and sampling months for each year in each bay segment.
+![Bottom water temperature and salinity trends from 1996 to 2021 for stations from the Fisheries Independent Monitoring program.
+Points show the average (95% confidence interval) across all stations in each bay segment and sampling months for each year.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppfimtrnds.png){#fig-suppfimtrnds}
@@
 
-![Water temperature and salinity trends from 2004 to 2022 for stations from the Pinellas County Department of Environmental Management in Old Tampa Bay.
-Points show the average (95% confidence interval) across all stations and sampling months for each year in each bay segment.](figs/supppincotrnds.png){#fig-supppincotrnds}
+![Bottom water temperature and salinity trends from 2004 to 2022 for stations from the Pinellas County Department of Environmental Management in Old Tampa Bay.
+Points show the average (95% confidence interval) across all stations in each bay segment and sampling months for each year.](figs/supppincotrnds.png){#fig-supppincotrnds}
 
@@
 ![Trends from 1998 to 2022 for temperature and salinity measurements at long-term monitoring stations in Tampa Bay.
-Results for (a) seasonal Kendall tests by station and monitoring location (top or bottom of the water column) are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr-1).
+Results for seasonal Kendall tests by station and monitoring location (top or bottom of the water column) are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr$^{-1}$).
 Summarized seasonal trends by month are shown for (b) top and (c) bottom measurements as the percent of stations in each bay segment with significant increasing (temperature) or decreasing (salinity) trends.
-Bay segment outlines are shown in (a); OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppkendall.png){#fig-suppkendall}
+Bay segment outlines are shown in (a); OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppkendall.png){#fig-suppkendall}
 
@@
 
-![Example of mixed effects models for the estimated number of days by year that temperature (red) or salinity (blue) were above or below thresholds of 30 degrees C or 25 psu, respectively, from 1998 to 2022.
-The bottom row (black) shows the number of days when both temperature and salinity were above or below the thresholds.
-The models included station as a random effect for each bay segment, with grey lines indicating individual station trends and thicker lines indicating the overall model fit.
+![Example of mixed effects models for the estimated number of days per year that temperature (red) or salinity (blue) were above or below thresholds of 30 $^\circ$C or 25 psu, respectively, from 1988 to 2022.
+The bottom row (black) shows the number of days when both temperature was above and salinity was below the thresholds.
+The models included station as a random effect for each bay segment, with grey lines indicating individual station trends, grey points as actual number of days, and thicker lines indicating the overall model fit.
 Slopes for significant models are shown in the bottom right of each facet.
@@
 \*\* p \< 0.005, \* p \< 0.05, - no model.
-OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
 
@@
 Columns for "Both" indicate the estimated number of days when both temperature and salinity were above or below critical thresholds.
-Note that not all models are significant (see table 1).
-\*\* p \< 0.005, \* p \< 0.05, - no model.
-OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
+Note that not all models are significant (see table 3).
+- no model.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
 
@@
 Columns for "Both" indicate the estimated number of days when both temperature and salinity were above or below critical thresholds.
-Note that not all models are significant (see table 1).
-\*\* p \< 0.005, \* p \< 0.05, - no model.
-OTB (northwest): Old Tampa Bay, HB (northeast): Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
+Note that not all models are significant (see table 4).
+- no model.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
 

### 2023-12-14 R/figs.R

@@ -277,7 +277,8 @@ thm <- theme_minimal() +
   theme(
     strip.placement = 'outside', 
     panel.grid.minor = element_blank(), 
-    axis.text.y = element_text(size = 12), 
+    axis.text.y = element_text(size = 11), 
+    axis.title.y = element_text(size = 10),
     legend.text = element_text(size= 12)
   )
 
@@ -462,7 +463,8 @@ thm <- theme_minimal() +
   theme(
     strip.placement = 'outside', 
     panel.grid.minor = element_blank(), 
-    axis.text.y = element_text(size = 12), 
+    axis.text.y = element_text(size = 11), 
+    axis.title.y = element_text(size = 10),
     legend.text = element_text(size= 12)
   )
 
@@ -613,7 +615,7 @@ p6 <- ggplot(toplo6, aes(x = yr, y = avev, group = loc, color = loc)) +
 
 p <- p1 + p2 + p3 + p4 + (p5 + p6 + plot_layout(ncol = 1, guides = 'collect')) + plot_layout(ncol = 1, heights = c(1, 1, 1, 1, 3)) & theme(legend.position = 'top')
 
-png(here('figs/suppmeteowqraw.png'), height = 8, width = 7, family = 'serif', units = 'in', res = 500)
+png(here('figs/suppmeteowqraw.png'), height = 7, width = 6.125, family = 'serif', units = 'in', res = 500)
 print(p)
 dev.off()
 

### 2023-12-15 manu-draft.qmd
add date ranges for each dataset to map
added cites to all supp figs and tabs in manu-draft

@@
 The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends.
-Model fit for each station was sufficient to calculate daily predictions to assess potential stressor metrics (see supplement, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models).
+Model fit for each station was sufficient to calculate daily predictions to assess potential stressor metrics (Figure S5, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models, Tables S1, S2).
 
@@
 
-Increasing water temperature and decreasing salinity generally followed the meteorological trends for all three *in situ* datasets (EPC, FIM, and PDEM, @fig-meteowqraw, @tbl-temptrndtab, @tbl-saltrndtab, figures in supplement).
+Increasing water temperature and decreasing salinity generally followed the meteorological trends for all three *in situ* datasets (EPC, FIM, and PDEM, @fig-meteowqraw, Tables [-@tbl-temptrndtab], [-@tbl-saltrndtab], Figures S1, S2, S3).
 Note that for tables [-@tbl-temptrndtab] and [-@tbl-saltrndtab], comparable time periods were evaluated between the datasets when possible given the different sample sizes, and therefore power, to detect trends.
@@
 Significant increasing temperature trends were also observed for the FIM dataset from 1996 to present in OTB and from 2004 to present in OTB and HB (@tbl-temptrndtab).
-No significant temperature or salinity trends were observed in the other bay segments for the FIM and PDEM datasets from 1996 to present nor from 2004 to present (see figures in supplement).
+No significant temperature or salinity trends were observed in the other bay segments for the FIM and PDEM datasets from 1996 to present nor from 2004 to present (Figures S2, S3).
 However, the EPC dataset showed significant increasing temperature across all time periods and bay segments.
@@
 
-The EPC dataset was also used to provide detailed information on station-level trends from 1975 to present (@fig-kendall).
+The EPC dataset was also used to provide detailed information on station-level trends from 1975 to present (@fig-kendall, see Figure S4 for 1998 to 2022).
 All stations had significantly increasing temperature and decreasing salinity from 1975 to present in both the top and bottom of the water column, excluding one station in HB that did not have significantly decreasing bottom salinity (Figure [-@fig-kendall]a).
@@
 Linear mixed-effects models showed varying trends in the potential stressor metrics for the number of days when temperature was above, salinity was below, or both occurred relative to different thresholds.
-All of the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed significantly increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (see supplement).
+All of the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed significantly increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (Table S3).
 The estimated slopes for the number of days when temperature was above 30 $^\circ$C were similar among bay segments, varying from 0.9 days per year in LTB to 1.1 days per year in MTB.
-Likewise, the average number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied from 8 (OTB, HB) to 21 (LTB) days in 1975 to 55 (HB) to 63 (LTB) days in 2022.
+Likewise, the average number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied from 8 (OTB, HB) to 21 (LTB) days in 1975 to 55 (HB) to 63 (LTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022).
 
-The salinity models were less similar between bay segments compared to the temperature models, primarily because of the natural variation in salinity along the bay's longitudinal axis (see supplement).
+The salinity models were less similar between bay segments compared to the temperature models, primarily because of the natural variation in salinity along the bay's longitudinal axis (Table S4).
 None of the bay segments had significantly increasing number of days per year when salinity was below 15 ppt.
@@
 The number of days per year when salinity was below 25 ppt in OTB increased by 2 days per year, whereas the number of days in HB increased by 0.9 days per year from 1975 to present.
-Likewise, the average number of days per year when salinity was continuously below 25 ppt at the beginning and end of the period of record was 128 (OTB) and 131 (HB) days in 1975 and 171 (HB) and 219 (OTB) days in 2022.
+Likewise, the average number of days per year when salinity was continuously below 25 ppt at the beginning and end of the period of record was 128 (OTB) and 131 (HB) days in 1975 and 171 (HB) and 219 (OTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022).
 
@@
 
-@fig-mixeff provides visual examples of the mixed-effects models for the estimated number of days over time for each bay segment when temperature was above 30$^\circ$, salinity was below 25 ppt, and when both occurred.
+@fig-mixeff provides visual examples of the mixed-effects models for the estimated number of days over time for each bay segment from 1975 to present when temperature was above 30$^\circ$, salinity was below 25 ppt, and when both occurred (see Figure S6 for 1998 to 2022).
 Temperature trends were similar among segments, whereas the number of days when salinity was below the threshold varied by proximity to the Gulf of Mexico (@tbl-mixdaytab).
@@
 Our models suggested that temperature and salinity are potentially associated with seagrass loss and the consistent long-term trend in temperature and salinity suggest that conditions are very likely to become more stressful in the future.
-Moreover, the GAMs used to model continuous temperature and salinity at each EPC monitoring station likely did not fully describe the current extent of stress given that extreme low or high values were often not captured by the daily predictions (see supplement).
+Moreover, the GAMs used to model continuous temperature and salinity at each EPC monitoring station likely did not fully describe the current extent of stress given that extreme low or high values were often not captured by the daily predictions (Figure S5).
 Regardless of the above limitations, the role of climate stressors on seagrass change is likely to continue in the future as suggested by our results.
@@
 ![Map of Tampa Bay and the three datasets used for trend analysis, including (a) bay segments, 2022 seagrass coverage (green), and transect starting points (red), (b) Environmental Protection Commission (EPC) long-term monitoring sites, (c) Fisheries Independent Monitoring (FIM) random sampling for near-shore seine hauls, and (d) OTB portion of Pinellas County Department of Environmental Management (PDEM) random sampling.
+Date ranges for each dataset are shown in the title.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/map.png){#fig-map}
@@
 
-![Conceptual stressor diagram demonstrating a two-dimensional niche space for temperature and salinity. Tampa Bay is trending towards the bottom right. ](figs/concept.png){#fig-concept}
+![Conceptual stressor diagram demonstrating a two-dimensional niche space for temperature and salinity. Tampa Bay is trending towards the bottom right.](figs/concept.png){#fig-concept}
 

### 2023-12-15 supplement.qmd
S prefix to manual tables in supp

@@
 
-Summary of mixed-effects models evaluating increases in the number of days each year from 1975 to 2022 when temperature or salinity were above or below critical thresholds, respectively.
+Table S3: Summary of mixed-effects models evaluating increases in the number of days each year from 1975 to 2022 when temperature or salinity were above or below critical thresholds, respectively.
 The slope estimates in the "Both" column indicate trends in the number of days when both temperature and salinity were above or below critical thresholds.
@@
 
-Summary of mixed-effects models evaluating increases in the number of days each year from 1998 to 2022 when temperature or salinity were above or below critical thresholds, respectively.
+Table S4: Summary of mixed-effects models evaluating increases in the number of days each year from 1998 to 2022 when temperature or salinity were above or below critical thresholds, respectively.
 The slope estimates in the "Both" column indicate trends in the number of days when both temperature and salinity were above or below critical thresholds.
@@
 
-Number of days at the beginning and end of the period of record when temperature or salinity were above or below critical thresholds, respectively.
+Table S5: Number of days at the beginning and end of the period of record when temperature or salinity were above or below critical thresholds, respectively.
 Values are the estimated mean number of days (plus standard error) from 1975 to 2022.
@@
 
-Number of days at the beginning and end of the period of record when temperature or salinity were above or below critical thresholds, respectively.
+Table S6: Number of days at the beginning and end of the period of record when temperature or salinity were above or below critical thresholds, respectively.
 Values are the estimated mean number of days (plus standard error) from 1998 to 2022.

### 2023-12-15 manu-draft.qmd
minor results edits

@@
 
-Linear mixed-effects models showed varying trends in the potential stressor metrics for the number of days when temperature was above, salinity was below, or both occurred relative to different thresholds.
+Linear mixed-effects models showed similar trends in each bay segment for the number of days when temperature was above different thresholds.
 All of the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed significantly increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (Table S3).
-The estimated slopes for the number of days when temperature was above 30 $^\circ$C were similar among bay segments, varying from 0.9 days per year in LTB to 1.1 days per year in MTB.
+The estimated slopes for the number of days when temperature was above 30 $^\circ$C varied from 0.9 days per year in LTB to 1.1 days per year in MTB.
 Likewise, the average number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied from 8 (OTB, HB) to 21 (LTB) days in 1975 to 55 (HB) to 63 (LTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022).
+Trends for the other two temperature thresholds were similar.
 
@@
 None of the bay segments had significantly increasing number of days per year when salinity was below 15 ppt.
-For MTB, only the number of days when salinity was below 25 ppt was significantly increasing at a rate of 0.7 days per year.
+The only significant model for MTB was the number of days when salinity was below 25 ppt, wich was increasing at a rate of 0.7 days per year.
 Both HB and OTB had significantly increasing number of days per year when salinity was below 20 and 25 ppt, although the slopes varied such that the rates in OTB were nearly double those in HB.
@@
 
-The number of days when both temperature was above a threshold and salinity was below a threshold also varied by bay segment.
-Four models were significant for OTB, seven for HB, four for MTB, and none for LTB.
+The number of days when both temperature was above and salinity was below a threshold also varied by bay segment.
+Four models were significant for OTB, seven for HB, four for MTB, and none for LTB (Table S3).
 The number of days when temperature was above 29 $^\circ$C and salinity was below 25 ppt had the largest slopes of 1.4, 1.2, and 0.7 days per year for OTB, HB, and MTB, respectively.
@@
 Temperature trends were similar among segments, whereas the number of days when salinity was below the threshold varied by proximity to the Gulf of Mexico (@tbl-mixdaytab).
-The number of days when both temperature was above and salinity was below the threshold generally followed the trends for the number of days when salinity was below the threshold, with the slopes varying by bay segment.
+The number of days when both temperature was above and salinity was below the threshold generally followed the trends for the number of days when salinity was below the threshold.
 These thresholds were used for comparison to seagrass changes described below, based primarily on the statistical strength of the trends and the variance of counts across stations within each bay segment (points in @fig-mixeff).
 That is, more restrictive thresholds did not provide sufficient counts of days per year to develop models and the chosen thresholds were based primarily on statistical considerations.
-The thresholds likely have minimal ecological significance, but rather provide a useful metric for tracking physical changes in the water column that are likely correlated with seagrass change.
 
@@
 
-Linear models to assess the potential effects of temperature and salinity on seagrass changes provided modest evidence that seagrass loss after 2016 was driven by climate stressors.
+Linear models to assess the potential effects of temperature and salinity on seagrass change provided modest evidence that seagrass loss after 2016 was driven by climate stressors.
 The most parsimonious model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence (`r epcmodsum1`).
@@
 However, the interaction between time period and temperature was not significant, suggesting no difference between time periods.
-The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure [-@fig-sgmod]a, @tbl-modtab).
+The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure \[-@fig-sgmod\]a, @tbl-modtab).
 The interaction between the temperature and salinity metric was included in the most parsimonious model (marginally significant, p \< 0.1), although the shape of the relationship of temperature with time period or salinity with time period did not noticeably change using different conditional values for salinity or temperature, respectively.
 The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (`r epcmodsum2`), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab).
+Note that Figure [@fig-sgmod] shows model predictions for OTB only, although the shape of the relationships do not change for other bay segments other than as additive shifts of the curves up or down (i.e., no interactions of bay segments with the other predictors were in the original models).
 
 The most parsimonious model for the FIM dataset also showed potential temperature and salinity associations with seagrass change after 2016, although explanatory power was lower than the EPC models likely due to larger spread of the data (`r fimmodsum`).
-Increases in temperature and decreases in salinity were both associated with reductions in percent cover after 2016, whereas no relationships were observed prior to 2016 (p \< 0.05 and p \< 0.1 for the interactions between temperature, salinity with time period, respectively, Figure [-@fig-sgmod]b, @tbl-modtab).
+Increases in temperature and decreases in salinity were both associated with reductions in percent cover after 2016, whereas no relationships were observed prior to 2016 (p \< 0.05 and p \< 0.1 for the interactions between temperature or salinity with time period, respectively, Figure [-@fig-sgmod]b, @tbl-modtab).
 The interaction between salinity and temperature was not included as a significant term in the final model.

### 2023-12-18 manu-draft.qmd
Minor methods text revision suggestions

@@
 First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at 45 stations in the major bay segments since the late 1970s (Figure [-@fig-map]b).
-These data are used for regulatory assessments for the nutrient TMDL in Tampa Bay.
+These data are used for regulatory assessments for the nutrient TMDL <!--# spell out --> in Tampa Bay.
 Water quality samples are collected at each station from surface water grabs (e.g., nutrients) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom.
@@
 The second dataset used to evaluate water quality trends was available from the Florida Fish and Wildlife Conservation Commission (FWC).
-The Fisheries Independent Monitoring (FIM) program administered by FWC provides monthly surveys of the entire nekton community in Tampa Bay, including species richness and abundance, using multiple survey gear types that target different habitats.
-A stratified sample design is used to target multiple habitats at different sites sampled each month.
-We used data from the 21.3 meter center-bag nearshore seine that targets shallow habitats where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present, Figure [-@fig-map]c).
-These data are collected by pulling the seine adjacent to the shore to sample approximately 140 m$^2$ of bay bottom [@schrandt2021].
-In addition to collecting fish and selected invertebrates, *in situ* physical measurements for water temperature and salinity are collected at the beginning and end of the seine haul, and typically at the surface and bottom.
-Only measurements at the beginning of the seine haul and from the bottom were used.
+The Fisheries Independent Monitoring (FIM) program administered by FWC provides monthly surveys of the entire nekton community in Tampa Bay, including species richness and abundance, using multiple gear types that target different habitats.
+A stratified sampling design is used to select sites for 21.3 meter center-bag <!--# Looking at the code - you pulled Gear 20, which would include both shoreline and offshore (>5m from shore) sets; if you want just offshore sets select by strata A and B only.
+If both deployments kept, suggest deleting nearshore --> seines that target shallow habitats (\<1.5 m) where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present, Figure [-@fig-map]c).
+Seines were deployed (1) parallel to bay shorelines and (2) offshore into the prevailing current at least 5 m from the shoreline to sample approximately 140 m$^2$ of bay bottom [@schrandt2021].
+In addition to collecting fish and selected invertebrates, *in situ* physical measurements for water temperature and salinity are collected at the bag, and at the surface and at 1 m intervals to the bottom.
+Only measurements from the bottom were used.
 Seagrass data are also provided for each site, with information on species and cover.
@@
 The third and final dataset evaluated was from the Pinellas County Department of Environmental Management (PDEM).
-Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES reporting.
+Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES <!--# spell out --> reporting.
 Data were obtained by request to PDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (Figure [-@fig-map]d, also available at <https://wateratlas.usf.edu/>).
@@
 
-All of the organizations that provided water quality datasets participate in the Southwest Florida Regional Ambient Monitoring Program.
+All of the organizations that provided water quality datasets participate in the Southwest Florida Regional Ambient Monitoring Program <!--# Does FIM participate in this?
+-->.
 This *ad hoc* group has routinely met quarterly to ensure similar standards and methods are used for the collection and processing of surface water quality monitoring data.
@@
 However, the interaction between time period and temperature was not significant, suggesting no difference between time periods.
-The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure \[-@fig-sgmod\]a, @tbl-modtab).
+The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure [-@fig-sgmod]a, @tbl-modtab).
 The interaction between the temperature and salinity metric was included in the most parsimonious model (marginally significant, p \< 0.1), although the shape of the relationship of temperature with time period or salinity with time period did not noticeably change using different conditional values for salinity or temperature, respectively.

### 2023-12-18 manu-draft.qmd
Abstract and Intro up to Line 74 for manu-draft.qmd
Initial edits up to line 74.

@@
 In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth.
-Changing physical conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be linked to climate stress.
+Changing physical water quality conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be linked to climate stress.
 Three datasets of varying sample design and coverage were evaluated to characterize physical environments in Tampa Bay and the potential relationships with seagrass change.
-Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 ppt per year.
+Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 PSU per year.
 These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay segments during the summer months.
-Simple linear models provided limited evidence that the recent seagrass decline is associated with hotter and fresher conditions.
-Trends in warming and increased precipitation in the region will continue, further creating suboptimal conditions for seagrasses in Tampa Bay.
-These results provide a motivating example for resource managers to consider the likelihood of reduced resilience of estuarine resources due to shifting ecological baselines from climate change.
-Conventional management approaches that focus on limiting nutrient loads should be continued, but the effectiveness may be limited and additional actions will be needed to maintain ecosystem health in the future.
+Simple linear models provided limited evidence that recent seagrass declines are associated with hotter and fresher conditions.
+Trends in warming and increased precipitation in the region are likely to continue, further creating suboptimal conditions for seagrasses in Tampa Bay.
+These results should compel resource managers to consider the likelihood that reduced resilience of estuarine resources due to shifting ecological baselines from climate change could complicate long-standing management paradigms.
+While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be limited and warrant additional, complementary interventions to maintain ecosystem health into the future.
 
@@
 Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021].
-These losses are comparable, if not more, than other critical coastal environments [@duarte2008a], which represents an existential threat to the natural and economic services that these habitats provide.
-Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @orth2006; @burkholder2007], whereas natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020].
-Management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].
+These losses are comparable, if not more significant, than other critical coastal environments [@duarte2008a], which represents an existential threat to the natural and economic services that these habitats provide.
+Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @orth2006; @burkholder2007].
+Furthermore, natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020].
+Contemporary management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].
 
-The sustained coverage or restoration of seagrasses in stressed environments requires management actions that create supportive conditions for growth.
-A long-standing approach adopted by numerous management entities has been the control of nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @greening2014].
+The sustained coverage or restoration of seagrasses in stressed coastal environments requires management actions that support growth and coverage expansion.
+A long-standing approach adopted by numerous management entities has been the control of external nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @greening2014].
 There are limited examples of successful recovery of seagrass through control of nutrient inputs alone, primarily because of the difficulty in identifying and regulating both point and diffuse non-point sources.
-Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014] and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in seagrass areal coverage through sustained and long-term reductions in nutrients.
+Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014] and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in seagrass areal coverage through sustained and long-term reductions in external nutrient loads.
 Cooperation among management, regulatory, public, and private sectors were critical aspects of both examples [@tango2016; @sherwood2016].
 In other cases, the reversal of seagrass losses through nutrient reductions alone may not be possible because of system hysteresis, where the path to recovery is not the same as the path to decline [@maxwell2016].
-Costly alternative management actions, in addition to nutrient reductions, are needed in these situations.
-Management actions are much more effective in estuaries that are stressed but not past critical thresholds that define an alternative state.
+Complementary management actions, in addition to nutrient reductions, are needed in these situations.
+Management actions with co=benefits are much more effective in estuaries that are stressed but not past critical thresholds that produce an alternative ecosystem state.
 

### 2024-01-08 manu-draft.qmd
abstract text change
intro text change
also consider "... losses are comparable to, if not greater than"?

@@
 
-Climate change is an existential threat to environmental resources that sustain human health and the environment.
+Climate change is an existential threat to natural resources that sustain human health and the environment.
 In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth.
@@
 Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021].
-These losses are comparable, if not more significant, than other critical coastal environments [@duarte2008a], which represents an existential threat to the natural and economic services that these habitats provide.
+These losses are comparable to, if not more significant than, other critical coastal environments [@duarte2008a], which represents an existential threat to the natural and economic services that these habitats provide.
 Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @orth2006; @burkholder2007].

### 2024-01-08 manu-draft.qmd
intro text changes

@@
 Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021].
-These losses are comparable to, if not more significant than, other critical coastal environments [@duarte2008a], which represents an existential threat to the natural and economic services that these habitats provide.
+These losses are comparable to, if not more significant than, other critical coastal environments [@duarte2008a], which threatens the natural and economic services that these habitats provide.
 Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @orth2006; @burkholder2007].
@@
 Complementary management actions, in addition to nutrient reductions, are needed in these situations.
-Management actions with co=benefits are much more effective in estuaries that are stressed but not past critical thresholds that produce an alternative ecosystem state.
+Management actions with co-benefits are much more effective in estuaries that are stressed but not past critical thresholds that produce an alternative ecosystem state.
 

### 2024-01-17 manu-draft.qmd
Minor text edits to Intro

@@
 
-The sustained coverage or restoration of seagrasses in stressed coastal environments requires management actions that support growth and coverage expansion.
+The sustained coverage or restoration of seagrasses in stressed coastal environments requires management actions that support vegetative growth, reproduction, and coverage expansion.
 A long-standing approach adopted by numerous management entities has been the control of external nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @greening2014].
@@
 This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay estuary over the last fifty years.
-Three datasets of varying sample design and temporal coverage were used to assess the primary hypothesis that conditions in Tampa Bay are trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges.
+Three datasets of varying sample design and temporal coverage were used to assess the primary hypothesis that Tampa Bay is trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges.
 This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring describing cover of individual species, and synoptic seagrass data collected with the water quality surveys.

### 2024-01-17 manu-draft.qmd
Revised SWFWMD flight window
Revised SG transect survey time period

@@
 The Southwest Florida Water Management District has estimated areal coverage of seagrasses approximately biennially since the late 1980s (Figure [-@fig-seagrasschg]a, available at <https://data-swfwmd.opendata.arcgis.com/>).
-These maps are created by photointerpretation of aerial images obtained at the end of the growing season, typically during November-December.
+These maps are created by photointerpretation of aerial images obtained at the end of the growing season, typically during November-February.
 No species information is provided.
@@
 For comparison to seagrass, the annual metrics were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends.
-Transect surveys are typically done in the fall at the end of the growing season.
+Transect surveys are typically done in the late summer during peak growing season.
 The average sampling date for each year across the subset of transects in each bay segment was estimated and the stressor metrics were referenced accordingly.

### 2024-01-17 manu-draft.qmd
Added Furman to Acknowledgments

@@
 
-This work benefited from constructive discussions with the Technical Advisory Committee of the Tampa Bay Estuary Program (TBEP), Mike Wessel, Ray Pribble, and Dave Tomasko.
+This work benefited from constructive discussions with the Technical Advisory Committee of the Tampa Bay Estuary Program (TBEP), Mike Wessel, Ray Pribble, Brad Furman, and Dave Tomasko.
 TBEP funding for this work stems from EPA Section 320 Grant Funds, and the TBEP's local government partners (Hillsborough, Manatee, Pasco, and Pinellas Counties; the Cities of Clearwater, St.

### 2024-01-17 manu-draft.qmd
various text edits

@@
 Tampa Bay is the largest open-water estuary in Florida covering 400 mi$^2$ (1,036 km$^2$) and the second largest in the Gulf of Mexico.
-The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee Rivers contributing a majority of freshwater inflow to the bay.
+The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee rivers contributing a majority of freshwater inflow to the bay.
 Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer.
@@
 Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality primarily from nutrient inputs from wastewater and stormwater [@greening2014].
-Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia Rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small tributaries [@tbep0623].
+Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small tributaries and manmade flood control conveyances [@tbep0623].
 Notably, Old Tampa Bay has restricted circulation from multiple land bridges that traverse the bay [@sherwood2015; @luther2022].
@@
 -->.
-This *ad hoc* group has routinely met quarterly to ensure similar standards and methods are used for the collection and processing of surface water quality monitoring data.
+This *ad hoc* group meets quarterly to ensure similar standards and methods are used for the collection and processing of surface water quality monitoring data.
 Split-samples evaluated by each organization are also compared to assess precision between different laboratories.
@@
 As Tampa Bay becomes hotter and fresher, similar load allocations may no longer be effectively assimilated.
-Strategies can be pursued to reduce allowable load inputs from regulated entities and to create additional resilience through more aggressive management strategies (e.g., additional stormwater controls, hydrological modifications).
+Strategies to mimic the pre-development flow of water or that further reduce allowable load inputs from regulated entities (e.g., additional stormwater controls, hydrological modifications) may be pursued to confer additional resilience and adaptive capacity.
 These considerations are especially critical for upper parts of Tampa Bay where a majority of seagrass loss has occurred and where temperature and salinity trends have been most pronounced.

### 2024-01-22 manu-draft.qmd
replace rainfall tia data with swfwmd data, reevaluate trend to rainy season #2

@@
 Two primary sources of data have been used to track seagrass change over time in Tampa Bay.
-The Southwest Florida Water Management District has estimated areal coverage of seagrasses approximately biennially since the late 1980s (Figure [-@fig-seagrasschg]a, available at <https://data-swfwmd.opendata.arcgis.com/>).
+The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s (Figure [-@fig-seagrasschg]a, available at <https://data-swfwmd.opendata.arcgis.com/>).
 These maps are created by photointerpretation of aerial images obtained at the end of the growing season, typically during November-February.
@@
 An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions related to long-term (multi-decadal) climate change.
-For comparison to water quality conditions, meteorological data describing air temperature and precipitation were obtained from Tampa International Airport where daily measurements have been collected since 1939.
-The *rnoaa* R package [@chamberlain23] was used to retrieve these data starting from 1975 when EPC monitoring began in Tampa Bay.
-Annual averages for air temperature and cumulative annual precipitation were calculated.
-Additionally, the Standardized Precipitation Index [SPI, @begueria2013] was estimated from the daily rainfall data to identify periods of time when rainfall significantly deviated from the long-term average [using the *spei* R package, @begueria23].
+For comparison to water quality conditions, daily air temperature data were obtained from Tampa International Airport [using the *rnoaa* R package, @chamberlain23] and monthly rainfall data for the Tampa Bay watershed were obtained from the SWFWMD (<https://www.swfwmd.state.fl.us/resources/data-maps/rainfall-summary-data-region>).
+Only rainfall data for the rainy season (June to September) were evaluated.
+Additionally, the Standardized Precipitation Index [SPI, @begueria2013] was estimated from the rainfall data to identify periods of time when rainfall significantly deviated from the long-term average [using the *spei* R package, @begueria23].
 Annual hydrologic loading data to Tampa Bay beginning in 1985 were also obtained for comparison to annual precipitation [@tbep0623].
@@
 prp <- speidat %>% 
-  select(yr, precip_mm) %>% 
+  select(yr, mo, precip_mm) %>%
+  filter(mo %in% 6:9) %>% 
   summarise(prptot = sum(precip_mm), .by = yr)
@@
 prpdif <- round(prpprd$fit[nrow(prp)]- prpprd$fit[1], 1)
+
+prp2 <- speidat %>% 
+  select(yr, mo, precip_mm) %>%
+  filter(mo %in% 6:8) %>% 
+  summarise(prptot = sum(precip_mm), .by = yr)
+prpmod2 <- lm(prptot ~ yr, prp2)
+prpslo2 <- round(coefficients(summary(prpmod2))[2, 1], 2)
+prppvl2 <- p_txt(coefficients(summary(prpmod2))[2, 4])
+prprsq2 <- round(summary(prpmod2)$adj.r.squared, 2)
+prpprd2 <- predict(prpmod2, se = T)
+prpstr2 <- paste0(round(prpprd2$fit[1], 1), ' (+/-', round(prpprd2$se.fit[1], 2), ' st. err.)')
+prpend2 <- paste0(round(prpprd2$fit[nrow(prp)], 1), ' (+/-', round(prpprd2$se.fit[nrow(prp)], 2), ' st. err.)')
+prpdif2 <- round(prpprd2$fit[nrow(prp)] - prpprd2$fit[1], 1)
 ```
 
-Long-term meteorological data from Tampa International Airport showed significantly increasing trends for air temperature and precipitation ([@fig-meteowqraw]).
+Long-term meteorological data showed increasing trends for air temperature and precipitation ([@fig-meteowqraw]).
 Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`).
 Mean annual air temperature in 1975 was `r airtempstr` $^\circ$C, whereas current mean annual air temperature in 2022 was `r airtempend` $^\circ$C, showing an overall increase in the period of record of `r airtempdif` $^\circ$C.
-Similarly, total precipitation has increased by `r prpslo` mm per year (`r prppvl`, $R^2$ `r prprsq`).
-Total precipitation in 1975 was `r prpstr` mm, whereas current total precipitation in 2022 was `r prpend` mm, showing an overall increase in the period of record of `r prpdif` mm.
+Similarly, total precipitation during the rainy season has increased by `r prpslo` mm per year, although the model was not significant (`r prppvl`, $R^2$ `r prprsq`).
+Removing September from the rainy season showed a significantly increasing trend of `r prpslo2` mm per year (`r prppvl2`, $R^2$ `r prprsq2`).
+Using this model, total precipitation in 1975 was `r prpstr2` mm, whereas current total precipitation in 2022 was `r prpend2` mm, showing an overall increase in the period of record of `r prpdif2` mm.
 Trends in total annual hydrologic load from 1985 to present followed total annual precipitation, although the linear model was not significant likely due to a shorter period of record.
-The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and the last ten years.
+The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and 2015-2020.
 
@@
 
-![Long-term air temperature, precipitation, hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1975 to 2022.
-Points for water temperature and salinity are colored by sampling location in the water column and show the average (95% confidence interval) across all stations in each bay segment and sampling months for each year.
+![Long-term air temperature, precipitation (Jun-Aug), hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1975 to 2022.
+Points for water temperature and salinity are colored by depth and shown as the average (95% confidence interval) across all stations in each bay segment and sampling months for each year.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/meteowqraw.png){#fig-meteowqraw}

### 2024-01-22 supplement.qmd
replace rainfall tia data with swfwmd data, reevaluate trend to rainy season #2

@@
 
-![Air temperature, precipitation, hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1998 to 2022.
-Points for water temperature and salinity are colored by sampling location in the water column and show the average (95% confidence interval) across all stations in each bay segment and sampling months for each year.
+![Air temperature, precipitation (Jun-Aug), hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1998 to 2022.
+Points for water temperature and salinity are colored by depth and shown as the average (95% confidence interval) across all stations in each bay segment and sampling months for each year.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppmeteowqraw.png){#fig-suppmeteowqraw}

### 2024-01-22 R/dat_proc.R

@@ -14,6 +14,7 @@ library(SPEI)
 library(rnoaa)
 library(dataRetrieval)
 library(readxl)
+library(zoo)
 
 source(here('R/funcs.R'))
 yrsel <- c(1998, 2022)
@@ -60,27 +61,68 @@ resdat <- res %>%
   unnest('data') %>% 
   unnest('ests') %>% 
   ungroup() %>% 
-  select(date, prcp, tmax, tmin) %>% 
+  select(date, tmax, tmin) %>% 
   mutate(
     yr = year(date),
     mo = month(date),
     tavg_c = (tmax + tmin) / 2,
     tavg_c = tavg_c / 10,
-    tavg_c = na.approx(tavg_c),
-    precip_mm = prcp / 10, 
-    precip_in = precip_mm * 0.0393701
+    tavg_c = na.approx(tavg_c)
   ) %>% 
-  select(date, yr, mo, tavg_c, precip_mm, precip_in)
+  select(date, yr, mo, tavg_c)
+
+# monthly rainfall data from swfwmd 
+# https://www.swfwmd.state.fl.us/resources/data-maps/rainfall-summary-data-region
+# file is from the link "USGS watershed"
+download.file(
+  'https://www4.swfwmd.state.fl.us/RDDataImages/surf.xlsx?_ga=2.186665249.868698214.1705929229-785009494.1704644825',
+  here('data-raw/swfwmdrainfall.xlsx'),
+  mode = 'wb'
+  )
+
+raindat <- readxl::excel_sheets(here('data-raw/swfwmdrainfall.xlsx')) %>% 
+  grep('jan|feb|mar|apr|may|jun|jul|aug|sep|oct|nov|dec', ., value = TRUE) %>% 
+  tibble(
+    mo = .
+  ) %>% 
+  nest(.by = mo) %>% 
+  mutate(
+    data = purrr::map(mo, function(mo){
+      
+      read_excel(here('data-raw/swfwmdrainfall.xlsx'), sheet = mo, skip = 1) %>% 
+        filter(Year %in% 1975:2022) %>% 
+        select(
+          yr = Year, 
+          precip_in = `Tampa Bay/Coastal Areas`
+        ) %>% 
+        mutate_all(as.numeric)
+
+    })
+  ) %>% 
+  unnest('data') %>% 
+  mutate(
+    mo = gsub('\\-usgsbsn$', '', mo),
+    mo = as.numeric(factor(mo,
[…]

### 2024-01-22 R/figs.R

@@ -266,10 +266,17 @@ hydrodat <- totanndat %>%
   ) 
 
 toplo1 <- speidat %>% 
-  select(yr, precip_mm, tavg_c) %>% 
+  select(yr, tavg_c) %>% 
+  summarise(
+    tavg_c = mean(tavg_c), 
+    .by = 'yr'
+  )
+
+toplo2 <- speidat %>% 
+  filter(mo %in% c(6:8)) %>% 
+  select(yr, precip_mm) %>% 
   summarise(
     precip_mm = sum(precip_mm), 
-    tavg_c = mean(tavg_c), 
     .by = 'yr'
   )
 
@@ -292,10 +299,10 @@ p1 <- ggplot(toplo1, aes(x = yr, y = tavg_c)) +
   ) +
   labs(
     x = NULL, 
-    y = 'Air temp. (\u00B0 C)'
+    y = 'Air temp. (\u00B0C)'
   )
 
-p2 <- ggplot(toplo1, aes(x = yr, y = precip_mm / 1e3)) + 
+p2 <- ggplot(toplo2, aes(x = yr, y = precip_mm / 1e3)) + 
   geom_line() + 
   geom_point() + 
   geom_smooth(method = 'lm', se = F, formula = y~x, color = 'blue') +
@@ -305,7 +312,7 @@ p2 <- ggplot(toplo1, aes(x = yr, y = precip_mm / 1e3)) +
   ) +
   labs(
     x = NULL, 
-    y = 'Precip. (m)'
+    y = 'Jun-Aug prec. (m)'
   )
 
 p3 <- ggplot(hydrodat, aes(x = year, y = hy_load)) +
@@ -319,7 +326,7 @@ p3 <- ggplot(hydrodat, aes(x = year, y = hy_load)) +
   ) +
   labs(
     x = NULL, 
-    y = expression(paste('Hyd. load (', 10^3, ' t/yr)'))
+    y = expression(paste('Hy. load (', 10^3, ' t/y)'))
   )
 
 p4 <- ggplot(speidat, aes(x = date, y = spi, fill = spisign)) + 
@@ -427,7 +434,7 @@ p6 <- ggplot(toplo6, aes(x = yr, y = avev, group = loc, color = loc)) +
     shape = NULL
   )
 
[…]

### 2024-01-23 manu-draft.qmd
text edits and minor comments in line with text

@@
 Changing physical water quality conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be linked to climate stress.
-Three datasets of varying sample design and coverage were evaluated to characterize physical environments in Tampa Bay and the potential relationships with seagrass change.
+Three datasets of varying sampling design and coverage were evaluated to characterize physicochemical environments in Tampa Bay and the potential relationships with seagrass change.
 Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 PSU per year.
-These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay segments during the summer months.
+These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay during the summer months.
 Simple linear models provided limited evidence that recent seagrass declines are associated with hotter and fresher conditions.
@@
 Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021].
-These losses are comparable to, if not more significant than, other critical coastal environments [@duarte2008a], which threatens the natural and economic services that these habitats provide.
-Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @orth2006; @burkholder2007].
+These losses are comparable to, if not more significant than, other interconnected critical coastal environments such as mangroves and salt marshes [@duarte2008a <!--# any more recent references?
+-->], which threatens the natural and economic services that these habitats provide.
+Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; <!--# Add Hall et al.
+1999, Florida bay seagrass dieoff and recovery --> @orth2006; @burkholder2007].
 Furthermore, natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020].
@@
 In addition to sea-level rise, the most anticipated effects of climate change in coastal environments are increasing temperature and alteration of precipitation patterns.
-These changes will profoundly alter physical habitats, creating suboptimal or uninhabitable conditions for many species [@madeira2012; @lefcheck2017; @hensel2023].
+These changes will profoundly alter physicochemical habitats, creating suboptimal or uninhabitable conditions for many species [@madeira2012; @lefcheck2017; @hensel2023].
 @lefcheck2017 evaluated interactive effects of water clarity and rising temperatures on seagrasses with over 30 years of data in Chesapeake Bay.
@@
 In Tampa Bay, Florida, seagrasses are a primary indicator of bay health that have been the focus of management efforts over the last thirty years [@sherwood2017].
-Through successful reduction of nitrogen loads by 2/3 pre-regulation, seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres [16,857 ha, @greening2006; @greening2014].
+Through successful reduction of nitrogen loads, seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres [16,857 ha, @greening2006; @greening2014].
 Seagrass distribution and abundance has been dynamic throughout this period, overall responding positively to increases in water clarity with nutrient load reductions, while also demonstrating more short-term variability in response to regional climate events [@greening2006].
@@
 This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay estuary over the last fifty years.
-Three datasets of varying sample design and temporal coverage were used to assess the primary hypothesis that Tampa Bay is trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges.
+Three datasets of varying sampling designs and temporal coverage were used to assess the primary hypothesis that Tampa Bay is trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges.
 This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring describing cover of individual species, and synoptic seagrass data collected with the water quality surveys.
@@
 Tampa Bay is the largest open-water estuary in Florida covering 400 mi$^2$ (1,036 km$^2$) and the second largest in the Gulf of Mexico.
-The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee rivers contributing a majority of freshwater inflow to the bay.
-Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer.
-The watershed is heavily developed and includes over 3 million people with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023].
+The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee rivers that contribute to a majority of freshwater inflow to the bay.
+Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer<!--# reference?
+-->.
+The watershed is heavily developed and includes over 3 million people <!--# reference economic valuation?
+-->with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023].
 The geology of the watershed is rich in phosphates and mining activities have greatly altered the landscape, with notable spills and releases of wastewater that have affected water quality and biological resources [@garrett2011; @beck2022a].
@@
 Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small tributaries and manmade flood control conveyances [@tbep0623].
-Notably, Old Tampa Bay has restricted circulation from multiple land bridges that traverse the bay [@sherwood2015; @luther2022].
+Notably, Old Tampa Bay has restricted circulation from multiple land bridges <!--# use causeways instead?
+-->that traverse the bay [@sherwood2015; @luther2022].
 Recurring seasonal harmful algal blooms of the dinoflagellete *Pyrodinium bahamense* have contributed to exceedances of the chlorophyll-a regulatory standard in Old Tampa Bay [@lopez2023].
@@
 First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at 45 stations in the major bay segments since the late 1970s (Figure [-@fig-map]b).
-These data are used for regulatory assessments for the nutrient TMDL <!--# spell out --> in Tampa Bay.
+These data are used for regulatory assessments for the nutrient Total Maximum Daily Load in Tampa Bay.
 Water quality samples are collected at each station from surface water grabs (e.g., nutrients) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom.
@@
 The third and final dataset evaluated was from the Pinellas County Department of Environmental Management (PDEM).
-Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES <!--# spell out --> reporting.
+Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has only been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES <!--# spell out --> reporting.
 Data were obtained by request to PDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (Figure [-@fig-map]d, also available at <https://wateratlas.usf.edu/>).
@@
 Water temperature and salinity trends using the EPC, FIM, and PDEM data were similarly evaluated by averaging the monthly data each year for each bay segment.
-Linear trends for these data were evaluated based on averages of all stations and months within each year and bay segment, which allowed for comparable statistical power between datasets with different sample designs.
+Linear trends for these data were evaluated based on averages of all stations and months within each year and bay segment, which allowed for comparable statistical power between datasets with different sampling designs.
 
@@
 Optimal temperature ranges are similar between species, with reduced growth observed at temperatures above 30$^\circ$ C [@zieman1975; @lewis1985].
+<!--# What about Syringodium?
+Found in deeper water, any info on temperature range -->
 
@@
 First, we developed metrics of potential temperature and salinity stress by quantifying the maximum number of continuous days each year when temperature was above or salinity was below a given threshold.
-This approach assumed that stress could be observed based on duration of exposure (i.e., maximum number of continuous days each year) relative to a threshold that may are may not be outside of the optimum range for seagrasses.
+This approach assumed that stress could be observed based on duration of exposure (i.e., maximum number of continuous days each year) relative to a threshold that may or may not be outside of the optimum range for seagrasses.
 These metrics were quantified from the monthly long-term observations in the EPC data.
@@
 
-For comparison to seagrass, the annual metrics were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends.
+For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends.
 Transect surveys are typically done in the late summer during peak growing season.
@@
 Similarly, total precipitation during the rainy season has increased by `r prpslo` mm per year, although the model was not significant (`r prppvl`, $R^2$ `r prprsq`).
-Removing September from the rainy season showed a significantly increasing trend of `r prpslo2` mm per year (`r prppvl2`, $R^2$ `r prprsq2`).
+Removing September from the rainy season showed a significantly increasing trend of `r prpslo2` mm per year (`r prppvl2`, $R^2$ = `r prprsq2`).
 Using this model, total precipitation in 1975 was `r prpstr2` mm, whereas current total precipitation in 2022 was `r prpend2` mm, showing an overall increase in the period of record of `r prpdif2` mm.
@@
 The strongest trends were observed for the EPC dataset, which had the longest record from 1975 to 2022.
-The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths and minimal stratification in Tampa Bay (@fig-meteowqraw).
+The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths <!--# depth differences among datasets might be useful here, shallow depths for FIM, larger range for other datasets?
+-->and minimal stratification in Tampa Bay (@fig-meteowqraw).
 Trends in water temperature were similar across bay segments with significant increases varying from 0.03 to 0.04 $^\circ$C per year, with a total increase of 1.4 (OTB) to 2.0 (HB) $^\circ$C from 1975 to present (@tbl-temptrndtab).
@@
 None of the bay segments had significantly increasing number of days per year when salinity was below 15 ppt.
-The only significant model for MTB was the number of days when salinity was below 25 ppt, wich was increasing at a rate of 0.7 days per year.
+The only significant model for MTB was the number of days when salinity was below 25 ppt, which was increasing at a rate of 0.7 days per year.
 Both HB and OTB had significantly increasing number of days per year when salinity was below 20 and 25 ppt, although the slopes varied such that the rates in OTB were nearly double those in HB.
@@
 The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (`r epcmodsum2`), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab).
-Note that Figure [@fig-sgmod] shows model predictions for OTB only, although the shape of the relationships do not change for other bay segments other than as additive shifts of the curves up or down (i.e., no interactions of bay segments with the other predictors were in the original models).
+Note that [@fig-sgmod] shows model predictions for OTB only, although the shape of the relationships do not change for other bay segments other than as additive shifts of the curves up or down (i.e., no interactions of bay segments with the other predictors were in the original models).
 
@@
 At a local scale, these changes have been demonstrated herein for Tampa Bay where the evaluation of long-term trends in water temperature and salinity showed clear changes mirrored by long-term changes in air temperature and precipitation.
-Over the previous fifty years, the Bay has consistently trended towards hotter and fresher conditions, where water temperature has approximately increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year.
+Over the previous fifty years, Tampa Bay has consistently trended towards hotter and fresher conditions, where water temperature has approximately increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year.
 These changes were demonstrated in three long-term datasets with different sampling methods and periods of record.
-Understandably, the trends were most consistently observed in the EPC dataset with the longest period of record covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season at locations farther from the mouth of the Bay.
-These long-term changes manifested into consistent trends in the stressor metrics, where the continuous number of days each year when temperature was above, salinity was below, or both crossed thresholds showed consistent increases across the period of record.
-Notably, bottom and surface trends for water temperature and salinity in the EPC dataset were similar, suggesting that any potential stress on biotic resources related to these factors are likely occurring throughout the water column.
+Understandably, the trends were most consistently observed in the dataset with the longest period of record (EPC) covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season at locations farther from the mouth of the Bay.
+These long-term changes manifested into consistent trends in the stressor metrics; the continuous number of days each year when temperature was above and/or salinity was below defined thresholds consistently increased across the period of record.
+<!--# Do the edits to the previous sentence still get at your meaning?
+-->Notably, bottom and surface trends for water temperature and salinity in the EPC dataset were similar, suggesting that any potential stress on biotic resources related to these factors are likely occurring throughout the water column.
+<!--# Suggest removing surface water quality comparisons -->
 
@@
 Temperature trends were similar for Tampa Bay (the EPC dataset was also used), although the power to detect trends decreased for shorter time periods, as verified herein with the FIM and PDEM datasets.
-@nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, which somewhat conflicts with our results showing distinct increases in the summer, particularly for the OTB and HB segments.
+@nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, which somewhat conflicts with our results showing distinct increases in the summer, particularly for the upper bay segments (OTB and HB).
 However, our assessment evaluated trends at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to the larger spatial area.
@@
 This suggests that other factors are influencing seagrass growth and the inclusion of light attenuation in the models was likely not critical for describing the roles of temperature and salinity.
-A final consideration for our models relates to how seagrasses may influence their environment, particularly for the EPC and PDEM datasets where temperature and salinity were measured at the same locations as seagrass.
+A final consideration for our models relates to how seagrasses may influence their environment, particularly for the EPC and FIM <!--# Should be FIM here, correct?
+--> datasets where temperature and salinity were measured at the same locations as seagrass.
 For example, temperature may simply be higher in locations where seagrasses are present, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses.
@@
 Additional biotic factors could be influencing regional patterns in seagrass growth.
-In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern [@jei2022; @scolaro2023].
+In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern \[ <!--# add Indian River paper?
+-->@jei2022; @scolaro2023\].
 Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present in Tampa Bay [@Beck20b].
@@
 Salinity changes with sea-level rise have caused numerous alterations of subtidal and nearshore habitats [@brinson1995; @white2017].
-In southwest Florida, the most common ecological example is migration of mangroves upland as porewater salinity and water levels have gradually increased over the past few decades [@borchert2018].
+In southwest Florida, the most common ecological example is the expansion of mangroves upland as porewater salinity and water levels have gradually increased over the past few decades [@borchert2018].
 Alteration of salinity regimes for surface and groundwater resources are also well documented.
@@
 @dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters.
-In Tampa Bay, the long-term trends of decreasing salinity suggest that the hydrologic load into the system has had a greater influence on salinity regimes than the effects of sea-level rise.
+In Tampa Bay, the long-term trends of decreasing salinity, especially in the upper bay segments, suggest that the hydrologic load into the system has had a greater influence on salinity regimes than the effects of sea-level rise.
 This hypothesis is supported by our assessment of precipitation patterns over time, where the long-term increase is inversely correlated with the decrease in salinity.
@@
 
-Area managers should consider how these climate-related stressors may alter the effectiveness of intervention activities aimed at protecting ecological resources in Tampa Bay.
+Natural resource managers should consider how these climate-related stressors may alter the effectiveness of intervention activities aimed at protecting ecological resources in Tampa Bay.
 Management actions that have historically been effective may have inadequate capacity to maintain ecosystem status due to changing resilience of ecological resources with climate change.
@@
 
-![Map of Tampa Bay and the three datasets used for trend analysis, including (a) bay segments, 2022 seagrass coverage (green), and transect starting points (red), (b) Environmental Protection Commission (EPC) long-term monitoring sites, (c) Fisheries Independent Monitoring (FIM) random sampling for near-shore seine hauls, and (d) OTB portion of Pinellas County Department of Environmental Management (PDEM) random sampling.
+![Map of Tampa Bay and the three datasets used for trend analysis, including (a) bay segments, 2022 seagrass coverage (green), and transect starting points (red), (b) Environmental Protection Commission (EPC) long-term monitoring sites, (c) Fisheries Independent Monitoring (FIM) random sampling for seine hauls, and (d) OTB portion of Pinellas County Department of Environmental Management (PDEM) random sampling.
 Date ranges for each dataset are shown in the title.
@@
 ![Long-term air temperature, precipitation (Jun-Aug), hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1975 to 2022.
-Points for water temperature and salinity are colored by depth and shown as the average (95% confidence interval) across all stations in each bay segment and sampling months for each year.
+Points for water temperature and salinity are colored by depth and shown as the average (95% confidence interval) across all Environmental Protection Commission (EPC) stations in each bay segment and sampling months for each year.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/meteowqraw.png){#fig-meteowqraw}

### 2024-01-25 manu-draft.qmd
edits from KFW

@@
 Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021].
-These losses are comparable to, if not more significant than, other interconnected critical coastal environments such as mangroves and salt marshes [@duarte2008a <!--# any more recent references?
--->], which threatens the natural and economic services that these habitats provide.
-Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; <!--# Add Hall et al.
-1999, Florida bay seagrass dieoff and recovery --> @orth2006; @burkholder2007].
+These losses are comparable to, if not more significant than, other interconnected critical coastal environments such as mangroves and salt marshes [@duarte2008a], which threatens the natural and economic services that these habitats provide.
+Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @hall1999; @orth2006; @burkholder2007].
 Furthermore, natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020].
@@
 The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee rivers that contribute to a majority of freshwater inflow to the bay.
-Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer<!--# reference?
--->.
+Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer [@morrison2006].
+El Ni\\\~{n}o weather events have also been associated with prolonged periods of heavy rainfall contributing to seagrass reductions through increased stormwater nutrient loads [@greening2006; @morrison2006].
 The watershed is heavily developed and includes over 3 million people <!--# reference economic valuation?
@@
 This suggests that other factors are influencing seagrass growth and the inclusion of light attenuation in the models was likely not critical for describing the roles of temperature and salinity.
-A final consideration for our models relates to how seagrasses may influence their environment, particularly for the EPC and FIM <!--# Should be FIM here, correct?
---> datasets where temperature and salinity were measured at the same locations as seagrass.
+A final consideration for our models relates to how seagrasses may influence their environment, particularly for the PDEM and FIM datasets where temperature and salinity were measured at the same locations as seagrass.
 For example, temperature may simply be higher in locations where seagrasses are present, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses.
@@
 Additional biotic factors could be influencing regional patterns in seagrass growth.
-In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern \[ <!--# add Indian River paper?
--->@jei2022; @scolaro2023\].
+In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern [@hall2022; @jei2022; @scolaro2023].
 Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present in Tampa Bay [@Beck20b].

### 2024-01-25 manu-draft.qmd
reference el nino fixes #2

@@
 Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer [@morrison2006].
-El Ni\\\~{n}o weather events have also been associated with prolonged periods of heavy rainfall contributing to seagrass reductions through increased stormwater nutrient loads [@greening2006; @morrison2006].
+El Niño weather events have also been associated with prolonged periods of heavy rainfall contributing to seagrass reductions through increased stormwater nutrient loads [@greening2006; @morrison2006].
 The watershed is heavily developed and includes over 3 million people <!--# reference economic valuation?

### 2024-01-25 manu-draft.qmd
addressing KFW comments, fixes #4 #5 #6 #7 #8 #9 #10 #11 #12 #13

@@
     number-sections: true
-title: "Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary"
+title: "Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary"
 bibliography: refs.bib
@@
 El Niño weather events have also been associated with prolonged periods of heavy rainfall contributing to seagrass reductions through increased stormwater nutrient loads [@greening2006; @morrison2006].
-The watershed is heavily developed and includes over 3 million people <!--# reference economic valuation?
--->with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023].
+The watershed is heavily developed and includes over 3 million people [@tbep0923] with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023].
 The geology of the watershed is rich in phosphates and mining activities have greatly altered the landscape, with notable spills and releases of wastewater that have affected water quality and biological resources [@garrett2011; @beck2022a].
@@
 Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small tributaries and manmade flood control conveyances [@tbep0623].
-Notably, Old Tampa Bay has restricted circulation from multiple land bridges <!--# use causeways instead?
--->that traverse the bay [@sherwood2015; @luther2022].
+Notably, Old Tampa Bay has restricted circulation from multiple land bridges associated with causeways that traverse the bay [@sherwood2015; @luther2022].
 Recurring seasonal harmful algal blooms of the dinoflagellete *Pyrodinium bahamense* have contributed to exceedances of the chlorophyll-a regulatory standard in Old Tampa Bay [@lopez2023].
@@
 Time of sampling can vary, although most samples are collected from mid-morning to early afternoon.
-All surface and bottom salinity and temperature measurements for each of the 45 monitoring stations were evaluated herein.
-Trends were assessed for both surface and bottom samples, as described below, whereas only the bottom measurements were used for comparison to seagrass trends.
 The data were obtained using the tbeptools R package that imports the data directly from a stable web address provided by the EPC [@beck2021].
@@
 The Fisheries Independent Monitoring (FIM) program administered by FWC provides monthly surveys of the entire nekton community in Tampa Bay, including species richness and abundance, using multiple gear types that target different habitats.
-A stratified sampling design is used to select sites for 21.3 meter center-bag <!--# Looking at the code - you pulled Gear 20, which would include both shoreline and offshore (>5m from shore) sets; if you want just offshore sets select by strata A and B only.
-If both deployments kept, suggest deleting nearshore --> seines that target shallow habitats (\<1.5 m) where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present, Figure [-@fig-map]c).
+A stratified sampling design is used to select sites for 21.3 meter center-bag seines that target shallow habitats (\<1.5 m) where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present, Figure [-@fig-map]c).
 Seines were deployed (1) parallel to bay shorelines and (2) offshore into the prevailing current at least 5 m from the shoreline to sample approximately 140 m$^2$ of bay bottom [@schrandt2021].
@@
 The third and final dataset evaluated was from the Pinellas County Department of Environmental Management (PDEM).
-Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has only been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES <!--# spell out --> reporting.
+Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has only been used in Tampa Bay since 2003 primarily to support robust statistical assessments for National Pollutant Discharge Elimination System reporting.
 Data were obtained by request to PDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (Figure [-@fig-map]d, also available at <https://wateratlas.usf.edu/>).
@@
 
-All of the organizations that provided water quality datasets participate in the Southwest Florida Regional Ambient Monitoring Program <!--# Does FIM participate in this?
--->.
+All of the organizations that provided water quality datasets participate in the Southwest Florida Regional Ambient Monitoring Program.
 This *ad hoc* group meets quarterly to ensure similar standards and methods are used for the collection and processing of surface water quality monitoring data.
@@
 The percentage of stations in each bay segment with significantly increasing temperature or decreasing salinity trends were evaluated for each month.
-All tests evaluated both surface and bottom measurements to assess potential differences by water depth.
 
@@
 Optimal temperature ranges are similar between species, with reduced growth observed at temperatures above 30$^\circ$ C [@zieman1975; @lewis1985].
-<!--# What about Syringodium?
-Found in deeper water, any info on temperature range -->
 
@@
 The strongest trends were observed for the EPC dataset, which had the longest record from 1975 to 2022.
-The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths <!--# depth differences among datasets might be useful here, shallow depths for FIM, larger range for other datasets?
--->and minimal stratification in Tampa Bay (@fig-meteowqraw).
+The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths and minimal stratification in Tampa Bay (@fig-meteowqraw).
+Additional analyses only considered bottom water trends given comparable results within the water column.
 Trends in water temperature were similar across bay segments with significant increases varying from 0.03 to 0.04 $^\circ$C per year, with a total increase of 1.4 (OTB) to 2.0 (HB) $^\circ$C from 1975 to present (@tbl-temptrndtab).
@@
 The EPC dataset was also used to provide detailed information on station-level trends from 1975 to present (@fig-kendall, see Figure S4 for 1998 to 2022).
-All stations had significantly increasing temperature and decreasing salinity from 1975 to present in both the top and bottom of the water column, excluding one station in HB that did not have significantly decreasing bottom salinity (Figure [-@fig-kendall]a).
+All stations had significantly increasing temperature and decreasing salinity from 1975 to present, excluding one station in HB that did not have significantly decreasing bottom salinity (Figure [-@fig-kendall]a).
 Seasonally, most stations had more significant trends in the summer, early fall period for both temperature and salinity (Figure [-@fig-kendall]b, c), although some variation was observed between bay segments.
@@
 Seasonal trends in salinity did not vary as much between segments, although OTB and LTB had more stations with significantly decreasing trends than the other bay segments.
-Again, differences between top or bottom salinity trends were not observed.
 
@@
 
-Linear mixed-effects models showed similar trends in each bay segment for the number of days when temperature was above different thresholds.
+Linear mixed-effects models for the EPC data showed similar trends in each bay segment for the number of days when temperature was above different thresholds.
 All of the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed significantly increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (Table S3).
@@
 These long-term changes manifested into consistent trends in the stressor metrics; the continuous number of days each year when temperature was above and/or salinity was below defined thresholds consistently increased across the period of record.
-<!--# Do the edits to the previous sentence still get at your meaning?
--->Notably, bottom and surface trends for water temperature and salinity in the EPC dataset were similar, suggesting that any potential stress on biotic resources related to these factors are likely occurring throughout the water column.
-<!--# Suggest removing surface water quality comparisons -->
+Notably, bottom and surface trends for water temperature and salinity in the EPC dataset were similar, suggesting that any potential stress on biotic resources related to these factors are likely occurring throughout the water column.
 
@@
 For the EPC model, the interaction term was not significant for temperature, whereas the interaction was marginally significant for the salinity metric such that a negative association was observed post-2016 as compared to pre-2016 where frequency occurrence of seagrass declined with the increasing number of days when salinity was below 25 ppt.
-Likewise, the FIM models had a significant interaction term for the association of the temperature with time period and a marginally significant interaction term for the association of salinity with time period, such that percent cover of seagrasses showed a decline with increasing temperature and decreasing salinity post-2016.
+Likewise, the FIM models had a significant interaction term for the association of temperature with time period and a marginally significant interaction term for the association of salinity with time period, such that percent cover of seagrasses showed a decline with increasing temperature and decreasing salinity post-2016.
 An important distinction between the EPC and FIM models is that the former evaluated the number of days above/below thresholds each year to quantify increases in annual stress associated with temperature and salinity, whereas the latter evaluated observed temperature and salinity values at the time of seagrass sampling.
 As such, both models attempted to describe the role of these stressors on potential seagrass change, but use different independent variables given the different sampling designs of each monitoring programs.
+The FIM model also used water quality measurements collected at the same locations where seagrasses were observed, whereas the EPC model provided a general characterization of seagrass transect data with water quality at stations farther from shore.
 These differences highlight challenges describing noisy relationships in long-term ecological datasets, while also demonstrating the utility of our weight-of-evidence approach to describe these relationships.
@@
 Additional biotic factors could be influencing regional patterns in seagrass growth.
-In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern [@hall2022; @jei2022; @scolaro2023].
+In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern [@hall2022; @jei2022; @brewton2023; @scolaro2023].
 Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present in Tampa Bay [@Beck20b].
-Competitive differences between seagrasses and macroalgae are poorly understood in these systems [but see @bell1997], in addition to insufficient data in Tampa Bay to clearly describe growth patterns and nutrient cycling related to macroalgae.
+Competitive differences between seagrasses and macroalgae are poorly understood in these systems [but see @bell1997; @brewton2023], in addition to insufficient data in Tampa Bay to clearly describe growth patterns and nutrient cycling related to macroalgae.
 Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth [@beck2022a; @scolaro2023; @tomasko2023], whereas the role of chronic nutrient loading and changing climatic conditions in promoting growth and interactions with seagrasses is less understood, particularly in recent years.
@@
 In Florida Bay, for example, widespread decline of *Thalassia testudinium* has been partially attributed to elevated salinity levels beyond the optimal range for the species [@hall2016].
-Although drought has been implicated in the hypersaline conditions observed in Florida Bay, sea level rise is expected to further alter salinity dynamics in the region.
+Sea level rise is further altering salinity dynamics in the region.
 @dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters.
@@
 
-![Map of Tampa Bay and the three datasets used for trend analysis, including (a) bay segments, 2022 seagrass coverage (green), and transect starting points (red), (b) Environmental Protection Commission (EPC) long-term monitoring sites, (c) Fisheries Independent Monitoring (FIM) random sampling for seine hauls, and (d) OTB portion of Pinellas County Department of Environmental Management (PDEM) random sampling.
+![Map of Tampa Bay and the three datasets used for trend analysis, including (a) bay segments, 2022 seagrass coverage (green), and transect starting points (black), (b) Environmental Protection Commission (EPC) long-term monitoring sites, (c) Fisheries Independent Monitoring (FIM) random sampling for seine hauls, and (d) OTB portion of Pinellas County Department of Environmental Management (PDEM) random sampling.
 Date ranges for each dataset are shown in the title.
@@
 ![Long-term air temperature, precipitation (Jun-Aug), hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1975 to 2022.
-Points for water temperature and salinity are colored by depth and shown as the average (95% confidence interval) across all Environmental Protection Commission (EPC) stations in each bay segment and sampling months for each year.
+The color shades for water temperature and salinity indicate sampling location and values shown are the averages (95% confidence interval) across all Environmental Protection Commission (EPC) stations in each bay segment and sampling months for each year.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/meteowqraw.png){#fig-meteowqraw}
@@
 
-![Trends from 1975 to 2022 for temperature and salinity measurements at long-term monitoring stations in Tampa Bay.
-Results for seasonal Kendall tests by station and monitoring location (top or bottom of the water column) are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr$^{-1}$).
-Summarized seasonal trends by month are shown for (b) top and (c) bottom measurements as the percent of stations in each bay segment with significant increasing (temperature) or decreasing (salinity) trends.
+![Trends from 1975 to 2022 for bottom water temperature and salinity measurements at long-term monitoring stations in Tampa Bay.
+Results for seasonal Kendall tests by station are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr$^{-1}$).
+Summarized seasonal trends by month are shown as (b) the percent of stations in each bay segment with significant increasing (temperature) or decreasing (salinity) trends.
 Bay segment outlines are shown in (a); OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/kendall.png){#fig-kendall}
@@
 
-![Example of mixed effects models for the estimated number of days per year that temperature (red) or salinity (blue) were above or below thresholds of 30 $^\circ$C or 25 psu, respectively.
+![Example of mixed effects models for the estimated number of days per year that temperature (red) or salinity (blue) were above or below thresholds of 30 $^\circ$C or 25 psu, respectively, from the EPC data.
 The bottom row (black) shows the number of days when both temperature was above and salinity was below the thresholds.

### 2024-01-25 supplement.qmd
addressing KFW comments, fixes #4 #5 #6 #7 #8 #9 #10 #11 #12 #13

@@
 ![Air temperature, precipitation (Jun-Aug), hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1998 to 2022.
-Points for water temperature and salinity are colored by depth and shown as the average (95% confidence interval) across all stations in each bay segment and sampling months for each year.
+The color shades for water temperature and salinity indicate sampling location and values shown are the averages (95% confidence interval) across all Environmental Protection Commission (EPC) stations in each bay segment and sampling months for each year.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppmeteowqraw.png){#fig-suppmeteowqraw}

### 2024-01-25 R/figs.R

@@ -76,7 +76,7 @@ thm <- theme(
 m1 <- ggplot() + 
   ggspatial::annotation_map_tile(zoom = 11, type = 'cartolight', cachedir = system.file("rosm.cache", package = "ggspatial")) +
   geom_sf(data = sgdat, fill = 'darkgreen', color = NA, inherit.aes = F) +
-  geom_sf(data = trnpts, color = 'tomato1', inherit.aes = F) +
+  geom_sf(data = trnpts, color = 'black', inherit.aes = F) +
   annotation_north_arrow(location = 'tl', style = north_arrow_orienteering(fill = c('black', 'black'), text_col = NA), 
                          height = unit(0.5, "cm"), width = unit(0.5, "cm")) +
   annotation_scale(location = 'br', text_cex = 1) +
@@ -354,7 +354,7 @@ toplo <- epcdata %>%
                  ),
     bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB'))
   ) %>% 
-  separate(var, c('var', 'loc')) %>% 
+  separate(var, c('var', 'loc'), remove = T) %>% 
   mutate(
     var = factor(var, levels = c('Temp', 'Sal'), labels = c('Water temp. (\u00B0C)', 'Salinity (ppt)')),
     loc = factor(loc, levels = c('Top', 'Bottom'))
@@ -401,7 +401,7 @@ p5 <- ggplot(toplo5, aes(x = yr, y = avev, group = loc, color = loc)) +
   geom_smooth(method = 'lm', se = F, formula = y ~ x) +
   # geom_line(data = mixmets %>% filter(var == 'Water temp. (\u00B0C)'), aes(y = prd)) +
   facet_grid(~ bay_segment) +
-  scale_color_manual(values = c( 'steelblue4', 'steelblue1')) +
+  scale_color_manual(values = c( 'red2', 'red4')) +
   thm +
   theme(
     strip.text = element_text(size = 12),
@@ -409,8 +409,8 @@ p5 <- ggplot(toplo5, aes(x = yr, y = avev, group = loc, color = loc)) +
   ) +
   labs(
     x = NULL, 
-    y = 'Water temp. (\u00B0C)', 
-    color = NULL, 
+    y = '\u00B0C', 
+    color = 'Water temp.', 
     shape = NULL
   )
 toplo6 <- toplo %>% 
[…]

### 2024-01-25 R/funcs.R

@@ -114,7 +114,7 @@ modplo_fun <- function(mod, xlab1, ylab1, subtitle1 = NULL, title1 = NULL,
   toplo2a <- toplo2$fit
   
   ylim <- c(min(toplo1a$visregLwr, toplo2a$visregLwr), max(toplo1a$visregUpr, toplo2a$visregUpr))
-  
+
   p1 <- ggplot(toplo1a, aes(x = temp)) + 
     geom_ribbon(aes(ymin = visregLwr, ymax = visregUpr), fill = 'red2', alpha = 0.5) + 
     geom_line(aes(y = visregFit), color = 'red2') + 

### 2024-01-25 manu-draft.qmd
add costa paper and hypothesis explanation, fixes #3

@@
 Three datasets of varying sampling designs and temporal coverage were used to assess the primary hypothesis that Tampa Bay is trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges.
+This hypothesis was generated from preliminary assessments of datasets used herein and discussions with the regional scientific and management community, despite information in the primary literature suggesting the contrary for salinity [e.g., @costa2023].
 This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring describing cover of individual species, and synoptic seagrass data collected with the water quality surveys.
@@
 
-Lastly, our result showing that salinity has decreased in Tampa Bay is generally contrary to expectations for coastal systems consistent with sea-level rise projections [@alarcon2024].
+Lastly, our result showing that salinity has decreased in Tampa Bay is generally contrary to expectations for coastal systems consistent with sea-level rise projections [@costa2023; @alarcon2024].
 Salinity changes with sea-level rise have caused numerous alterations of subtidal and nearshore habitats [@brinson1995; @white2017].

### 2024-01-25 supplement.qmd
fix colors for all plots to keep temp, sal similar

@@
     number-sections: true
-title: "*Supplement* Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary"
+title: "*Supplement* Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary"
 bibliography: refs.bib

### 2024-01-26 manu-draft.qmd
added ES edits

@@
       correspondence: false
+  - Bradley T Furman:
+      email: Bradley.Furman@myfwc.com
+      institute: [FWRI]
+      correspondence: false
+  - David J Karlen: 
+      email: karlen@epchc.org
+      institute: [EPCHC]
+      correspondence: false
+  - Chris Pratt: 
+      email: prattc@epchc.org
+      institute: [EPCHC]
+      correspondence: false
   - Edward T Sherwood:
@@
   - TBEP: Tampa Bay Estuary Program, 263 13th Ave S., Suite 350, St. Petersburg, Florida 33701 USA
+  - FWRI: Florida Fish and Wildlife Conservation Commission, Florida Fish and Wildlife Research Institute, 100 8th Avenue SE, St. Petersburg, FL 33701, USA
+  - EPCHC: Environmental Protection Commission of Hillsborough County, 3629 Queen Palm Drive, Tampa, FL 33619, USA
 
@@
 Changing physical water quality conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be linked to climate stress.
-Three datasets of varying sampling design and coverage were evaluated to characterize physicochemical environments in Tampa Bay and the potential relationships with seagrass change.
+Three ambient water quality datasets of varying sampling design and coverage were evaluated to characterize physicochemical environments in Tampa Bay and the potential relationships with seagrass change.
 Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 PSU per year.
@@
 Trends in warming and increased precipitation in the region are likely to continue, further creating suboptimal conditions for seagrasses in Tampa Bay.
-These results should compel resource managers to consider the likelihood that reduced resilience of estuarine resources due to shifting ecological baselines from climate change could complicate long-standing management paradigms.
-While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be limited and warrant additional, complementary interventions to maintain ecosystem health into the future.
+These results should compel resource managers to consider the likelihood that reduced resilience of estuarine resources due to shifting ecological baselines driven by additional climate change drivers will complicate long-standing management paradigms.
+While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be confounded by other cliamte change driveres and warrant additional, complementary interventions to maintain ecosystem health into the future.
 
@@
 The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale.
-Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023].
+Seagrasses are fundamental indicators of coastal ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023].
 Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021].
@@
 The sustained coverage or restoration of seagrasses in stressed coastal environments requires management actions that support vegetative growth, reproduction, and coverage expansion.
-A long-standing approach adopted by numerous management entities has been the control of external nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @greening2014].
+A long-standing approach adopted by numerous management entities has been the control of external nutrient inputs in systems where excessive algal growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @han2014; @greening2014].
 There are limited examples of successful recovery of seagrass through control of nutrient inputs alone, primarily because of the difficulty in identifying and regulating both point and diffuse non-point sources.
@@
 Complementary management actions, in addition to nutrient reductions, are needed in these situations.
-Management actions with co-benefits are much more effective in estuaries that are stressed but not past critical thresholds that produce an alternative ecosystem state.
+Management actions with other ecosystem co-benefits are much more effective in estuaries that are stressed but not past critical thresholds that produce an alternative ecosystem state.
 
@@
 The effects of decreased water clarity for seagrasses in shallow habitats were exacerbated by acute warming related to climate change, demonstrating a concerning synergy of multiple stressors.
+@moreno-marin2018 produced similar results using a multifactorial experiment that considered temperature, light, and nutrient (nitrogen) availability for North Sea eelgrass (*Zostera marina*).
 Species shifts are also expected to occur as changing physical conditions decrease the competitive advantages of historically abundant species.
@@
 In Tampa Bay, Florida, seagrasses are a primary indicator of bay health that have been the focus of management efforts over the last thirty years [@sherwood2017].
-Through successful reduction of nitrogen loads, seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres [16,857 ha, @greening2006; @greening2014].
+Through successful reduction of external nitrogen loads, seagrasses have recovered from a low, system-wide coverage in the 1980s to an all-time high in 2016 of 41,655 acres [16,857 ha, @greening2006; @greening2014].
 Seagrass distribution and abundance has been dynamic throughout this period, overall responding positively to increases in water clarity with nutrient load reductions, while also demonstrating more short-term variability in response to regional climate events [@greening2006].
 Since 2016, seagrass areal cover has decreased by nearly 1/3 despite relatively stable water quality with adequate light penetration that has been historically shown to support growth [@tbep0696; @Beck20a].
-Factors that have influenced this recent decline are unknown and the effects of climate change on physical conditions have been implicated as potential drivers.
+Factors that have influenced this recent decline are unknown and the effects of climate change drivers on physical water quality conditions have been implicated as potential stressors.
 Following global trends, recent work has demonstrated a broad long-term trend of increasing water temperature in Tampa Bay, although at a relatively coarse scale [@nickerson2023].
@@
 This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay estuary over the last fifty years.
-Three datasets of varying sampling designs and temporal coverage were used to assess the primary hypothesis that Tampa Bay is trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges.
+Three datasets of varying sampling designs and temporal coverage were used to assess the primary hypothesis that Tampa Bay is trending towards hotter and fresher conditions that are likely stressing seagrasses beyond their optimal tolerance ranges, particularly in upper bay segments where marginal light environments exist.
 This hypothesis was generated from preliminary assessments of datasets used herein and discussions with the regional scientific and management community, despite information in the primary literature suggesting the contrary for salinity [e.g., @costa2023].
-This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring describing cover of individual species, and synoptic seagrass data collected with the water quality surveys.
+This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring describing cover of individual species, and synoptic seagrass data collected with routine biotic and water quality surveys.
 Specific focus was on the periods prior to and after 2016, when seagrasses recovered baywide and then declined despite water quality conditions remaining relatively stable.
 Our primary goal was to relate recent seagrass changes to climate stressors to provide a motivating example for the management community in the Tampa Bay region that a shifting ecological baseline presents new challenges beyond the known success story of long-term bay recovery.
-The results are also provided as a cautionary example of the subtle but increasing role of climate-related stressors on the resiliency of estuarine systems at the tropical-temperate boundary, prompting a reassessment of existing management paradigms to accommodate current and anticipated future conditions.
+The results are also provided as a cautionary example of the subtle but increasing role of climate-related stressors on the resiliency of estuarine systems at the tropical-temperate boundary, necessitating the need to resassess existing management paradigms and adaptively manage to accommodate current and anticipated future conditions.
 
@@
 Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer [@morrison2006].
-El Niño weather events have also been associated with prolonged periods of heavy rainfall contributing to seagrass reductions through increased stormwater nutrient loads [@greening2006; @morrison2006].
+El Niño weather events have also been associated with prolonged periods of heavy rainfall contributing to seagrass reductions through increased stormwater nutrient loads [@schmidt2002; @greening2006; @morrison2006].
 The watershed is heavily developed and includes over 3 million people [@tbep0923] with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023].
@@
 Tampa Bay is divided into distinct sub-segments defined by physical and natural boundaries to assist with water quality management activities [@lewis1985]: Old Tampa Bay (OTB) in the northwest; Hillsborough Bay (HB) in the northeast; Middle Tampa Bay (MTB); and Lower Tampa Bay (LTB) that connects to the Gulf of Mexico (Figure [-@fig-map]a).
-Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality primarily from nutrient inputs from wastewater and stormwater [@greening2014].
+Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality primarily from direct external nutrient inputs from wastewater and stormwater [@greening2014].
 Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small tributaries and manmade flood control conveyances [@tbep0623].
@@
 All bay segments are relatively shallow, with a baywide mean depth of approximately 3 m.
-Light penetration typically reaches bottom habitats under current conditions, although seagrasses were historically limited by high phytoplankton production that affected light environments prior to wastewater regulation, particularly in Old Tampa Bay and Hillsborough Bay [@greening2014; @johansson15].
+Light penetration typically reaches bottom habitats under current conditions, although seagrasses were historically limited by high phytoplankton production that affected light environments prior to wastewater and stormwater regulation, particularly in Old Tampa Bay and Hillsborough Bay [@greening2014; @johansson15].
 
@@
 
-The long-term recovery of seagrass habitats in Tampa Bay is a nationally-recognized success story that demonstrates application of a successful management paradigm through the National Estuary Program [@greening2006; @greening2014; @sherwood2017].
-From 1988 to 2016, seagrasses increased 79% to 41,655 acres (16,857 ha), surpassing the regional management goal of restoring coverage to 95% of a benchmark 1950s coverage (Figure [-@fig-seagrasschg]a).
-Throughout this same period, nitrogen load estimates decreased by about 2/3 from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration projects in the watershed [@greening2014; @beck2019].
+The long-term recovery of seagrass habitats in Tampa Bay is a nationally-recognized success story that demonstrates application of a successful management paradigm through the EPA-administered National Estuary Program [@greening2006; @greening2014; @sherwood2017].
+From 1988 to 2016, seagrasses increased 79% to 41,655 acres (16,857 ha), surpassing the regional management goal of restoring coverage to 95% of a benchmark 1950s coverage.
+The greatest areal coverage expansions were observed in the upper bay segments (OTB, HB, and MTB; Figure [-@fig-seagrasschg]a).
+Throughout this same period, external nitrogen load estimates decreased by about 2/3 from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed [@greening2014; @beck2019].
 These reductions in nutrient loadings resulted in large scale reductions in chlorophyll concentrations and light attenuation in the water column, creating favorable environments for seagrass growth.
-The most dramatic improvements in seagrass cover were observed in Old Tampa Bay where coverage increased by 122% or 4,465 acres (2,477 ha) from 1988 estimates to a total of 11,247 acres (4,511 ha).
-Similar gains were observed in Middle Tampa Bay where seagrass cover increased by 86% or 4,465 acres (1,807 ha) to a total of 9,652 acres (3,906 ha) and Hillsborough Bay where cover increased from nearly zero acres to a total of 2,007 acres (810 ha).
-Seagrasses have generally been stable over time in Lower Tampa Bay.
 
 From 2016 to present, dramatic losses of seagrasses have been observed in Tampa Bay, despite nutrient loading and light attenuation remaining relatively stable (Figure [-@fig-seagrasschg]a).
-Total cover in Tampa Bay has decreased by 28% from the 2016 peak by 11,518 acres (4,661 ha) to a total of 30,137 acres (12,196 ha).
+Total cover in Tampa Bay has decreased by 28% (11,518 acres/4,661 ha) from the 2016 peak to a total baywide coverage of 30,137 acres (12,196 ha).
 Losses have been most pronounced in Old Tampa Bay (62%; 6,963 acres/2,818 ha loss) and Hillsborough Bay (80%; 1,599 acres/647 ha loss).
@@
 Annual transect surveys have been conducted since 1998 at 62 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017].
-This dataset provides species information on abundance, cover, frequency occurrence, and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed.
+This dataset provides species information on cover-abundance, frequency occurrence, and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed.
 Although the areal maps provide the standard for assessment of restoration goals, the transect data allow inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined.
@@
 First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at 45 stations in the major bay segments since the late 1970s (Figure [-@fig-map]b).
-These data are used for regulatory assessments for the nutrient Total Maximum Daily Load in Tampa Bay.
-Water quality samples are collected at each station from surface water grabs (e.g., nutrients) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom.
+These data are used for regulatory assessments for the federally-recognized nutrient Total Maximum Daily Load and state Reasonable Assurance determination in Tampa Bay.
+Water quality samples are collected at each station from surface water grabs (e.g., nutrients, biological, and chemical constituents) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom.
+Most analyses herein used only bottom water measurements given the shallow depth and mixed water column of most of Tampa Bay [@weisberg2006].
 Time of sampling can vary, although most samples are collected from mid-morning to early afternoon.
+Compared to the additional datasets described below, the monitoring stations are generally in deeper water beyond where seagrasses occur along the shallow margins of the bay.
 The data were obtained using the tbeptools R package that imports the data directly from a stable web address provided by the EPC [@beck2021].
@@
 A stratified sampling design is used to select sites for 21.3 meter center-bag seines that target shallow habitats (\<1.5 m) where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present, Figure [-@fig-map]c).
-Seines were deployed (1) parallel to bay shorelines and (2) offshore into the prevailing current at least 5 m from the shoreline to sample approximately 140 m$^2$ of bay bottom [@schrandt2021].
+Seines were deployed (1) parallel to bay shorelines in shallow waters (\<1.5 m) and (2) offshore into the prevailing current at least 5 m from the shoreline to sample approximately 140 m$^2$ of bay bottom [@schrandt2021].
 In addition to collecting fish and selected invertebrates, *in situ* physical measurements for water temperature and salinity are collected at the bag, and at the surface and at 1 m intervals to the bottom.
@@
 Sites exclusively with macroalgae were not included in the analysis.
-All FIM data were provided from FWC staff by request.
+All FIM data were provided from FWC staff upon request.
 
 The third and final dataset evaluated was from the Pinellas County Department of Environmental Management (PDEM).
-Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has only been used in Tampa Bay since 2003 primarily to support robust statistical assessments for National Pollutant Discharge Elimination System reporting.
-Data were obtained by request to PDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (Figure [-@fig-map]d, also available at <https://wateratlas.usf.edu/>).
+Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has only been used in Tampa Bay since 2003 primarily to support robust statistical assessments for regulatory stormwater permit reporting.
+Data were obtained by request to PDEM staff for the western portion of Old Tampa Bay where sampling occurred from 2003 to present (Figure [-@fig-map]d, also available at <https://wateratlas.usf.edu/>).
 We focused primarily on OTB for the analysis of the PDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments.
 Four distinct spatial zones in OTB are used to stratify the random selection of sample points that typically include 4 sample points per month in each zone.
-Water quality samples at each site are similar to those collected by EPC, where only bottom measurements for salinity and temperature were retained for analysis.
+Water quality samples at each site are similar to those collected by EPC but can occur in shallower locations.
 Seagrass presence/absence is also recorded at each site and all sites were defined as "seagrass" if only seagrass species were identified (any with macroalgae were excluded) and "no seagrass" if bare sediment was observed.
 
-All of the organizations that provided water quality datasets participate in the Southwest Florida Regional Ambient Monitoring Program.
-This *ad hoc* group meets quarterly to ensure similar standards and methods are used for the collection and processing of surface water quality monitoring data.
+All of the organizations that provided water quality datasets participate in the Southwest Florida Regional Ambient Monitoring Program [@sherwood2016].
+This *ad hoc* group meets quarterly to ensure similar standards and methods are used for the collection and processing of surface water quality monitoring data in the region.
 Split-samples evaluated by each organization are also compared to assess precision between different laboratories.
@@
 This assessment provided an indication of the extent of change in Tampa Bay as context for understanding potential relationships with seagrass change.
-An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions related to long-term (multi-decadal) climate change.
-For comparison to water quality conditions, daily air temperature data were obtained from Tampa International Airport [using the *rnoaa* R package, @chamberlain23] and monthly rainfall data for the Tampa Bay watershed were obtained from the SWFWMD (<https://www.swfwmd.state.fl.us/resources/data-maps/rainfall-summary-data-region>).
-Only rainfall data for the rainy season (June to September) were evaluated.
-Additionally, the Standardized Precipitation Index [SPI, @begueria2013] was estimated from the rainfall data to identify periods of time when rainfall significantly deviated from the long-term average [using the *spei* R package, @begueria23].
+An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions related to long-term (multi-decadal) climate change drivers.
+For comparison to water quality conditions, daily air temperature (Tampa International Airport [TIA] National Weather Service site) and precipitation (SWFWMD area-weighted watershed summaries) were used to characterize regional conditions for the period of record covered by the water quality samples (i.e., 1975 to present for the EPC data).
+The rnoaa R package [@chamberlain23] was used to obtain the TIA temperature data.
+Regional precipitation summaries were obtained directly from the SWFWMD (https://www.swfwmd.state.fl.us/resources/data-maps/rainfall-summary-data-region).
+Only rainfall data for the wet season (June to September) were evaluated for trends, whereas the complete record was used to calculate the Standardized Precipitation Index [SPI, @begueria2013] to identify periods of time when rainfall significantly deviated from the long-term average [using the *spei* R package, @begueria23].
 Annual hydrologic loading data to Tampa Bay beginning in 1985 were also obtained for comparison to annual precipitation [@tbep0623].
@@
 The change per year was also evaluated for each parameter based on the slope estimates returned by each test.
-Kendall tests were also used to evaluate changes over time for each month across years to determine when the trends were most pronounced seasonally, e.g., all January estimates across years, all February estimates, etc.
+Kendall tests were also used to evaluate changes over time for each month across years to determine when the trends were most pronounced seasonally (e.g., all January estimates across years, all February estimates, etc.).
 The percentage of stations in each bay segment with significantly increasing temperature or decreasing salinity trends were evaluated for each month.
@@
 A fundamental challenge describing niche space is identifying the boundaries for optimal conditions.
-In Tampa Bay, the three dominant seagrass species are *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinum* (turtle grass) [@lewis1985].
+In Tampa Bay, three dominant seagrass species occur: *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinum* (turtle grass) [@lewis1985].
 These species co-occur often in mixed beds throughout the bay, although some differences in abundance are observed across salinity ranges.
@@
 Reported salinity ranges for each of these species varies depending on location, season, and other co-occurring factors like temperature [@phillips1960; @mcmillan1967; @zieman1975; @lewis1985], although most studies place lower limits of salinity in the range of 15-25 ppt.
-Optimal temperature ranges are similar between species, with reduced growth observed at temperatures above 30$^\circ$ C [@zieman1975; @lewis1985].
+Optimal temperature ranges are similar between these temperate-tropical species, with reduced growth observed at temperatures above 30$^\circ$ C [@zieman1975; @lewis1985].
 
-Because of the uncertainty in defining thresholds for optimal temperature and salinity ranges, multiple thresholds were evaluated to describe the potential for stress and how it may be related to changes in seagrass.
+Because of the uncertainty in defining *in situ* thresholds for optimal temperature and salinity ranges, multiple thresholds were evaluated to describe the potential for stress and how it may be related to changes in seagrass.
 Distinctions were not made between species, primarily due to lack of consensus between studies and likely site-specific ranges that affect seagrass growth in Tampa Bay.
@@
 The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends.
-Model fit for each station was sufficient to calculate daily predictions to assess potential stressor metrics (Figure S5, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models, Tables S1, S2).
+Model fit for each station was considered sufficient to calculate daily predictions to assess potential stressor metrics (Figure S5, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models, Tables S1, S2).
 
@@
 Generalized Linear regression Models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre-2016, decline post-2016).
-Two models were evaluated, one with the temperature and salinity metrics together and another with the both metric given that the latter was a combined metric of the former.
+Two models were evaluated, one with the temperature and salinity metrics together and another with the both metric.
 Interactions were included between temperature, salinity, and time period or the both metric and time period, depending on the model.
@@
 
-Additional models were constructed for the FIM and PDEM datasets to evaluate seagrass change relative to temperature and salinity.
+Separate models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change.
 These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs.
@@
 Using this model, total precipitation in 1975 was `r prpstr2` mm, whereas current total precipitation in 2022 was `r prpend2` mm, showing an overall increase in the period of record of `r prpdif2` mm.
-Trends in total annual hydrologic load from 1985 to present followed total annual precipitation, although the linear model was not significant likely due to a shorter period of record.
+Increasing trends in total annual hydrologic load from 1985 to present mirrored total annual precipitation, although the linear model was not significant.
 The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and 2015-2020.
@@
 The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths and minimal stratification in Tampa Bay (@fig-meteowqraw).
-Additional analyses only considered bottom water trends given comparable results within the water column.
 Trends in water temperature were similar across bay segments with significant increases varying from 0.03 to 0.04 $^\circ$C per year, with a total increase of 1.4 (OTB) to 2.0 (HB) $^\circ$C from 1975 to present (@tbl-temptrndtab).
-Salinity trends were similar between bay segments, although overall salinity was of course higher for bay segments closer to the Gulf of Mexico.
-Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, with a total decrease of 2.7 (OTB) and 2.0 (LTB) ppt from 1975 to present.
-As expected, intra-annual variability of water temperature was much higher than salinity following the larger seasonal variation in temperature (@fig-meteowqraw).
+Salinity trends were similar between bay segments, although overall salinity was predictably higher for bay segments closer to the Gulf of Mexico.
+Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, with a total decrease of 2.7 (OTB) and 2.1 (LTB) ppt from 1975 to present.
+As expected, intra-annual variability of water temperature was much higher than salinity consistent with larger seasonal variation in temperature (@fig-meteowqraw).
 This variation may also be related to differences among the monitoring stations in each segment because the averages combine monthly and spatial variation.
@@
 
-The salinity models were less similar between bay segments compared to the temperature models, primarily because of the natural variation in salinity along the bay's longitudinal axis (Table S4).
+The salinity models were less similar between bay segments compared to the temperature models, primarily because of the natural variation in salinity along the bay's longitudinal axis (Tables S3, S4).
 None of the bay segments had significantly increasing number of days per year when salinity was below 15 ppt.
@@
 @fig-mixeff provides visual examples of the mixed-effects models for the estimated number of days over time for each bay segment from 1975 to present when temperature was above 30$^\circ$, salinity was below 25 ppt, and when both occurred (see Figure S6 for 1998 to 2022).
-Temperature trends were similar among segments, whereas the number of days when salinity was below the threshold varied by proximity to the Gulf of Mexico (@tbl-mixdaytab).
+Temperature trends were similar among segments, whereas the number of days when salinity was below the threshold decreased with proximity to the Gulf of Mexico (@tbl-mixdaytab).
 The number of days when both temperature was above and salinity was below the threshold generally followed the trends for the number of days when salinity was below the threshold.
 These thresholds were used for comparison to seagrass changes described below, based primarily on the statistical strength of the trends and the variance of counts across stations within each bay segment (points in @fig-mixeff).
-That is, more restrictive thresholds did not provide sufficient counts of days per year to develop models and the chosen thresholds were based primarily on statistical considerations.
+That is, more restrictive thresholds did not provide sufficient counts of days per year to more rigorously develop seagrass response models and the chosen thresholds were based primarily on statistical considerations.
 
@@
 
-Linear models to assess the potential effects of temperature and salinity on seagrass change provided modest evidence that seagrass loss after 2016 was driven by climate stressors.
+Linear models to assess the potential effects of temperature and salinity on seagrass change provided modest evidence that seagrass loss after 2016 was influenced by climate stressors.
 The most parsimonious model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence (`r epcmodsum1`).
@@
 The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure [-@fig-sgmod]a, @tbl-modtab).
-The interaction between the temperature and salinity metric was included in the most parsimonious model (marginally significant, p \< 0.1), although the shape of the relationship of temperature with time period or salinity with time period did not noticeably change using different conditional values for salinity or temperature, respectively.
+The interaction between the temperature and salinity metric was included in the most parsimonious model (marginally significant, p \< 0.1), although the direction of the relationship of temperature with time period or salinity with time period did not noticeably change using different conditional values for salinity or temperature, respectively.
 The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (`r epcmodsum2`), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab).
@@
 Increases in temperature and decreases in salinity were both associated with reductions in percent cover after 2016, whereas no relationships were observed prior to 2016 (p \< 0.05 and p \< 0.1 for the interactions between temperature or salinity with time period, respectively, Figure [-@fig-sgmod]b, @tbl-modtab).
-The interaction between salinity and temperature was not included as a significant term in the final model.
+The interaction between salinity and temperature was not included as a significant term in the final FIM model.
 Lastly, the most parsimonious model for the PDEM dataset only included time period as a marginally significant variable and the overall model fit was not significant (@tbl-modtab).
@@
 Understandably, the trends were most consistently observed in the dataset with the longest period of record (EPC) covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season at locations farther from the mouth of the Bay.
-These long-term changes manifested into consistent trends in the stressor metrics; the continuous number of days each year when temperature was above and/or salinity was below defined thresholds consistently increased across the period of record.
+These long-term changes manifested into consistent trends in select stressor metrics; the continuous number of days each year when temperature was above and/or salinity was below defined thresholds consistently increased across the period of record.
 Notably, bottom and surface trends for water temperature and salinity in the EPC dataset were similar, suggesting that any potential stress on biotic resources related to these factors are likely occurring throughout the water column.
@@
 @nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, which somewhat conflicts with our results showing distinct increases in the summer, particularly for the upper bay segments (OTB and HB).
-However, our assessment evaluated trends at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to the larger spatial area.
+However, our assessment evaluated trends at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to their larger spatial area and model domain.
 These differences are understandable given the analysis goals and methods for data synthesis.
@@
 
-Our relatively simple modeling approach provided some evidence that climate-related stressors can explain the recent loss in seagrass in Tampa Bay.
-The models did not provide a consistent, nor statistically powerful, explanation that increasing temperature and decreasing salinity were key drivers.
-However, evaluating all models together does suggest a pattern that demonstrates the value of considering multiple datasets and models to explain noisy ecological patterns.
+Our relatively simple modeling approach provided some evidence that climate-related stressors impart some effect on recent seagrass losses in Tampa Bay.
+The models did not provide a consistent, nor statistically powerful, explanation that increasing temperature and decreasing salinity were key (or the sole) drivers.
+However, evaluating all models together does suggest a pattern that demonstrates the value of considering multiple datasets and models to explain noisy patterns and compounding ecological processes.
 Model results for the EPC and FIM datasets both suggested that increasing temperature and decreasing salinity were associated with potential seagrass loss post-2016, described primarily using separate interaction terms of temperature or salinity with time period.
@@
 This does not suggest that these factors are unimportant, both currently and in the future.
-Our models suggested that temperature and salinity are potentially associated with seagrass loss and the consistent long-term trend in temperature and salinity suggest that conditions are very likely to become more stressful in the future.
-Moreover, the GAMs used to model continuous temperature and salinity at each EPC monitoring station likely did not fully describe the current extent of stress given that extreme low or high values were often not captured by the daily predictions (Figure S5).
-Regardless of the above limitations, the role of climate stressors on seagrass change is likely to continue in the future as suggested by our results.
+Our models suggested that temperature and salinity are potentially associated with seagrass loss and the consistent long-term trend in temperature and salinity imply that these conditions are likely to become more stressful in the future, further confounding seagrass recovery efforts in the region.
+Moreover, the GAMs used to model continuous temperature and salinity at each EPC monitoring station likely did not fully describe the current extent of stress given that extreme low or high values were not often captured by the daily modeled predictions (Figure S5).
+Regardless of the above limitations, climate stressors are likely to affect Tampa Bay's seagrass ecology in the future, as suggested by our results.
 
@@
 This relates to the inclusion of additional drivers as well as issues related to correlation/causation.
-For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, nitrogen in particular, in affecting light environments for seagrass growth [@greening2014; @sherwood2017].
+For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, external nitrogen inputs in particular, in affecting light environments for seagrass growth [@greening2014; @sherwood2017].
 As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences after accounting for light attenuation could additionally be explained by temperature or salinity.
@@
 Thus, time period was necessary to control for these confounding relationships.
-Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay [OTB in particular where the most losses have been seen, @tbep0696; @Beck20a].
+Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay, particularly in OTB where the most loss occurred [@tbep0696; @Beck20a].
 This suggests that other factors are influencing seagrass growth and the inclusion of light attenuation in the models was likely not critical for describing the roles of temperature and salinity.
 A final consideration for our models relates to how seagrasses may influence their environment, particularly for the PDEM and FIM datasets where temperature and salinity were measured at the same locations as seagrass.
-For example, temperature may simply be higher in locations where seagrasses are present, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses.
+For example, temperature may simply be higher in locations where seagrasses are present and can absorb solar radiation, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses.
 This explanation cannot be ruled out with the existing datasets, although the trend analyses and models suggest that climate-related stressors driving seagrass loss is a more likely scenario.
+This is especially true for water temperature trends captured by the EPC dataset which includes deeper, fixed sites adjacent to shallow seagrass flats.
 
@@
 Losses have been observed throughout southwest Florida during this time period, including Sarasota Bay directly south of Tampa Bay and Charlotte Harbor further south.
-These regional losses suggest that large-scale stressors are driving these changes, leading us to our initial hypothesis that climate-related stressors could explain the change in Tampa Bay.
+These regional losses suggest that large-scale stressors are driving these changes, leading us to our initial hypothesis that climate-related stressors could partially explain the change in Tampa Bay.
 Based on our results, the losses elsewhere may potentially be explained by temperature and salinity and are worth exploring in other southwest Florida coastal regions and estuaries where long-term datasets exist.
-Additional factors that could explain these changes are also likely co-occurring with climate stress, some of which are unique to Tampa Bay and others that are likely more pervasive.
+Additional factors that could explain these changes are also likely co-occurring with climate stress, some of which are unique to Tampa Bay and others that are more likely pervasive.
 For Tampa Bay, annual summer/fall blooms of the toxic dinoflagellate *Pyrodinium bahamense* have occurred in OTB since 2008 [@usup1994; @lopez2023] and the specific relationships of these blooms with seagrass change is unclear, although the expectation is that seagrass growth may be limited by the degradation of the light environment with algal growth.
-These blooms are exacerbated by the hydrologic conditions in OTB that contribute to relatively longer residence times [@phlips2006; @lopez2021].
-The effect of warming temperature and decreasing salinity in OTB will further complicate the understanding of how these blooms manifest and persist each year [e.g., @stelling2023].
+These blooms are exacerbated by the hydrologic conditions in OTB that contribute to relatively longer water residence times [@phlips2006; @lopez2021].
+The effect of warming temperature and decreasing salinity in OTB will further complicate the understanding of how these blooms manifest and persist each year [e.g., @stelling2023], and ultimately contribute to changes in the light environment affecting seagrass resources in this bay segment.
 
 Additional biotic factors could be influencing regional patterns in seagrass growth.
-In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern [@hall2022; @jei2022; @brewton2023; @scolaro2023].
-Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present in Tampa Bay [@Beck20b].
-Competitive differences between seagrasses and macroalgae are poorly understood in these systems [but see @bell1997; @brewton2023], in addition to insufficient data in Tampa Bay to clearly describe growth patterns and nutrient cycling related to macroalgae.
-Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth [@beck2022a; @scolaro2023; @tomasko2023], whereas the role of chronic nutrient loading and changing climatic conditions in promoting growth and interactions with seagrasses is less understood, particularly in recent years.
+In Tampa Bay and elsewhere, enhanced macroalgal production may also influence seagrass growth and recovery in Florida estuaries [@hall2022; @jei2022; @brewton2023; @scolaro2023].
+Attached macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present in Tampa Bay [@Beck20b].
+Competitive differences between seagrasses and macroalgae are poorly understood in these systems [but see @bell1997; @brewton2023], in addition to insufficient macroalgae data in Tampa Bay that cannot clearly describe seasonal growth and distribution patterns and nutrient cycling.
+Discrete pollutant loading events in Tampa Bay have been documented to promote both phytoplankton and macroalgae growth [@beck2022a; @scolaro2023; @tomasko2023].
+The role that evolving nutrient loading and changing climatic conditions may have on Tampa Bay's primary producers – particularly algal and seagrass growth and interactions in recent years – is less understood.
 Finally, additional research has focused on how diseases and pathogens can influence seagrass growth patterns in Florida [@robblee1991; @duffin2021; @vanbogaert2018].
@@
 Existing research has primarily focused on describing spatial patterns, past die-off events, or immunology of these pathogens [@robblee1991; @duffin2021].
+More research should be directed towards the influence of climate stressors on seagrass pathogen vulnerability.
 
@@
 An evaluation of each dataset showed a clear pattern of increasing temperature and decreasing salinity mirrored by long-term changes in air temperature and precipitation, suggesting that Tampa Bay has become hotter and fresher with the trends likely continuing in the future.
-Simple regression models provided weak, but partially-supporting evidence that these changes can be linked to recent seagrass losses since 2016.
-Our models suggested that rising temperatures and decreasing salinities are additively affecting seagrasses, rather than multiplicatively as evidenced by lack of significant interactions of these stressors.
+Simple regression models provided weak, but partially-supporting evidence, that these changes can be linked to recent seagrass losses since 2016.
+Our models suggested that rising temperatures and decreasing salinities are additively affecting seagrasses, rather than multiplicatively, as evidenced by lack of significant interactions of these stressors.
 Future analyses may show more significant associations between physical habitat conditions and seagrass change as the trends are very likely to continue to push seagrasses further outside of their tolerance ranges.
-These analyses should be supported by additional data, particularly high-resolution continuous monitoring data that provide a more precise assessment of stress at multiple time-scales.
+These analyses should be supported by additional data collection efforts, particularly high-resolution continuous monitoring data that provide a more precise assessment of diurnal stress across multiple time-scales.
 
 Natural resource managers should consider how these climate-related stressors may alter the effectiveness of intervention activities aimed at protecting ecological resources in Tampa Bay.
-Management actions that have historically been effective may have inadequate capacity to maintain ecosystem status due to changing resilience of ecological resources with climate change.
-For example, nitrogen load reductions have been effective at restoring Tampa Bay, primarily due to wastewater load allocations considered to be within the nutrient assimilative capacity of Tampa Bay [@greening2006; @greening2014].
-As Tampa Bay becomes hotter and fresher, similar load allocations may no longer be effectively assimilated.
-Strategies to mimic the pre-development flow of water or that further reduce allowable load inputs from regulated entities (e.g., additional stormwater controls, hydrological modifications) may be pursued to confer additional resilience and adaptive capacity.
+Management actions that have historically been effective may have inadequate capacity to maintain ecosystem status due to changing resilience of ecological resources affected by climate change.
+For example, nitrogen load reductions have been effective at restoring seagrass in Tampa Bay, primarily due to wastewater load allocations being constrained to the historical nutrient assimilative capacity of Tampa Bay [@greening2006; @greening2014].
+As Tampa Bay becomes hotter and fresher, similar load allocations may no longer be effectively assimilated and algal and seagrass ecology dynamics may shift.
+Strategies that mimic or restore pre-development hydrology or that further reduce allowable load inputs from regulated entities (e.g., additional stormwater controls, hydrological modifications) may be needed to confer additional resilience and adaptive capacity for seagrass in Tampa Bay.
 These considerations are especially critical for upper parts of Tampa Bay where a majority of seagrass loss has occurred and where temperature and salinity trends have been most pronounced.
-Reversal of recent trends may be more likely to occur if aggressive actions and controls are pursued sooner rather than later, given the challenges of restoring degraded systems due to hysteresis.
+Reversal of recent trends may be more likely to occur if aggressive actions and controls are pursued sooner rather than later, given the challenges of restoring degraded estuarine ecosystems due to hysteresis.
 
@@
 
-This work benefited from constructive discussions with the Technical Advisory Committee of the Tampa Bay Estuary Program (TBEP), Mike Wessel, Ray Pribble, Brad Furman, and Dave Tomasko.
+This work benefited from constructive discussions with the Technical Advisory Committee of the Tampa Bay Estuary Program (TBEP), in particular Mike Wessel, Ray Pribble, and Dave Tomasko.
 TBEP funding for this work stems from EPA Section 320 Grant Funds, and the TBEP's local government partners (Hillsborough, Manatee, Pasco, and Pinellas Counties; the Cities of Clearwater, St.
 Petersburg, and Tampa; Tampa Bay Water; and the Southwest Florida Water Management District) through contributions to the operating budget.
-We are indebted to the hundreds of field and technical staff for collecting and maintaining the valuable long-term monitoring data used in this study.
+We are indebted to the community of field and technical staff that have routinely collected and maintained the valuable long-term monitoring data used in this study.
 
@@
 The models included station as a random effect for each bay segment, with grey lines indicating individual station trends, grey points as actual number of days, and thicker lines indicating the overall model fit.
-Slopes for significant models are shown in the bottom right of each facet.
+Slopes for significant models are shown in the bottom left of each facet.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/mixeff.png){#fig-mixeff}
@@
 ![Predicted results from generalized linear models evaluating seagrass changes versus temperature or salinity predictors for the (a) EPC and (b) FIM datasets.
+Results are shown only for OTB, although the shape of the relationships do not change between bay segments.
 The predictors for (a) were the number of days when temperature was above the threshold and the number of days when salinity was below the threshold.

### 2024-01-26 supplement.qmd
added ES edits

--- 2024-01-25 supplement.qmd
@@
 The models included station as a random effect for each bay segment, with grey lines indicating individual station trends, grey points as actual number of days, and thicker lines indicating the overall model fit.
-Slopes for significant models are shown in the bottom right of each facet.
+Slopes for significant models are shown in the bottom left of each facet.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppmixeff.png){#fig-suppmixeff}

### 2024-01-26 R/figs.R

@@ -1187,9 +1187,13 @@ p <- ggplot(toplo1, aes(x = yr, y = cnt)) +
   geom_point(size = 0.7, alpha = 0.6, color = 'darkgrey') + 
   geom_line(aes(y = prd, group = station), color = 'darkgrey', linewidth = 0.5) + 
   geom_line(data = toplo2, aes(y = prd, col = thrtyp), show.legend = F, linewidth = 1.5) + 
-  geom_label(data = toplo3, aes(x = 2022, y = 0, label = slo, col = thrtyp), show.legend = F, 
-             vjust = 0, hjust = 1, fontface = 'italic', size = 3, label.r = unit(0, "lines"),
-             label.padding = unit(0.1, "lines")) +
+  geom_text(data = subset(toplo3, thrtyp == 'Temperature > 30'), aes(x = 1975, y = -10, label = slo, col = thrtyp), show.legend = F, 
+             vjust = 0, hjust = 0, fontface = 'italic', size = 3) +
+  geom_text(data = subset(toplo3, thrtyp == 'Salinity < 25'), aes(x = 1975, y = -30, label = slo, col = thrtyp), show.legend = F, 
+            vjust = 0, hjust = 0, fontface = 'italic', size = 3) +
+  geom_text(data = subset(toplo3, thrtyp == 'Both'), aes(x = 1975, y = -12, label = slo, col = thrtyp), show.legend = F, 
+            vjust = 0, hjust = 0, fontface = 'italic', size = 3) +
+  
   scale_color_manual(values = c('red2', 'dodgerblue2', 'black')) +
   # coord_cartesian(ylim = c(-10, NA)) +
   facet_grid(thrtyp ~ bay_segment, scales = 'free_y') + 
@@ -1237,9 +1241,12 @@ p <- ggplot(toplo1, aes(x = yr, y = cnt)) +
   geom_point(size = 0.7, alpha = 0.6, color = 'darkgrey') + 
   geom_line(aes(y = prd, group = station), color = 'darkgrey', linewidth = 0.5) + 
   geom_line(data = toplo2, aes(y = prd, col = thrtyp), show.legend = F, linewidth = 1.5) + 
-  geom_label(data = toplo3, aes(x = 2022, y = 0, label = slo, col = thrtyp), show.legend = F, 
-             vjust = 0, hjust = 1, fontface = 'italic', size = 3, label.r = unit(0, "lines"),
-             label.padding = unit(0.1, "lines")) +
[…]

### 2024-01-28 manu-draft.qmd
shorten text, remove hydro load from figs

@@
 These results should compel resource managers to consider the likelihood that reduced resilience of estuarine resources due to shifting ecological baselines driven by additional climate change drivers will complicate long-standing management paradigms.
-While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be confounded by other cliamte change driveres and warrant additional, complementary interventions to maintain ecosystem health into the future.
+While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be confounded by other cliamte change drivers and warrant additional, complementary interventions to maintain ecosystem health into the future.
 
@@
 Complementary management actions, in addition to nutrient reductions, are needed in these situations.
-Management actions with other ecosystem co-benefits are much more effective in estuaries that are stressed but not past critical thresholds that produce an alternative ecosystem state.
 
@@
 Species shifts are also expected to occur as changing physical conditions decrease the competitive advantages of historically abundant species.
-Again in Chesapeake Bay, the abundance of widgeongrass (*Ruppia maritima*) has been mediated by nutrient reductions, while also outcompeting and replacing formerly dominant eelgrass (*Zostera marina*) given higher tolerance to increased temperatures [@hensel2023].
+Again in Chesapeake Bay, the abundance of widgeongrass (*Ruppia maritima*) has been mediated by nutrient reductions, while also outcompeting and replacing formerly dominant *Z.
+marina* given higher tolerance to increased temperatures [@hensel2023].
 Changing frequency and severity of precipitation patterns may further alter the distribution and abundance of seagrasses by pushing species outside of optimal salinity tolerance ranges [@rasheed2011; @webster2021].
@@
 Seagrass distribution and abundance has been dynamic throughout this period, overall responding positively to increases in water clarity with nutrient load reductions, while also demonstrating more short-term variability in response to regional climate events [@greening2006].
-Since 2016, seagrass areal cover has decreased by nearly 1/3 despite relatively stable water quality with adequate light penetration that has been historically shown to support growth [@tbep0696; @Beck20a].
+Since 2016, seagrass areal cover has decreased by nearly 1/3 despite relatively stable water quality [@tbep0696; @Beck20a].
 Factors that have influenced this recent decline are unknown and the effects of climate change drivers on physical water quality conditions have been implicated as potential stressors.
@@
 Changing salinity has not been well-described, nor have potential links of changing salinity and temperature with recent seagrass change.
-Tampa Bay is rich with historical data that can be used to evaluate long-term trends and to develop models that can explain relationships of seagrass change with water quality conditions.
+Tampa Bay is rich with historical data that can be used to evaluate long-term trends.
 This information can fill a critical knowledge gap that can inform regional management activities, while also demonstrating the confounding effects of climate change with ecosystem response to conventional stressors.
@@
 This hypothesis was generated from preliminary assessments of datasets used herein and discussions with the regional scientific and management community, despite information in the primary literature suggesting the contrary for salinity [e.g., @costa2023].
-This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring describing cover of individual species, and synoptic seagrass data collected with routine biotic and water quality surveys.
+The analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring describing cover of individual species, and synoptic seagrass data collected with routine biotic and water quality surveys.
 Specific focus was on the periods prior to and after 2016, when seagrasses recovered baywide and then declined despite water quality conditions remaining relatively stable.
-Our primary goal was to relate recent seagrass changes to climate stressors to provide a motivating example for the management community in the Tampa Bay region that a shifting ecological baseline presents new challenges beyond the known success story of long-term bay recovery.
-The results are also provided as a cautionary example of the subtle but increasing role of climate-related stressors on the resiliency of estuarine systems at the tropical-temperate boundary, necessitating the need to resassess existing management paradigms and adaptively manage to accommodate current and anticipated future conditions.
 
@@
 The geology of the watershed is rich in phosphates and mining activities have greatly altered the landscape, with notable spills and releases of wastewater that have affected water quality and biological resources [@garrett2011; @beck2022a].
-Important subtidal habitats include seagrasses, hard bottom, tidal flats, and oyster reefs, where a majority of management efforts have historically focused on restoring and maintaining seagrass cover [@sherwood2017].
-Additional native habitats include intertidal wetlands (mangroves, salt marshes, salt barrens) and pine forests, oak hammocks, and freshwater wetlands present in upland habitats [@Robison20].
-Losses of native uplands and potentially restorable habitats to development in the watershed from 1990 to 2020 have been estimated at 188,429 acres [76,254 ha, @beck2023].
 
@@
 All bay segments are relatively shallow, with a baywide mean depth of approximately 3 m.
-Light penetration typically reaches bottom habitats under current conditions, although seagrasses were historically limited by high phytoplankton production that affected light environments prior to wastewater and stormwater regulation, particularly in Old Tampa Bay and Hillsborough Bay [@greening2014; @johansson15].
+Light penetration typically reaches bottom habitats under current conditions, although seagrasses were historically limited by high phytoplankton production that affected light environments[@greening2014; @johansson15].
 
@@
 Throughout this same period, external nitrogen load estimates decreased by about 2/3 from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed [@greening2014; @beck2019].
-These reductions in nutrient loadings resulted in large scale reductions in chlorophyll concentrations and light attenuation in the water column, creating favorable environments for seagrass growth.
 
@@
 First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at 45 stations in the major bay segments since the late 1970s (Figure [-@fig-map]b).
-These data are used for regulatory assessments for the federally-recognized nutrient Total Maximum Daily Load and state Reasonable Assurance determination in Tampa Bay.
 Water quality samples are collected at each station from surface water grabs (e.g., nutrients, biological, and chemical constituents) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom.
@@
 The second dataset used to evaluate water quality trends was available from the Florida Fish and Wildlife Conservation Commission (FWC).
-The Fisheries Independent Monitoring (FIM) program administered by FWC provides monthly surveys of the entire nekton community in Tampa Bay, including species richness and abundance, using multiple gear types that target different habitats.
+The Fisheries Independent Monitoring (FIM) program administered by FWC provides monthly surveys of the entire nekton community in Tampa Bay, including species richness and abundance, using multiple gear types that target different habitats [@schrandt2021].
 A stratified sampling design is used to select sites for 21.3 meter center-bag seines that target shallow habitats (\<1.5 m) where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present, Figure [-@fig-map]c).
-Seines were deployed (1) parallel to bay shorelines in shallow waters (\<1.5 m) and (2) offshore into the prevailing current at least 5 m from the shoreline to sample approximately 140 m$^2$ of bay bottom [@schrandt2021].
 In addition to collecting fish and selected invertebrates, *in situ* physical measurements for water temperature and salinity are collected at the bag, and at the surface and at 1 m intervals to the bottom.
@@
 The third and final dataset evaluated was from the Pinellas County Department of Environmental Management (PDEM).
-Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has only been used in Tampa Bay since 2003 primarily to support robust statistical assessments for regulatory stormwater permit reporting.
 Data were obtained by request to PDEM staff for the western portion of Old Tampa Bay where sampling occurred from 2003 to present (Figure [-@fig-map]d, also available at <https://wateratlas.usf.edu/>).
 We focused primarily on OTB for the analysis of the PDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments.
-Four distinct spatial zones in OTB are used to stratify the random selection of sample points that typically include 4 sample points per month in each zone.
 Water quality samples at each site are similar to those collected by EPC but can occur in shallower locations.
@@
 An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions related to long-term (multi-decadal) climate change drivers.
-For comparison to water quality conditions, daily air temperature (Tampa International Airport [TIA] National Weather Service site) and precipitation (SWFWMD area-weighted watershed summaries) were used to characterize regional conditions for the period of record covered by the water quality samples (i.e., 1975 to present for the EPC data).
+For comparison to water quality conditions, daily air temperature (Tampa International Airport \[TIA\] National Weather Service site) and precipitation (SWFWMD area-weighted watershed summaries) were used to characterize regional conditions for the period of record covered by the water quality samples (i.e., 1975 to present for the EPC data).
 The rnoaa R package [@chamberlain23] was used to obtain the TIA temperature data.
@@
 Only rainfall data for the wet season (June to September) were evaluated for trends, whereas the complete record was used to calculate the Standardized Precipitation Index [SPI, @begueria2013] to identify periods of time when rainfall significantly deviated from the long-term average [using the *spei* R package, @begueria23].
-Annual hydrologic loading data to Tampa Bay beginning in 1985 were also obtained for comparison to annual precipitation [@tbep0623].
-All climate and loading data were evaluated annually with simple linear regression trends to assess change over time.
+All climate data were evaluated annually with simple linear regression trends to assess change over time.
 Water temperature and salinity trends using the EPC, FIM, and PDEM data were similarly evaluated by averaging the monthly data each year for each bay segment.
-Linear trends for these data were evaluated based on averages of all stations and months within each year and bay segment, which allowed for comparable statistical power between datasets with different sampling designs.
 
@@
 Seasonal Kendall trend tests were used to evaluate the monotonic change for temperature and salinity from 1975 to present at each water quality station [@hirsch1982; @millard13].
-The change per year was also evaluated for each parameter based on the slope estimates returned by each test.
 Kendall tests were also used to evaluate changes over time for each month across years to determine when the trends were most pronounced seasonally (e.g., all January estimates across years, all February estimates, etc.).
@@
 
-The second goal of the analysis was to evaluate if seagrass changes were linked to long-term changes in water temperature and salinity, with particular attention on differences between bay segments and the time periods before and after 2016 (pre/post seagrass recovery).
+The second goal of the analysis was to evaluate if seagrass changes were linked to long-term changes in water temperature and salinity.
 The conceptual model for evaluating these changes describes the niche space where seagrass growth and reproduction is hypothesized to be greatest within optimal ranges for forcing factors that are present in the environment [@hutchinson1957; @vandermeer1972].
@@
 To quantify daily counts each year, a continuous prediction of temperature and salinity over time at each of 45 stations was estimated using Generalized Additive Models (GAMs) fit to temperature or salinity with a single predictor for decimal year [@wood17].
-The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends.
 Model fit for each station was considered sufficient to calculate daily predictions to assess potential stressor metrics (Figure S5, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models, Tables S1, S2).
@@
 Stressor metrics were further aggregated across stations in each bay segment using a mixed-effects regression model where the annual stressor counts for stations in a bay segment were fit against year (1975 to 2022) using a random intercept for station [@zuur2009a; @bates15].
-This produced an overall assessment of how the stressor metrics have changed over time by bay segment, including counts for temperature, salinity, and both relative to the thresholds.
+This produced an overall assessment of how the stressor metrics have changed over time by bay segment.
 
@@
 
-For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends.
-Transect surveys are typically done in the late summer during peak growing season.
-The average sampling date for each year across the subset of transects in each bay segment was estimated and the stressor metrics were referenced accordingly.
-For example, if the average transect date for a bay segment was September 15th in 2009 and Oct 1st in 2010, the 2010 stressor counts were based on all days between the two average dates.
-This was repeated for all years from 1998 to 2022 when transect data were available.
-Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment.
+For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends above.
+Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment from 1998 to 2022.
 The stressor metrics were compared to frequency occurrence (all species) each year by bay segment.
@@
 Interactions with time period and a main effect for bay segment were also included as above.
-GLMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated (no bay segment interaction) and the seagrass response variable was presence/absence (i.e., a binomial distribution was used) to model the probability of observing seagrass.
+GLMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data and the seagrass response variable was presence/absence (i.e., a binomial distribution was used) to model the probability of observing seagrass.
 Model input data were further subset to include only months from July to November to describe seagrasses during the growing season and to reduce potential seasonal effects.
@@
 Mean annual air temperature in 1975 was `r airtempstr` $^\circ$C, whereas current mean annual air temperature in 2022 was `r airtempend` $^\circ$C, showing an overall increase in the period of record of `r airtempdif` $^\circ$C.
-Similarly, total precipitation during the rainy season has increased by `r prpslo` mm per year, although the model was not significant (`r prppvl`, $R^2$ `r prprsq`).
+Similarly, total precipitation during the rainy season has increased by `r prpslo` mm per year, although the model was not significant (`r prppvl`, $R^2$ = `r prprsq`).
 Removing September from the rainy season showed a significantly increasing trend of `r prpslo2` mm per year (`r prppvl2`, $R^2$ = `r prprsq2`).
 Using this model, total precipitation in 1975 was `r prpstr2` mm, whereas current total precipitation in 2022 was `r prpend2` mm, showing an overall increase in the period of record of `r prpdif2` mm.
-Increasing trends in total annual hydrologic load from 1985 to present mirrored total annual precipitation, although the linear model was not significant.
 The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and 2015-2020.
@@
 Note that for tables [-@tbl-temptrndtab] and [-@tbl-saltrndtab], comparable time periods were evaluated between the datasets when possible given the different sample sizes, and therefore power, to detect trends.
-The strongest trends were observed for the EPC dataset, which had the longest record from 1975 to 2022.
-The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths and minimal stratification in Tampa Bay (@fig-meteowqraw).
+The strongest trends were observed for the EPC dataset.
+The top and bottom water temperature or salinity changes were similar across bay segments (@fig-meteowqraw).
 Trends in water temperature were similar across bay segments with significant increases varying from 0.03 to 0.04 $^\circ$C per year, with a total increase of 1.4 (OTB) to 2.0 (HB) $^\circ$C from 1975 to present (@tbl-temptrndtab).
-Salinity trends were similar between bay segments, although overall salinity was predictably higher for bay segments closer to the Gulf of Mexico.
+Salinity trends were also similar between bay segments, although overall salinity was predictably higher for bay segments closer to the Gulf of Mexico.
 Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, with a total decrease of 2.7 (OTB) and 2.1 (LTB) ppt from 1975 to present.
-As expected, intra-annual variability of water temperature was much higher than salinity consistent with larger seasonal variation in temperature (@fig-meteowqraw).
-This variation may also be related to differences among the monitoring stations in each segment because the averages combine monthly and spatial variation.
+
 Significant increasing temperature trends were also observed for the FIM dataset from 1996 to present in OTB and from 2004 to present in OTB and HB (@tbl-temptrndtab).
@@
 Seasonally, most stations had more significant trends in the summer, early fall period for both temperature and salinity (Figure [-@fig-kendall]b, c), although some variation was observed between bay segments.
-Temperature trends were more often observed in the summer, early fall for the upper bay segments (OTB, HB), whereas the lower bay segments (MTB, LTB) had more seasonally homogeneous trends.
+Temperature trends were more often observed in the summer or early fall for the upper bay segments (OTB, HB), whereas the lower bay segments (MTB, LTB) had more seasonally homogeneous trends.
 Seasonal trends in salinity did not vary as much between segments, although OTB and LTB had more stations with significantly decreasing trends than the other bay segments.
@@
 
-The number of days when both temperature was above and salinity was below a threshold also varied by bay segment.
-Four models were significant for OTB, seven for HB, four for MTB, and none for LTB (Table S3).
+The number of days when both temperature was above and salinity was below a threshold also varied by bay segment (Table S3).
 The number of days when temperature was above 29 $^\circ$C and salinity was below 25 ppt had the largest slopes of 1.4, 1.2, and 0.7 days per year for OTB, HB, and MTB, respectively.
@@
 The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure [-@fig-sgmod]a, @tbl-modtab).
-The interaction between the temperature and salinity metric was included in the most parsimonious model (marginally significant, p \< 0.1), although the direction of the relationship of temperature with time period or salinity with time period did not noticeably change using different conditional values for salinity or temperature, respectively.
+The interaction between the temperature and salinity metric was included in the most parsimonious model (marginally significant, p \< 0.1), although the strength of the relationship of temperature or salinity with time period did not noticeably change using different conditional values for salinity or temperature, respectively.
 The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (`r epcmodsum2`), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab).
-Note that [@fig-sgmod] shows model predictions for OTB only, although the shape of the relationships do not change for other bay segments other than as additive shifts of the curves up or down (i.e., no interactions of bay segments with the other predictors were in the original models).
+Note that [@fig-sgmod] shows model predictions for OTB only, although the shape of the relationships do not change for other bay segments other than as additive shifts of the curves up or down.
 
@@
 These long-term changes manifested into consistent trends in select stressor metrics; the continuous number of days each year when temperature was above and/or salinity was below defined thresholds consistently increased across the period of record.
-Notably, bottom and surface trends for water temperature and salinity in the EPC dataset were similar, suggesting that any potential stress on biotic resources related to these factors are likely occurring throughout the water column.
 
@@
 @nickerson2023 evaluated sea surface trends at a larger spatial scale for Tampa Bay, the West Florida Continental Shelf, and the adjacent Gulf of Mexico.
-Temperature trends were similar for Tampa Bay (the EPC dataset was also used), although the power to detect trends decreased for shorter time periods, as verified herein with the FIM and PDEM datasets.
+Temperature trends were similar to those herein for Tampa Bay (the EPC dataset was also used).
 @nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, which somewhat conflicts with our results showing distinct increases in the summer, particularly for the upper bay segments (OTB and HB).
 However, our assessment evaluated trends at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to their larger spatial area and model domain.
-These differences are understandable given the analysis goals and methods for data synthesis.
 Our results showing increases in temperature and decreases in salinity for OTB and HB for the summer months are likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.
@@
 Model results for the EPC and FIM datasets both suggested that increasing temperature and decreasing salinity were associated with potential seagrass loss post-2016, described primarily using separate interaction terms of temperature or salinity with time period.
-For the EPC model, the interaction term was not significant for temperature, whereas the interaction was marginally significant for the salinity metric such that a negative association was observed post-2016 as compared to pre-2016 where frequency occurrence of seagrass declined with the increasing number of days when salinity was below 25 ppt.
-Likewise, the FIM models had a significant interaction term for the association of temperature with time period and a marginally significant interaction term for the association of salinity with time period, such that percent cover of seagrasses showed a decline with increasing temperature and decreasing salinity post-2016.
-An important distinction between the EPC and FIM models is that the former evaluated the number of days above/below thresholds each year to quantify increases in annual stress associated with temperature and salinity, whereas the latter evaluated observed temperature and salinity values at the time of seagrass sampling.
+For the EPC model, the interaction term was not significant for temperature, whereas the interaction was marginally significant for the salinity metric, such that a negative association between seagrass and salinity stress was observed post-2016 as compared to pre-2016.
+Likewise, the FIM models had a significant interaction term for the association of temperature with time period and a marginally significant interaction term for the association of salinity with time period.
+An important distinction between the EPC and FIM models is that the former evaluated the number of days above/below thresholds each year to quantify annual temperature or salinity stress, whereas the latter evaluated observed temperature and salinity values at the time of seagrass sampling.
 As such, both models attempted to describe the role of these stressors on potential seagrass change, but use different independent variables given the different sampling designs of each monitoring programs.
-The FIM model also used water quality measurements collected at the same locations where seagrasses were observed, whereas the EPC model provided a general characterization of seagrass transect data with water quality at stations farther from shore.
 These differences highlight challenges describing noisy relationships in long-term ecological datasets, while also demonstrating the utility of our weight-of-evidence approach to describe these relationships.
@@
 An additional caveat of our models was the use of "thresholds" to define potential stressor metrics for temperature and salinity on an annual time scale.
-Our choice to use 30 $^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed, as shown by the mixed-effects linear models for each bay segment.
+Our choice to use 30 $^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed.
 That is, sufficient change and variation in the independent variables for the models of seagrass change were needed to statistically describe potential relationships.
 These thresholds are likely not ecologically significant since most of the reported values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity [@phillips1960; @mcmillan1967; @zieman1975; @lirman2003].
-Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values (i.e., higher temperature, lower salinity), conditions in Tampa Bay in recent years are generally within the ecological niche for seagrasses.
+Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values, conditions in Tampa Bay in recent years are generally within the ecological niche for seagrasses.
 This does not suggest that these factors are unimportant, both currently and in the future.
-Our models suggested that temperature and salinity are potentially associated with seagrass loss and the consistent long-term trend in temperature and salinity imply that these conditions are likely to become more stressful in the future, further confounding seagrass recovery efforts in the region.
-Moreover, the GAMs used to model continuous temperature and salinity at each EPC monitoring station likely did not fully describe the current extent of stress given that extreme low or high values were not often captured by the daily modeled predictions (Figure S5).
-Regardless of the above limitations, climate stressors are likely to affect Tampa Bay's seagrass ecology in the future, as suggested by our results.
+Our models suggested that temperature and salinity are potentially associated with seagrass loss and the consistent long-term trend in both imply that these conditions are likely to become more stressful in the future, further confounding seagrass recovery efforts in the region.
 
-Additional limitations of our model may relate to an incomplete description of factors influencing seagrass growth.
+Additional limitations of our models may relate to an incomplete description of factors influencing seagrass growth.
 This relates to the inclusion of additional drivers as well as issues related to correlation/causation.
-For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, external nitrogen inputs in particular, in affecting light environments for seagrass growth [@greening2014; @sherwood2017].
+For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of external nitrogen inputs in affecting light environments for seagrass growth [@greening2014; @sherwood2017].
 As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences after accounting for light attenuation could additionally be explained by temperature or salinity.
@@
 Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay, particularly in OTB where the most loss occurred [@tbep0696; @Beck20a].
-This suggests that other factors are influencing seagrass growth and the inclusion of light attenuation in the models was likely not critical for describing the roles of temperature and salinity.
 A final consideration for our models relates to how seagrasses may influence their environment, particularly for the PDEM and FIM datasets where temperature and salinity were measured at the same locations as seagrass.
-For example, temperature may simply be higher in locations where seagrasses are present and can absorb solar radiation, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses.
+For example, temperature may simply be lower in locations where seagrasses are present and can absorb solar radiation, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses.
 This explanation cannot be ruled out with the existing datasets, although the trend analyses and models suggest that climate-related stressors driving seagrass loss is a more likely scenario.
@@
 These regional losses suggest that large-scale stressors are driving these changes, leading us to our initial hypothesis that climate-related stressors could partially explain the change in Tampa Bay.
-Based on our results, the losses elsewhere may potentially be explained by temperature and salinity and are worth exploring in other southwest Florida coastal regions and estuaries where long-term datasets exist.
+Based on our results, the losses elsewhere may potentially be explained by temperature and salinity and are worth exploring in other southwest Florida coastal regions where long-term datasets exist.
 Additional factors that could explain these changes are also likely co-occurring with climate stress, some of which are unique to Tampa Bay and others that are more likely pervasive.
@@
 Attached macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present in Tampa Bay [@Beck20b].
-Competitive differences between seagrasses and macroalgae are poorly understood in these systems [but see @bell1997; @brewton2023], in addition to insufficient macroalgae data in Tampa Bay that cannot clearly describe seasonal growth and distribution patterns and nutrient cycling.
+Competitive differences between seagrasses and macroalgae are poorly understood in these systems [but see @bell1997; @brewton2023], in addition to insufficient macroalgae data in Tampa Bay that cannot clearly describe seasonal growth, distribution patterns, and nutrient cycling.
 Discrete pollutant loading events in Tampa Bay have been documented to promote both phytoplankton and macroalgae growth [@beck2022a; @scolaro2023; @tomasko2023].
-The role that evolving nutrient loading and changing climatic conditions may have on Tampa Bay's primary producers – particularly algal and seagrass growth and interactions in recent years – is less understood.
+The role that evolving nutrient loading and changing climatic conditions may have on Tampa Bay's primary producers – particularly algal and seagrass growth and interactions in recent years – is now well understood.
 Finally, additional research has focused on how diseases and pathogens can influence seagrass growth patterns in Florida [@robblee1991; @duffin2021; @vanbogaert2018].
@@
 In Tampa Bay, the long-term trends of decreasing salinity, especially in the upper bay segments, suggest that the hydrologic load into the system has had a greater influence on salinity regimes than the effects of sea-level rise.
-This hypothesis is supported by our assessment of precipitation patterns over time, where the long-term increase is inversely correlated with the decrease in salinity.
+This hypothesis is supported by our assessment of precipitation patterns over time, where the long-term increase is inversely associated with the decrease in salinity.
 
@@
 Our models suggested that rising temperatures and decreasing salinities are additively affecting seagrasses, rather than multiplicatively, as evidenced by lack of significant interactions of these stressors.
-Future analyses may show more significant associations between physical habitat conditions and seagrass change as the trends are very likely to continue to push seagrasses further outside of their tolerance ranges.
+Future analyses may show more significant associations between physicochemical habitat conditions and seagrass change as the trends are very likely to continue to push seagrasses further outside of their tolerance ranges.
 These analyses should be supported by additional data collection efforts, particularly high-resolution continuous monitoring data that provide a more precise assessment of diurnal stress across multiple time-scales.
@@
 Management actions that have historically been effective may have inadequate capacity to maintain ecosystem status due to changing resilience of ecological resources affected by climate change.
-For example, nitrogen load reductions have been effective at restoring seagrass in Tampa Bay, primarily due to wastewater load allocations being constrained to the historical nutrient assimilative capacity of Tampa Bay [@greening2006; @greening2014].
+For example, nitrogen load reductions have been effective at restoring seagrass in Tampa Bay [@greening2006; @greening2014].
 As Tampa Bay becomes hotter and fresher, similar load allocations may no longer be effectively assimilated and algal and seagrass ecology dynamics may shift.
@@
 
-![Long-term air temperature, precipitation (Jun-Aug), hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1975 to 2022.
+![Long-term air temperature, precipitation (Jun-Aug), Standard Precipitation Index (SPI), water temperature, and salinity trends from 1975 to 2022.
 The color shades for water temperature and salinity indicate sampling location and values shown are the averages (95% confidence interval) across all Environmental Protection Commission (EPC) stations in each bay segment and sampling months for each year.

### 2024-01-28 R/figs.R

@@ -315,19 +315,19 @@ p2 <- ggplot(toplo2, aes(x = yr, y = precip_mm / 1e3)) +
     y = 'Jun-Aug prec. (m)'
   )
 
-p3 <- ggplot(hydrodat, aes(x = year, y = hy_load)) +
-  geom_line() + 
-  geom_point() + 
-  geom_smooth(method = 'lm', se = F, formula = y~x, color = 'blue') +
-  coord_cartesian(xlim = range(speidat$yr)) + 
-  thm +
-  theme(
-    axis.text.x = element_blank()
-  ) +
-  labs(
-    x = NULL, 
-    y = expression(paste('Hy. load (', 10^3, ' t/y)'))
-  )
+# p3 <- ggplot(hydrodat, aes(x = year, y = hy_load)) +
+#   geom_line() + 
+#   geom_point() + 
+#   geom_smooth(method = 'lm', se = F, formula = y~x, color = 'blue') +
+#   coord_cartesian(xlim = range(speidat$yr)) + 
+#   thm +
+#   theme(
+#     axis.text.x = element_blank()
+#   ) +
+#   labs(
+#     x = NULL, 
+#     y = expression(paste('Hy. load (', 10^3, ' t/y)'))
+#   )
 
 p4 <- ggplot(speidat, aes(x = date, y = spi, fill = spisign)) + 
   geom_col(width = 35) +
@@ -434,9 +434,9 @@ p6 <- ggplot(toplo6, aes(x = yr, y = avev, group = loc, color = loc)) +
     shape = NULL
   )
 
-p <- p1 + p2 + p3 + p4 + (p5 + p6 + plot_layout(ncol = 1, guides = 'collect')) + plot_layout(ncol = 1, heights = c(1, 1, 1, 1, 2.75)) & theme(legend.position = 'top')
+p <- p1 + p2 + p4 + (p5 + p6 + plot_layout(ncol = 1, guides = 'collect')) + plot_layout(ncol = 1, heights = c(1, 1, 1, 2.75)) & theme(legend.position = 'top')
 
-png(here('figs/meteowqraw.png'), height = 8, width = 7, family = 'serif', units = 'in', res = 500)
+png(here('figs/meteowqraw.png'), height = 7.5, width = 7, family = 'serif', units = 'in', res = 500)
 print(p)
 dev.off()
 
[…]

### 2024-01-29 manu-draft.qmd
add orcid to author yaml

@@
       email: mbeck@tbep.org
+      orcid: 0000-0002-4996-0059
       institute: [TBEP]
@@
       email: kfwalia@tbep.org
+      orcid: 0000-0002-6032-0570
       institute: [TBEP]
@@
       email: sscolaro@tbep.org
+      orcid: 0000-0003-3220-3737
       institute: [TBEP]
@@
       email: mburke@tbep.org
+      orcid: 0000-0002-3847-6501      
       institute: [TBEP]
@@
       email: Bradley.Furman@myfwc.com
+      orcid: 0000-0002-9284-8388
       institute: [FWRI]
@@
       email: karlen@epchc.org
+      orcid: 0000-0002-1432-9099
       institute: [EPCHC]
@@
       email: esherwood@tbep.org
+      orcid: 0000-0001-5330-302X
       institute: [TBEP]

### 2024-02-01 manu-draft.qmd
better characerization of precip trends in text

@@
 prpdif2 <- round(prpprd2$fit[nrow(prp)] - prpprd2$fit[1], 1)
+
+seponly <- speidat %>% 
+  select(yr, mo, precip_mm) %>%
+  filter(mo %in% 9)
+sepmod <- lm(precip_mm ~ yr, seponly)
+seppvl <- p_txt(coefficients(summary(sepmod))[2, 4])
+seprsq <- round(summary(sepmod)$adj.r.squared, 2)
+dryonly <- speidat %>% 
+  select(yr, mo, precip_mm) %>%
+  filter(!mo %in% c(6:9)) %>% 
+  summarise(prptot = sum(precip_mm), .by = yr)
+drymod <- lm(prptot ~ yr, dryonly)
+drypvl <- p_txt(coefficients(summary(drymod))[2, 4])
+dryrsq <- round(summary(drymod)$adj.r.squared, 2)
+apronly <- speidat %>% 
+  select(yr, mo, precip_mm) %>%
+  filter(mo %in% 4)
+aprmod <- lm(precip_mm ~ yr, apronly)
+aprpvl <- p_txt(coefficients(summary(aprmod))[2, 4])
+aprrsq <- round(summary(aprmod)$adj.r.squared, 2)
 ```
@@
 Mean annual air temperature in 1975 was `r airtempstr` $^\circ$C, whereas current mean annual air temperature in 2022 was `r airtempend` $^\circ$C, showing an overall increase in the period of record of `r airtempdif` $^\circ$C.
-Similarly, total precipitation during the rainy season has increased by `r prpslo` mm per year, although the model was not significant (`r prppvl`, $R^2$ = `r prprsq`).
-Removing September from the rainy season showed a significantly increasing trend of `r prpslo2` mm per year (`r prppvl2`, $R^2$ = `r prprsq2`).
+Similarly, total precipitation during the rainy season has increased by `r prpslo` mm per year, although the model was not significant (`r prppvl`).
+Removing September from the rainy season showed a significantly increasing trend of `r prpslo2` mm per year (`r prppvl2`, $R^2$ = `r prprsq2`; September alone did not have a significant trend, `r seppvl`).
 Using this model, total precipitation in 1975 was `r prpstr2` mm, whereas current total precipitation in 2022 was `r prpend2` mm, showing an overall increase in the period of record of `r prpdif2` mm.
+Notably, trends for the dry season (October through May) were not significant (`r drypvl`), except April which had a significantly increasing trend (`r aprpvl`, $R^2$ = `r aprrsq`).
 The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and 2015-2020.

### 2024-02-02 manu-draft.qmd
slight word edit

@@
 Removing September from the rainy season showed a significantly increasing trend of `r prpslo2` mm per year (`r prppvl2`, $R^2$ = `r prprsq2`; September alone did not have a significant trend, `r seppvl`).
-Using this model, total precipitation in 1975 was `r prpstr2` mm, whereas current total precipitation in 2022 was `r prpend2` mm, showing an overall increase in the period of record of `r prpdif2` mm.
+Using this model, mean precipitation in 1975 was `r prpstr2` mm, whereas current mean precipitation in 2022 was `r prpend2` mm, showing an overall increase in the period of record of `r prpdif2` mm.
 Notably, trends for the dry season (October through May) were not significant (`r drypvl`), except April which had a significantly increasing trend (`r aprpvl`, $R^2$ = `r aprrsq`).

### 2024-02-05 manu-draft.qmd
add EPC comments

@@
 These datasets were evaluated individually to assess trends and relationships with seagrass change to provide a weight-of-evidence approach for potential causal relationships driving the recent decline.
-First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at 45 stations in the major bay segments since the late 1970s (Figure [-@fig-map]b).
+First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at fixed stations in the major bay segments since the early 1970s (Figure [-@fig-map]b).
+The 45 stations with the longest and most complete temporal record from 1975 to present were used herein.
 Water quality samples are collected at each station from surface water grabs (e.g., nutrients, biological, and chemical constituents) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom.
@@
 An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions related to long-term (multi-decadal) climate change drivers.
-For comparison to water quality conditions, daily air temperature (Tampa International Airport \[TIA\] National Weather Service site) and precipitation (SWFWMD area-weighted watershed summaries) were used to characterize regional conditions for the period of record covered by the water quality samples (i.e., 1975 to present for the EPC data).
+For comparison to water quality conditions, daily air temperature (Tampa International Airport \[TIA\] National Weather Service site) and precipitation (SWFWMD area-weighted watershed summaries) were used to characterize regional conditions for the most consistent period of record covered by the water quality samples (i.e., 1975 to present for the EPC data).
 The rnoaa R package [@chamberlain23] was used to obtain the TIA temperature data.
@@
 An important distinction between the EPC and FIM models is that the former evaluated the number of days above/below thresholds each year to quantify annual temperature or salinity stress, whereas the latter evaluated observed temperature and salinity values at the time of seagrass sampling.
-As such, both models attempted to describe the role of these stressors on potential seagrass change, but use different independent variables given the different sampling designs of each monitoring programs.
+As such, both models attempted to describe the role of these stressors on potential seagrass change, but use different independent variables given the different sampling designs of each monitoring program.
 These differences highlight challenges describing noisy relationships in long-term ecological datasets, while also demonstrating the utility of our weight-of-evidence approach to describe these relationships.

### 2024-02-06 manu-draft.qmd
working on BF revisions

@@
       correspondence: false
+  - Chris Anastasiou: 
+      email: chris.anastasiou@swfwmd.state.fl.us
+      institute: [SWFWMD]
+      correspondence: false      
   - Edward T Sherwood:
@@
   - EPCHC: Environmental Protection Commission of Hillsborough County, 3629 Queen Palm Drive, Tampa, FL 33619, USA
+  - SWFWMD: Southwest Florida Water Management District, 7601 US Highway 301 North, Tampa, FL 33637, USA
 
@@
 
-Climate change is an existential threat to natural resources that sustain human health and the environment.
+Seagrasses have long been a focal point for management efforts aimed at restoring ecosystem health in estuaries worldwide.
 In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth.
@@
 Three ambient water quality datasets of varying sampling design and coverage were evaluated to characterize physicochemical environments in Tampa Bay and the potential relationships with seagrass change.
-Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 PSU per year.
+Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 ppt per year.
 These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay during the summer months.
-Simple linear models provided limited evidence that recent seagrass declines are associated with hotter and fresher conditions.
+Simple linear models provided a weight-of-evidence that recent seagrass declines are somewhat associated with hotter and fresher conditions.
 Trends in warming and increased precipitation in the region are likely to continue, further creating suboptimal conditions for seagrasses in Tampa Bay.
@@
 Furthermore, natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020].
-Contemporary management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].
+Contemporary management actions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].
 
-The sustained coverage or restoration of seagrasses in stressed coastal environments requires management actions that support vegetative growth, reproduction, and coverage expansion.
+The sustained coverage or restoration of seagrasses in coastal environments requires environmental conditions that support vegetative growth, reproduction, and coverage expansion.
 A long-standing approach adopted by numerous management entities has been the control of external nutrient inputs in systems where excessive algal growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @han2014; @greening2014].
@@
 Climate change has complicated the understanding of ecosystem response to conventional stressors, presenting new challenges and expectations for how ecological resources will respond to management actions [@statham2012a; @sherwood2013].
-In addition to sea-level rise, the most anticipated effects of climate change in coastal environments are increasing temperature and alteration of precipitation patterns.
+In addition to sea-level rise, the most anticipated effects of climate change in coastal environments are increased temperature and altered precipitation patterns.
 These changes will profoundly alter physicochemical habitats, creating suboptimal or uninhabitable conditions for many species [@madeira2012; @lefcheck2017; @hensel2023].
 @lefcheck2017 evaluated interactive effects of water clarity and rising temperatures on seagrasses with over 30 years of data in Chesapeake Bay.
-The effects of decreased water clarity for seagrasses in shallow habitats were exacerbated by acute warming related to climate change, demonstrating a concerning synergy of multiple stressors.
+The effects of decreased water clarity for seagrasses in shallow habitats were exacerbated by acute warming related to climate change, demonstrating a concerning synergy of stressors most likely to affect Tampa Bay.
 @moreno-marin2018 produced similar results using a multifactorial experiment that considered temperature, light, and nutrient (nitrogen) availability for North Sea eelgrass (*Zostera marina*).
 Species shifts are also expected to occur as changing physical conditions decrease the competitive advantages of historically abundant species.
-Again in Chesapeake Bay, the abundance of widgeongrass (*Ruppia maritima*) has been mediated by nutrient reductions, while also outcompeting and replacing formerly dominant *Z.
-marina* given higher tolerance to increased temperatures [@hensel2023].
-Changing frequency and severity of precipitation patterns may further alter the distribution and abundance of seagrasses by pushing species outside of optimal salinity tolerance ranges [@rasheed2011; @webster2021].
+In Chesapeake Bay, the abundance of widgeongrass (*Ruppia maritima*) has responded positively to nutrient reduction and is replacing the formerly dominant but now heat-stressed [*Z.
+marina*, @hensel2023].
+Changing frequency and severity of precipitation patterns may further alter salinity regimes and with it the distribution and abundance of seagrasses throughout the estuary [@rasheed2011; @webster2021].
 These changes may produce hypersaline conditions under periods of prolonged drought or more freshwater conditions with increased storm events.
-Seagrasses may respond differently under individual or multiple stressors [@lefcheck2017; @zhang2023], suggesting place-based assessments of response to climate change provide valuable insights for managers responsible for maintaining ecological resources.
+Seagrass species may respond differently under individual or multiple stressors and each watershed will respond differently to climate change [@lefcheck2017; @zhang2023], suggesting place-based empirical assessments will be needed to properly inform management decisions.
 
-In Tampa Bay, Florida, seagrasses are a primary indicator of bay health that have been the focus of management efforts over the last thirty years [@sherwood2017].
+In Tampa Bay, Florida, seagrasses are a primary indicator of bay health and have been the focus of management efforts for the last thirty years [@sherwood2017].
 Through successful reduction of external nitrogen loads, seagrasses have recovered from a low, system-wide coverage in the 1980s to an all-time high in 2016 of 41,655 acres [16,857 ha, @greening2006; @greening2014].
@@
 Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality primarily from direct external nutrient inputs from wastewater and stormwater [@greening2014].
-Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small tributaries and manmade flood control conveyances [@tbep0623].
+Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small channelized tributaries and manmade flood control conveyances [@tbep0623].
 Notably, Old Tampa Bay has restricted circulation from multiple land bridges associated with causeways that traverse the bay [@sherwood2015; @luther2022].
@@
 The second goal of the analysis was to evaluate if seagrass changes were linked to long-term changes in water temperature and salinity.
-The conceptual model for evaluating these changes describes the niche space where seagrass growth and reproduction is hypothesized to be greatest within optimal ranges for forcing factors that are present in the environment [@hutchinson1957; @vandermeer1972].
-In the simplest form, this can conceptualized as a bell curve with optimal conditions defined within a range of values for a single parameter (e.g., minimum and maximum temperatures where a species is typically observed), where reduced growth or mortality is observed outside of these ranges.
+The conceptual model for evaluating these changes describes the fundamental niche space where seagrass growth and reproduction is hypothesized to be greatest within optimal ranges for forcing factors that are present in the environment [@hutchinson1957; @vandermeer1972].
+In the simplest form, this can conceptualized as a bell curve with optimal conditions defined within a range of values for a single parameter, where reduced growth or mortality is observed outside of these ranges.
 Because both water temperature and salinity were evaluated, the same model can be conceptualized in two-dimensional space ([@fig-concept]).
@@
 Based on the results of the trend tests, we hypothesized that seagrasses are likely stressed by both high temperature and low salinity (bottom right, @fig-concept).
-Although the optimal niche space can be defined in multiple dimensions for many parameters, we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have generally not been limiting for growth in recent years.
+Although the fundamental niche space can be defined in multiple dimensions for many parameters, we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have been relatively stable in recent years [@tbep0323].
 
 A fundamental challenge describing niche space is identifying the boundaries for optimal conditions.
-In Tampa Bay, three dominant seagrass species occur: *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinum* (turtle grass) [@lewis1985].
+In Tampa Bay, three dominant seagrass species occur: *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinum* (turtle grass) [@lewis1985; @phillips88].
 These species co-occur often in mixed beds throughout the bay, although some differences in abundance are observed across salinity ranges.
@@
 Because of the uncertainty in defining *in situ* thresholds for optimal temperature and salinity ranges, multiple thresholds were evaluated to describe the potential for stress and how it may be related to changes in seagrass.
-Distinctions were not made between species, primarily due to lack of consensus between studies and likely site-specific ranges that affect seagrass growth in Tampa Bay.
+Distinctions were not made between species, primarily due to lack of consensus between studies and likely site-specific ranges that affect seagrass growth in Tampa Bay, as well as challenges of modeling fundamental and realized niche spaces between competing species [@araújo2006].
 First, we developed metrics of potential temperature and salinity stress by quantifying the maximum number of continuous days each year when temperature was above or salinity was below a given threshold.
@@
 The stressor metrics were compared to frequency occurrence (all species) each year by bay segment.
-Generalized Linear regression Models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre-2016, decline post-2016).
+Generalized Linear regression Models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre 2016, decline post 2016).
 Two models were evaluated, one with the temperature and salinity metrics together and another with the both metric.
@@
 All stations had significantly increasing temperature and decreasing salinity from 1975 to present, excluding one station in HB that did not have significantly decreasing bottom salinity (Figure [-@fig-kendall]a).
-Seasonally, most stations had more significant trends in the summer, early fall period for both temperature and salinity (Figure [-@fig-kendall]b, c), although some variation was observed between bay segments.
+Seasonally, most bay segments had more significant stations during the summer, early fall period for both temperature and salinity (Figure [-@fig-kendall]b, c), although some variation was observed throughout the bay.
 Temperature trends were more often observed in the summer or early fall for the upper bay segments (OTB, HB), whereas the lower bay segments (MTB, LTB) had more seasonally homogeneous trends.
@@
 Linear mixed-effects models for the EPC data showed similar trends in each bay segment for the number of days when temperature was above different thresholds.
-All of the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed significantly increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (Table S3).
+All of the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (Table S3).
 The estimated slopes for the number of days when temperature was above 30 $^\circ$C varied from 0.9 days per year in LTB to 1.1 days per year in MTB.
-Likewise, the average number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied from 8 (OTB, HB) to 21 (LTB) days in 1975 to 55 (HB) to 63 (LTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022).
+Likewise, the mean number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied from 8 (OTB, HB) to 21 (LTB) days in 1975 to 55 (HB) to 63 (LTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022).
 Trends for the other two temperature thresholds were similar.
 
-The salinity models were less similar between bay segments compared to the temperature models, primarily because of the natural variation in salinity along the bay's longitudinal axis (Tables S3, S4).
+The salinity models were less similar between bay segments compared to the temperature models, primarily because of the natural salinity gradient along the bay's longitudinal axis (Tables S3, S4).
 None of the bay segments had significantly increasing number of days per year when salinity was below 15 ppt.
@@
 The number of days per year when salinity was below 25 ppt in OTB increased by 2 days per year, whereas the number of days in HB increased by 0.9 days per year from 1975 to present.
-Likewise, the average number of days per year when salinity was continuously below 25 ppt at the beginning and end of the period of record was 128 (OTB) and 131 (HB) days in 1975 and 171 (HB) and 219 (OTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022).
+Likewise, the mean number of days per year when salinity was continuously below 25 ppt at the beginning and end of the period of record was 128 (OTB) and 131 (HB) days in 1975 and 171 (HB) and 219 (OTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022).
 
@@
 
-Global increases in temperature and precipitation related to climate change have contributed to widespread alteration of the structure and functioning of natural environments [@osland2015; @oliver2018].
-At a local scale, these changes have been demonstrated herein for Tampa Bay where the evaluation of long-term trends in water temperature and salinity showed clear changes mirrored by long-term changes in air temperature and precipitation.
-Over the previous fifty years, Tampa Bay has consistently trended towards hotter and fresher conditions, where water temperature has approximately increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year.
+Global increases in temperature and altered precipitation patterns related to climate change have had measurable affects on the structure and functioning of a wide range of natural environments [@osland2015; @oliver2018].
+For Tampa Bay, these changes have been demonstrated using long-term trends in water temperature and salinity, which mirrored long-term changes in air temperature and precipitation.
+Over the previous fifty years, Tampa Bay has gotten hotter and fresher; water temperature has increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year.
 These changes were demonstrated in three long-term datasets with different sampling methods and periods of record.
-Understandably, the trends were most consistently observed in the dataset with the longest period of record (EPC) covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season at locations farther from the mouth of the Bay.
-These long-term changes manifested into consistent trends in select stressor metrics; the continuous number of days each year when temperature was above and/or salinity was below defined thresholds consistently increased across the period of record.
+Understandably, the trends were most cearly observed in the dataset with the longest period of record (EPC), covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season at locations farther from the mouth of the Bay.
+These long-term changes manifested into consistent trends in known seagrass stressors; the continuous number of days increased each year when temperature, salinity, or both crossed thresholds.
 
-Similar regional, long-term changes in coastal waters and estuaries have been observed by others [@nickerson2023; @shiir].
+Similar regional, long-term changes in coastal waters and estuaries have been observed by others [@carlson2018; r@nickerson2023; @shiir].
 @nickerson2023 evaluated sea surface trends at a larger spatial scale for Tampa Bay, the West Florida Continental Shelf, and the adjacent Gulf of Mexico.
 Temperature trends were similar to those herein for Tampa Bay (the EPC dataset was also used).
-@nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, which somewhat conflicts with our results showing distinct increases in the summer, particularly for the upper bay segments (OTB and HB).
-However, our assessment evaluated trends at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to their larger spatial area and model domain.
+@nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, although they rightfully acknowledge the sensitivity of their results to conditions at the start and end of the time series.
+Our assessment evaluated non-parametric trends (i.e., Kendall tests, less sensitive to outliers) at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to their larger spatial area and model domain.
 Our results showing increases in temperature and decreases in salinity for OTB and HB for the summer months are likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.
@@
 The upper limit of our warming estimate for Tampa Bay is comparable.
+Notably, @carlson2018 suggest a link betwen historical seagrass losses in Florida Bay and rapid warming in shallow areas with low surface reflectance.
 
@@
 The models did not provide a consistent, nor statistically powerful, explanation that increasing temperature and decreasing salinity were key (or the sole) drivers.
-However, evaluating all models together does suggest a pattern that demonstrates the value of considering multiple datasets and models to explain noisy patterns and compounding ecological processes.
-Model results for the EPC and FIM datasets both suggested that increasing temperature and decreasing salinity were associated with potential seagrass loss post-2016, described primarily using separate interaction terms of temperature or salinity with time period.
-For the EPC model, the interaction term was not significant for temperature, whereas the interaction was marginally significant for the salinity metric, such that a negative association between seagrass and salinity stress was observed post-2016 as compared to pre-2016.
+However, evaluating all models together as weight-of-evidence suggests there is value in considering multiple datasets and models to interpret noisy patterns and compounding ecological processes.
+Model results for the EPC and FIM datasets both suggested that increasing temperature and decreasing salinity were associated with potential seagrass loss post 2016, described primarily using separate interaction terms of temperature or salinity with time period.
+For the EPC model, the interaction term was not significant for temperature, whereas the interaction was marginally significant for the salinity metric, such that a negative association between seagrass and salinity stress was observed post 2016 as compared to pre 2016.
 Likewise, the FIM models had a significant interaction term for the association of temperature with time period and a marginally significant interaction term for the association of salinity with time period.
@@
 As such, both models attempted to describe the role of these stressors on potential seagrass change, but use different independent variables given the different sampling designs of each monitoring program.
-These differences highlight challenges describing noisy relationships in long-term ecological datasets, while also demonstrating the utility of our weight-of-evidence approach to describe these relationships.
+These differences highlight challenges describing autecological relationships in long-term datasets, while also demonstrating the utility of our weight-of-evidence approach to describe such relationships.
 
@@
 As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences after accounting for light attenuation could additionally be explained by temperature or salinity.
-Our use of time period (pre-/post-2016) as a categorical variable partially addressed this issue.
+Our use of time period (pre/post 2016) as a categorical variable partially addressed this issue.
 Modeling seagrass change as related to temperature or salinity for the entire record would have shown a spurious correlation of both with seagrass given the long-term recovery of seagrass.
@@
 
-![Example of mixed effects models for the estimated number of days per year that temperature (red) or salinity (blue) were above or below thresholds of 30 $^\circ$C or 25 psu, respectively, from the EPC data.
+![Example of mixed effects models for the estimated number of days per year that temperature (red) or salinity (blue) were above or below thresholds of 30 $^\circ$C or 25 ppt, respectively, from the EPC data.
 The bottom row (black) shows the number of days when both temperature was above and salinity was below the thresholds.

### 2024-02-07 manu-draft.qmd
BF revisions done

@@
 This does not suggest that these factors are unimportant, both currently and in the future.
-Our models suggested that temperature and salinity are potentially associated with seagrass loss and the consistent long-term trend in both imply that these conditions are likely to become more stressful in the future, further confounding seagrass recovery efforts in the region.
+Extreme temperature or precipitation events acting individually or in combination are likely captured by the trends in stressor metrics using these conservative thresholds, i.e., an increase in a bay segment median number of days also suggests extremes are increasing given the variation around these summary metrics (@fig-mixeff).
+Further, our models suggested that temperature and salinity are at least associated with seagrass loss and, if so, long-term trends in both are set to amplify their effect.
 
-Additional limitations of our models may relate to an incomplete description of factors influencing seagrass growth.
-This relates to the inclusion of additional drivers as well as issues related to correlation/causation.
+Additional limitations of our models may relate to an incomplete description of factors influencing seagrass growth, such as the inclusion of additional drivers and an incomplete or overly simplified causal network.
 For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of external nitrogen inputs in affecting light environments for seagrass growth [@greening2014; @sherwood2017].
 As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences after accounting for light attenuation could additionally be explained by temperature or salinity.
-Our use of time period (pre/post 2016) as a categorical variable partially addressed this issue.
-Modeling seagrass change as related to temperature or salinity for the entire record would have shown a spurious correlation of both with seagrass given the long-term recovery of seagrass.
+However, our use of time period (pre/post 2016) as a categorical variable indirectly addressed this issue.
+Modeling seagrass change as related to temperature or salinity for the entire record would have shown a spurious correlation of both with seagrass given the long-term recovery of seagrass, i.e., modeling challenges related to correlated predictors [@fourqurean2003].
 Thus, time period was necessary to control for these confounding relationships.
@@
 A final consideration for our models relates to how seagrasses may influence their environment, particularly for the PDEM and FIM datasets where temperature and salinity were measured at the same locations as seagrass.
-For example, temperature may simply be lower in locations where seagrasses are present and can absorb solar radiation, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses.
-This explanation cannot be ruled out with the existing datasets, although the trend analyses and models suggest that climate-related stressors driving seagrass loss is a more likely scenario.
+For example, temperature may simply be lower in locations where seagrasses are present and can absorb solar radiation, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses [@carlson2018].
+This explanation cannot be ruled out with the existing datasets, although the trend analyses and models suggest that climate-related stressors are a more likely scenario.
 This is especially true for water temperature trends captured by the EPC dataset which includes deeper, fixed sites adjacent to shallow seagrass flats.
@@
 The seagrass loss in Tampa Bay since 2016 is a notable phenomenon that is not limited to our study area [@lizcano-sandoval2022].
-Losses have been observed throughout southwest Florida during this time period, including Sarasota Bay directly south of Tampa Bay and Charlotte Harbor further south.
-These regional losses suggest that large-scale stressors are driving these changes, leading us to our initial hypothesis that climate-related stressors could partially explain the change in Tampa Bay.
-Based on our results, the losses elsewhere may potentially be explained by temperature and salinity and are worth exploring in other southwest Florida coastal regions where long-term datasets exist.
+Losses have been observed throughout southwest Florida during this time period, including Sarasota Bay directly south of Tampa Bay and Charlotte Harbor further south [@tomasko2020].
+These regional losses suggest that large-scale stressors are driving these changes, supporting our initial hypothesis that climate-related stressors could partially explain the change in Tampa Bay.
+Based on our results, the losses elsewhere may potentially be explained by temperature and salinity and are worth exploring in other southwest Florida coastal regions where long-term datasets exist [@tomasko2005].
 Additional factors that could explain these changes are also likely co-occurring with climate stress, some of which are unique to Tampa Bay and others that are more likely pervasive.
@@
 These blooms are exacerbated by the hydrologic conditions in OTB that contribute to relatively longer water residence times [@phlips2006; @lopez2021].
-The effect of warming temperature and decreasing salinity in OTB will further complicate the understanding of how these blooms manifest and persist each year [e.g., @stelling2023], and ultimately contribute to changes in the light environment affecting seagrass resources in this bay segment.
+The effect of warming temperature and decreasing salinity in OTB will further complicate the understanding of how these blooms manifest and persist each year [@koch2007; @stelling2023], and ultimately contribute to changes in the light environment affecting seagrass resources in this bay segment.
 
 Additional biotic factors could be influencing regional patterns in seagrass growth.
-In Tampa Bay and elsewhere, enhanced macroalgal production may also influence seagrass growth and recovery in Florida estuaries [@hall2022; @jei2022; @brewton2023; @scolaro2023].
+In Tampa Bay and elsewhere, enhanced macroalgal production has been a recent concern [@hall2022; @jei2022; @brewton2023; @scolaro2023].
 Attached macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present in Tampa Bay [@Beck20b].
-Competitive differences between seagrasses and macroalgae are poorly understood in these systems [but see @bell1997; @brewton2023], in addition to insufficient macroalgae data in Tampa Bay that cannot clearly describe seasonal growth, distribution patterns, and nutrient cycling.
+Competitive differences between seagrasses and macroalgae are poorly understood in these systems [but see @bell1997; @taplin2005; @brewton2023], in addition to insufficient macroalgae data in Tampa Bay that cannot clearly describe seasonal growth, distribution patterns, and nutrient cycling.
 Discrete pollutant loading events in Tampa Bay have been documented to promote both phytoplankton and macroalgae growth [@beck2022a; @scolaro2023; @tomasko2023].
@@
 
-Lastly, our result showing that salinity has decreased in Tampa Bay is generally contrary to expectations for coastal systems consistent with sea-level rise projections [@costa2023; @alarcon2024].
-Salinity changes with sea-level rise have caused numerous alterations of subtidal and nearshore habitats [@brinson1995; @white2017].
-In southwest Florida, the most common ecological example is the expansion of mangroves upland as porewater salinity and water levels have gradually increased over the past few decades [@borchert2018].
-Alteration of salinity regimes for surface and groundwater resources are also well documented.
-In Florida Bay, for example, widespread decline of *Thalassia testudinium* has been partially attributed to elevated salinity levels beyond the optimal range for the species [@hall2016].
-Sea level rise is further altering salinity dynamics in the region.
+Lastly, our result showing that salinity has decreased in Tampa Bay is contrary to expectations for how sea-level rise will affect coastal systems [@costa2023; @alarcon2024], as salinity increases with sea-level rise have already caused numerous alterations of subtidal and nearshore habitats [@brinson1995; @white2017].
+In southwest Florida, the most common ecological example is the upland expansion of mangroves in response to increased porewater salinity and water levels over the past few decades [@borchert2018].
+Alteration of salinity regimes for surface and groundwater resources have been well documented.
+In Florida Bay, for example, widespread decline of *T.
+testudinium* has been attributed to altered hydrology and drought-induced hypersaline conditions, and sea level rise is expected to further modify salinity dynamics in the region [@hall2016].
 @dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters.
-In Tampa Bay, the long-term trends of decreasing salinity, especially in the upper bay segments, suggest that the hydrologic load into the system has had a greater influence on salinity regimes than the effects of sea-level rise.
+In Tampa Bay, the long-term trends of decreasing salinity, especially in the upper bay segments, suggest that the hydrologic loading has had a greater influence on salinity regimes than the effects of sea-level rise.
 This hypothesis is supported by our assessment of precipitation patterns over time, where the long-term increase is inversely associated with the decrease in salinity.
@@
 Simple regression models provided weak, but partially-supporting evidence, that these changes can be linked to recent seagrass losses since 2016.
-Our models suggested that rising temperatures and decreasing salinities are additively affecting seagrasses, rather than multiplicatively, as evidenced by lack of significant interactions of these stressors.
+Our models suggested that rising temperatures and decreasing salinities have had additive rather than multiplicative effects on seagrass, as evidenced by lack of significant interactions in models involving both stressors.
 Future analyses may show more significant associations between physicochemical habitat conditions and seagrass change as the trends are very likely to continue to push seagrasses further outside of their tolerance ranges.
 These analyses should be supported by additional data collection efforts, particularly high-resolution continuous monitoring data that provide a more precise assessment of diurnal stress across multiple time-scales.
+Morpohological or physiological measurements at the individual level could also provide early indications of heat and osmotic stress.
 
 Natural resource managers should consider how these climate-related stressors may alter the effectiveness of intervention activities aimed at protecting ecological resources in Tampa Bay.
-Management actions that have historically been effective may have inadequate capacity to maintain ecosystem status due to changing resilience of ecological resources affected by climate change.
+Management actions that have historically been effective may not be able to maintain ecosystem resilience to climatic change.
 For example, nitrogen load reductions have been effective at restoring seagrass in Tampa Bay [@greening2006; @greening2014].
-As Tampa Bay becomes hotter and fresher, similar load allocations may no longer be effectively assimilated and algal and seagrass ecology dynamics may shift.
+As Tampa Bay becomes hotter and fresher, current nutrient loads may no longer be effectively assimilated and algal and seagrass ecology dynamics may shift.
 Strategies that mimic or restore pre-development hydrology or that further reduce allowable load inputs from regulated entities (e.g., additional stormwater controls, hydrological modifications) may be needed to confer additional resilience and adaptive capacity for seagrass in Tampa Bay.
-These considerations are especially critical for upper parts of Tampa Bay where a majority of seagrass loss has occurred and where temperature and salinity trends have been most pronounced.
-Reversal of recent trends may be more likely to occur if aggressive actions and controls are pursued sooner rather than later, given the challenges of restoring degraded estuarine ecosystems due to hysteresis.
+These considerations are especially critical for upper parts of Tampa Bay where a majority of seagrass loss has occurred and where temperature and salinity trends appear most pronounced.
+Reversal of recent trends may be more likely to occur if aggressive actions and controls are pursued sooner rather than later, given the challenges of restoring these long-lived foundation species once lost, ongoing development in the watershed, and the current climate trajectory.
 

### 2024-02-08 manu-draft.qmd
working on CA revisions

@@
       correspondence: false
-  - Chris Pratt: 
+  - Christopher Pratt: 
       email: prattc@epchc.org
@@
       correspondence: false
-  - Chris Anastasiou: 
+  - Christopher J Anastasiou: 
       email: chris.anastasiou@swfwmd.state.fl.us
@@
 These results should compel resource managers to consider the likelihood that reduced resilience of estuarine resources due to shifting ecological baselines driven by additional climate change drivers will complicate long-standing management paradigms.
-While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be confounded by other cliamte change drivers and warrant additional, complementary interventions to maintain ecosystem health into the future.
+While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be confounded by climate change drivers and warrant additional, complementary interventions to maintain ecosystem health into the future.
 
@@
 In addition to sea-level rise, the most anticipated effects of climate change in coastal environments are increased temperature and altered precipitation patterns.
-These changes will profoundly alter physicochemical habitats, creating suboptimal or uninhabitable conditions for many species [@madeira2012; @lefcheck2017; @hensel2023].
+These changes will profoundly alter physicochemical habitats, creating suboptimal or uninhabitable conditions for many species [@madeira2012; @lefcheck2017; @hammer2018; @hensel2023].
 @lefcheck2017 evaluated interactive effects of water clarity and rising temperatures on seagrasses with over 30 years of data in Chesapeake Bay.
-The effects of decreased water clarity for seagrasses in shallow habitats were exacerbated by acute warming related to climate change, demonstrating a concerning synergy of stressors most likely to affect Tampa Bay.
+The environmental stress on seagrasses from acute warming related to climate change was compounded by stress from poor light environments in shallow waters, demonstrating a concerning synergy of stressors most likely to affect Tampa Bay.
 @moreno-marin2018 produced similar results using a multifactorial experiment that considered temperature, light, and nutrient (nitrogen) availability for North Sea eelgrass (*Zostera marina*).
@@
 Three datasets of varying sampling designs and temporal coverage were used to assess the primary hypothesis that Tampa Bay is trending towards hotter and fresher conditions that are likely stressing seagrasses beyond their optimal tolerance ranges, particularly in upper bay segments where marginal light environments exist.
-This hypothesis was generated from preliminary assessments of datasets used herein and discussions with the regional scientific and management community, despite information in the primary literature suggesting the contrary for salinity [e.g., @costa2023].
-The analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring describing cover of individual species, and synoptic seagrass data collected with routine biotic and water quality surveys.
+This hypothesis was generated from preliminary assessments of datasets used herein and discussions with the regional scientific and management community.
+While other studies have suggested negative effects from climate change in the form of increased salinity [e.g., @costa2023], lower salinity may be a stressor for seagrass in Tampa Bay.
+The analysis was supported by 30-year seagrass datasets including aerial surveys of total seagrass distribution, annual transect monitoring describing species-specific percent cover, and synoptic seagrass data collected with routine biotic and water quality surveys.
 Specific focus was on the periods prior to and after 2016, when seagrasses recovered baywide and then declined despite water quality conditions remaining relatively stable.
@@
 The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee rivers that contribute to a majority of freshwater inflow to the bay.
-Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer [@morrison2006].
+The climate of Tampa Bay is subtropical with warm humid summers and cool less humid winters [@garcia2023].
+Unique to Tampa Bay and the entire central Florida peninsula is that this region is within a transition zone from more temperate weather to the north, similar to the rest of the southeastern United States, and more tropical weather to the south [@morrison2006].
 El Niño weather events have also been associated with prolonged periods of heavy rainfall contributing to seagrass reductions through increased stormwater nutrient loads [@schmidt2002; @greening2006; @morrison2006].
@@
 Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality primarily from direct external nutrient inputs from wastewater and stormwater [@greening2014].
-Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small channelized tributaries and manmade flood control conveyances [@tbep0623].
-Notably, Old Tampa Bay has restricted circulation from multiple land bridges associated with causeways that traverse the bay [@sherwood2015; @luther2022].
+Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia Rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small channelized tributaries and manmade flood control conveyances [@tbep0623].
+Notably, Old Tampa Bay has restricted circulation from multiple land bridges associated with causeways that traverse the bay, causing relatively longer residence times and accumulation of legacy pollutant loads compared to the other bay segments [@sherwood2015; @luther2022].
 Recurring seasonal harmful algal blooms of the dinoflagellete *Pyrodinium bahamense* have contributed to exceedances of the chlorophyll-a regulatory standard in Old Tampa Bay [@lopez2023].
@@
 
-The long-term recovery of seagrass habitats in Tampa Bay is a nationally-recognized success story that demonstrates application of a successful management paradigm through the EPA-administered National Estuary Program [@greening2006; @greening2014; @sherwood2017].
+The long-term recovery of seagrass habitats in Tampa Bay since the 1980s is a nationally recognized success story that demonstrates application of a successful management paradigm through the EPA-administered National Estuary Program [@greening2006; @greening2014; @sherwood2017].
 From 1988 to 2016, seagrasses increased 79% to 41,655 acres (16,857 ha), surpassing the regional management goal of restoring coverage to 95% of a benchmark 1950s coverage.
-The greatest areal coverage expansions were observed in the upper bay segments (OTB, HB, and MTB; Figure [-@fig-seagrasschg]a).
-Throughout this same period, external nitrogen load estimates decreased by about 2/3 from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed [@greening2014; @beck2019].
+Though Tampa Bay was far from pristine in the 1950s, aerial imagery was sufficient to estimate a relatively unimpacted condition for seagrass coverage.
+Since that time, the greatest areal coverage expansions were observed in the upper bay segments (OTB, HB, and MTB; Figure [-@fig-seagrasschg]a).
+External nitrogen load estimates decreased by about 2/3 by the 1980s from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed [@greening2014; @beck2019].
 
-From 2016 to present, dramatic losses of seagrasses have been observed in Tampa Bay, despite nutrient loading and light attenuation remaining relatively stable (Figure [-@fig-seagrasschg]a).
-Total cover in Tampa Bay has decreased by 28% (11,518 acres/4,661 ha) from the 2016 peak to a total baywide coverage of 30,137 acres (12,196 ha).
+From 2016 to present, dramatic seagrass loss has been observed in Tampa Bay, despite nutrient loading and light attenuation remaining relatively stable (Figure [-@fig-seagrasschg]a).
+Total cover in Tampa Bay has decreased by 28% (11,518 acres/4,661 ha) from the 2016 peak to a total baywide coverage of 30,137 acres (12,196 ha) in 2022.
 Losses have been most pronounced in Old Tampa Bay (62%; 6,963 acres/2,818 ha loss) and Hillsborough Bay (80%; 1,599 acres/647 ha loss).
 The current estimate for Old Tampa Bay of 4,183 acres (1,693 ha) is the lowest ever recorded in that bay segment since mapping efforts began in the 1980s.
-Coverage in Middle Tampa Bay decreased by 20% (1,926 acres/779 ha loss), whereas coverage in Lower Tampa Bay has remained stable, with only a 2% loss that is likely within the range of mapping error for the coverage estimates.
+Coverage in Middle Tampa Bay decreased by 20% (1,926 acres/779 ha loss), whereas coverage in Lower Tampa Bay has remained stable, with only a 2% loss which is close to the mapping error.
 
@@
 The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s (Figure [-@fig-seagrasschg]a, available at <https://data-swfwmd.opendata.arcgis.com/>).
-These maps are created by photointerpretation of aerial images obtained at the end of the growing season, typically during November-February.
-No species information is provided.
-A more detailed, but spatially-specific, data source is the Tampa Bay Interagency Seagrass Monitoring Program (Figure [-@fig-seagrasschg]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>).
+These maps are created from aerial images collected specifically to map seagrass and are acquired during a flight window from December and February.
+The maps are created by photointerpretation of image signatures coupled with a robust field verification and accuracy assessment.
+The maps provide a spatial estimate of seagrass cover at the landscape scale and is irrespective of species.
+Complementary to the SWFWMD seagrass mapping program is the Tampa Bay Interagency Seagrass Monitoring Program (Figure [-@fig-seagrasschg]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>).
 Annual transect surveys have been conducted since 1998 at 62 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017].
@@
 The first goal of the analysis was to describe spatial and temporal trends in water temperature and salinity using the three water quality datasets described above.
-This assessment provided an indication of the extent of change in Tampa Bay as context for understanding potential relationships with seagrass change.
+This assessment provided an indication of the extent of water quality change in Tampa Bay as context for understanding potential relationships with seagrass change.
 An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions related to long-term (multi-decadal) climate change drivers.
@@
 In Tampa Bay, three dominant seagrass species occur: *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinum* (turtle grass) [@lewis1985; @phillips88].
+Other less common species include *Ruppia maritima* (widgeon grass) and *Halophila engelmanii* (star grass), where the former is often mapped during wet years in the upper bay segments from the aerial surveys.
 These species co-occur often in mixed beds throughout the bay, although some differences in abundance are observed across salinity ranges.
@@
 Interactions with time period and a main effect for bay segment were also included as above.
-GLMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data and the seagrass response variable was presence/absence (i.e., a binomial distribution was used) to model the probability of observing seagrass.
+GLMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data.
+The seagrass response variable was presence/absence (i.e., a binomial distribution was used) to model the probability of observing seagrasses.
 Model input data were further subset to include only months from July to November to describe seagrasses during the growing season and to reduce potential seasonal effects.
@@
   filter(mo %in% 6:9) %>% 
-  summarise(prptot = sum(precip_mm), .by = yr)
+  summarise(prptot = sum(precip_mm), .by = yr) %>% 
+  filter(yr > 1975)
 prpmod <- lm(prptot ~ yr, prp)
@@
 The top and bottom water temperature or salinity changes were similar across bay segments (@fig-meteowqraw).
-Trends in water temperature were similar across bay segments with significant increases varying from 0.03 to 0.04 $^\circ$C per year, with a total increase of 1.4 (OTB) to 2.0 (HB) $^\circ$C from 1975 to present (@tbl-temptrndtab).
+Trends in water temperature were similar across bay segments with significant increases for all for all four bay segments varying from 0.03 to 0.04 $^\circ$C per year, with a total change from 1.4 (OTB) to 2.0 (HB) $^\circ$C across the period of record from 1975 to 2022 (@tbl-temptrndtab).
 Salinity trends were also similar between bay segments, although overall salinity was predictably higher for bay segments closer to the Gulf of Mexico.
@@
 
-Similar regional, long-term changes in coastal waters and estuaries have been observed by others [@carlson2018; r@nickerson2023; @shiir].
+Similar regional, long-term changes in coastal waters and estuaries have been observed by others [@carlson2018; @nickerson2023; @shiir].
 @nickerson2023 evaluated sea surface trends at a larger spatial scale for Tampa Bay, the West Florida Continental Shelf, and the adjacent Gulf of Mexico.
@@
 The upper limit of our warming estimate for Tampa Bay is comparable.
-Notably, @carlson2018 suggest a link betwen historical seagrass losses in Florida Bay and rapid warming in shallow areas with low surface reflectance.
+Notably, @carlson2018 suggest a link between historical seagrass losses in Florida Bay and rapid warming in shallow areas with low surface reflectance.
 

### 2024-02-08 manu-draft.qmd
reproc all data with sal mid replace for 1975

@@
 Linear mixed-effects models for the EPC data showed similar trends in each bay segment for the number of days when temperature was above different thresholds.
-All of the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (Table S3).
+All the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (Table S3).
 The estimated slopes for the number of days when temperature was above 30 $^\circ$C varied from 0.9 days per year in LTB to 1.1 days per year in MTB.
@@
 That is, sufficient change and variation in the independent variables for the models of seagrass change were needed to statistically describe potential relationships.
-These thresholds are likely not ecologically significant since most of the reported values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity [@phillips1960; @mcmillan1967; @zieman1975; @lirman2003].
+The reported threshold values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity [@phillips1960; @mcmillan1967; @zieman1975; @lirman2003].
 Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values, conditions in Tampa Bay in recent years are generally within the ecological niche for seagrasses.
 This does not suggest that these factors are unimportant, both currently and in the future.
-Extreme temperature or precipitation events acting individually or in combination are likely captured by the trends in stressor metrics using these conservative thresholds, i.e., an increase in a bay segment median number of days also suggests extremes are increasing given the variation around these summary metrics (@fig-mixeff).
-Further, our models suggested that temperature and salinity are at least associated with seagrass loss and, if so, long-term trends in both are set to amplify their effect.
+Extreme temperature or precipitation events acting individually or in combination are likely captured by the trends in stressor metrics using these lower thresholds, i.e., an increase in a bay segment median number of days also suggests extremes are increasing given the variation around these summary metrics (@fig-mixeff).
+Our thresholds may also be indicative of the potential for chronic sublethal effects of stress on seagrasses, reducing their resilience to other stressors.
+Regardless, our models suggested that temperature and salinity are at least associated with seagrass loss and, if so, long-term trends in both are set to amplify their effect.
 

### 2024-02-08 R/dat_proc.R

@@ -201,7 +201,7 @@ epcdat %>%
 
 # get daily temp predictions from GAM files ---------------------------------------------------
 
-fls <- list.files(here('data'), pattern = 'temp', full.names = T)
+fls <- list.files(here('data'), pattern = 'tempbot|temptop', full.names = T)
 obs <- gsub('\\.RData$', '', basename(fls))
 
 tempprd <- tibble(obs = obs) %>% 
@@ -279,7 +279,7 @@ epcdat %>%
 
 # get daily sali predictions from GAM files ---------------------------------------------------
 
-fls <- list.files(here('data'), pattern = 'sali', full.names = T)
+fls <- list.files(here('data'), pattern = 'salibot|salitop', full.names = T)
 obs <- gsub('\\.RData$', '', basename(fls))
 
 saliprd <- tibble(obs = obs) %>% 
@@ -600,7 +600,7 @@ mixmods <- thrdat %>%
       if(inherits(mod, 'try-error'))
         return(NA)
       
-      effs <- estimate_means(mod, at = 'yr = c(1976, 2022)')
+      effs <- estimate_means(mod, at = 'yr = c(1975, 2022)')
       
       summod <- summary(mod)$coefficients
       

### 2024-02-08 R/tabs.R

@@ -6,13 +6,22 @@ library(tbeptools)
 
 source(here('R/funcs.R'))
 
+# use mid salinity for 1975, bottom missing
+data(epcdata)
+epcdat <- epcdata %>% 
+  mutate(
+    Sal_Bottom_ppth = case_when(
+      yr == 1975 & is.na(Sal_Bottom_ppth) ~ Sal_Mid_ppth, 
+      T ~ Sal_Bottom_ppth
+    )
+  )
+
 # linear trend summaries ----------------------------------------------------------------------
 
-data(epcdata)
 data(fimsgtempdat)
 data(pincotemp)
 
-epctmp <- epcdata %>% 
+epctmp <- epcdat %>% 
   select(bay_segment, epchc_station, SampleTime, yr, matches('Top|Bottom')) %>% 
   filter(yr < 2023) %>% 
   pivot_longer(names_to = 'var', values_to = 'val', matches('Top|Bottom')) %>% 
@@ -69,7 +78,7 @@ pincotmp <- pincotemp %>%
 
 trnds <- bind_rows(epctmp, fimtmp, pincotmp) %>% 
   group_nest(org, bay_segment, var) %>% 
-  crossing(yrstr = c(1976, 1996, 2004)) %>% 
+  crossing(yrstr = c(1975, 1996, 2004)) %>% 
   mutate(
     i = 1:n(),
     data = purrr::pmap(list(i, yrstr, data), function(i, yrstr, data){

### 2024-02-08 R/tabs.R
redo figs, tables with new gams

@@ -3,6 +3,7 @@ library(here)
 library(mgcv)
 library(broom)
 library(tbeptools)
+library(flextable)
 
 source(here('R/funcs.R'))
 
@@ -14,7 +15,8 @@ epcdat <- epcdata %>%
       yr == 1975 & is.na(Sal_Bottom_ppth) ~ Sal_Mid_ppth, 
       T ~ Sal_Bottom_ppth
     )
-  )
+  ) %>% 
+  filter(yr < 2023)
 
 # linear trend summaries ----------------------------------------------------------------------
 
@@ -305,9 +307,9 @@ supp1mixtab <- totab %>%
   merge_at(i = 1, j = c(2:3), part = 'header') %>% 
   merge_at(i = 1, j = c(4:6), part = 'header') %>% 
   set_header_labels(i = 2, values = c('Bay Segment', 'Temperature', 'Salinity', 'Temperature', 'Salinity', 'Both')) %>% 
-  padding(padding = 0, part = 'all') %>% 
-  font(part = 'all', fontname = 'Times New Roman') %>% 
-  fontsize(size = 9, part = 'body')
+  flextable::padding(padding = 0, part = 'all') %>% 
+  flextable::font(part = 'all', fontname = 'Times New Roman') %>% 
+  flextable::fontsize(size = 9, part = 'body')
 
 save(supp1mixtab, file = here('tabs/supp1mixtab.RData'))
 

### 2024-02-08 manu-draft.qmd
redo figs, tables with new gams
more CA edits

@@
 Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 ppt per year.
-These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay during the summer months.
+These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay.
 Simple linear models provided a weight-of-evidence that recent seagrass declines are somewhat associated with hotter and fresher conditions.
@@
 Water quality samples are collected at each station from surface water grabs (e.g., nutrients, biological, and chemical constituents) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom.
-Most analyses herein used only bottom water measurements given the shallow depth and mixed water column of most of Tampa Bay [@weisberg2006].
-Time of sampling can vary, although most samples are collected from mid-morning to early afternoon.
+Most analyses herein used only bottom water measurements given the shallow depth and mixed water column of most of Tampa Bay [@weisberg2006], although 1975 bottom salinity used middle water column sampling since the former was not available until the following year.
+Most samples are collected from mid-morning to early afternoon.
 Compared to the additional datasets described below, the monitoring stations are generally in deeper water beyond where seagrasses occur along the shallow margins of the bay.
@@
 The top and bottom water temperature or salinity changes were similar across bay segments (@fig-meteowqraw).
-Trends in water temperature were similar across bay segments with significant increases for all for all four bay segments varying from 0.03 to 0.04 $^\circ$C per year, with a total change from 1.4 (OTB) to 2.0 (HB) $^\circ$C across the period of record from 1975 to 2022 (@tbl-temptrndtab).
+Trends in water temperature were similar across bay segments with significant increases for all for all four bay segments varying from 0.03 to 0.04 $^\circ$C per year, with a total change from 1.3 (OTB) to 1.7 (HB) $^\circ$C across the period of record from 1975 to 2022 (@tbl-temptrndtab).
 Salinity trends were also similar between bay segments, although overall salinity was predictably higher for bay segments closer to the Gulf of Mexico.
-Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, with a total decrease of 2.7 (OTB) and 2.1 (LTB) ppt from 1975 to present.
+Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, with a total decrease of 2.6 (OTB) and 1.6 (LTB) ppt from 1975 to present.
 
@@
 The EPC dataset was also used to provide detailed information on station-level trends from 1975 to present (@fig-kendall, see Figure S4 for 1998 to 2022).
-All stations had significantly increasing temperature and decreasing salinity from 1975 to present, excluding one station in HB that did not have significantly decreasing bottom salinity (Figure [-@fig-kendall]a).
+All stations had significantly increasing temperature and decreasing salinity from 1975 to present, although several stations in HB did not have significant salinity trends, the exception being the northernmost stations near the mouth of the Hillsborough Rive (Figure [-@fig-kendall]a).
 Seasonally, most bay segments had more significant stations during the summer, early fall period for both temperature and salinity (Figure [-@fig-kendall]b, c), although some variation was observed throughout the bay.
@@
 Linear mixed-effects models for the EPC data showed similar trends in each bay segment for the number of days when temperature was above different thresholds.
-All the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (Table S3).
-The estimated slopes for the number of days when temperature was above 30 $^\circ$C varied from 0.9 days per year in LTB to 1.1 days per year in MTB.
-Likewise, the mean number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied from 8 (OTB, HB) to 21 (LTB) days in 1975 to 55 (HB) to 63 (LTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022).
-Trends for the other two temperature thresholds were similar.
+All the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed increasing trends for each bay segment, with the largest slope of 1.5 days per year in OTB when temperature was above 29 $^\circ$C (Table S3).
+The estimated slopes for the number of days when temperature was above 30 $^\circ$C varied from less than 1 day per year in LTB to 1.1 days per year in MTB.
+Likewise, the mean number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied between bay segments, although most had a general increase of 50 days each year across the bay segments for the period of record (Table S5, see Table S6 for 1998 to 2022).
+The increase in the number days each year when temperatures were above 29 or 31 $^\circ$C was similar.
 
@@
 None of the bay segments had significantly increasing number of days per year when salinity was below 15 ppt.
-The only significant model for MTB was the number of days when salinity was below 25 ppt, which was increasing at a rate of 0.7 days per year.
-Both HB and OTB had significantly increasing number of days per year when salinity was below 20 and 25 ppt, although the slopes varied such that the rates in OTB were nearly double those in HB.
-The number of days per year when salinity was below 25 ppt in OTB increased by 2 days per year, whereas the number of days in HB increased by 0.9 days per year from 1975 to present.
-Likewise, the mean number of days per year when salinity was continuously below 25 ppt at the beginning and end of the period of record was 128 (OTB) and 131 (HB) days in 1975 and 171 (HB) and 219 (OTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022).
+The only significantly increasing model for MTB was the number of days when salinity was below 25 ppt, which was increasing at a rate of 0.5 days per year.
+Both HB and OTB had significantly increasing number of days per year when salinity was below 20 and 25 ppt, although the slopes varied such that the rates in OTB were more than twice those in HB.
+Some salinity models had significantly decreasing trends, although the total number of days at the beginning and end of the period of record were neglible (Table S5).
+Overall, OTB showed the largest increase in the number of days each year when salinity was below a theshold, particularly for 25 ppt, where the change was 86 days per year from 1975 to 2022 (130 days to 216 days per year, Table S5, see Table S6 for 1998 to 2022).
 
 The number of days when both temperature was above and salinity was below a threshold also varied by bay segment (Table S3).
-The number of days when temperature was above 29 $^\circ$C and salinity was below 25 ppt had the largest slopes of 1.4, 1.2, and 0.7 days per year for OTB, HB, and MTB, respectively.
-Likewise, the average number of days when both temperature was above 29 $^\circ$C and salinity was below 25 ppt from the beginning to the end of the period of record varied from 6, 12, and 1 day(s) in 1975 to 72, 68, and 34 days in 2022 for OTB, HB, and MTB, respectively.
+The number of days when temperature was above 29 $^\circ$C and salinity was below 25 ppt had the largest slopes of 1.4, 1.1, and 0.7 days per year for OTB, HB, and MTB, respectively.
+Likewise, the average number of days when both temperature was above 29 $^\circ$C and salinity was below 25 ppt from the beginning to the end of the period showed the greatest increase for OTB of 68 days (4 days to 72 days per year from 1975 to 2022).
 
@@
 These changes were demonstrated in three long-term datasets with different sampling methods and periods of record.
-Understandably, the trends were most cearly observed in the dataset with the longest period of record (EPC), covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season at locations farther from the mouth of the Bay.
+Understandably, the trends were most clearly observed in the dataset with the longest period of record (EPC), covering nearly fifty years of monthly observations.
 These long-term changes manifested into consistent trends in known seagrass stressors; the continuous number of days increased each year when temperature, salinity, or both crossed thresholds.
@@
 Our assessment evaluated non-parametric trends (i.e., Kendall tests, less sensitive to outliers) at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to their larger spatial area and model domain.
-Our results showing increases in temperature and decreases in salinity for OTB and HB for the summer months are likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.
+Our results showing increases in temperature and decreases in salinity in the summer months, most notably for OTB and northern stations of HB [@fig-kendall], are likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.
 These upper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to more rapid warming with shallower depths.

### 2024-02-08 supplement.qmd
shorten text, remove hydro load from figs
more CA edits

--- 2024-01-25 supplement.qmd
@@
       correspondence: false
+  - Bradley T Furman:
+      email: Bradley.Furman@myfwc.com
+      institute: [FWRI]
+      correspondence: false
+  - David J Karlen: 
+      email: karlen@epchc.org
+      institute: [EPCHC]
+      correspondence: false
+  - Christopher Pratt: 
+      email: prattc@epchc.org
+      institute: [EPCHC]
+      correspondence: false
+  - Christopher J Anastasiou: 
+      email: chris.anastasiou@swfwmd.state.fl.us
+      institute: [SWFWMD]
+      correspondence: false    
   - Edward T Sherwood:
@@
   - TBEP: Tampa Bay Estuary Program, 263 13th Ave S., Suite 350, St. Petersburg, Florida 33701 USA
+  - FWRI: Florida Fish and Wildlife Conservation Commission, Florida Fish and Wildlife Research Institute, 100 8th Avenue SE, St. Petersburg, FL 33701, USA
+  - EPCHC: Environmental Protection Commission of Hillsborough County, 3629 Queen Palm Drive, Tampa, FL 33619, USA
+  - SWFWMD: Southwest Florida Water Management District, 7601 US Highway 301 North, Tampa, FL 33637, USA
 
@@
 
-![Air temperature, precipitation (Jun-Aug), hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1998 to 2022.
+![Air temperature, precipitation (Jun-Aug), Standard Precipitation Index (SPI), water temperature, and salinity trends from 1998 to 2022.
 The color shades for water temperature and salinity indicate sampling location and values shown are the averages (95% confidence interval) across all Environmental Protection Commission (EPC) stations in each bay segment and sampling months for each year.

### 2024-02-09 manu-draft.qmd
CA revisions done

@@
 Three ambient water quality datasets of varying sampling design and coverage were evaluated to characterize physicochemical environments in Tampa Bay and the potential relationships with seagrass change.
-Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 ppt per year.
+Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.04 $^\circ$C per year and salinity decreasing by 0.04 - 0.06 ppt per year, translating to an increase of 1.3 to 1.7 $^\circ$C and a decrease of 1.6 to 2.6 ppt over the last fifty years.
 These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay.
@@
 
-The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale.
+The monitoring and management of seagrasses in coastal environments has received substantial attention on a global scale.
 Seagrasses are fundamental indicators of coastal ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023].
@@
 Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality primarily from direct external nutrient inputs from wastewater and stormwater [@greening2014].
-Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia Rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small channelized tributaries and manmade flood control conveyances [@tbep0623].
+Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia Rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small, channelized tributaries and manmade flood control conveyances [@tbep0623].
 Notably, Old Tampa Bay has restricted circulation from multiple land bridges associated with causeways that traverse the bay, causing relatively longer residence times and accumulation of legacy pollutant loads compared to the other bay segments [@sherwood2015; @luther2022].
@@
 All bay segments are relatively shallow, with a baywide mean depth of approximately 3 m.
-Light penetration typically reaches bottom habitats under current conditions, although seagrasses were historically limited by high phytoplankton production that affected light environments[@greening2014; @johansson15].
+Light penetration typically reaches bottom habitats under current conditions, although seagrasses were historically limited by high phytoplankton production that affected light environments [@greening2014; @johansson15].
 
@@
 The conceptual model for evaluating these changes describes the fundamental niche space where seagrass growth and reproduction is hypothesized to be greatest within optimal ranges for forcing factors that are present in the environment [@hutchinson1957; @vandermeer1972].
-In the simplest form, this can conceptualized as a bell curve with optimal conditions defined within a range of values for a single parameter, where reduced growth or mortality is observed outside of these ranges.
+In the simplest form, this can be conceptualized as a bell curve with optimal conditions defined within a range of values for a single parameter, where reduced growth or mortality is observed outside of these ranges.
 Because both water temperature and salinity were evaluated, the same model can be conceptualized in two-dimensional space ([@fig-concept]).
@@
 These species co-occur often in mixed beds throughout the bay, although some differences in abundance are observed across salinity ranges.
-Shoal grass is tolerant of a wide range of salinity [@lirman2003], but is more abundant in oligo/mesohaline portions of Tampa Bay.
+Shoal grass is tolerant of a wide range of salinity [@lirman2003] but is more abundant in oligo/mesohaline portions of Tampa Bay.
 Conversely, turtle grass is less tolerant of low salinity and is more abundant in more euryhaline conditions near the mouth of Tampa Bay.
@@
 The estimated slopes for the number of days when temperature was above 30 $^\circ$C varied from less than 1 day per year in LTB to 1.1 days per year in MTB.
-Likewise, the mean number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied between bay segments, although most had a general increase of 50 days each year across the bay segments for the period of record (Table S5, see Table S6 for 1998 to 2022).
+Likewise, the mean number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied between bay segments, although most had an approximate increase of 50 days each year for the period of record (@tbl-mixdaytab, Table S5, see Table S6 for 1998 to 2022).
 The increase in the number days each year when temperatures were above 29 or 31 $^\circ$C was similar.
@@
 Some salinity models had significantly decreasing trends, although the total number of days at the beginning and end of the period of record were neglible (Table S5).
-Overall, OTB showed the largest increase in the number of days each year when salinity was below a theshold, particularly for 25 ppt, where the change was 86 days per year from 1975 to 2022 (130 days to 216 days per year, Table S5, see Table S6 for 1998 to 2022).
+Overall, OTB showed the largest increase in the number of days each year when salinity was below a theshold, particularly for 25 ppt, where the change was 86 days per year from 1975 to 2022 (130 days to 216 days per year, @tbl-mixdaytab, Table S5, see Table S6 for 1998 to 2022).
 
@@
 
-Global increases in temperature and altered precipitation patterns related to climate change have had measurable affects on the structure and functioning of a wide range of natural environments [@osland2015; @oliver2018].
+Global increases in temperature and altered precipitation patterns related to climate change have had measurable effects on the structure and functioning of a wide range of natural environments [@osland2015; @oliver2018].
 For Tampa Bay, these changes have been demonstrated using long-term trends in water temperature and salinity, which mirrored long-term changes in air temperature and precipitation.
-Over the previous fifty years, Tampa Bay has gotten hotter and fresher; water temperature has increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year.
+Tampa Bay has gotten hotter and fresher; water temperature has increased by 0.03 - 0.04 $^\circ$C per year and salinity has decreased by 0.04 - 0.06 ppt per year, translating to an increase of 1.3 to 1.7 $^\circ$C and a decrease of 1.6 to 2.6 ppt over the past fifty years.
 These changes were demonstrated in three long-term datasets with different sampling methods and periods of record.
@@
 Our assessment evaluated non-parametric trends (i.e., Kendall tests, less sensitive to outliers) at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to their larger spatial area and model domain.
-Our results showing increases in temperature and decreases in salinity in the summer months, most notably for OTB and northern stations of HB [@fig-kendall], are likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.
+Our results showing increases in temperature and decreases in salinity in the summer, early fall, most notably for OTB and northern stations of HB [@fig-kendall], are likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.
 These upper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to more rapid warming with shallower depths.
+The significant reduction in salinity for LTB is also of note, perhaps related to gravitational circulation patterns that export lower salinity water from upstream in the main shipping channels [@weisberg2006].
 Additionally, @shiir provided a recent assessment of a 2023 heatwave in south Florida, supported by a 20-year trend assessment that suggested estuaries were warming at nearly double the rate of the Gulf of Mexico.
@@
 Discrete pollutant loading events in Tampa Bay have been documented to promote both phytoplankton and macroalgae growth [@beck2022a; @scolaro2023; @tomasko2023].
-The role that evolving nutrient loading and changing climatic conditions may have on Tampa Bay's primary producers – particularly algal and seagrass growth and interactions in recent years – is now well understood.
+The role that evolving nutrient loading and changing climatic conditions may have on Tampa Bay's primary producers – particularly algal and seagrass growth and interactions in recent years – is not well understood.
 Finally, additional research has focused on how diseases and pathogens can influence seagrass growth patterns in Florida [@robblee1991; @duffin2021; @vanbogaert2018].
@@
 Start year describes the range of the trend test to the present year (2022).
-The starting value is the estimated temperature at the start year and the end value is the estimated temperature at 2022.
+The starting value is the estimated temperature at the start year and the end value is the estimated temperature in 2022.
 Total change is the difference between the two.
@@
 Start year describes the range of the trend test to the present year (2022).
-The starting value is the estimated salinity at the start year and the end value is the estimated salinity at 2022.
+The starting value is the estimated salinity at the start year and the end value is the estimated salinity in 2022.
 Total change is the difference between the two.

### 2024-02-09 manu-draft.qmd
starting cover letter

@@
 The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure [-@fig-sgmod]a, @tbl-modtab).
-The interaction between the temperature and salinity metric was included in the most parsimonious model (marginally significant, p \< 0.1), although the strength of the relationship of temperature or salinity with time period did not noticeably change using different conditional values for salinity or temperature, respectively.
+The interaction between the temperature and salinity metric was included in the most parsimonious model, although it was not significant.
 The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (`r epcmodsum2`), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab).

### 2024-02-09 cover-letter.qmd
cover letter done

---
format: 
  pdf:
    include-in-header:
      text: '\pagenumbering{gobble}'

editor: visual
---

\today

Dr. Paul Montagna, Dr. Linda Deegan, Editors-in-Chief\
Estuaries and Coasts

\vfill

Dear Dr. Montagna and Dr. Deegan,

\vspace{0.5cm}

Enclosed please find our manuscript, titled \`Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary', to be considered as an original article in Estuaries and Coasts.

The long-term recovery of Tampa Bay is a nationally recognized success story that demonstrates how the USEPA National Estuary Program is a valuable model for collaborative management.  Through large-scale reductions in external nutrient loading, seagrasses have recovered in the bay, reaching an all-time high in 2016.  Since that time, seagrasses have declined by nearly 1/3 and questions remain regarding the role of climate-related stressors in contributing to this decline.  We leverage multiple data sources to describe changing water temperature and salinity in the bay and relate these trends to the recent seagrass loss from 2016 to present. No such comprehensive assessment has previously been conducted for Tampa Bay and our results provide context for how climate stressors can confound existing management models, particularly related to nutrient management.

This work has not been published elsewhere, nor is it under consideration for publication anywhere else. Each author has contributed substantially to this work. All authors have contributed to the writing and review of the original draft. Analysis was conducted by the first author, with interpretation of results supported by all other co-authors. All authors have approved the final draft and the submitted product was prepared according to instructions for authors provided on the journal's website.

We greatly appreciate the opportunity to publish this work in Estuaries and Coasts and are confident readers will consider it a valuable contribution. Please feel free to contact us directly should additional information be needed.

\vspace{0.5cm}

Regards,

Marcus W. Beck, Tampa Bay Estuary Program, St. Petersburg, Florida, USA, [mbeck@tbep.org](mailto:mbeck@tbep.org)

\vfill

### 2024-02-09 R/tabs.R

@@ -307,9 +307,9 @@ supp1mixtab <- totab %>%
   merge_at(i = 1, j = c(2:3), part = 'header') %>% 
   merge_at(i = 1, j = c(4:6), part = 'header') %>% 
   set_header_labels(i = 2, values = c('Bay Segment', 'Temperature', 'Salinity', 'Temperature', 'Salinity', 'Both')) %>% 
-  flextable::padding(padding = 0, part = 'all') %>% 
-  flextable::font(part = 'all', fontname = 'Times New Roman') %>% 
-  flextable::fontsize(size = 9, part = 'body')
+  padding(padding = 0, part = 'all') %>% 
+  font(part = 'all', fontname = 'Times New Roman') %>% 
+  fontsize(size = 9, part = 'body')
 
 save(supp1mixtab, file = here('tabs/supp1mixtab.RData'))
 

### 2024-02-09 manu-draft.qmd
ESCO submitted

@@
 csl: estuaries-and-coasts.csl
+crossref: 
+  fig-title: Fig.
 author:
@@
 While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be confounded by climate change drivers and warrant additional, complementary interventions to maintain ecosystem health into the future.
+
+*Key words*: climate change; seagrass; salinity; temperature; Tampa Bay
+
+## Statements and Declarations {.unnumbered}
+
+The authors have no competing interests to declare that are relevant to the content of this article.
 

### 2024-03-31 manu-draft.qmd
starting ESCO revision

@@
 *Key words*: climate change; seagrass; salinity; temperature; Tampa Bay
-
-## Statements and Declarations {.unnumbered}
-
-The authors have no competing interests to declare that are relevant to the content of this article.
 
@@
 The top and bottom water temperature or salinity changes were similar across bay segments (@fig-meteowqraw).
-Trends in water temperature were similar across bay segments with significant increases for all for all four bay segments varying from 0.03 to 0.04 $^\circ$C per year, with a total change from 1.3 (OTB) to 1.7 (HB) $^\circ$C across the period of record from 1975 to 2022 (@tbl-temptrndtab).
+Trends in water temperature were similar across bay segments with significant increases for all four bay segments varying from 0.03 to 0.04 $^\circ$C per year, with a total change from 1.3 (OTB) to 1.7 (HB) $^\circ$C across the period of record from 1975 to 2022 (@tbl-temptrndtab).
 Salinity trends were also similar between bay segments, although overall salinity was predictably higher for bay segments closer to the Gulf of Mexico.
@@
 
-## Acknowledgments
-
-This work benefited from constructive discussions with the Technical Advisory Committee of the Tampa Bay Estuary Program (TBEP), in particular Mike Wessel, Ray Pribble, and Dave Tomasko.
-TBEP funding for this work stems from EPA Section 320 Grant Funds, and the TBEP's local government partners (Hillsborough, Manatee, Pasco, and Pinellas Counties; the Cities of Clearwater, St.
-Petersburg, and Tampa; Tampa Bay Water; and the Southwest Florida Water Management District) through contributions to the operating budget.
-We are indebted to the community of field and technical staff that have routinely collected and maintained the valuable long-term monitoring data used in this study.
-
 {{< pagebreak >}}
@@
 The bottom row (black) shows the number of days when both temperature was above and salinity was below the thresholds.
-The models included station as a random effect for each bay segment, with grey lines indicating individual station trends, grey points as actual number of days, and thicker lines indicating the overall model fit.
+The models included station as a random effect for each bay segment, with grey lines indicating individual station trends, grey points as actual number of days for each station, and thicker lines indicating the overall model fit.
 Slopes for significant models are shown in the bottom left of each facet.
@@
 
+## Data Availability {.unnumbered}
+
+All data and analysis code for this manuscript is available on GitHub at <https://github.com/tbep-tech/temp-manu>.
+A preprint of an earlier version of this manuscript is available at <https://doi.org/10.21203/rs.3.rs-3946855/v1>.
+
+## Acknowledgments {.unnumbered}
+
+This work benefited from constructive discussions with the Technical Advisory Committee of the Tampa Bay Estuary Program (TBEP), in particular Mike Wessel, Ray Pribble, and Dave Tomasko.
+We are thankful for constructive comments provided by the associate editor and two anonymous reviewers.
+We are indebted to the community of field and technical staff that have routinely collected and maintained the valuable long-term monitoring data used in this study.
+
+## Statements and Declarations {.unnumbered}
+
+### Author Contributions {.unnumbered}
+
+Conceptualization: Marcus W Beck, Kerry Flaherty-Walia, Sheila Scolaro, Maya C Burke, Edward T Sherwood; Data Curation: Bradley T Furman, David J Karlen, Christopher J Anastasiou, Christopher Pratt; Methodology: Marcus W Beck, Kerry Flaherty-Walia, Sheila Scolaro, Maya C Burke, Edward T Sherwood; Formal analysis and investigation: Marcus W Beck; Writing - original draft preparation: all authors; Writing - review and editing: all authors; Funding acquisition: Edward T Sherwood.
+
+### Ethics Declaration {.unnumbered}
+
+The authors have no competing interests to declare that are relevant to the content of this article.
+
+### Funding {.unnumbered}
+
+This study was funded by EPA Section 320 Grant Funds, and the TBEP's local government partners (Hillsborough, Manatee, Pasco, and Pinellas Counties; the Cities of Clearwater, St.
+Petersburg, and Tampa; Tampa Bay Water; and the Southwest Florida Water Management District) through contributions to the operating budget.
+
 ## References {.unnumbered}

### 2024-03-31 R/figs.R

@@ -208,17 +208,17 @@ p1 <- ggplot(toplo1, aes(x = factor(year), y = acres)) +
   theme(panel.grid.minor =element_blank(),
         panel.grid.major.x =element_blank(),
         # plot.background = element_rect(fill = NA, color = NA),
-        axis.text.y = element_text(colour = 'black'),
-        axis.text.x = element_text(colour = 'black', angle = 45, size = 6, hjust = 1),
+        axis.text.y = element_text(colour = 'black', size = 9),
+        axis.text.x = element_text(colour = 'black', angle = 60, size = 9, hjust = 1),
         strip.background = element_blank(),
-        # strip.text = element_text(hjust = 0, size = 13),
+        strip.text = element_text(size = 11),
         legend.position = 'none'
   ) +
   labs(
     y = 'Seagrass Coverage (x1,000 acres)',
     x = NULL,
     color = NULL,
-    subtitle = '(a) Coverage changes for all species by bay segment',
+    title = '(a) Coverage changes for all species by bay segment',
     caption = expression(italic('Source: Southwest Florida Water Management District'))
   )
 
@@ -241,23 +241,26 @@ p2 <- ggplot(toplo2, aes(x = yr, y = foest, color = Savspecies)) +
   theme_bw() + 
   theme(
     legend.position = 'top',
+    strip.text = element_text(size = 11),
     strip.background = element_blank(), 
     panel.grid.minor = element_blank(),
     panel.grid.major.x = element_blank(),
-    axis.text.x = element_text(colour = 'black', angle = 45, size = 6, hjust = 1),
+    legend.text = element_text(size = 11),
+    axis.text.y = element_text(colour = 'black', size = 9),
+    axis.text.x = element_text(colour = 'black', angle = 60, size = 9, hjust = 1),
   ) + 
   coord_cartesian(ylim = c(0, 1)) + 
   labs(
     y = 'Frequency Occurrence', 
     x = NULL, 
     color = NULL,
-    subtitle = '(b) Frequency occurrence changes by species by bay segment',
+    title = '(b) Frequency occurrence changes by species by bay segment',
     caption = expression(italic('Source: Interagency Seagrass Monitoring Program'))
   )
 
 p <- p1 + p2 + plot_layout(ncol = 1)
 
-png(here('figs/seagrasschg.png'), height = 6, width = 8, family = 'serif', units = 'in', res = 300)
[…]

### 2024-04-01 manu-draft.qmd
data summary table added, more revisions

@@
 load(file = here('data/gamfit.RData'))
+load(file = here('tabs/dattab.RData'))
 load(file = here('tabs/temptrndtab.RData'))
@@
 
-Two primary sources of data have been used to track seagrass change over time in Tampa Bay.
+Two primary sources of data have been used to track seagrass change over time in Tampa Bay (@tbl-dattab).
 The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s (Figure [-@fig-seagrasschg]a, available at <https://data-swfwmd.opendata.arcgis.com/>).
@@
 Annual transect surveys have been conducted since 1998 at 62 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017].
-This dataset provides species information on cover-abundance, frequency occurrence, and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed.
-Although the areal maps provide the standard for assessment of restoration goals, the transect data allow inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined.
+This dataset provides species information on cover-abundance, frequency occurrence [number of sample points with seagrass divided by total points on a transect, as in @sherwood2017], and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed.
+Although the areal maps provide the standard for assessment of restoration goals, the maps are produced every other year.
+The transect data are collected each year, allowing inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined.
 As such, the transect data were used below for comparison with temperature and salinity changes for the major bay segments.
@@
 
-Several datasets with distinct sample designs are available to assess long-term changes in water temperature and salinity in Tampa Bay.
+Several datasets with distinct sample designs are available to assess long-term changes in water temperature and salinity in Tampa Bay (@tbl-dattab).
 These datasets were evaluated individually to assess trends and relationships with seagrass change to provide a weight-of-evidence approach for potential causal relationships driving the recent decline.
@@
 An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions related to long-term (multi-decadal) climate change drivers.
-For comparison to water quality conditions, daily air temperature (Tampa International Airport \[TIA\] National Weather Service site) and precipitation (SWFWMD area-weighted watershed summaries) were used to characterize regional conditions for the most consistent period of record covered by the water quality samples (i.e., 1975 to present for the EPC data).
+For comparison to water quality conditions, daily air temperature (Tampa International Airport [TIA] National Weather Service site) and precipitation (SWFWMD area-weighted watershed summaries) were used to characterize regional conditions for the most consistent period of record covered by the water quality samples (i.e., 1975 to present for the EPC data, @tbl-dattab).
 The rnoaa R package [@chamberlain23] was used to obtain the TIA temperature data.
@@
 Separate models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change.
-These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs.
+These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location).
 GLMs were used for the FIM data to evaluate seagrass percent cover using continuous temperature and salinity as independent variables.
@@
 ```{r}
+#| label: tbl-dattab
+#| tbl-cap: "Summary of datasets used for trend analyses and comparisons with seagrass datasets.
+Temporal and spatial scales are those used for analysis.
+SWFWMD: Southwest Florida Water Management District."
+dattab
+```
+
+```{r}
 #| label: tbl-temptrndtab

### 2024-04-01 manu-draft.qmd
add la and chl to epccmbdat, working on gam scratch

@@
 
-Significant increasing temperature trends were also observed for the FIM dataset from 1996 to present in OTB and from 2004 to present in OTB and HB (@tbl-temptrndtab).
+Significant increasing temperature trends were also observed for the FIM dataset from 1996 to present in OTB and from 2004 to present in HB (@tbl-temptrndtab).
 No significant temperature or salinity trends were observed in the other bay segments for the FIM and PDEM datasets from 1996 to present nor from 2004 to present (Figures S2, S3).

### 2024-04-01 R/dat_proc.R

@@ -953,9 +953,125 @@ thrtrndat <- thralltrndat %>%
   ) %>% 
   rename(yr = trnyr)
 
+##
+# get annual light attenuation for average transect year by bay segment
+
+data(epcdata)
+
+# subset secchi data from epcdata, expand to all days to join with transect dates
+epcsub <- epcdata %>% 
+  select(bay_segment, epchc_station, SampleTime, chla, sd_m) %>% 
+  mutate(
+    date = as.Date(SampleTime)
+  ) %>% 
+  select(-SampleTime) %>% 
+  arrange(bay_segment, epchc_station, date) %>% 
+  filter(year(date) > 1995) %>% # minimum date for ave transecs is 1997-10-24
+  group_by(bay_segment, epchc_station) %>% 
+  complete(date = seq.Date(min(date) ,max(date), by = 'day')) %>% 
+  ungroup()
+
+# mean month of tranect sampling in each bay segment
+transectocc <- anlz_transectocc(transect) 
+trnptssub <- trnpts %>%
+  st_set_geometry(NULL) %>%
+  select(
+    Transect = TRAN_ID,
+    bay_segment
+  ) %>%
+  filter(bay_segment %in% c('OTB', 'HB', 'MTB', 'LTB'))
+trndts <- transectocc %>%
+  ungroup() %>% 
+  filter(Savspecies == 'total') %>%
+  inner_join(trnptssub, by = 'Transect') %>%
+  mutate(
+    year = year(Date)
+  ) %>% 
+  summarise(
+    date = ymd(median(Date)), # use ymd because averaging dates leaves decimals 
+    .by = c('bay_segment', 'year')
+  ) %>% 
+  mutate(trndt = paste0('trndt', 1:n()), .by = 'bay_segment') %>% 
+  select(-year)
+
+# join transect date data with expanded secchi data, then make secchi monthly again
+epcsub <- epcsub %>% 
+  arrange(bay_segment, epchc_station, date) %>% 
+  left_join(trndts, by = c('bay_segment', 'date')) %>% 
+  group_by(bay_segment, epchc_station) %>% 
+  fill(trndt, .direction = 'up') %>% 
+  ungroup() %>% 
+  filter(!is.na(trndt)) %>% # counts occurring after max date in trnds
[…]

### 2024-04-01 R/gam_scratch.R

@@ -1,0 +1,34 @@
+library(mgcv)
+
+data(epccmbdat)
+
+tomod <- epccmbdat %>% 
+  # select(-bbave, -chla) %>% 
+  filter(!bay_segment %in% c('LTB', 'MTB', 'OTB'))
+
+mod <- gam(total ~ s(yr) + s(la) + s(temp) + s(sal), data = tomod, method = 'REML')
+
+draw(mod)
+toplo <- smooth_estimates(mod) %>% 
+  add_confint() %>% 
+  select(.smooth, .estimate, yr, la, temp, sal, .lower_ci, .upper_ci) %>% 
+  pivot_longer(cols = c(yr, la, temp, sal), names_to = 'var', values_to = 'value') %>% 
+  filter(!is.na(value))
+
+ggplot(toplo, aes(x = value, y = .estimate)) + 
+  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2) +
+  geom_line()+ 
+  facet_wrap(~var, scales = 'free')
+
+### repeat down to here but for each bay segment model
+
+# slc <- data_slice(mod, 
+#                     yr = evenly(yr, n = 3), 
+#                     temp = evenly(temp, n = 100)
+#                     )
+# toplo <- fitted_values(mod, data = slc, scale = 'response')
+# 
+# ggplot(toplo, aes(x = temp, y = .fitted)) + 
+#   geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2) +
+#   geom_line()+ 
+#   facet_wrap(~yr)

### 2024-04-03 revision/gam_exploration.qmd
better gam scratch and la from monthly, initial gam exploration html

---
title: "GAM exploration"
format: 
  html:
    code-fold: true
editor: visual

filters:
   - lightbox
lightbox: auto

execute:
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(mgcv)
library(gratia)
library(patchwork)
library(here)
library(broom)
library(knitr)

source(here('R/funcs.R'))

load(file = here('data/epccmbdat.RData'))
load(file = here('data/fimsgtempdat.RData'))
load(file = here('data/pincotemp.RData'))
```

The following is an exploratory analysis of potential seagrass response over time as a function of time, light attenuation, temperature, and salinity. Three different datasets are used, each with different sampling designs. GAMs are used to assess potential non-linear responses using the sum of additive smoothing functions. Lower Tampa Bay is not modeled in any of the datasets given minimal seagrass loss.

## EPC

The EPC data were compared to bay segment averages of seagrass transect frequency occurrence data at the annual scale. Light attenuation was estimated from all monitoring stations using Secchi depth and combined as a bay segment average. Temperature and salinity are expressed as the estimated number of days each when when temperature was above or salinity was below a threshold. The model formula is as follows:

$$
\begin{aligned}
Freq. Occ. \sim s(year, by = bay\_segment) + s(la, by = bay\_segment) + 
\\
s(temp, by = bay\_segment) + s(salinity, by = bay\_segment)
\end{aligned}
$$

```{r}
tomod <- epccmbdat %>% 
  filter(!bay_segment %in% c('LTB'))

##
# single smoothers only, temp, sal

mod <- gam(total ~ s(yr, by = bay_segment) + s(la, by = bay_segment) + s(temp, by = bay_segment) + s(sal, by = bay_segment), data = tomod, method = 'REML')

mod %>% 
  tidy() %>% 
  mutate(
    star = p_ast2(p.value)
  ) %>% 
  kable(digits = 3)
```

```{r}
smmod <- summary(mod)
smtab <- tibble(
  R.sq = smmod$r.sq,
  `Deviance explained` = paste0(round(100*smmod$dev.expl, 0), '%')
)
kable(smtab, digits = 2, align = 'l')
```

```{r}
#| fig-width: 8
#| fig-height: 10
draw(mod, ncol = 3)
```

The same model is created but the "both" metric is used instead of temperature and salinity.

$$
\begin{aligned}
Freq. Occ. \sim s(year, by = bay\_segment) + s(la, by = bay\_segment) + 
\\
s(both, by = bay\_segment)
\end{aligned}
$$

```{r}
##
# single smoothers only, temp, sal

mod <- gam(total ~ s(yr, by = bay_segment) + s(la, by = bay_segment) + s(both, by = bay_segment), data = tomod, method = 'REML')

mod %>% 
  tidy() %>% 
  mutate(
    star = p_ast2(p.value)
  ) %>% 
  kable(digits = 3)
```

```{r}
smmod <- summary(mod)
smtab <- tibble(
  R.sq = smmod$r.sq,
  `Deviance explained` = paste0(round(100*smmod$dev.expl, 0), '%')
)
kable(smtab, digits = 2, align = 'l')
```

```{r}
#| fig-width: 8
#| fig-height: 7.5
draw(mod, ncol = 3)
```

#### Conclusions

There's some weirdness with these models that don't align with expecations.  For example, increasing light attenuation is associated with increasing seagrass frequency occurrence.  Additionally, increases of the stressor metrics are also often associated with increasing seagrass frequency occurence. Perhaps we can attribute this to the different in sampling location between the EPC stations and the seagrass transects, i.e., the stations are in deeper water and may not reflect conditions at the seagrass beds.

### 2024-04-03 R/dat_proc.R

@@ -1371,12 +1371,6 @@ fimsgtempdat <- fimtempdat %>%
     lon = st_coordinates(.)[, 1], 
     lat = st_coordinates(.)[, 2], 
     yr = year(date), 
-    la = case_when(
-      bay_segment %in% "OTB" ~ 1.49 / secchi_m,
-      bay_segment %in% "HB" ~ 1.61 / secchi_m,
-      bay_segment %in% "MTB" ~ 1.49 / secchi_m,
-      bay_segment %in% "LTB" ~ 1.84 / secchi_m
-    ), 
     bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB'))
   ) %>% 
   st_set_geometry(NULL) %>% 
@@ -1403,6 +1397,7 @@ pincotemp <- tmp %>%
     level = Level, 
     temp  = Temp_Water, 
     sal = Salinity, 
+    chla = `Chl-a`,
     secchi_m = Secchi,
     secchi_on_bottom = Secchi_q,
     SAV, 
@@ -1420,7 +1415,6 @@ pincotemp <- tmp %>%
       SAV %in% c('T', 'Y') ~ 1, 
       T ~ NaN
     ), 
-    secchi_m = as.numeric(secchi_m),
     level = factor(level, levels = c('Surface', 'Middle', 'Bottom')), 
     levelint = as.numeric(level), 
     lon = ifelse(sign(as.numeric(lon)) == 1, -1 * as.numeric(lon), as.numeric(lon)), 
@@ -1431,7 +1425,7 @@ pincotemp <- tmp %>%
     )
   ) %>% 
   filter(SAV %in% c(0, 1)) %>% 
-  mutate_at(vars(lat, lon, temp, sal), as.numeric) %>% 
+  mutate_at(vars(lat, lon, temp, sal, secchi_m, chla), as.numeric) %>% 
   filter(lat < 35) # one outlier at 72
 
 # get salinity, temperature, secchi by lowest level (not always bottom)
@@ -1447,7 +1441,7 @@ saltempsec <- pincotemp %>%
 # identify sites w/ and w/o seagrass (sav 1 could also include macro)
 # allsg sites are those with only sg species found (will be zero if any macro found)
 sav <- pincotemp %>% 
-  select(-temp, -sal, -secchi_m, -secchi_on_bottom, -level, -levelint) %>% 
[…]

### 2024-04-03 R/gam_scratch.R

@@ -11,7 +11,7 @@ tomod <- epccmbdat %>%
   filter(!bay_segment %in% c('LTB'))
 
 ##
-# single smoothers only
+# single smoothers only, temp, sal
 
 mod <- gam(total ~ s(yr, by = bay_segment) + s(la, by = bay_segment) + s(temp, by = bay_segment) + s(sal, by = bay_segment), data = tomod, method = 'REML')
 
@@ -40,70 +40,33 @@ ggplot(toplo, aes(x = value, y = .estimate, color = bay_segment, fill = bay_segm
   geom_line()+ 
   facet_wrap(~var, scales = 'free')
 
-## 
-# interaction terms
+##
+# single smoothers only, both
 
-mod <- gam(total ~ te(la, yr, by = bay_segment) + te(temp, yr, by = bay_segment) + te(sal, yr, by = bay_segment), data = tomod, method = 'REML')
+mod <- gam(total ~ s(yr, by = bay_segment) + s(la, by = bay_segment) + s(both, by = bay_segment), data = tomod, method = 'REML')
 
-salslc <- data_slice(mod,
-                    yr = c(2000, 2020),
-                    sal = evenly(sal, n = 100), 
-                    bay_segment = evenly(bay_segment)
-                    )
-toploslc <- smooth_estimates(mod, select = 'te(sal,yr)', data = salslc, partial_match = T) %>% 
-  add_confint() %>% 
-  filter(!((sal < min(tomod[tomod$bay_segment == 'HB', 'sal']) | sal > max(tomod[tomod$bay_segment == 'HB', 'sal'])) & bay_segment == 'HB')) %>%
-  filter(!((sal < min(tomod[tomod$bay_segment == 'OTB', 'sal']) | sal > max(tomod[tomod$bay_segment == 'OTB', 'sal'])) & bay_segment == 'OTB')) %>%
-  filter(!((sal < min(tomod[tomod$bay_segment == 'MTB', 'sal']) | sal > max(tomod[tomod$bay_segment == 'MTB', 'sal'])) & bay_segment == 'MTB')) 
-
-p1 <- ggplot(toploslc, aes(x = sal, y = .estimate, color = factor(yr), fill = factor(yr), group = yr)) + 
-  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2, color = NA) +
-  geom_line()+ 
[…]

### 2024-04-04 revision/gam_exploration.qmd
complete gam exploration doc, rework gam_scratch to match doc

@@
     code-fold: true
+    toc: true
+    code-tools: true
 editor: visual
@@
   message: false
+  eval: true
 ---
@@
 
-The following is an exploratory analysis of potential seagrass response over time as a function of time, light attenuation, temperature, and salinity.
+The following is an exploratory analysis of potential seagrass response related to time, light attenuation, temperature, and salinity.
 Three different datasets are used, each with different sampling designs.
-GAMs are used to assess potential non-linear responses using the sum of additive smoothing functions.
-Lower Tampa Bay is not modeled in any of the datasets given minimal seagrass loss.
+GAMs are used to assess potential non-linear responses in seagrass change using the sum of additive smoothing functions for different predictors.
+Lower Tampa Bay is not modeled in any of the models given minimal seagrass loss.
+Model selection techniques are not considered for simplicity and the results are evaluated based on the percent deviance explained and "partial effects" of each smoother, where the smoother is a proxy for the effect of the variable on the response while considering a more comprehensive set of predictors.
+These models were developed in response to reviewer comments on including additional predictors important for characterizing change, namely time and light attenuation.
+
+## Quick summary
+
+All models had much more explained variance (32-94%) compared to earlier linear regression models.
+The EPC models returned confusing results, which could be from differences in sampling location for water quality (mid-bay) vs seagrass transect sampling.
+The FIM models for nearshore provide the most compelling results, particularly for temperature in HB and salinity in MTB.
+There are also some notable differences by season in the FIM models.
+The PDEM models are inconsistent, although a significant interaction term for temperature and season at stratum E4 (most southern portion of OTB) suggests increasing temperature in the winter is associated with lower seagrass frequency occurrence.
+In general, trends in OTB are not explained well by temperature or salinity, whereas HB and MTB follow expectations for some of the models.
 
@@
 The EPC data were compared to bay segment averages of seagrass transect frequency occurrence data at the annual scale.
-Light attenuation was estimated from all monitoring stations using Secchi depth and combined as a bay segment average.
-Temperature and salinity are expressed as the estimated number of days each when when temperature was above or salinity was below a threshold.
+Light attenuation was estimated from all monitoring stations using Secchi depth and combined as a bay segment average, as is done for TBEP reporting to FDEP.
+Temperature and salinity are expressed as the estimated number of days each year when when temperature was above or salinity was below a threshold, where the expectation is that total frequency occurrence will decrease as both increase.
 The model formula is as follows:
@@
 \begin{aligned}
-Freq. Occ. \sim s(year, by = bay\_segment) + s(la, by = bay\_segment) + 
+Freq. Occ. \sim s(year, by = bay\_segment, bs = 'tp') + s(la, by = bay\_segment, bs = 'tp') + 
 \\
-s(temp, by = bay\_segment) + s(salinity, by = bay\_segment)
+s(temp, by = bay\_segment, bs = 'tp') + s(salinity, by = bay\_segment, bs = 'tp')
 \end{aligned}
@@
 mod <- gam(total ~ s(yr, by = bay_segment) + s(la, by = bay_segment) + s(temp, by = bay_segment) + s(sal, by = bay_segment), data = tomod, method = 'REML')
+```
 
+Smoother summary:
+
+```{r}
 mod %>% 
@@
 ```
+
+Model fit summary:
 
@@
 
-The same model is created but the "both" metric is used instead of temperature and salinity.
+The same model is created but the "both" metric is used instead of temperature and salinity, which is the number of days each year when both temperature was above or salinity was below thresholds.
+The expectation is that total frequency occurrence will decrease as the metric increases, either linearly (additive effect) or non-linearly (multiplicative effect).
+The model formula is as follows:
 
@@
 \begin{aligned}
-Freq. Occ. \sim s(year, by = bay\_segment) + s(la, by = bay\_segment) + 
+Freq. Occ. \sim s(year, by = bay\_segment, bs = 'tp') + s(la, by = bay\_segment, bs = 'tp') + 
 \\
-s(both, by = bay\_segment)
+s(both, by = bay\_segment, bs = 'tp')
 \end{aligned}
@@
 mod <- gam(total ~ s(yr, by = bay_segment) + s(la, by = bay_segment) + s(both, by = bay_segment), data = tomod, method = 'REML')
+```
 
+Smoother summary:
+
+```{r}
 mod %>% 
@@
 
-#### Conclusions
+### Conclusions
 
-There's some weirdness with these models that don't align with expecations.
-For example, increasing light attenuation is associated with increasing seagrass frequency occurrence.
-Additionally, increases of the stressor metrics are also often associated with increasing seagrass frequency occurence.
-Perhaps we can attribute this to the different in sampling location between the EPC stations and the seagrass transects, i.e., the stations are in deeper water and may not reflect conditions at the seagrass beds.
+There's some weirdness with these models that don't align with expectations.
+For example, increasing light attenuation is associated with increasing seagrass frequency occurrence (MTB).
+Additionally, increases of the stressor metrics are also often associated with increasing seagrass frequency occurrence.
+These results could be from a discrepancy between sampling location of the EPC stations and the seagrass transects, i.e., the stations are in deeper water and may not reflect conditions at the seagrass beds in more shallow waters.
+The association of the salinity metric with seagrasses in MTB is as expected, although the smoother is not significant.
+However, this is consistent with the FIM models.
+
+## FIM
+
+The FIM models are different from the EPC models in several ways:
+
+* Seagrass and stressor data are collected at the same locations possibly allowing for a more consistent comparison
+* The stressor data (light attenuation, temperature, salinity) are observed data, so the expectation of how seagrass will response varies from the EPC data and by stressor, i.e., seagrass will decrease with temperature and light attenuation, but increase with salinity (fresh conditions are more stressful)
+* Monthly observations are available, allowing a seasonal component in the model. 
+* Secchi data is used for light attenuation (as for EPC), but more than half of the observations are on the bottom.
+Attempts were made to calculate monthly medians using Kaplan-Meier survival curves, but data were insufficient to do so.
+
+The FIM data are averaged at the annual and monthly time steps.
+The model is more complex than the previous and includes smoothers for year, month, light attenuation, temperature, and salinity and interaction smoothers (`ti`) for the predictions with month, allowing the effects of the predictors to vary seasonally.
+Different smoothers are also constructed for each bay segment.
+The model formula is as follows:
+
+$$
+\begin{aligned}
+Coverage \sim s(year, by = bay\_segment, bs = 'tp') + s(la, by = bay\_segment, bs = 'tp') + 
+\\
+s(temp, by = bay\_segment, bs = 'tp') + s(salinity, by = bay\_segment, bs = 'tp') +
+\\
+s(month, bs = 'cc', by = bay\_segment) + 
+\\
+ti(year, month, bs = c('tp', 'cc'), by = bay\_segment) + 
+\\
+ti(la, month, bs = c('tp', 'cc'), by = bay\_segment) + 
+\\
+ti(temp, month, bs = c('tp', 'cc'), by = bay\_segment) +
+\\
+ti(salinity, month, bs = c('tp', 'cc'), by = bay\_segment)
+\end{aligned}
+$$
+
+```{r}
+tomod <- fimsgtempdat %>% 
+  filter(!bay_segment %in% c('LTB')) %>% 
+  select(date, sgcov, sgpres, yr, temp, sal, secchi_m, secchi_on_bottom, bay_segment) %>% 
+  na.omit() %>%
+  mutate(mo = month(date)) %>% 
+  # filter(month(date) %in% c(6:9)) %>%
+  summarise(
+    sgcov = mean(sgcov),
+    sgpres = mean(sgpres),
+    temp = mean(temp),
+    sal = mean(sal),
+    secchi_m = mean(secchi_m),
+    .by = c(bay_segment, yr, mo)
+  ) %>% 
+  mutate(
+    la = dplyr::case_when(
+      bay_segment %in% "OTB" ~ 1.49 / secchi_m,
+      bay_segment %in% "HB" ~ 1.61 / secchi_m,
+      bay_segment %in% "MTB" ~ 1.49 / secchi_m,
+      bay_segment %in% "LTB" ~ 1.84 / secchi_m
+    )
+  )
+
+mod <- gam(sgcov ~ s(yr, by = bay_segment) + s(la, by = bay_segment) + s(temp, by = bay_segment) + s(sal, by = bay_segment) + s(mo, bs = 'cc', by = bay_segment) + ti(yr, mo, bs = c('tp', 'cc'), by = bay_segment) + ti(la, mo, bs = c('tp', 'cc'), by = bay_segment) + ti(temp, mo, bs = c('tp', 'cc'), by = bay_segment) + ti(sal, mo, bs = c('tp', 'cc'), by = bay_segment), knots = list(doy = c(1, 12)), data = tomod, method = 'REML')
+```
+
+Smoother summary:
+
+```{r}
+mod %>% 
+  tidy() %>% 
+  mutate(
+    star = p_ast2(p.value)
+  ) %>% 
+  kable(digits = 3)
+```
+
+Model fit summary:
+
+```{r}
+smmod <- summary(mod)
+smtab <- tibble(
+  R.sq = smmod$r.sq,
+  `Deviance explained` = paste0(round(100*smmod$dev.expl, 0), '%')
+)
+kable(smtab, digits = 2, align = 'l')
+```
+
+Select smoothers:
+
+```{r}
+#| fig-width: 8
+#| fig-height: 3
+draw(mod, select = 's(temp)', ncol = 3, partial_match = TRUE)
+draw(mod, select = 's(sal)', ncol = 3, partial_match = TRUE)
+```
+```{r}
+#| fig-width: 10
+#| fig-height: 4
+draw(mod, select = 'ti(temp,mo)', ncol = 3, partial_match = TRUE)
+draw(mod, select = 'ti(sal,mo)', ncol = 3, partial_match = TRUE)
+```
+
+Model slices for interaction terms:
+
+```{r}
+#| fig-width: 7
+#| fig-height: 3
+tempslc <- smooth_estimates(mod, select = 'ti(temp,mo)', partial_match = T) %>% 
+  add_confint() %>% 
+  filter(mo %in% c(1, 7)) %>% 
+  filter(bay_segment %in% c('HB', 'MTB'))
+ggplot(tempslc, aes(x = temp, y = .estimate, color = factor(mo), fill = factor(mo), group = mo)) +
+  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2, color = NA) +
+  geom_line() +
+  facet_wrap(~bay_segment, scales = 'free_y') + 
+  labs(
+    x = 'Temperature',
+    y = 'Partial effect',
+    color = 'Month',
+    fill = 'Month'
+  )
+```
+```{r}
+#| fig-width: 4
+#| fig-height: 3
+salslc <- smooth_estimates(mod, select = 'ti(sal,mo)', partial_match = T) %>% 
+  add_confint() %>% 
+  filter(mo %in% c(3, 9)) %>% 
+  filter(bay_segment %in% c('MTB'))
+ggplot(salslc, aes(x = sal, y = .estimate, color = factor(mo), fill = factor(mo), group = mo)) +
+  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2, color = NA) +
+  geom_line() +
+  facet_wrap(~bay_segment, scales = 'free_y') + 
+  labs(
+    x = 'Salinity',
+    y = 'Partial effect',
+    color = 'Month',
+    fill = 'Month'
+  )
+```
+
+### Conclusions
+
+The FIM models are more promising as expected relationships with temperature and salinity are observed, particularly for MTB and HB.
+There are also interesting interactions by month suggesting the effects vary seasonally, e.g., the temperature effect in HB seems to be driven by July temperatures and the salinity effect in MTB seems to be pronounced in the dry season (March).
+
+## PDEM
+
+The PDEM models are similar to the FIM models except only OTB is evaluated.
+Effects are also evaluated differently for four sampling strata in OTB (E1-E4, somewhat north to south), allowing for a more precise spatial assessment of seagrass relationships.
+The data are averaged for each year, month, and stratum.
+Compared to the FIM data, fewer Secchi measurements were on the bottom given the broader depth coverage (see Figure 1 in the manuscript).
+Additionally, PDEM only records seagrass presence/absence, but this was converted to a strata-wide frequency occurrence metric as the number of points with seagrass divided by the total points in each stratum.
+
+The PDEM model is the same as for FIM, except separate smoothers are estimated for each stratum instead of bay segment. The model formula is as follows:
+
+$$
+\begin{aligned}
+Freq. Occ. \sim s(year, by = stratum, bs = 'tp') + s(la, by = stratum, bs = 'tp') + 
+\\
+s(temp, by = stratum, bs = 'tp') + s(salinity, by = stratum, bs = 'tp') +
+\\
+s(month, bs = 'cc', by = stratum) + 
+\\
+ti(year, month, bs = c('tp', 'cc'), by = stratum) + 
+\\
+ti(la, month, bs = c('tp', 'cc'), by = stratum) + 
+\\
+ti(temp, month, bs = c('tp', 'cc'), by = stratum) +
+\\
+ti(salinity, month, bs = c('tp', 'cc'), by = stratum)
+\end{aligned}
+$$
+```{r}
+
+tomod <- pincotemp %>% 
+  filter(!bay_segment %in% c('LTB')) %>% 
+  select(date, allsg, yr, site, temp, sal, secchi_m, secchi_on_bottom, bay_segment) %>% 
+  na.omit() %>%
+  mutate(
+    mo = month(date),
+    site = factor(site)
+    ) %>% 
+  summarise(
+    allsg = sum(allsg) / length(allsg),
+    temp = mean(temp),
+    sal = mean(sal),
+    secchi_m = mean(secchi_m),
+    .by = c(yr, mo, site)
+  ) %>% 
+  mutate(
+    la = 1.49 / secchi_m
+  )
+
+mod <- gam(allsg ~ s(yr, by = site) + s(la, by = site) + s(temp, by = site) + s(sal, by = site) + s(mo, bs = 'cc', by = site) + ti(yr, mo, bs = c('tp', 'cc'), by = site) + ti(la, mo, bs = c('tp', 'cc'), by = site) + ti(temp, mo, bs = c('tp', 'cc'), by = site) + ti(sal, mo, bs = c('tp', 'cc'), by = site), knots = list(doy = c(1, 12)), data = tomod, method = 'REML')
+```
+
+Smoother summary:
+
+```{r}
+mod %>% 
+  tidy() %>% 
+  mutate(
+    star = p_ast2(p.value)
+  ) %>% 
+  kable(digits = 3)
+```
+
+Model fit summary:
+
+```{r}
+smmod <- summary(mod)
+smtab <- tibble(
+  R.sq = smmod$r.sq,
+  `Deviance explained` = paste0(round(100*smmod$dev.expl, 0), '%')
+)
+kable(smtab, digits = 2, align = 'l')
+```
+
+Select smoothers:
+
+```{r}
+#| fig-width: 8
+#| fig-height: 3
+draw(mod, select = 's(temp)', ncol = 4, partial_match = TRUE)
+draw(mod, select = 's(sal)', ncol = 4, partial_match = TRUE)
+```
+```{r}
+#| fig-width: 10
+#| fig-height: 4
+draw(mod, select = 'ti(temp,mo)', ncol = 4, partial_match = TRUE)
+draw(mod, select = 'ti(sal,mo)', ncol = 4, partial_match = TRUE)
+```
+
+Model slices for interaction terms:
+
+```{r}
+#| fig-width: 7
+#| fig-height: 3
+tempslc <- smooth_estimates(mod, select = 'ti(temp,mo)', partial_match = T) %>% 
+  add_confint() %>% 
+  filter(mo %in% c(1, 7)) %>% 
+  filter(site %in% c('E1', 'E4'))
+ggplot(tempslc, aes(x = temp, y = .estimate, color = factor(mo), fill = factor(mo), group = mo)) +
+  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2, color = NA) +
+  geom_line() +
+  facet_wrap(~site, scales = 'free_y') + 
+  labs(
+    x = 'Temperature',
+    y = 'Partial effect',
+    color = 'Month',
+    fill = 'Month'
+  )
+```
+```{r}
+#| fig-width: 4
+#| fig-height: 3
+salslc <- smooth_estimates(mod, select = 'ti(sal,mo)', partial_match = T) %>% 
+  add_confint() %>% 
+  filter(mo %in% c(3, 9)) %>% 
+  filter(site %in% c('E4'))
+ggplot(salslc, aes(x = sal, y = .estimate, color = factor(mo), fill = factor(mo), group = mo)) +
+  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2, color = NA) +
+  geom_line() +
+  facet_wrap(~site, scales = 'free_y') + 
+  labs(
+    x = 'Salinity',
+    y = 'Partial effect',
+    color = 'Month',
+    fill = 'Month'
+  )
+```
+
+### Conclusions
+
+None of the main effects for temperature or salinity are significant, although Some of the interaction terms by stratum are informative.
+The model slices for salinity at stratum E4 and temperature at E1 are unclear, but temperature at E4 is associated with lower frequency occurrence of seagrass in January.

### 2024-04-04 R/gam_scratch.R

@@ -2,75 +2,58 @@ library(tidyverse)
 library(mgcv)
 library(gratia)
 library(patchwork)
+library(here)
+library(broom)
+library(knitr)
 
-# EPC eval ------------------------------------------------------------------------------------
+source(here('R/funcs.R'))
 
-data(epccmbdat)
+load(file = here('data/epccmbdat.RData'))
+load(file = here('data/fimsgtempdat.RData'))
+load(file = here('data/pincotemp.RData'))
+
+# EPC -----------------------------------------------------------------------------------------
 
 tomod <- epccmbdat %>% 
   filter(!bay_segment %in% c('LTB'))
 
-##
-# single smoothers only, temp, sal
-
 mod <- gam(total ~ s(yr, by = bay_segment) + s(la, by = bay_segment) + s(temp, by = bay_segment) + s(sal, by = bay_segment), data = tomod, method = 'REML')
 
-toplo <- smooth_estimates(mod) %>% 
-  add_confint() %>% 
-  # add_residuals(data = ., model = mod)
-  select(.smooth, .estimate, yr, la, temp, sal, .lower_ci, .upper_ci) %>% 
-  pivot_longer(cols = c(yr, la, temp, sal), names_to = 'var', values_to = 'value') %>% 
-  filter(!is.na(value)) %>% 
+mod %>% 
+  tidy() %>% 
   mutate(
-    bay_segment = gsub('^.*segment', '', .smooth)
+    star = p_ast2(p.value)
   ) %>% 
-  filter(!((value < min(tomod[tomod$bay_segment == 'HB', 'sal']) | value > max(tomod[tomod$bay_segment == 'HB', 'sal'])) & bay_segment == 'HB' & var == 'sal')) %>%
-  filter(!((value < min(tomod[tomod$bay_segment == 'HB', 'temp']) | value > max(tomod[tomod$bay_segment == 'HB', 'temp'])) & bay_segment == 'HB' & var == 'temp')) %>%
-  filter(!((value < min(tomod[tomod$bay_segment == 'HB', 'la']) | value > max(tomod[tomod$bay_segment == 'HB', 'la'])) & bay_segment == 'HB' & var == 'la')) %>%
[…]

### 2024-04-23 manu-draft.qmd
revision to seagrass change description

@@
 Coverage in Middle Tampa Bay decreased by 20% (1,926 acres/779 ha loss), whereas coverage in Lower Tampa Bay has remained stable, with only a 2% loss which is close to the mapping error.
+As such, trajectories of recovery and decline have varied by bay segment in magnitude and timing of the change, although a consistent decline baywide has been observed since 2016.
+Change by species has been most notable for *Halodule wrightii* (shoal grass) in OTB and HB, whereas some losses have been observed for *Thalassia testudinum* (turtle grass) in LTB, likely related to dominance of each species across the salinity gradient and proximity to hydrologic inputs [@lewis1985].
 
@@
 A fundamental challenge describing niche space is identifying the boundaries for optimal conditions.
-In Tampa Bay, three dominant seagrass species occur: *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinum* (turtle grass) [@lewis1985; @phillips88].
+In Tampa Bay, three dominant seagrass species occur: *H.
+wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinum* (turtle grass) [@lewis1985; @phillips88].
 Other less common species include *Ruppia maritima* (widgeon grass) and *Halophila engelmanii* (star grass), where the former is often mapped during wet years in the upper bay segments from the aerial surveys.
@@
 All models excluded Lower Tampa Bay because of minimal seagrass change over time.
+Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model.
 
@@
 For example, the parasitic slime mold *Labryinthula* spp.
-that causes seagrass wasting disease has been known to infect *Thalassia testudinium* in Tampa Bay [@blakesley2001], although it is unclear if these infections have had large-scale, population-level effects.
+that causes seagrass wasting disease has been known to infect *Thalassia testudinum* in Tampa Bay [@blakesley2001], although it is unclear if these infections have had large-scale, population-level effects.
 Existing research has primarily focused on describing spatial patterns, past die-off events, or immunology of these pathogens [@robblee1991; @duffin2021].
@@
 In Florida Bay, for example, widespread decline of *T.
-testudinium* has been attributed to altered hydrology and drought-induced hypersaline conditions, and sea level rise is expected to further modify salinity dynamics in the region [@hall2016].
+testudinum* has been attributed to altered hydrology and drought-induced hypersaline conditions, and sea level rise is expected to further modify salinity dynamics in the region [@hall2016].
 @dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters.

### 2024-04-30 manu-draft.qmd
more revisions

@@
 Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.04 $^\circ$C per year and salinity decreasing by 0.04 - 0.06 ppt per year, translating to an increase of 1.3 to 1.7 $^\circ$C and a decrease of 1.6 to 2.6 ppt over the last fifty years.
+Additionally, the number of days when temperature was above 30 $^\circ$C or salinity was below 25 ppt has increased on average across all bay segments by 48 and 37 days, respectively, since 1975.
 These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay.
@@
 These results should compel resource managers to consider the likelihood that reduced resilience of estuarine resources due to shifting ecological baselines driven by additional climate change drivers will complicate long-standing management paradigms.
-While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be confounded by climate change drivers and warrant additional, complementary interventions to maintain ecosystem health into the future.
+While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be confounded by climate change drivers and warrant additional, complementary interventions and continuous monitoring data to support ecosystem health into the future.
 
@@
 Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment from 1998 to 2022.
-The stressor metrics were compared to frequency occurrence (all species) each year by bay segment.
+Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the annual summaries.
+As such, the annual stressor metrics were compared to frequency occurrence (all species) each year by bay segment.
 Generalized Linear regression Models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre 2016, decline post 2016).
@@
 All models excluded Lower Tampa Bay because of minimal seagrass change over time.
-Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model.
+Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model and that the majority of seagrass loss in recent years was dominated by *H.
+wrightii* in the upper bay segments (Figure [-@fig-seagrasschg]b).
 
@@
 Similarly, total precipitation during the rainy season has increased by `r prpslo` mm per year, although the model was not significant (`r prppvl`).
-Removing September from the rainy season showed a significantly increasing trend of `r prpslo2` mm per year (`r prppvl2`, $R^2$ = `r prprsq2`; September alone did not have a significant trend, `r seppvl`).
+Removing September from the rainy season showed a significantly increasing trend of `r prpslo2` mm per year (`r prppvl2`, $R^2$ = `r prprsq2`; September alone did not have a significant trend, `r seppvl`), suggesting precipitation increases were driven by the earlier months (June - August).
 Using this model, mean precipitation in 1975 was `r prpstr2` mm, whereas current mean precipitation in 2022 was `r prpend2` mm, showing an overall increase in the period of record of `r prpdif2` mm.
@@
 The estimated slopes for the number of days when temperature was above 30 $^\circ$C varied from less than 1 day per year in LTB to 1.1 days per year in MTB.
-Likewise, the mean number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied between bay segments, although most had an approximate increase of 50 days each year for the period of record (@tbl-mixdaytab, Table S5, see Table S6 for 1998 to 2022).
+Likewise, the mean number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record were similar between bay segments, with an average increase of 48 days across the bay segments for the period of record (@tbl-mixdaytab, Table S5, see Table S6 for 1998 to 2022).
 The increase in the number days each year when temperatures were above 29 or 31 $^\circ$C was similar.
@@
 Some salinity models had significantly decreasing trends, although the total number of days at the beginning and end of the period of record were neglible (Table S5).
-Overall, OTB showed the largest increase in the number of days each year when salinity was below a theshold, particularly for 25 ppt, where the change was 86 days per year from 1975 to 2022 (130 days to 216 days per year, @tbl-mixdaytab, Table S5, see Table S6 for 1998 to 2022).
+Overall, OTB showed the largest increase in the number of days each year when salinity was below a threshold, particularly for 25 ppt, where the change was 86 days per year from 1975 to 2022 (130 days to 216 days per year, @tbl-mixdaytab, Table S5, see Table S6 for 1998 to 2022).
+Across all bay segments, the average increase in the number of days salinity was below 25 ppt from 1975 to 2022 was 36 days, although a distinct gradient towards the mouth of the bay was observed.
 
@@
 Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values, conditions in Tampa Bay in recent years are generally within the ecological niche for seagrasses.
-This does not suggest that these factors are unimportant, both currently and in the future.
+This is especially true for *H.
+wrightii* that had the greatest changes over the period of record and is tolerant of a wide range of salinity.
+However, this does not suggest that these factors are unimportant, both currently and in the future.
 Extreme temperature or precipitation events acting individually or in combination are likely captured by the trends in stressor metrics using these lower thresholds, i.e., an increase in a bay segment median number of days also suggests extremes are increasing given the variation around these summary metrics (@fig-mixeff).
@@
 These analyses should be supported by additional data collection efforts, particularly high-resolution continuous monitoring data that provide a more precise assessment of diurnal stress across multiple time-scales.
+Ongoing work in OTB using continuous data loggers in shallow areas will provide insights into short-term diurnal changes as potential acute temperature stress on seagrass (see <https://tbep-tech.github.io/otb-temp/tempeval>).
 Morpohological or physiological measurements at the individual level could also provide early indications of heat and osmotic stress.

### 2024-05-03 manu-draft.qmd
see gam scratch, these models tell a good story

@@
 ### Links to seagrass
+
+```{r}
+# secchi on bottom is na issue
+# also averaging all models by year in the bay segment for comparable sample size, fim/pdem limited to july - nov to minimize seasonal effect and when transect surveys completed for comparability
+# < 17% for epc, much higher for fim/pdem so la not included, but can show la relatively stable in a plot
+# epc is bonk becuase of disconnect between sites
+# fim is good, makes sense, but otb is weird
+# pdem temp makes sense for otb
+```
 

### 2024-05-03 R/gam_scratch.R

@@ -17,68 +17,7 @@ load(file = here('data/pincotemp.RData'))
 tomod <- epccmbdat %>% 
   filter(!bay_segment %in% c('LTB'))
 
-mod <- gam(total ~ s(yr, by = bay_segment) + s(la, by = bay_segment) + s(temp, by = bay_segment) + s(sal, by = bay_segment), data = tomod, method = 'REML')
-
-mod %>% 
-  tidy() %>% 
-  mutate(
-    star = p_ast2(p.value)
-  ) %>% 
-  kable(digits = 3)
-
-smmod <- summary(mod)
-smtab <- tibble(
-  R.sq = smmod$r.sq,
-  `Deviance explained` = paste0(round(100*smmod$dev.expl, 0), '%')
-)
-kable(smtab, digits = 2, align = 'l')
-
-draw(mod, ncol = 3)
-
-mod <- gam(total ~ s(yr, by = bay_segment) + s(la, by = bay_segment) + s(both, by = bay_segment), data = tomod, method = 'REML')
-
-mod %>% 
-  tidy() %>% 
-  mutate(
-    star = p_ast2(p.value)
-  ) %>% 
-  kable(digits = 3)
-
-smmod <- summary(mod)
-smtab <- tibble(
-  R.sq = smmod$r.sq,
-  `Deviance explained` = paste0(round(100*smmod$dev.expl, 0), '%')
-)
-kable(smtab, digits = 2, align = 'l')
-
-draw(mod, ncol = 3)
-
-# FIM -----------------------------------------------------------------------------------------
-
-tomod <- fimsgtempdat %>% 
-  filter(!bay_segment %in% c('LTB')) %>% 
-  select(date, sgcov, sgpres, yr, temp, sal, secchi_m, secchi_on_bottom, bay_segment) %>% 
-  na.omit() %>%
-  mutate(mo = month(date)) %>% 
-  # filter(month(date) %in% c(6:9)) %>%
-  summarise(
-    sgcov = mean(sgcov),
-    sgpres = mean(sgpres),
-    temp = mean(temp),
-    sal = mean(sal),
-    secchi_m = mean(secchi_m),
-    .by = c(bay_segment, yr, mo)
-  ) %>% 
-  mutate(
-    la = dplyr::case_when(
-      bay_segment %in% "OTB" ~ 1.49 / secchi_m,
-      bay_segment %in% "HB" ~ 1.61 / secchi_m,
[…]

### 2024-05-03 manu-draft.qmd
light attenuation added to fig 2, caption updated

@@
 
-![Seagrass changes over time in Tampa Bay for (a) areal coverage 1988 - 2022 from mapping and (b) frequency occurrence of major species 1998 - 2022 from annual transect monitoring.
+![Seagrass changes over time in Tampa Bay for (a) areal coverage 1988 - 2022 from mapping, (b) frequency occurrence of major species 1998 - 2022 from annual transect monitoring, and (c) mean annual light attenuation from 1970 to 2022.
 Changes are shown for major bay segments.
-Red lines in (a) show approximate capacity of seagrass coverage based on the baywide target of 40,000 acres.
+Red lines in (a) show approximate capacity of seagrass coverage based on the baywide target of 40,000 acres and red lines in (c) show the approximate linear trend for the period of record covered by the transect data in (b).
+The dashed line in (c) shows the light attenuation threshold for each bay segment identified as supportive for seagrass growth.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/seagrasschg.png){#fig-seagrasschg}

### 2024-05-03 revision/gam_exploration.qmd
light attenuation added to fig 2, caption updated

@@
 Model selection techniques are not considered for simplicity and the results are evaluated based on the percent deviance explained and "partial effects" of each smoother, where the smoother is a proxy for the effect of the variable on the response while considering a more comprehensive set of predictors.
-These models were developed in response to reviewer comments on including additional predictors important for characterizing change, namely time and light attenuation.
 
@@
 
-* Seagrass and stressor data are collected at the same locations possibly allowing for a more consistent comparison
-* The stressor data (light attenuation, temperature, salinity) are observed data, so the expectation of how seagrass will response varies from the EPC data and by stressor, i.e., seagrass will decrease with temperature and light attenuation, but increase with salinity (fresh conditions are more stressful)
-* Monthly observations are available, allowing a seasonal component in the model. 
-* Secchi data is used for light attenuation (as for EPC), but more than half of the observations are on the bottom.
+-   Seagrass and stressor data are collected at the same locations possibly allowing for a more consistent comparison
+-   The stressor data (light attenuation, temperature, salinity) are observed data, so the expectation of how seagrass will response varies from the EPC data and by stressor, i.e., seagrass will decrease with temperature and light attenuation, but increase with salinity (fresh conditions are more stressful)
+-   Monthly observations are available, allowing a seasonal component in the model.
+-   Secchi data is used for light attenuation (as for EPC), but more than half of the observations are on the bottom.
 Attempts were made to calculate monthly medians using Kaplan-Meier survival curves, but data were insufficient to do so.
@@
 ```
+
 ```{r}
@@
 $$
+
 ```{r}

### 2024-05-03 R/figs.R

@@ -258,9 +258,46 @@ p2 <- ggplot(toplo2, aes(x = yr, y = foest, color = Savspecies)) +
     caption = expression(italic('Source: Interagency Seagrass Monitoring Program'))
   )
 
-p <- p1 + p2 + plot_layout(ncol = 1)
+dat <- anlz_avedat(epcdata)$ann %>% 
+  filter(var == 'mean_la') %>% 
+  filter(yr < 2023) %>% 
+  mutate(
+    bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB'))
+  )
 
-png(here('figs/seagrasschg.png'), height = 6, width = 9, family = 'serif', units = 'in', res = 300)
+lns <- targets %>% 
+  filter(bay_segment %in% c('OTB', 'HB', 'MTB', 'LTB')) %>%
+  mutate(bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB'))) %>% 
+  select(bay_segment, la_thresh)
+
+p3 <- ggplot(dat, aes(x = yr, y = val)) + 
+  geom_line() +
+  geom_point() +
+  geom_smooth(data = filter(dat, yr >= 1998), method = 'lm', formula = y ~ x, se = F, color = 'tomato1') +
+  geom_hline(data = lns, aes(yintercept = la_thresh), color = 'blue', linetype = 'dashed') +
+  facet_wrap(~bay_segment, ncol = 4) + 
+  theme_bw() + 
+  theme(
+    legend.position = 'top',
+    strip.text = element_text(size = 11),
+    strip.background = element_blank(), 
+    panel.grid.minor = element_blank(),
+    panel.grid.major.x = element_blank(),
+    legend.text = element_text(size = 11),
+    axis.text.y = element_text(colour = 'black', size = 9),
+    axis.text.x = element_text(colour = 'black', angle = 60, size = 9, hjust = 1),
+  ) + 
+  labs(
+    y = expression("Light Att. (m " ^-1 *")"),
+    x = NULL, 
+    color = NULL,
+    title = '(c) Mean annual light attenuation bay segment',
+    caption = expression(italic('Source: Environmental Protection Commission of Hillsborough County'))
+  ) 
+
+p <- p1 + p2 + p3 + plot_layout(ncol = 1)
+
+png(here('figs/seagrasschg.png'), height = 9, width = 9, family = 'serif', units = 'in', res = 300)
 print(p)
 dev.off()
 

### 2024-05-06 manu-draft.qmd
added new methods text for gams, object for secchi on bottom

@@
 load(file = here('data/sgmods.RData'))
+load(file = here('data/onb.RData'))
 ```
@@
 This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay estuary over the last fifty years.
-Three datasets of varying sampling designs and temporal coverage were used to assess the primary hypothesis that Tampa Bay is trending towards hotter and fresher conditions that are likely stressing seagrasses beyond their optimal tolerance ranges, particularly in upper bay segments where marginal light environments exist.
+Three datasets of varying sampling designs and temporal coverage were used to assess the primary hypothesis that Tampa Bay is trending towards hotter and fresher conditions that are likely stressing seagrasses beyond their optimal tolerance ranges, particularly in upper bay segments that are more shallow and receive a majority of hydrologic inflow.
 This hypothesis was generated from preliminary assessments of datasets used herein and discussions with the regional scientific and management community.
@@
 All bay segments are relatively shallow, with a baywide mean depth of approximately 3 m.
-Light penetration typically reaches bottom habitats under current conditions, although seagrasses were historically limited by high phytoplankton production that affected light environments [@greening2014; @johansson15].
+Light penetration typically reaches bottom habitats under current conditions (Figure [-@fig-seagrasschg]c), although seagrasses were historically limited by high phytoplankton production that affected light environments [@greening2014; @johansson15].
 
@@
 Since that time, the greatest areal coverage expansions were observed in the upper bay segments (OTB, HB, and MTB; Figure [-@fig-seagrasschg]a).
-External nitrogen load estimates decreased by about 2/3 by the 1980s from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed [@greening2014; @beck2019].
+Light environments have improved (Figure [-@fig-seagrasschg]c) through reduction of external nitrogen loadings by about 2/3 by the 1980s from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed \[@greening2014; @beck2019\].
 
-From 2016 to present, dramatic seagrass loss has been observed in Tampa Bay, despite nutrient loading and light attenuation remaining relatively stable (Figure [-@fig-seagrasschg]a).
+From 2016 to present, dramatic seagrass loss has been observed in Tampa Bay, despite light environments remaining supportive of growth (Figure [-@fig-seagrasschg]c).
 Total cover in Tampa Bay has decreased by 28% (11,518 acres/4,661 ha) from the 2016 peak to a total baywide coverage of 30,137 acres (12,196 ha) in 2022.
@@
 Based on the results of the trend tests, we hypothesized that seagrasses are likely stressed by both high temperature and low salinity (bottom right, @fig-concept).
-Although the fundamental niche space can be defined in multiple dimensions for many parameters, we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have been relatively stable in recent years [@tbep0323].
+Although the fundamental niche space can be defined in multiple dimensions for many parameters, we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have been sufficient for growth in recent years (Figure [-@fig-seagrasschg]c).
 
@@
 As such, the annual stressor metrics were compared to frequency occurrence (all species) each year by bay segment.
-Generalized Linear regression Models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre 2016, decline post 2016).
+GAMs were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, or both.
+Additional predictors included year and light attenuation as estimated from Secchi depth [@tbep0696].
+For 1998 to 2022, `r onb$epc` of the Secchi observations were recorded on the bottom.
+A single smooth term was used for each predictor using a thing plate regression spline, including a tensor product smoother term that evaluated the potential interacting effects of each predictor with year.
 Two models were evaluated, one with the temperature and salinity metrics together and another with the both metric.
-Interactions were included between temperature, salinity, and time period or the both metric and time period, depending on the model.
-Bay segment was included as a main effect without interactions.
-The most parsimonious models were identified by forward and backward selection of the independent variables and comparison with AIC values [@sakamoto1986; @venables2002b].
 All models excluded Lower Tampa Bay because of minimal seagrass change over time.
@@
 These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location).
-GLMs were used for the FIM data to evaluate seagrass percent cover using continuous temperature and salinity as independent variables.
-Interactions with time period and a main effect for bay segment were also included as above.
-GLMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data.
-The seagrass response variable was presence/absence (i.e., a binomial distribution was used) to model the probability of observing seagrasses.
-Model input data were further subset to include only months from July to November to describe seagrasses during the growing season and to reduce potential seasonal effects.
+GAMs were used for the FIM data to evaluate seagrass percent cover using continuous temperature and salinity as independent variables.
+Interactions with year for each predictor variable and separate smoothed terms for each bay segment were also included as above.
+GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data.
+Neither the FIM nor PDEM models used light attenuation as a predictor variable given that a large percentage of Secchi observations were measured on the bottom (`r onb$fim` and `r onb$pinco`, respectively), providing further support that light environments have not been limiting for seagrasses.
+Model input data were further subset to include only months from July to October to describe seagrasses during the growing season and to reduce potential seasonal effects.
+Lastly, all data were averaged annually for the monthly subsets for comparability of sample size (i.e., power) with the EPC models.
+For the PDEM models, presence/absence was converted to frequency occurrence as the number of sites with seagrass in a year divided by the total number of sites and all data were subset to less than 2 meters to better characterize locations where seagrass occurs.
+
+Add equations.....
 
@@
 ```{r}
-epcmodsum1 <- modtxt_fun(sgmods$epcmod1)
-epcmodsum2 <- modtxt_fun(sgmods$epcmod2)
-fimmodsum <- modtxt_fun(sgmods$fimmod)
+# epcmodsum1 <- modtxt_fun(sgmods$epcmod1)
+# epcmodsum2 <- modtxt_fun(sgmods$epcmod2)
+# fimmodsum <- modtxt_fun(sgmods$fimmod)
 ```
@@
 Linear models to assess the potential effects of temperature and salinity on seagrass change provided modest evidence that seagrass loss after 2016 was influenced by climate stressors.
-The most parsimonious model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence (`r epcmodsum1`).
+The most parsimonious model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence ().
 An interaction of the metrics with time period was also observed, with notable differences by predictor (Figure [-@fig-sgmod]a, @tbl-modtab).
@@
 The interaction between the temperature and salinity metric was included in the most parsimonious model, although it was not significant.
-The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (`r epcmodsum2`), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab).
+The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab).
 Note that [@fig-sgmod] shows model predictions for OTB only, although the shape of the relationships do not change for other bay segments other than as additive shifts of the curves up or down.
 
-The most parsimonious model for the FIM dataset also showed potential temperature and salinity associations with seagrass change after 2016, although explanatory power was lower than the EPC models likely due to larger spread of the data (`r fimmodsum`).
+The most parsimonious model for the FIM dataset also showed potential temperature and salinity associations with seagrass change after 2016, although explanatory power was lower than the EPC models likely due to larger spread of the data ().
 Increases in temperature and decreases in salinity were both associated with reductions in percent cover after 2016, whereas no relationships were observed prior to 2016 (p \< 0.05 and p \< 0.1 for the interactions between temperature or salinity with time period, respectively, Figure [-@fig-sgmod]b, @tbl-modtab).
@@
 Thus, time period was necessary to control for these confounding relationships.
-Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay, particularly in OTB where the most loss occurred [@tbep0696; @Beck20a].
+Additionally, light attenuation has improved over the period of record and within the limits estimated to be supportive of seagrass growth in Tampa Bay [@tbep0400], particularly in OTB where the most loss occurred (Figure -[@fig-seagrasschg]c).
 A final consideration for our models relates to how seagrasses may influence their environment, particularly for the PDEM and FIM datasets where temperature and salinity were measured at the same locations as seagrass.
@@
 Red lines in (a) show approximate capacity of seagrass coverage based on the baywide target of 40,000 acres and red lines in (c) show the approximate linear trend for the period of record covered by the transect data in (b).
-The dashed line in (c) shows the light attenuation threshold for each bay segment identified as supportive for seagrass growth.
+The dashed line in (c) shows the light attenuation threshold to support seagrass growth in each bay segment.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/seagrasschg.png){#fig-seagrasschg}

### 2024-05-06 R/dat_proc.R

@@ -1471,41 +1471,95 @@ pincotemp <-  saltempsec %>%
 
 save(pincotemp, file = here('data/pincotemp.RData'))
 
+
+# secchi on bottom for each dataset -----------------------------------------------------------
+
+data(epcdata, package = 'tbeptools')
+load(file = here::here('data/fimsgtempdat.RData'))
+load(file = here::here('data/pincotemp.RData'))
+
+epconb <- epcdata %>% 
+  filter(yr >= 1998 & yr <= 2022) %>%
+  summarise(per = 100 * mean(grepl('>', sd_q))) %>%
+  round(0) %>% 
+  paste0('%')
+fimonb <- fimsgtempdat %>% 
+  filter(month(date) %in% c(7:10)) %>%
+  summarise(per = 100 * sum(secchi_on_bottom, na.rm = T) / n()) %>%
+  round(0) %>% 
+  paste0('%')
+pincoonb <- pincotemp %>% 
+  filter(month(date) %in% c(7:10)) %>%
+  filter(depth_m < 2) %>% 
+  summarise(per = 100 * sum(secchi_on_bottom, na.rm = T) / n()) %>%
+  round(0) %>% 
+  paste0('%')
+
+onb <- list(epc = epconb, fim = fimonb, pinco = pincoonb)
+
+save(onb, file = here::here('data/onb.RData'))
+
 # seagrass declines models, epc, fim, and pinco -----------------------------------------------
 
 load(file  = here('data/epccmbdat.RData'))
 load(file = here('data/fimsgtempdat.RData'))
 load(file = here('data/pincotemp.RData'))
 
-modprep <- function(dat){
+tomod <- epccmbdat %>% 
+  filter(!bay_segment %in% c('LTB'))
 
-  dat %>% 
-    filter(!bay_segment %in% c('LTB')) %>% 
-    mutate(
-      yrcat = case_when(
-        yr <= 2016 ~ 'pre', 
-        yr > 2016 ~ 'post'
-      ), 
-      yrcat = factor(yrcat, levels = c('pre', 'post'), 
-                     labels = c('Recovery (pre - 2016)', 'Decline (post - 2016)'))
[…]

### 2024-05-06 R/gam_scratch.R

@@ -65,7 +65,7 @@ tomod <- fimsgtempdat %>%
   select(date, sgcov, sgpres, yr, temp, sal, secchi_m, secchi_on_bottom, bay_segment) %>% 
   na.omit() %>%
   mutate(mo = month(date)) %>% 
-  filter(month(date) %in% c(7:11)) %>%
+  filter(month(date) %in% c(7:10)) %>%
   summarise(
     sgcov = mean(sgcov),
     sgpres = mean(sgpres),
@@ -138,14 +138,14 @@ ggplot(salslc, aes(x = sal, y = .estimate, color = factor(mo), fill = factor(mo)
 # PDEM ----------------------------------------------------------------------------------------
 
 tomod <- pincotemp %>% 
-  filter(!bay_segment %in% c('LTB')) %>% 
+  filter(depth_m < 2) %>% 
   select(date, allsg, yr, site, temp, sal, secchi_m, secchi_on_bottom, bay_segment) %>% 
   na.omit() %>%
   mutate(
     mo = month(date),
     site = factor(site)
   ) %>% 
-  filter(mo >= 7 & mo <= 11) %>% 
+  filter(mo >= 7 & mo <= 10) %>% 
   summarise(
     allsg = sum(allsg) / length(allsg),
     temp = mean(temp),

### 2024-05-07 manu-draft.qmd
add supp gam figs and tabs

@@
 Since that time, the greatest areal coverage expansions were observed in the upper bay segments (OTB, HB, and MTB; Figure [-@fig-seagrasschg]a).
-Light environments have improved (Figure [-@fig-seagrasschg]c) through reduction of external nitrogen loadings by about 2/3 by the 1980s from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed \[@greening2014; @beck2019\].
+Light environments have improved (Figure [-@fig-seagrasschg]c) through reduction of external nitrogen loadings by about 2/3 by the 1980s from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed [@greening2014; @beck2019].
 
@@
 
-```{r}
-# secchi on bottom is na issue
-# also averaging all models by year in the bay segment for comparable sample size, fim/pdem limited to july - nov to minimize seasonal effect and when transect surveys completed for comparability
-# < 17% for epc, much higher for fim/pdem so la not included, but can show la relatively stable in a plot
-# epc is bonk becuase of disconnect between sites
-# fim is good, makes sense, but otb is weird
-# pdem temp makes sense for otb
-```
-
 For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends above.
@@
 For 1998 to 2022, `r onb$epc` of the Secchi observations were recorded on the bottom.
-A single smooth term was used for each predictor using a thing plate regression spline, including a tensor product smoother term that evaluated the potential interacting effects of each predictor with year.
-Two models were evaluated, one with the temperature and salinity metrics together and another with the both metric.
+A single smooth term was used for each predictor using a thin plate regression spline, including a tensor product interaction term with a cubic regression spline that evaluated the potential interacting effects of each predictor with year.
+Two models were evaluated, one with the temperature and salinity metrics together (@eq-epcmod1) and another with the both metric (@eq-epcmod2).
 All models excluded Lower Tampa Bay because of minimal seagrass change over time.
@@
 wrightii* in the upper bay segments (Figure [-@fig-seagrasschg]b).
+The GAM equations in mgcv R package notation [@wood17] are below, where $s$ is an smoother function for a single predictor and $ti$ is the tensor product interaction between predictors.
+All GAMs were fit using restricted maximum likelihood evaluation [@wood11].
 
+$$
+\begin{aligned}
+Seagrass \sim s(year, by = bay \ segment) + s(light \ att, by = bay \ segment) + 
+\\
+s(temperature, by = bay \ segment) + s(salinity, by = bay \ segment) + 
+\\
+ti(light \ att, \ year, by = bay \ segment) + ti(temperature, year, by = bay \ segment) +
+\\ 
+ti(salinity, year, by = bay \ segment)
+\end{aligned}
+$$ {#eq-epcmod1}
+
+$$
+\begin{aligned}
+Seagrass \sim s(year, by = bay \ segment) + s(light \ att, by = bay \ segment) + 
+\\
+s(both, by = bay \ segment) +
+\\
+ti(light \ att, \ year, by, bay \ segment) + ti(both, year, by = bay \ segment)
+\end{aligned}
+$$ {#eq-epcmod2}
 Separate models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change.
 These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location).
-GAMs were used for the FIM data to evaluate seagrass percent cover using continuous temperature and salinity as independent variables.
-Interactions with year for each predictor variable and separate smoothed terms for each bay segment were also included as above.
+GAMs were used for the FIM data to evaluate seagrass percent cover using year, temperature, and salinity as independent variables.
+Separate smoothed terms for each predictor and iteractions with year for each predictor for each bay segment were included as above.
 GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data.
@@
 For the PDEM models, presence/absence was converted to frequency occurrence as the number of sites with seagrass in a year divided by the total number of sites and all data were subset to less than 2 meters to better characterize locations where seagrass occurs.
+The GAM equations for the FIM (@eq-fimmod) and PDEM (@eq-pincomod) datasets were as follows.
 
-Add equations.....
+$$
+\begin{aligned}
+Seagrass \sim s(year, by = bay \ segment) + s(temperature, by = bay \ segment) + 
+\\
+s(salinity, by = bay \ segment) + 
+\\
+ti(temperature, year, by = bay \ segment) + ti(salinity, year, by = bay \ segment)
+\end{aligned}
+$$ {#eq-fimmod}
+
+$$
+\begin{aligned}
+Seagrass \sim s(year) + s(temperature) + s(salinity) + 
+\\
+ti(temperature, year) + ti(salinity, year)
+\end{aligned}
+$$ {#eq-pincomod}
 
@@
 # fimmodsum <- modtxt_fun(sgmods$fimmod)
+# secchi on bottom is na issue
+# also averaging all models by year in the bay segment for comparable sample size, fim/pdem limited to july - nov to minimize seasonal effect and when transect surveys completed for comparability
+# < 17% for epc, much higher for fim/pdem so la not included, but can show la relatively stable in a plot
+# epc is bonk becuase of disconnect between sites
+# fim is good, makes sense, but otb is weird
+# pdem temp makes sense for otb
 ```
 
-Linear models to assess the potential effects of temperature and salinity on seagrass change provided modest evidence that seagrass loss after 2016 was influenced by climate stressors.
+GAMs to assess the potential effects of temperature and salinity on seagrass change provided modest evidence that seagrass change was influenced by climate stressors.
 The most parsimonious model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence ().

### 2024-05-07 supplement.qmd
add supp gam figs and tabs

@@
 load(file = here('tabs/supp2daytab.RData'))
+load(file = here('tabs/suppepcmod1tab.RData'))
+load(file = here('tabs/suppepcmod2tab.RData'))
+load(file = here('tabs/suppfimmodtab.RData'))
+load(file = here('tabs/supppincomodtab.RData'))
 ```
@@
 
-<!-- ![Relationship between water temperature and discharge in the Hillsborough River (USGS gage 02303000) over time.
-The relationship is shown for February (dry month) and August (wet month) grouped by historic (1975-1999) and contemporary (2000-2022) time periods.
-Regression lines with 95% confidence intervals for each temporal grouping are shown.](figs/suppflowtemp.png){#fig-suppflowtemp} -->
-
-<!-- {{< pagebreak >}} -->
-
 ![Examples of Generalized Additive Models of bottom water temperature and salinity for station 66 in Old Tampa Bay.
@@
 {{< pagebreak >}}
+
+![Partial effects of smoothers by bay segment from a Generalized Additive model used to describe seagrass change relative to year, light attenuation, the number of days each year when temperature was above 30 $^\circ$C, and the number of days each year when salinity was below 25 ppt.
+The EPC data were used to develop the model.
+Partial effects describe the modeled association between each predictor and seagrass frequency occurrence after accounting for the effects of the other predictors.
+The model also included smoothed interaction terms with year (not shown).
+See Table S7 for additional model fit statistics.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.
+n = 75, R.sq.
+0.93, Deviance explained 97\%](figs/suppepcmod1.png){#fig-suppepcmod1}
+{{< pagebreak >}}
+
+![Partial effects of smoothers by bay segment from a Generalized Additive model used to describe seagrass change relative to year, light attenuation, and the number of days each year when both temperature was above 30 $^\circ$C and salinity was below 25 ppt.
+The EPC data were used to develop the model.
+Partial effects describe the modeled association between each predictor and seagrass frequency occurrence after accounting for the effects of the other predictors.
+The model also included smoothed interaction terms with year (not shown).
+See Table S8 for additional model fit statistics.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.
+n = 75, R.sq.
+0.93, Deviance explained 97%](figs/suppepcmod2.png){#fig-suppepcmod2}
+{{< pagebreak >}}
+
+![Partial effects of smoothers by bay segment from a Generalized Additive model used to describe seagrass change relative to year, temperature ($^\circ$C), and salinity (ppt).
+The FIM data were used to develop the model.
+Partial effects describe the modeled association between each predictor and mean seagrass coverage (0-100) after accounting for the effects of the other predictors.
+The model also included smoothed interaction terms with year (not shown).
+See Table S9 for additional model fit statistics.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.
+n = 81, R.sq.
+0.81, Deviance explained 89%](figs/suppfimmod.png){#fig-suppfimmod}
+{{< pagebreak >}}
+
+![Partial effects of smoothers from a Generalized Additive model used to describe seagrass change relative to year, temperature ($^\circ$C), and salinity (ppt).
+The PDEM data in Old Tampa Bay were used to develop the model.
+Partial effects describe the modeled association between each predictor and seagrass frequency occurrence after accounting for the effects of the other predictors.
+The model also included smoothed interaction terms with year (not shown).
+See Table S10 for additional model fit statistics.
+n = 19, R.sq.
+0.69, Deviance explained 81%](figs/supppincomod.png){#fig-supppincomod}
 
@@
 ```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: tbl-suppepcmod1tab
+suppepcmod1tab
+```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: tbl-suppepcmod2tab
+suppepcmod2tab
+```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: tbl-suppfimmodtab
+suppfimmodtab
+```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: tbl-supppincomodtab
+supppincomodtab
+```

### 2024-05-07 R/figs.R

@@ -19,6 +19,7 @@ library(car)
 library(mgcv)
 library(mixmeta)
 library(maps)
+library(gratia)
 
 source(here('R/funcs.R'))
 
@@ -1404,137 +1405,67 @@ png(here('figs/suppflowtemp.png'), height = 4, width = 6, family = 'serif', unit
 print(p)
 dev.off()
 
-# # ports data threshold counts -----------------------------------------------------------------
-# 
-# load(file = here('data/portsdat.RData'))
-# 
-# # thresholds to count by year, temp is above
-# thrs <- tibble(
-#   temp = c(29, 30, 31)
-# )
-# 
-# # get T/F vectors for predictions above/below thresholds by day
-# portscnt <- portsdat %>%
-#   group_nest(name) %>%
-#   crossing(thrs) %>%
-#   mutate(
-#     data = purrr::pmap(list(data, temp), function(data, temp){
-# 
-#       out <- data %>%
-#         mutate(
-#           cnt = temp_c >= temp
-#         )
-# 
-#       return(out)
-# 
-#     })
-#   ) %>%
-#   unnest('data') %>%
-#   summarise(
-#     runcnt = runfunc(cnt),
-#     sumcnt = sum(cnt),
-#     .by = c(name, yr, temp)
-#   )
-# 
-# # ggplot(portscnt, aes(x = yr, y = runcnt, col = factor(temp))) +
-# #   geom_point() +
-# #   geom_smooth(se = F, method = 'lm', formula = y ~ x) +
-# #   facet_wrap(~name)
-# 
-# ggplot(portsdat, aes(x = date, y = temp_c)) +
-#   geom_line() +
-#   facet_wrap(~name, ncol = 1)
-# 
-# ##
-# # epc stations closes to gages (verified from maps)
-# 
-# epccnts <- tibble(
-#     fl = c('MTB_28_tempbot', 'LTB_90_temptop', 'HB_52_temptop', 'OTB_36_temptop'),
-#     name = c('stpete', 'manatee', 'mckay', 'oldporttampa')
-#   ) %>% 
-#   group_nest(name) %>% 
-#   mutate(
-#     mod = purrr::map(data, function(x){
-#   
-#       load(file = here(paste0('data/', x, '.RData')))
-#       
-#       out <- get(x[[1]])
-#       
-#       return(out)
-#       
-#     }),
[…]

### 2024-05-07 R/tabs.R

@@ -457,3 +457,22 @@ supp2daytab <- totab %>%
 
 save(supp2daytab, file = here('tabs/supp2daytab.RData'))
 
+# gam summaries -------------------------------------------------------------------------------
+
+load(file = here::here('data/sgmods.RData'))
+
+suppepccap1 <- 'Summary of smoother terms in the Generalized Additive Model used to evaluate seagrass response for the EPC data in relation to temperature and salinity stress metrics, with additional smoothers for year and light attenuation.  Separate smoothers were fit for each bay segment.  $s$ = invidual smoother, $ti$ = interaction term. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.'  
+suppepcmod1tab <- gam_table(sgmods$epcmod1, cap = suppepccap1)
+save(suppepcmod1tab, file = here('tabs/suppepcmod1tab.RData'))
+
+suppepccap2 <- 'Summary of smoother terms in the Generalized Additive Model used to evaluate seagrass response for the EPC data in relation to "both" stressors (both temperuture above and salinity below thresholds), with additional smoothers for year and light attenuation.  Separate smoothers were fit for each bay segment.  $s$ = invidual smoother, $ti$ = interaction term. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.'  
+suppepcmod2tab <- gam_table(sgmods$epcmod2, cap = suppepccap2)
+save(suppepcmod2tab, file = here('tabs/suppepcmod2tab.RData'))
+
+suppfimcap <- 'Summary of smoother terms in the Generalized Additive Model used to evaluate seagrass response for the FIM data in relation to temperature, salinity, and year.  Separate smoothers were fit for each bay segment.  $s$ = invidual smoother, $ti$ = interaction term. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.'  
+suppfimmodtab <- gam_table(sgmods$fimmod, cap = suppfimcap)
+save(suppfimmodtab, file = here('tabs/suppfimmodtab.RData'))
+
+supppincocap <- 'Summary of smoother terms in the Generalized Additive Model used to evaluate seagrass response for the PDEM data in relation to temperature and salinity, salinity, and year.  $s$ = invidual smoother, $ti$ = interaction term.'  
+supppincomodtab <- gam_table(sgmods$pincomod, cap = supppincocap)
+save(supppincomodtab, file = here('tabs/supppincomodtab.RData'))

### 2024-05-07 manu-draft.qmd
gam summary plot main text added

@@
 
-GAMs to assess the potential effects of temperature and salinity on seagrass change provided modest evidence that seagrass change was influenced by climate stressors.
-The most parsimonious model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence ().
+GAMs to assess the potential effects of year, light attenuation, temperature and salinity on seagrass change provided modest evidence that seagrass change was influenced by climate stressors.
+The model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence ().
 An interaction of the metrics with time period was also observed, with notable differences by predictor (Figure [-@fig-sgmod]a, @tbl-modtab).
@@
 
-![Predicted results from generalized linear models evaluating seagrass changes versus temperature or salinity predictors for the (a) EPC and (b) FIM datasets.
-Results are shown only for OTB, although the shape of the relationships do not change between bay segments.
-The predictors for (a) were the number of days when temperature was above the threshold and the number of days when salinity was below the threshold.
-The predictors for (b) were measured temperature and salinity at the bottom of the water column.
-Time periods were seagrass recovery prior to 2016 and seagrass decline after 2016.
-Shaded areas are 95% confidence intervals.
-Points are observed data that include aggregated effects of bay segment included in the models.](figs/sgmods.png){#fig-sgmod}
+![Predicted results from Generalized Additive Models evaluating seagrass changes versus year, temperature, and salinity for the (a) FIM (first three columns evaluating mean annual percent cover from 0-100, n = 81, R.sq.
+0.81, deviance explained 89%) and (b) PDEM (right column evaluating annual frequency occurrence from 0-1 in OTB only, n = 19, R.sq.
+0.69, deviance explained 81%) datasets.
+Partial effects describe the modeled association between each predictor and seagrass change after accounting for the effects of the other predictors.
+The models also included smoothed interaction terms with year (not shown).
+See Tables S9 and S10 for additional model fit statistics.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/sgmods.png){#fig-sgmod}
 

### 2024-05-07 supplement.qmd
gam summary plot main text added

@@
 
-![Partial effects of smoothers by bay segment from a Generalized Additive model used to describe seagrass change relative to year, temperature ($^\circ$C), and salinity (ppt).
-The FIM data were used to develop the model.
-Partial effects describe the modeled association between each predictor and mean seagrass coverage (0-100) after accounting for the effects of the other predictors.
-The model also included smoothed interaction terms with year (not shown).
-See Table S9 for additional model fit statistics.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.
-n = 81, R.sq.
-0.81, Deviance explained 89%](figs/suppfimmod.png){#fig-suppfimmod}
-{{< pagebreak >}}
-
-![Partial effects of smoothers from a Generalized Additive model used to describe seagrass change relative to year, temperature ($^\circ$C), and salinity (ppt).
-The PDEM data in Old Tampa Bay were used to develop the model.
-Partial effects describe the modeled association between each predictor and seagrass frequency occurrence after accounting for the effects of the other predictors.
-The model also included smoothed interaction terms with year (not shown).
-See Table S10 for additional model fit statistics.
-n = 19, R.sq.
-0.69, Deviance explained 81%](figs/supppincomod.png){#fig-supppincomod}
-
 ## Tables {.unnumbered}

### 2024-05-07 R/figs.R

@@ -1318,50 +1318,11 @@ dev.off()
 
 # seagrass change plots -----------------------------------------------------------------------
 
-load(file = here('data/sgmods.RData'))
+load(file = here::here('data/sgmods.RData'))
 
-epcmod1 <- sgmods$epcmod1
-fimmod <- sgmods$fimmod
-pincomod <- sgmods$pincomod
+p <- gamplo_fun(sgmods)
 
-##
-# epc model plot
-
-epcplo <- modplo_fun(
-  mod = epcmod1, 
-  xlab1 = '# days temperature > 30 \u00B0C', 
-  ylab1 = 'Frequency occurrence', 
-  xlab2 = '# days salinity < 25 ppt',
-  title1 = '(a) EPC', 
-  ismetric = T
-)
-
-## 
-# fim model plot
-
-fimplo <- modplo_fun(
-  mod = fimmod, 
-  xlab1 = 'Temperature (\u00B0C)', 
-  ylab1 = '% cover', 
-  xlab2 = 'Salinity (ppt)',
-  title1 = '(b) FIM'
-)
-
-# ## 
-# # pinco model plot, only yrcat significant
-# 
-# pincoplo <- modplo_fun(
-#   mod = pincomod, 
-#   xlab1 = 'Temperature (\u00B0C)', 
-#   ylab1 = 'Probability of occurrence', 
-#   xlab2 = 'Salinity (ppt)', 
-#   isbinom = T,
-#   title1 = '(c) PDEM'
-# )
-
-p <- epcplo / fimplo 
-
-png(here('figs/sgmods.png'), height = 6, width = 7, family = 'serif', units = 'in', res = 300)
+png(here('figs/sgmods.png'), width = 8, height = 6, units = 'in', res = 300, family = 'serif')
 print(p)
 dev.off()
 
@@ -1413,11 +1374,11 @@ mod <- sgmods$epcmod1
 smths <- c('s(yr)', 's(la)', 's(temp)', 's(sal)')
 labels <- c('year', 'light att (m-1)', '# days temperature > 30 \u00B0C', '# days salinity < 25 ppt')
 cols <- c('grey20', 'bisque4', 'red2', 'dodgerblue2')
-toplo <- gamplo_fun(mod, smths, labels, cols)
+toplo <- suppgamplo_fun(mod, smths, labels, cols)
 
 p <- toplo$plos[[1]] / toplo$plos[[2]] / toplo$plos[[3]] / toplo$plos[[4]]  
 
[…]

### 2024-05-07 R/tabs.R

@@ -26,7 +26,7 @@ totab <- tibble(
   Description = c('Seagrass coverage in acres', 'Seagrass frequency occurrence by species', 'Water quality monitoring samples', 'Nearshore temperature and salinity measurements, seagrass species and cover', 'Water quality and seagrass presence/absence', 'Air temperature', 'Area-weighted precipitation for the wet season (June-September) for the Tampa Bay watershed'),
   Temporal = c('1988 - 2022, biennial', '1999 - 2022, annual', '1975 - 2022, monthly', '1996 to 2022, monthly', '2003 - 2022, monthly', '1975 - 2022, annual', '1975-2022, annual'),
   Spatial = c('Whole bay', 'Whole bay, 62 transects', 'Whole bay, fixed sites', 'Whole bay nearshore, stratified random sites', 'Old Tampa Bay, stratified random sites', '27.979$^\\circ$N, 82.535$^\\circ$W', 'Whole watershed'),
-  Analysis = c('Biennial trends by bay segment, visual only', 'Annual trends by bay segment and species, comparison with temperature, salinity, and light attenuation as stressor metrics or observed data at annual scale', 'Trends in annual change and seasonal Kendall tests, estimate of stressor metrics as number of days above/below threshold', 'Trends in annual observed temperature, salinity, comparison to seagrass % cover', 'Trends in annual observed temperature, salinity, comparison to seagrass presence/absence', 'Annual trend', 'Annual trend')
+  Analysis = c('Biennial trends by bay segment, visual only', 'Annual trends by bay segment and species, comparison with temperature, salinity, and light attenuation as stressor metrics or observed data at annual scale', 'Trends in annual change and seasonal Kendall tests, estimate of stressor metrics as number of days above/below threshold', 'Trends in annual observed temperature, salinity, comparison to annual seagrass % cover', 'Trends in annual observed temperature, salinity, comparison to annual seagrass frequency occurrence', 'Annual trend', 'Annual trend')
 )
 
 dattab <- knitr::kable(totab)#, col.names = c('Bay Segment', 'Threshold', 'Slope', 'Start', 'End'))

### 2024-05-08 manu-draft.qmd
more gam revisions, results done, working on discuss

@@
 load(file = here('tabs/mixdaytab.RData'))
-load(file = here('tabs/modtab.RData'))
 load(file = here('data/sgmods.RData'))
+load(file = here('data/sgmodsum.RData'))
 load(file = here('data/onb.RData'))
@@
 Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment from 1998 to 2022.
-Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the annual summaries.
-As such, the annual stressor metrics were compared to frequency occurrence (all species) each year by bay segment.
+Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the transect year summaries.
+As such, the stressor metrics were compared to frequency occurrence (all species) each year by bay segment.
 GAMs were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, or both.
@@
 wrightii* in the upper bay segments (Figure [-@fig-seagrasschg]b).
-The GAM equations in mgcv R package notation [@wood17] are below, where $s$ is an smoother function for a single predictor and $ti$ is the tensor product interaction between predictors.
+The GAM equations in mgcv R package notation [@wood17] are below, where $s$ is a smoother function for a single predictor and $ti$ is the tensor product interaction between predictors.
 All GAMs were fit using restricted maximum likelihood evaluation [@wood11].
@@
 \end{aligned}
-$$ {#eq-epcmod2}
-Separate models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change.
+$$ {#eq-epcmod2} models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change.
 These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location).
-GAMs were used for the FIM data to evaluate seagrass percent cover using year, temperature, and salinity as independent variables.
-Separate smoothed terms for each predictor and iteractions with year for each predictor for each bay segment were included as above.
+GAMs were used for the FIM data to evaluate annual average seagrass percent cover using year, temperature, and salinity as independent variables.
+Separate smoothed terms for each predictor and interactions with year for each predictor for each bay segment were included as above.
 GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data.
@@
 For the PDEM models, presence/absence was converted to frequency occurrence as the number of sites with seagrass in a year divided by the total number of sites and all data were subset to less than 2 meters to better characterize locations where seagrass occurs.
-The GAM equations for the FIM (@eq-fimmod) and PDEM (@eq-pincomod) datasets were as follows.
+The GAM equations for the FIM (@eq-fimmod) and PDEM (@eq-pincomod) datasets, respectively, were as follows.
 
@@
 ```{r}
-# epcmodsum1 <- modtxt_fun(sgmods$epcmod1)
-# epcmodsum2 <- modtxt_fun(sgmods$epcmod2)
-# fimmodsum <- modtxt_fun(sgmods$fimmod)
-# secchi on bottom is na issue
-# also averaging all models by year in the bay segment for comparable sample size, fim/pdem limited to july - nov to minimize seasonal effect and when transect surveys completed for comparability
-# < 17% for epc, much higher for fim/pdem so la not included, but can show la relatively stable in a plot
-# epc is bonk becuase of disconnect between sites
-# fim is good, makes sense, but otb is weird
-# pdem temp makes sense for otb
+sgmodtxt <- modtxt_fun(sgmodsum)
 ```
 
-GAMs to assess the potential effects of year, light attenuation, temperature and salinity on seagrass change provided modest evidence that seagrass change was influenced by climate stressors.
-The model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence ().
-An interaction of the metrics with time period was also observed, with notable differences by predictor (Figure [-@fig-sgmod]a, @tbl-modtab).
-For the period prior to 2016, frequency occurrence of seagrass increased as the number of days where temperature was above 30$^\circ$ increased, whereas no relationship was observed after 2016.
-However, the interaction between time period and temperature was not significant, suggesting no difference between time periods.
-The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure [-@fig-sgmod]a, @tbl-modtab).
-The interaction between the temperature and salinity metric was included in the most parsimonious model, although it was not significant.
-The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab).
-Note that [@fig-sgmod] shows model predictions for OTB only, although the shape of the relationships do not change for other bay segments other than as additive shifts of the curves up or down.
+GAMs to assess the potential effects of year, light attenuation, temperature and salinity on seagrass change provided some evidence that seagrass change was influenced by climate stressors, particularly for the FIM and PDEM datasets.
+The model for the EPC data (`r sgmodtxt$epcmod1`) had notable associations with year for the OTB and MTB bay segments and for light attenuation for the HB bay segment (@fig-sgmod1, Table S7).
+Associations of seagrass change with the temperature and salinity metrics were weak and somewhat contrary to expectation (i.e., an increasing partial effect on seagrass as the stressor metrics increased).
+The EPC model using the "both" stressor metric was similar (`r sgmodtxt$epcmod2`, Figure S7, Table S8), suggesting no added benefit of evaluating times when both temperature was above and salinity was below the threshold.
 
-The most parsimonious model for the FIM dataset also showed potential temperature and salinity associations with seagrass change after 2016, although explanatory power was lower than the EPC models likely due to larger spread of the data ().
-Increases in temperature and decreases in salinity were both associated with reductions in percent cover after 2016, whereas no relationships were observed prior to 2016 (p \< 0.05 and p \< 0.1 for the interactions between temperature or salinity with time period, respectively, Figure [-@fig-sgmod]b, @tbl-modtab).
-The interaction between salinity and temperature was not included as a significant term in the final FIM model.
-Lastly, the most parsimonious model for the PDEM dataset only included time period as a marginally significant variable and the overall model fit was not significant (@tbl-modtab).
+The GAMs for the FIM and PDEM datasets had notable associations with temperature and salinity stress, where an increase in temperature and a decrease in salinity was often associated with a decline in seagrass depending on bay segment.
+The FIM model (`r sgmodtxt$fimmod`), Figure \[-@fig-sgmod2\]a, Table S9) showed a slight decrease in seagrass with increasing temperature for HB and MTB, with a similar change as salinity decreased.
+The PDEM model (`r sgmodtxt$pincomod`, Figure \[-@fig-sgmod2\]b, Table S10) that evaluated only OTB showed a distinct decrease with seagrass as temperature increased, but no association with salinity.
+Some of the interaction terms of the predictors with year suggested the relationships with seagrass may have changed over time, but the results were not consistent between bay segments or models (Tables S7 to S10).
 
@@
 Tampa Bay has gotten hotter and fresher; water temperature has increased by 0.03 - 0.04 $^\circ$C per year and salinity has decreased by 0.04 - 0.06 ppt per year, translating to an increase of 1.3 to 1.7 $^\circ$C and a decrease of 1.6 to 2.6 ppt over the past fifty years.
+The number of days each year when temperature was above 30 $^\circ$C or salinity was below 25 ppt has also increased consistently, with an average increase baywide of 48 and 37 days, respectively, since 1975.
 These changes were demonstrated in three long-term datasets with different sampling methods and periods of record.
@@
 Our assessment evaluated non-parametric trends (i.e., Kendall tests, less sensitive to outliers) at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to their larger spatial area and model domain.
-Our results showing increases in temperature and decreases in salinity in the summer, early fall, most notably for OTB and northern stations of HB [@fig-kendall], are likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.
+Our results showing increases in temperature and decreases in salinity in the summer, early fall, most notably for OTB and northern stations of HB (@fig-kendall), are likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.
 These upper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to more rapid warming with shallower depths.
-The significant reduction in salinity for LTB is also of note, perhaps related to gravitational circulation patterns that export lower salinity water from upstream in the main shipping channels [@weisberg2006].
+The significant reduction in salinity for LTB is also of note, perhaps related to gravitational circulation patterns that export lower salinity water from upstream in the main shipping channels \[@weisberg2006\].
 Additionally, @shiir provided a recent assessment of a 2023 heatwave in south Florida, supported by a 20-year trend assessment that suggested estuaries were warming at nearly double the rate of the Gulf of Mexico.
@@
 Our relatively simple modeling approach provided some evidence that climate-related stressors impart some effect on recent seagrass losses in Tampa Bay.
-The models did not provide a consistent, nor statistically powerful, explanation that increasing temperature and decreasing salinity were key (or the sole) drivers.
+The models did not provide a consistent explanation that increasing temperature and decreasing salinity were key (or the sole) drivers.
 However, evaluating all models together as weight-of-evidence suggests there is value in considering multiple datasets and models to interpret noisy patterns and compounding ecological processes.
-Model results for the EPC and FIM datasets both suggested that increasing temperature and decreasing salinity were associated with potential seagrass loss post 2016, described primarily using separate interaction terms of temperature or salinity with time period.
-For the EPC model, the interaction term was not significant for temperature, whereas the interaction was marginally significant for the salinity metric, such that a negative association between seagrass and salinity stress was observed post 2016 as compared to pre 2016.
-Likewise, the FIM models had a significant interaction term for the association of temperature with time period and a marginally significant interaction term for the association of salinity with time period.
-An important distinction between the EPC and FIM models is that the former evaluated the number of days above/below thresholds each year to quantify annual temperature or salinity stress, whereas the latter evaluated observed temperature and salinity values at the time of seagrass sampling.
+Model results for the FIM and PDEM datasets both suggested that increasing temperature and decreasing salinity were associated with potential seagrass loss in recent years (@fig-sgmod2).
+For the EPC model, none of the stressor metrics were associated with seagrass change.
+An important distinction between the EPC and other models is that the former evaluated the number of days above/below thresholds each year to quantify annual temperature or salinity stress, whereas the latter evaluated observed temperature and salinity values at the time of seagrass sampling.
 As such, both models attempted to describe the role of these stressors on potential seagrass change, but use different independent variables given the different sampling designs of each monitoring program.
@@
 
-![Predicted results from Generalized Additive Models evaluating seagrass changes versus year, temperature, and salinity for the (a) FIM (first three columns evaluating mean annual percent cover from 0-100, n = 81, R.sq.
-0.81, deviance explained 89%) and (b) PDEM (right column evaluating annual frequency occurrence from 0-1 in OTB only, n = 19, R.sq.
-0.69, deviance explained 81%) datasets.
+![Partial effects of smoothers (rows) by bay segment (columns) from a Generalized Additive Model used to describe seagrass change relative to year, light attenuation, the number of days each year when temperature was above 30 $^\circ$C, and the number of days each year when salinity was below 25 ppt.
+The EPC data were used for the independent variables in the model.
+Partial effects describe the modeled association between each predictor and seagrass frequency occurrence after accounting for the effects of the other predictors.
+The model also included smoothed interaction terms with year (not shown).
+See Table S7 for additional model fit statistics.
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.
+`r modtxt_fun(sgmodsum)$epcmod1`](figs/sgmod1.png){#fig-sgmod1}
+
+{{< pagebreak >}}
+
+![Partial effects of smoothers (rows) by bay segment (columns) from Generalized Additive Models evaluating seagrass changes versus year, temperature, and salinity for the (a) FIM (first three columns evaluating mean annual percent cover from 0-100, `r modtxt_fun(sgmodsum)$fimmod`) and (b) PDEM (right column evaluating annual frequency occurrence from 0-1 in OTB only, `r modtxt_fun(sgmodsum)$pincomod`) datasets.
 Partial effects describe the modeled association between each predictor and seagrass change after accounting for the effects of the other predictors.
@@
 See Tables S9 and S10 for additional model fit statistics.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/sgmods.png){#fig-sgmod}
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/sgmod2.png){#fig-sgmod2}
 
@@
 
-```{r}
-#| label: tbl-modtab
-#| tbl-cap: 'Summary of the most parsimonious Generalized Linear Models (GLMs) developed to evaluate the effects of temperature and salinity variables on seagrass change between time periods.
-Models were developed for different datasets described in the model column.
-The significant intercept term for each model is not shown and described factor conditions not explicitly shown by the remaining terms (e.g., OTB and pre-2016).
-The model term estimates include the standard error in parentheses.
-Note that the "Temp" and "Sal" terms for the EPC models describe metrics that measure the number of days when temperature was above or salinity was below a threshold value, whereas the terms in all other models describe observed temperature or salinity (see text for description).
-** p < 0.005, * p < 0.05, $^.$ p < 0.1.
-EPC: Environmental Protection Commission, FIM: Fisheries Independent Monitoring, PDEM: Pinellas County Department of Environmental Management.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.'
-modtab
-```
-
-{{< pagebreak >}}
-
 ## Data Availability {.unnumbered}

### 2024-05-08 supplement.qmd
more gam revisions, results done, working on discuss

@@
 
+source(here('R/funcs.R'))
+
+load(file = here('data/sgmodsum.RData'))
 load(file = here('tabs/supptempfittab.RData'))
@@
 
-![Partial effects of smoothers by bay segment from a Generalized Additive model used to describe seagrass change relative to year, light attenuation, the number of days each year when temperature was above 30 $^\circ$C, and the number of days each year when salinity was below 25 ppt.
-The EPC data were used to develop the model.
-Partial effects describe the modeled association between each predictor and seagrass frequency occurrence after accounting for the effects of the other predictors.
-The model also included smoothed interaction terms with year (not shown).
-See Table S7 for additional model fit statistics.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.
-n = 75, R.sq.
-0.93, Deviance explained 97\%](figs/suppepcmod1.png){#fig-suppepcmod1}
-{{< pagebreak >}}
-
-![Partial effects of smoothers by bay segment from a Generalized Additive model used to describe seagrass change relative to year, light attenuation, and the number of days each year when both temperature was above 30 $^\circ$C and salinity was below 25 ppt.
+![Partial effects of smoothers (rows) by bay segment (columns) from a Generalized Additive model used to describe seagrass change relative to year, light attenuation, and the number of days each year when both temperature was above 30 $^\circ$C and salinity was below 25 ppt.
 The EPC data were used to develop the model.
@@
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.
-n = 75, R.sq.
-0.93, Deviance explained 97%](figs/suppepcmod2.png){#fig-suppepcmod2}
+`r modtxt_fun(sgmodsum)$epcmod2`](figs/suppsgmod3.png){#fig-suppsgmod3}
+
 {{< pagebreak >}}

### 2024-05-08 R/dat_proc.R

@@ -1565,3 +1565,26 @@ sgmods <- list(epcmod1 = epcmod1, epcmod2 = epcmod2, fimmod = fimmod, pincomod =
 
 save(sgmods, file = here('data/sgmods.RData'))
 
+
+# sg mod gam performance ----------------------------------------------------------------------
+
+load(file = here("data/sgmods.RData"))
+
+sgmodsum <- sgmods %>% 
+  lapply(function(x){
+    
+    summod <- summary(x)
+    resdf <- sprintf('%.2f', round(summod$residual.df, 2))
+    rsq <- sprintf('%.2f', round(summod$r.sq, 2))
+    dev <- round(100 * summod$dev.expl, 0)
+    n <- nrow(x$model)
+    
+    out <- tibble(n = n, resdf = resdf, rsq = rsq, dev = dev)
+    
+    return(out)
+    
+  })
+
+save(sgmodsum, file = here('data/sgmodsum.RData'))
+
+

### 2024-05-08 R/figs.R

@@ -1316,71 +1316,35 @@ png(here('figs/suppmixeff.png'), height = 6, width = 7, family = 'serif', units 
 print(p)
 dev.off()
 
-# seagrass change plots -----------------------------------------------------------------------
-
-load(file = here::here('data/sgmods.RData'))
-
-p <- gamplo_fun(sgmods)
-
-png(here('figs/sgmods.png'), width = 8, height = 6, units = 'in', res = 300, family = 'serif')
-print(p)
-dev.off()
-
-# supp hillsborough flow v temp over time -----------------------------------------------------
-
-load(file = here('data/gagedat.RData'))
-
-toplo <- gagedat %>%
-  filter(name == 'Hillsborough') %>%
-  pivot_wider(names_from = 'var', values_from = 'val') %>%
-  na.omit() %>%
-  mutate(
-    yrgroup = ifelse(year < 2000, '1975 - 1999', '2000 - 2022'),
-    mo = month(date, label = T, abbr = F)
-  ) %>%
-  # filter(flow_m3d < quantile(flow_m3d, 0.9)) %>%
-  filter(mo %in% c('February', 'August')) #%>%
-# filter(temp_c > 1)
-
-p <- ggplot(toplo, aes(x = flow_m3d / 1e6, y = temp_c, group = yrgroup, fill = yrgroup, color = yrgroup)) +
-  geom_point(size = 0.6, show.legend = F) +
-  scale_x_log10() +
-  facet_grid(~mo) +
-  theme_minimal() + 
-  scale_color_manual(values = c('tomato1', 'darkred')) +
-  scale_fill_manual(values = c('tomato1', 'darkred')) +
-  theme(
-    legend.position = 'top', 
-    panel.grid.minor = element_blank(), 
-    strip.text = element_text(size = 11)
-  ) +
-  geom_smooth(se = T, method = 'lm', formula = y ~ x) +
-  labs(
-    x = expression(paste('Hillsborough River discharge (', 10^6~m^3, ' / yr)')), 
-    y = 'Water temp. (\u00B0 C)',
-    color = 'Time period',
-    fill = 'Time period'
-  )
-
-png(here('figs/suppflowtemp.png'), height = 4, width = 6, family = 'serif', units = 'in', res = 600)
-print(p)
-dev.off()
-
[…]

### 2024-05-08 R/tabs.R

@@ -18,7 +18,6 @@ epcdat <- epcdata %>%
   ) %>% 
   filter(yr < 2023)
 
-
 # dataset summaries ---------------------------------------------------------------------------
 
 totab <- tibble(
@@ -257,42 +256,6 @@ suppsalifittab <- fittab %>%
 save(supptempfittab, file = here('tabs/supptempfittab.RData'))
 save(suppsalifittab, file = here('tabs/suppsalifittab.RData'))
 
-
-# glm performance -----------------------------------------------------------------------------
-
-load(file = here("data/sgmods.RData"))
-
-# epcmod1
-modtab <- sgmods %>% 
-  lapply(tidy) %>% 
-  enframe() %>% 
-  unnest(value) %>%
-  filter(term != '(Intercept)') %>% 
-  mutate(p.value = p_ast2(p.value)) %>% 
-  mutate_if(is.numeric, formatC, digits = 2) %>% 
-  mutate(
-    std.error = paste0('(', std.error, ')'), 
-    name = case_when(
-      name == 'epcmod1' ~ 'EPC 1', 
-      name == 'epcmod2' ~ 'EPC 2',
-      name == 'fimmod' ~ 'FIM',
-      name == 'pincomod' ~ 'PDEM'
-    ), 
-    term = gsub('temp', 'Temp', term), 
-    term = gsub('sal', 'Sal', term),
-    term = gsub('both', 'Both', term),
-    term = gsub('yrcatDecline \\(post - 2016\\)', 'Time_post', term),
-    term = gsub('yrcatDecline \\(post - 2016\\)', 'Time_post', term), 
-    term = gsub('bay_segment', 'Baysegment_', term),
-    name = ifelse(duplicated(name), '', name)
-  ) %>% 
-  unite('estimate', estimate, std.error, sep = ' ') %>% 
-  unite('statistic', statistic, p.value, sep = '') %>%
-  select(Model = name, Term = term, Estimate = estimate, `t-statistic` = statistic) %>% 
-  knitr::kable()
-
-save(modtab, file = here('tabs/modtab.RData'))
-
 # supp1 mixef mod summary table ----------------------------------------------------------------
 
 load(file = here('data/mixmods.RData'))
@@ -460,19 +423,20 @@ save(supp2daytab, file = here('tabs/supp2daytab.RData'))
 # gam summaries -------------------------------------------------------------------------------
 
 load(file = here::here('data/sgmods.RData'))
+load(file = here::here('data/sgmodsum.RData'))
 
[…]

### 2024-05-08 manu-draft.qmd
discussion updated for gams, fixes #14

@@
 These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay.
-Simple linear models provided a weight-of-evidence that recent seagrass declines are somewhat associated with hotter and fresher conditions.
+Generalized Additive Models provided a weight-of-evidence that recent seagrass declines are somewhat associated with hotter and fresher conditions.
 Trends in warming and increased precipitation in the region are likely to continue, further creating suboptimal conditions for seagrasses in Tampa Bay.
@@
 The analysis was supported by 30-year seagrass datasets including aerial surveys of total seagrass distribution, annual transect monitoring describing species-specific percent cover, and synoptic seagrass data collected with routine biotic and water quality surveys.
-Specific focus was on the periods prior to and after 2016, when seagrasses recovered baywide and then declined despite water quality conditions remaining relatively stable.
 
@@
 The GAMs for the FIM and PDEM datasets had notable associations with temperature and salinity stress, where an increase in temperature and a decrease in salinity was often associated with a decline in seagrass depending on bay segment.
-The FIM model (`r sgmodtxt$fimmod`), Figure \[-@fig-sgmod2\]a, Table S9) showed a slight decrease in seagrass with increasing temperature for HB and MTB, with a similar change as salinity decreased.
-The PDEM model (`r sgmodtxt$pincomod`, Figure \[-@fig-sgmod2\]b, Table S10) that evaluated only OTB showed a distinct decrease with seagrass as temperature increased, but no association with salinity.
+The FIM model (`r sgmodtxt$fimmod`), Figure [-@fig-sgmod2]a, Table S9) showed a slight decrease in seagrass with increasing temperature for HB and MTB, with a similar change as salinity decreased.
+The PDEM model (`r sgmodtxt$pincomod`, Figure [-@fig-sgmod2]b, Table S10) that evaluated only OTB showed a distinct decrease with seagrass as temperature increased, but no association with salinity.
 Some of the interaction terms of the predictors with year suggested the relationships with seagrass may have changed over time, but the results were not consistent between bay segments or models (Tables S7 to S10).
@@
 These upper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to more rapid warming with shallower depths.
-The significant reduction in salinity for LTB is also of note, perhaps related to gravitational circulation patterns that export lower salinity water from upstream in the main shipping channels \[@weisberg2006\].
+The significant reduction in salinity for LTB is also of note, perhaps related to gravitational circulation patterns that export lower salinity water from upstream in the main shipping channels [@weisberg2006].
 Additionally, @shiir provided a recent assessment of a 2023 heatwave in south Florida, supported by a 20-year trend assessment that suggested estuaries were warming at nearly double the rate of the Gulf of Mexico.
@@
 An important distinction between the EPC and other models is that the former evaluated the number of days above/below thresholds each year to quantify annual temperature or salinity stress, whereas the latter evaluated observed temperature and salinity values at the time of seagrass sampling.
+Additionally, the EPC models evaluated water quality changes from data collected at relatively deeper locations than where seagrass typically grows (Figure [-@fig-map]b), suggesting a potential disconnect in relating the two.
 As such, both models attempted to describe the role of these stressors on potential seagrass change, but use different independent variables given the different sampling designs of each monitoring program.
@@
 For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of external nitrogen inputs in affecting light environments for seagrass growth [@greening2014; @sherwood2017].
-As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences after accounting for light attenuation could additionally be explained by temperature or salinity.
-However, our use of time period (pre/post 2016) as a categorical variable indirectly addressed this issue.
-Modeling seagrass change as related to temperature or salinity for the entire record would have shown a spurious correlation of both with seagrass given the long-term recovery of seagrass, i.e., modeling challenges related to correlated predictors [@fourqurean2003].
-Thus, time period was necessary to control for these confounding relationships.
-Additionally, light attenuation has improved over the period of record and within the limits estimated to be supportive of seagrass growth in Tampa Bay [@tbep0400], particularly in OTB where the most loss occurred (Figure -[@fig-seagrasschg]c).
+Our inclusion of light attenuation in the EPC models was meant to account for how the light environment may be influencing seagrass growth, in addition to climate-related stressors.
+However, light attenuation has improved over the period of record and within the limits estimated to be supportive of seagrass growth in Tampa Bay [@tbep0400], particularly in OTB where the most loss occurred (Figure -[@fig-seagrasschg]c).
+Additional water quality parameters could be included in the models to provide further evidence that light-limitation is not the present driver for seagrass change (e.g., nitrogen loading, chlorophyll-a, color), although nutrient management for the benefit of seagrass growth will likely continue to be a dominant management paradigm for Tampa Bay.
 A final consideration for our models relates to how seagrasses may influence their environment, particularly for the PDEM and FIM datasets where temperature and salinity were measured at the same locations as seagrass.
@@
 An evaluation of each dataset showed a clear pattern of increasing temperature and decreasing salinity mirrored by long-term changes in air temperature and precipitation, suggesting that Tampa Bay has become hotter and fresher with the trends likely continuing in the future.
-Simple regression models provided weak, but partially-supporting evidence, that these changes can be linked to recent seagrass losses since 2016.
-Our models suggested that rising temperatures and decreasing salinities have had additive rather than multiplicative effects on seagrass, as evidenced by lack of significant interactions in models involving both stressors.
+GAMs provided partially-supporting evidence that these changes can be linked to recent seagrass losses.
 Future analyses may show more significant associations between physicochemical habitat conditions and seagrass change as the trends are very likely to continue to push seagrasses further outside of their tolerance ranges.

### 2024-05-08 manu-draft.qmd
minor edit to table 1, add water to title

@@
     number-sections: true
-title: "Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary"
+title: "Hot and fresh: evidence of climate-related suboptimal water conditions for seagrass in a large Gulf coast estuary"
 bibliography: refs.bib
@@
 Temporal and spatial scales are those used for analysis.
-SWFWMD: Southwest Florida Water Management District."
+SWFWMD: Southwest Florida Water Management District, EPC: Environmental Protection Commission of Hillsborough County, FIM: Florida Fish and Wildlife Comission, Fisheries Independent Monitoring, PDEM: Pinellas County Department of Environmental Management."
 dattab
 ```
+
+{{< pagebreak >}}
 

### 2024-05-08 supplement.qmd
minor edit to table 1, add water to title

@@
     number-sections: true
-title: "*Supplement* Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary"
+title: "*Supplement* Hot and fresh: evidence of climate-related suboptimal water conditions for seagrass in a large Gulf coast estuary"
 bibliography: refs.bib

### 2024-05-09 manu-draft.qmd
p val info removed from all figs, tabs

@@
 Results for seasonal Kendall tests by station are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr$^{-1}$).
-Summarized seasonal trends by month are shown as (b) the percent of stations in each bay segment with significant increasing (temperature) or decreasing (salinity) trends.
+Summarized seasonal trends by month are shown as (b) the average magnitude of change (slope) per month for stations in each bay segment for temperature and salinity.
 Bay segment outlines are shown in (a); OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/kendall.png){#fig-kendall}
@@
 The models included station as a random effect for each bay segment, with grey lines indicating individual station trends, grey points as actual number of days for each station, and thicker lines indicating the overall model fit.
-Slopes for significant models are shown in the bottom left of each facet.
+Slopes are shown in the bottom left of each facet.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/mixeff.png){#fig-mixeff}
@@
 Note that PDEM includes data only for the western portion of Old Tampa Bay from 2004 to present and EPC is the only dataset beginning prior to 1996.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
-** p < 0.005, * p < 0.05"
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay."
 temptrndtab
@@
 Note that PDEM includes data only for the western portion of Old Tampa Bay from 2004 to present and EPC is the only dataset beginning prior to 1996.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
-** p < 0.005, * p < 0.05"
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay."
 saltrndtab
@@
 #| tbl-cap: "Summary of mixed-effects models evaluating increases in the number of days each year from 1975 to 2022 when temperature was above 30 $^\\circ$C, salinity was below 25 ppt, or both temperature and salinity were above/below the thresholds.
-The start and end columns show the estimated number of days at the beginning and end of the period of record when temperature or salinity were above or below the thresholds.
-Values are the estimated mean number of days (plus standard error) from 1975 and 2022.
-** p < 0.005, * p < 0.05.
+The start and end columns show the estimated number of days (mean, standard error) in 1975 and 2022 when temperature or salinity were above or below the thresholds.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay."

### 2024-05-09 supplement.qmd
p val info removed from all figs, tabs

@@
 Results for seasonal Kendall tests by station and monitoring location (top or bottom of the water column) are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr$^{-1}$).
-Summarized seasonal trends by month are shown for (b) top and (c) bottom measurements as the percent of stations in each bay segment with significant increasing (temperature) or decreasing (salinity) trends.
+Summarized seasonal trends by month are shown as (b) the average magnitude of change (slope) per month for all stations in each bay segment for temperature and salinity.
 Bay segment outlines are shown in (a); OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppkendall.png){#fig-suppkendall}
@@
 The models included station as a random effect for each bay segment, with grey lines indicating individual station trends, grey points as actual number of days, and thicker lines indicating the overall model fit.
-Slopes for significant models are shown in the bottom left of each facet.
+Slopes are shown in the bottom left of each facet.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppmixeff.png){#fig-suppmixeff}
@@
 The slope estimates in the "Both" column indicate trends in the number of days when both temperature and salinity were above or below critical thresholds.
-\*\* p \< 0.005, \* p \< 0.05, - no model.
+- no model.
 OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.
@@
 Columns for "Both" indicate the estimated number of days when both temperature and salinity were above or below critical thresholds.
-Note that not all models are significant (see table 3).
 - no model.
@@
 Columns for "Both" indicate the estimated number of days when both temperature and salinity were above or below critical thresholds.
-Note that not all models are significant (see table 4).
 - no model.

### 2024-05-09 R/dat_proc.R

@@ -716,10 +716,6 @@ mixmodprds <- thrdat %>%
         
         summod <- summary(mod)$coefficients
         
-        pvl <- summod['yr', 'Pr(>|t|)']
-        if(pvl >= 0.05)
-          return('')
-        
         out <- summod['yr', 'Estimate'] %>%
           round(2) %>%
           as.character()
@@ -795,10 +791,6 @@ suppmixmodprds <- thrdat %>%
       if(!is.null(mod)){
         
         summod <- summary(mod)$coefficients
-        
-        pvl <- summod['yr', 'Pr(>|t|)']
-        if(pvl >= 0.05)
-          return('')
         
         out <- summod['yr', 'Estimate'] %>%
           round(2) %>%

### 2024-05-09 R/figs.R

@@ -800,7 +800,10 @@ ktresplo <- ktres %>%
           ),
           cnt = n(),
           nsigper = round(100 * nsig / cnt, 0),
-          perlab = ifelse(nsigper == 0, '', as.character(abs(nsigper)))
+          perlab = ifelse(nsigper == 0, '', as.character(abs(nsigper))), 
+          aveslo = median(slos, na.rm = T), 
+          aveslolab = gsub('^0', '', as.character(round(aveslo, 2))),
+          aveslolab = gsub('^\\-0', '-', aveslolab)
         )
       
     })
@@ -808,12 +811,13 @@ ktresplo <- ktres %>%
   select(-data) %>% 
   unnest('sum')
 
-p2 <- ggplot(ktresplo, aes(x = mo, y = bay_segment, fill = nsigper)) + 
+p2 <- ggplot(ktresplo, aes(x = mo, y = bay_segment, fill = aveslo)) + 
   geom_tile(color = 'darkgrey') +
-  geom_text(aes(label = perlab), color = 'white', fontface = 'bold') +
+  geom_text(aes(label = aveslolab, size = abs(aveslo)), color = 'white', fontface = 'bold') +
   scale_x_discrete(expand = c(0, 0)) + 
   scale_y_discrete(expand = c(0, 0)) + 
-  scale_fill_gradientn(leglab, limits = c(-100, 100), colors = c('blue', 'grey', 'tomato1')) +
+  scale_size_continuous(range = c(1, 5)) +
+  scale_fill_gradientn(leglab, limits = c(-0.12, 0.12), colors = c('blue', 'grey', 'tomato1')) +
   facet_wrap(~ varsimp, ncol = 1, scales = 'free_x') + 
   theme(
     strip.background = element_blank(), 
@@ -824,7 +828,7 @@ p2 <- ggplot(ktresplo, aes(x = mo, y = bay_segment, fill = nsigper)) +
   labs(
     x = NULL, 
     y = 'Bay segment',
-    title = '(b) % stations with significant trends by month'
+    title = '(b) Average magnitude of change by month'
   )
     
 leglab <- expression(paste(yr^{-1}))
@@ -860,11 +864,11 @@ pthm <- theme_bw(base_family = 'serif') +
 p1 <- ggplot() + 
   annotation_map_tile('cartolight', zoom = 11) +
   geom_sf(data = tbseg, inherit.aes = F) +
[…]

### 2024-05-09 R/tabs.R

@@ -4,6 +4,7 @@ library(mgcv)
 library(broom)
 library(tbeptools)
 library(flextable)
+library(knitr)
 
 source(here('R/funcs.R'))
 
@@ -144,7 +145,6 @@ totab <- trnds %>%
     bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB')), 
     org = factor(org, levels = c('EPC', 'FIM', 'PDEM'))
   ) %>% 
-  unite('slo', slo, pval, sep = '') %>% 
   select(
     var, 
     `Start year` = yrstr,
@@ -204,11 +204,11 @@ totab <- mixmods %>%
                     labels = c(paste('Temperature >', tmpthr), paste('Salinity <', salthr), 'Both')),
     bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB'))
   ) %>% 
-  unite('slo', slo, pvl, sep = '') %>% 
+  select(-pvl) %>% 
   unite('yrstr', yrstr, yrstrse, sep = ' ') %>%
   unite('yrend', yrend, yrendse, sep = ' ') %>% 
   mutate(
-    slo = ifelse(slo == 'NANA', '-', slo),
+    slo = ifelse(is.na(slo), '-', slo),
     yrstr = ifelse(grepl('NA', yrstr), '-', yrstr),
     yrend = ifelse(grepl('NA', yrend), '-', yrend)
   ) %>%
@@ -261,9 +261,8 @@ save(suppsalifittab, file = here('tabs/suppsalifittab.RData'))
 load(file = here('data/mixmods.RData'))
 
 totab <- mixmods %>% 
-  select(-yrstr, -yrstrse, -yrend, -yrendse) %>% 
+  select(-yrstr, -yrstrse, -yrend, -yrendse, -pvl) %>% 
   mutate(
-    pvl = p_ast(pvl), 
     slo = as.character(round(slo, 2)), 
     salithr = gsub('^.*_', '', salithr),
     tempthr = gsub('^.*_', '', tempthr), 
@@ -271,9 +270,8 @@ totab <- mixmods %>%
                      labels = c('Temperature', 'Salinity', 'Both')),
     bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB'))
   ) %>% 
-  unite('slo', slo, pvl, sep = '') %>% 
   mutate(
[…]

### 2024-05-09 manu-draft.qmd
revisions finished, need to proof, fixes #16, #17

@@
 library(tibble)
+library(tidyr)
 
@@
 load(file = here('data/gamfit.RData'))
+load(file = here('data/lintrnds.RData'))
 load(file = here('tabs/dattab.RData'))
@@
 These changes may produce hypersaline conditions under periods of prolonged drought or more freshwater conditions with increased storm events.
-Seagrass species may respond differently under individual or multiple stressors and each watershed will respond differently to climate change [@lefcheck2017; @zhang2023], suggesting place-based empirical assessments will be needed to properly inform management decisions.
+Seagrass species may respond differently under individual or multiple stressors and each watershed will respond differently to climate change [@hall2016; @lefcheck2017; @zhang2023], suggesting place-based empirical assessments will be needed to properly inform management decisions.
 
@@
 
-Two primary sources of data have been used to track seagrass change over time in Tampa Bay (@tbl-dattab).
+Two primary sources of data have been used to track seagrass change in Tampa Bay (@tbl-dattab).
 The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s (Figure [-@fig-seagrasschg]a, available at <https://data-swfwmd.opendata.arcgis.com/>).
@@
 Regional precipitation summaries were obtained directly from the SWFWMD (https://www.swfwmd.state.fl.us/resources/data-maps/rainfall-summary-data-region).
-Only rainfall data for the wet season (June to September) were evaluated for trends, whereas the complete record was used to calculate the Standardized Precipitation Index [SPI, @begueria2013] to identify periods of time when rainfall significantly deviated from the long-term average [using the *spei* R package, @begueria23].
+Only rainfall data for the wet season (June to September) were evaluated for trends, whereas the complete record was used to calculate the Standardized Precipitation Index [SPI, @begueria2013] to identify periods of time when rainfall deviated from the long-term average [using the *spei* R package, @begueria23].
 All climate data were evaluated annually with simple linear regression trends to assess change over time.
@@
 Kendall tests were also used to evaluate changes over time for each month across years to determine when the trends were most pronounced seasonally (e.g., all January estimates across years, all February estimates, etc.).
-The percentage of stations in each bay segment with significantly increasing temperature or decreasing salinity trends were evaluated for each month.
 
@@
 Counts of the maximum continuous number of days each year that temperature was above or salinity was below a threshold were obtained from the daily GAM predictions.
-This was done at each of the 45 stations in the EPC data using temperature thresholds of 29, 30, and 31$^\circ$ C and salinity thresholds of 15, 20, and 25 ppt.
+This was done at each of the 45 stations in the EPC data using temperature thresholds of 29, 30, and 31 $^\circ$C and salinity thresholds of 15, 20, and 25 ppt.
 The number of days when both temperature was above and salinity was below the thresholds was also estimated as a combined potential stress measure.
@@
 airtempmod <- lm(temp ~ yr, airtemp)
+airtempsiz <- paste('n =', nrow(airtempmod$model))
 airtempslo <- round(coefficients(summary(airtempmod))[2, 1], 2)
@@
 prpmod <- lm(prptot ~ yr, prp)
+prpsiz <- paste('n =', nrow(prpmod$model))
 prpslo <- round(coefficients(summary(prpmod))[2, 1], 2)
@@
 prpmod2 <- lm(prptot ~ yr, prp2)
+prpsiz2 <- paste('n =', nrow(prpmod2$model))
 prpslo2 <- round(coefficients(summary(prpmod2))[2, 1], 2)
@@
 
-seponly <- speidat %>% 
-  select(yr, mo, precip_mm) %>%
-  filter(mo %in% 9)
-sepmod <- lm(precip_mm ~ yr, seponly)
-seppvl <- p_txt(coefficients(summary(sepmod))[2, 4])
-seprsq <- round(summary(sepmod)$adj.r.squared, 2)
 dryonly <- speidat %>% 
@@
 drymod <- lm(prptot ~ yr, dryonly)
+drysiz <- paste('n =', nrow(drymod$model))
 drypvl <- p_txt(coefficients(summary(drymod))[2, 4])
 dryrsq <- round(summary(drymod)$adj.r.squared, 2)
-apronly <- speidat %>% 
-  select(yr, mo, precip_mm) %>%
-  filter(mo %in% 4)
-aprmod <- lm(precip_mm ~ yr, apronly)
-aprpvl <- p_txt(coefficients(summary(aprmod))[2, 4])
-aprrsq <- round(summary(aprmod)$adj.r.squared, 2)
+dryslo <- round(abs(coefficients(summary(drymod)))[2, 1], 2)
 ```
@@
 Long-term meteorological data showed increasing trends for air temperature and precipitation ([@fig-meteowqraw]).
-Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`).
+Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtempsiz`, `r airtemppvl`, $R^2$ `r airtemprsq`).
 Mean annual air temperature in 1975 was `r airtempstr` $^\circ$C, whereas current mean annual air temperature in 2022 was `r airtempend` $^\circ$C, showing an overall increase in the period of record of `r airtempdif` $^\circ$C.
-Similarly, total precipitation during the rainy season has increased by `r prpslo` mm per year, although the model was not significant (`r prppvl`).
-Removing September from the rainy season showed a significantly increasing trend of `r prpslo2` mm per year (`r prppvl2`, $R^2$ = `r prprsq2`; September alone did not have a significant trend, `r seppvl`), suggesting precipitation increases were driven by the earlier months (June - August).
-Using this model, mean precipitation in 1975 was `r prpstr2` mm, whereas current mean precipitation in 2022 was `r prpend2` mm, showing an overall increase in the period of record of `r prpdif2` mm.
-Notably, trends for the dry season (October through May) were not significant (`r drypvl`), except April which had a significantly increasing trend (`r aprpvl`, $R^2$ = `r aprrsq`).
-The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and 2015-2020.
+Similarly, total precipitation during the rainy season has increased by `r prpslo` mm per year, although the trend was weak (`r prpsiz`, `r prppvl`, $R^2$ = `r prprsq`).
+Removing September from the rainy season showed a slightly larger trend of `r prpslo2` mm per year and slightly more powerful trend model (`r prpsiz2`, `r prppvl2`, $R^2$ = `r prprsq2`), suggesting precipitation increases were driven by the earlier months (June - August).
+Using this model, mean precipitation for June to August in 1975 was `r prpstr2` mm, whereas current mean precipitation for June to August in 2022 was `r prpend2` mm, showing an overall increase in the period of record of `r prpdif2` mm.
+Notably, rainfall during the dry season (October through May) has decreased slightly over time at `r dryslo` mm per year, although the trend model was weak (`r drysiz`, `r drypvl`, $R^2$ = `r dryrsq`) The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and 2015-2020.
 
 Increasing water temperature and decreasing salinity generally followed the meteorological trends for all three *in situ* datasets (EPC, FIM, and PDEM, @fig-meteowqraw, Tables [-@tbl-temptrndtab], [-@tbl-saltrndtab], Figures S1, S2, S3).
-Note that for tables [-@tbl-temptrndtab] and [-@tbl-saltrndtab], comparable time periods were evaluated between the datasets when possible given the different sample sizes, and therefore power, to detect trends.
-The strongest trends were observed for the EPC dataset.
+Note that for tables [-@tbl-temptrndtab] and [-@tbl-saltrndtab], comparable time periods were evaluated between the datasets when possible given the different sample sizes, and therefore power, to detect trends (starting year `r lintrndsn_txt(lintrndsn)`).
+The strongest trends were observed for the EPC dataset based on the standard errors for the slope estimates.
 The top and bottom water temperature or salinity changes were similar across bay segments (@fig-meteowqraw).
-Trends in water temperature were similar across bay segments with significant increases for all four bay segments varying from 0.03 to 0.04 $^\circ$C per year, with a total change from 1.3 (OTB) to 1.7 (HB) $^\circ$C across the period of record from 1975 to 2022 (@tbl-temptrndtab).
+Trends in water temperature were similar across bay segments with large increases for all four bay segments varying from 0.03 to 0.04 $^\circ$C per year, with a total change from 1.3 (OTB) to 1.7 (HB) $^\circ$C across the period of record from 1975 to 2022 (@tbl-temptrndtab).
 Salinity trends were also similar between bay segments, although overall salinity was predictably higher for bay segments closer to the Gulf of Mexico.
-Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, with a total decrease of 2.6 (OTB) and 1.6 (LTB) ppt from 1975 to present.
-
-Significant increasing temperature trends were also observed for the FIM dataset from 1996 to present in OTB and from 2004 to present in HB (@tbl-temptrndtab).
-No significant temperature or salinity trends were observed in the other bay segments for the FIM and PDEM datasets from 1996 to present nor from 2004 to present (Figures S2, S3).
-However, the EPC dataset showed significant increasing temperature across all time periods and bay segments.
-Salinity trends from the EPC dataset were not significant from 1996 to present, but were significantly decreasing from 2004 to present in all bay segments (@tbl-saltrndtab).
+Only Old Tampa Bay and Lower Tampa Bay had notable decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, with a total decrease of 2.6 (OTB) and 1.6 (LTB) ppt from 1975 to present.
+Increasing temperature trends and decreasing salinity trends were also observed for the FIM and PDEM dataset, although confidence in the slope estimates (larger standard errors) generally decreased for shorter time periods of assessment.
 
 The EPC dataset was also used to provide detailed information on station-level trends from 1975 to present (@fig-kendall, see Figure S4 for 1998 to 2022).
-All stations had significantly increasing temperature and decreasing salinity from 1975 to present, although several stations in HB did not have significant salinity trends, the exception being the northernmost stations near the mouth of the Hillsborough Rive (Figure [-@fig-kendall]a).
-Seasonally, most bay segments had more significant stations during the summer, early fall period for both temperature and salinity (Figure [-@fig-kendall]b, c), although some variation was observed throughout the bay.
-Temperature trends were more often observed in the summer or early fall for the upper bay segments (OTB, HB), whereas the lower bay segments (MTB, LTB) had more seasonally homogeneous trends.
-Seasonal trends in salinity did not vary as much between segments, although OTB and LTB had more stations with significantly decreasing trends than the other bay segments.
+All stations had increasing temperature and decreasing salinity from 1975 to present, although several stations in HB did not have weak salinity trends (Figure [-@fig-kendall]a).
+Seasonally, most bay segments had more stronger trends in the early fall/winter periods for both temperature and salinity (Figure [-@fig-kendall]b), although some variation was observed throughout the bay.
+Temperature trends were generally stronger in the fall for OTB, whereas the remaining bay segments showed the strongest trends in the winter (February).
+Seasonal trends in salinity showed the largest decreases in the fall following the rainy season, with trends being especially strong in OTB.
+Small increases in salinity were observed in the spring for all but LTB.
 
@@
 Likewise, the mean number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record were similar between bay segments, with an average increase of 48 days across the bay segments for the period of record (@tbl-mixdaytab, Table S5, see Table S6 for 1998 to 2022).
-The increase in the number days each year when temperatures were above 29 or 31 $^\circ$C was similar.
+The increase in the number of days each year when temperatures were above 29 or 31 $^\circ$C was similar.
 
 The salinity models were less similar between bay segments compared to the temperature models, primarily because of the natural salinity gradient along the bay's longitudinal axis (Tables S3, S4).
-None of the bay segments had significantly increasing number of days per year when salinity was below 15 ppt.
-The only significantly increasing model for MTB was the number of days when salinity was below 25 ppt, which was increasing at a rate of 0.5 days per year.
-Both HB and OTB had significantly increasing number of days per year when salinity was below 20 and 25 ppt, although the slopes varied such that the rates in OTB were more than twice those in HB.
-Some salinity models had significantly decreasing trends, although the total number of days at the beginning and end of the period of record were neglible (Table S5).
+None of the bay segments had strong trends in the number of days per year when salinity was below 15 ppt.
+Both OTB and HB had models showing an increasing number of days when salinity was below 20 or 25 ppt, whereas MTB only showed an increase in the number of days when salinity was below 25 ppt.
+LTB had no notable salinity models.
+Some salinity models had decreasing trends, although the total number of days at the beginning and end of the period of record were neglible (Table S5).
 Overall, OTB showed the largest increase in the number of days each year when salinity was below a threshold, particularly for 25 ppt, where the change was 86 days per year from 1975 to 2022 (130 days to 216 days per year, @tbl-mixdaytab, Table S5, see Table S6 for 1998 to 2022).
@@
 Our assessment evaluated non-parametric trends (i.e., Kendall tests, less sensitive to outliers) at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to their larger spatial area and model domain.
-Our results showing increases in temperature and decreases in salinity in the summer, early fall, most notably for OTB and northern stations of HB (@fig-kendall), are likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.
+Our results showing increases in temperature and decreases in salinity in the fall, winter period provided a finer-scale comparison, where trends were most notable for OTB and northern stations of HB (@fig-kendall),  likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.
 These upper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to more rapid warming with shallower depths.
@@
 testudinum* has been attributed to altered hydrology and drought-induced hypersaline conditions, and sea level rise is expected to further modify salinity dynamics in the region [@hall2016].
+However, elevated summertime temperatures were also implicated in the decline.
 @dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters.
@@
 GAMs provided partially-supporting evidence that these changes can be linked to recent seagrass losses.
-Future analyses may show more significant associations between physicochemical habitat conditions and seagrass change as the trends are very likely to continue to push seagrasses further outside of their tolerance ranges.
+Future analyses may show stronger associations between physicochemical habitat conditions and seagrass change as the trends are very likely to continue to push seagrasses further outside of their tolerance ranges.
 These analyses should be supported by additional data collection efforts, particularly high-resolution continuous monitoring data that provide a more precise assessment of diurnal stress across multiple time-scales.
@@
 #| label: tbl-temptrndtab
-#| tbl-cap: "Bottom temperature trends ($^\\circ$C) by bay segments and datasets.
-Start year describes the range of the trend test to the present year (2022).
+#| tbl-cap: !expr 'paste0("Bottom temperature trends ($^\\circ$C) by bay segments and datasets.
+Change per year is the model slope with standard error in parentheses.
+Start year describes the range of the trend test to 2022 (", lintrndsn_txt(lintrndsn), ").
 The starting value is the estimated temperature at the start year and the end value is the estimated temperature in 2022.
@@
 Note that PDEM includes data only for the western portion of Old Tampa Bay from 2004 to present and EPC is the only dataset beginning prior to 1996.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay."
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.")'
 temptrndtab
@@
 #| label: tbl-saltrndtab
-#| tbl-cap: "Bottom salinity trends (ppt) by bay segments and datasets.
-Start year describes the range of the trend test to the present year (2022).
+#| tbl-cap: !expr 'paste0("Bottom salinity trends (ppt) by bay segments and datasets.
+Change per year is the model slope with standard error in parentheses.
+Start year describes the range of the trend test to 2022 (", lintrndsn_txt(lintrndsn), ").
 The starting value is the estimated salinity at the start year and the end value is the estimated salinity in 2022.
@@
 Note that PDEM includes data only for the western portion of Old Tampa Bay from 2004 to present and EPC is the only dataset beginning prior to 1996.
-OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay."
+OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.")'
 saltrndtab

### 2024-05-09 R/dat_proc.R

@@ -115,10 +115,10 @@ raindat <- readxl::excel_sheets(here('data-raw/swfwmdrainfall.xlsx')) %>%
     })
   ) %>% 
   unnest('data') %>% 
-  mutate(
-    precip_in = rowSums(select(., -mo, -yr), na.rm = TRUE)
-  ) %>%
-  select(mo, yr, precip_in) %>% 
+  # mutate(
+  #   precip_in = rowSums(select(., -mo, -yr), na.rm = TRUE)
+  # ) %>%
+  select(mo, yr, precip_in = tampacoastal_in) %>% 
   mutate(
     mo = gsub('\\-usgsbsn$', '', mo),
     mo = as.numeric(factor(mo,
@@ -1579,4 +1579,112 @@ sgmodsum <- sgmods %>%
 
 save(sgmodsum, file = here('data/sgmodsum.RData'))
 
+# linear temp, sal trends all data ------------------------------------------------------------
 
+data(fimsgtempdat)
+data(pincotemp)
+
+epctmp <- epcdat %>% 
+  select(bay_segment, epchc_station, SampleTime, yr, matches('Top|Bottom')) %>% 
+  filter(yr < 2023) %>% 
+  pivot_longer(names_to = 'var', values_to = 'val', matches('Top|Bottom')) %>% 
+  mutate(
+    var = factor(var, 
+                 levels = c(c("Sal_Top_ppth", "Sal_Bottom_ppth", "Temp_Water_Top_degC", "Temp_Water_Bottom_degC"
+                 )), 
+                 labels = c("Sal_Top", "Sal_Bottom", "Temp_Top", "Temp_Bottom")
+    ),
+    bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB'))
+  ) %>% 
+  separate(var, c('var', 'loc')) %>% 
+  mutate(
+    var = factor(var, levels = c('Temp', 'Sal'), labels = c('temp', 'sal'))
+  ) %>% 
+  filter(loc %in% 'Bottom') %>% 
+  filter(!is.na(val)) %>% 
+  summarise(
+    avev = mean(val, na.rm = T),
+    .by = c('bay_segment', 'yr', 'var') 
+  ) %>% 
+  mutate(
+    org = 'EPC'
+  )
+
+fimtmp <- fimsgtempdat %>% 
+  select(date, temp, sal, bay_segment) %>% 
+  mutate(
+    yr = year(date), 
+    mo = month(date)
+  ) %>% 
[…]

### 2024-05-09 R/funcs.R

@@ -274,3 +274,19 @@ epcgamplo_fun <- function(mod, smths, labels, cols){
   
 }
 
+# get n text for linear temp, sal trends by starting year
+lintrndsn_txt <- function(lintrndsn){
+ 
+  out <- unique(lintrnds[, c('yrstr', 'n')]) %>% 
+    na.omit() %>% 
+    mutate(
+      yrstr = paste0(yrstr, ': '),
+      n = paste('n =', n)
+    ) %>% 
+    unite('n', yrstr, n, sep = ' ') %>% 
+    pull(n) %>% 
+    paste(collapse = ', ') 
+  
+  return(out)
+  
+}

### 2024-05-09 R/tabs.R

@@ -25,7 +25,7 @@ totab <- tibble(
   Dataset = c('SWFWMD aerial maps', 'Transect data', 'EPC', 'FIM', 'PDEM', 'Tampa International Airport', 'SWFWMD precipitation'),
   Description = c('Seagrass coverage in acres', 'Seagrass frequency occurrence by species', 'Water quality monitoring samples', 'Nearshore temperature and salinity, seagrass species and cover', 'Water quality and seagrass presence/absence', 'Air temperature', 'Area-weighted precipitation for the wet season (June-September) for the Tampa Bay watershed'),
   Temporal = c('1988 - 2022, biennial', '1999 - 2022, annual', '1975 - 2022, monthly', '1996 to 2022, monthly', '2003 - 2022, monthly', '1975 - 2022, annual', '1975-2022, annual'),
-  Spatial = c('Whole bay', 'Whole bay, 62 transects', 'Whole bay, fixed sites', 'Whole bay nearshore, stratified random sites', 'Old Tampa Bay, stratified random sites', '27.979$^\\circ$N, 82.535$^\\circ$W', 'Whole watershed'),
+  Spatial = c('Whole bay', 'Whole bay, 62 transects', 'Whole bay, fixed sites', 'Whole bay shallow, stratified random sites', 'Old Tampa Bay, stratified random sites', '27.979$^\\circ$N, 82.535$^\\circ$W', 'Whole watershed'),
   Analysis = c('Biennial trends by bay segment, visual only', 'Annual trends by bay segment and species, comparison with temperature, salinity, and light attenuation as stressor metrics or observed data at annual scale', 'Trends in annual change and seasonal Kendall tests, estimate of stressor metrics as number of days above/below threshold', 'Trends in annual observed temperature, salinity, comparison to annual seagrass % cover', 'Trends in annual observed temperature, salinity, comparison to annual seagrass frequency occurrence', 'Annual trend', 'Annual trend')
 )
 
@@ -35,116 +35,19 @@ save(dattab, file = here('tabs/dattab.RData'))
 
 # linear trend summaries ----------------------------------------------------------------------
 
[…]

### 2024-05-10 manu-draft.qmd
remove ramp paragraph, fixes #18

@@
 Seagrass presence/absence is also recorded at each site and all sites were defined as "seagrass" if only seagrass species were identified (any with macroalgae were excluded) and "no seagrass" if bare sediment was observed.
-
-All of the organizations that provided water quality datasets participate in the Southwest Florida Regional Ambient Monitoring Program [@sherwood2016].
-This *ad hoc* group meets quarterly to ensure similar standards and methods are used for the collection and processing of surface water quality monitoring data in the region.
-Split-samples evaluated by each organization are also compared to assess precision between different laboratories.
-As such, the water quality measurements used herein are considered comparable, relative to the different sampling designs used by each program.
 

### 2024-05-10 manu-draft.qmd
clarify pdem bottom temp, sal used, clarification in figs, tab captions, fixes #19

@@
 Water quality samples at each site are similar to those collected by EPC but can occur in shallower locations.
+Only bottom temperature and salinity were used for analysis.
 Seagrass presence/absence is also recorded at each site and all sites were defined as "seagrass" if only seagrass species were identified (any with macroalgae were excluded) and "no seagrass" if bare sediment was observed.
@@
 A single smooth term was used for each predictor using a thin plate regression spline, including a tensor product interaction term with a cubic regression spline that evaluated the potential interacting effects of each predictor with year.
-Two models were evaluated, one with the temperature and salinity metrics together (@eq-epcmod1) and another with the both metric (@eq-epcmod2).
+Two models were evaluated, one with the bottom temperature and salinity metrics together (@eq-epcmod1) and another with the both metric (@eq-epcmod2).
 All models excluded Lower Tampa Bay because of minimal seagrass change over time.
@@
 \end{aligned}
-$$ {#eq-epcmod2} models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change.
-These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location).
+$$ {#eq-epcmod2}
+
+Separate models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change.
+These models used direct measurements of bottom salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location).
 GAMs were used for the FIM data to evaluate annual average seagrass percent cover using year, temperature, and salinity as independent variables.
@@
 
-![Example of mixed effects models for the estimated number of days per year that temperature (red) or salinity (blue) were above or below thresholds of 30 $^\circ$C or 25 ppt, respectively, from the EPC data.
+![Example of mixed effects models for the estimated number of days per year that bottom temperature (red) or salinity (blue) were above or below thresholds of 30 $^\circ$C or 25 ppt, respectively, from the EPC data.
 The bottom row (black) shows the number of days when both temperature was above and salinity was below the thresholds.
@@
 
-![Partial effects of smoothers (rows) by bay segment (columns) from a Generalized Additive Model used to describe seagrass change relative to year, light attenuation, the number of days each year when temperature was above 30 $^\circ$C, and the number of days each year when salinity was below 25 ppt.
+![Partial effects of smoothers (rows) by bay segment (columns) from a Generalized Additive Model used to describe seagrass change relative to year, light attenuation, the number of days each year when bottom temperature was above 30 $^\circ$C, and the number of days each year when bottom salinity was below 25 ppt.
 The EPC data were used for the independent variables in the model.
@@
 
-![Partial effects of smoothers (rows) by bay segment (columns) from Generalized Additive Models evaluating seagrass changes versus year, temperature, and salinity for the (a) FIM (first three columns evaluating mean annual percent cover from 0-100, `r modtxt_fun(sgmodsum)$fimmod`) and (b) PDEM (right column evaluating annual frequency occurrence from 0-1 in OTB only, `r modtxt_fun(sgmodsum)$pincomod`) datasets.
+![Partial effects of smoothers (rows) by bay segment (columns) from Generalized Additive Models evaluating seagrass changes versus year, bottom temperature, and bottom salinity for the (a) FIM (first three columns evaluating mean annual percent cover from 0-100, `r modtxt_fun(sgmodsum)$fimmod`) and (b) PDEM (right column evaluating annual frequency occurrence from 0-1 in OTB only, `r modtxt_fun(sgmodsum)$pincomod`) datasets.
 Partial effects describe the modeled association between each predictor and seagrass change after accounting for the effects of the other predictors.
@@
 #| label: tbl-mixdaytab
-#| tbl-cap: "Summary of mixed-effects models evaluating increases in the number of days each year from 1975 to 2022 when temperature was above 30 $^\\circ$C, salinity was below 25 ppt, or both temperature and salinity were above/below the thresholds.
+#| tbl-cap: "Summary of mixed-effects models evaluating increases in the number of days each year from 1975 to 2022 when bottom temperature was above 30 $^\\circ$C, bottom salinity was below 25 ppt, or both temperature and salinity were above/below the thresholds.
 The start and end columns show the estimated number of days (mean, standard error) in 1975 and 2022 when temperature or salinity were above or below the thresholds.

### 2024-05-10 supplement.qmd
clarify pdem bottom temp, sal used, clarification in figs, tab captions, fixes #19

@@
 
-![Trends from 1998 to 2022 for temperature and salinity measurements at long-term monitoring stations in Tampa Bay.
+![Trends from 1998 to 2022 for bottom temperature and salinity measurements at long-term monitoring stations in Tampa Bay.
 Results for seasonal Kendall tests by station and monitoring location (top or bottom of the water column) are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr$^{-1}$).
@@
 
-![Example of mixed effects models for the estimated number of days per year that temperature (red) or salinity (blue) were above or below thresholds of 30 $^\circ$C or 25 psu, respectively, from 1988 to 2022.
+![Example of mixed effects models for the estimated number of days per year that bottom temperature (red) or salinity (blue) were above or below thresholds of 30 $^\circ$C or 25 psu, respectively, from 1988 to 2022.
 The bottom row (black) shows the number of days when both temperature was above and salinity was below the thresholds.
@@
 
-![Partial effects of smoothers (rows) by bay segment (columns) from a Generalized Additive model used to describe seagrass change relative to year, light attenuation, and the number of days each year when both temperature was above 30 $^\circ$C and salinity was below 25 ppt.
+![Partial effects of smoothers (rows) by bay segment (columns) from a Generalized Additive model used to describe seagrass change relative to year, light attenuation, and the number of days each year when both bottom temperature was above 30 $^\circ$C and bottom salinity was below 25 ppt.
 The EPC data were used to develop the model.
@@
 
-Table S3: Summary of mixed-effects models evaluating increases in the number of days each year from 1975 to 2022 when temperature or salinity were above or below critical thresholds, respectively.
+Table S3: Summary of mixed-effects models evaluating increases in the number of days each year from 1975 to 2022 when bottom temperature or salinity were above or below critical thresholds, respectively.
 The slope estimates in the "Both" column indicate trends in the number of days when both temperature and salinity were above or below critical thresholds.
@@
 
-Table S4: Summary of mixed-effects models evaluating increases in the number of days each year from 1998 to 2022 when temperature or salinity were above or below critical thresholds, respectively.
+Table S4: Summary of mixed-effects models evaluating increases in the number of days each year from 1998 to 2022 when bottom temperature or salinity were above or below critical thresholds, respectively.
 The slope estimates in the "Both" column indicate trends in the number of days when both temperature and salinity were above or below critical thresholds.
@@
 
-Table S5: Number of days at the beginning and end of the period of record when temperature or salinity were above or below critical thresholds, respectively.
+Table S5: Number of days at the beginning and end of the period of record when bottom temperature or salinity were above or below critical thresholds, respectively.
 Values are the estimated mean number of days (plus standard error) from 1975 to 2022.
@@
 
-Table S6: Number of days at the beginning and end of the period of record when temperature or salinity were above or below critical thresholds, respectively.
+Table S6: Number of days at the beginning and end of the period of record when bottom temperature or salinity were above or below critical thresholds, respectively.
 Values are the estimated mean number of days (plus standard error) from 1998 to 2022.

### 2024-05-10 R/tabs.R

@@ -324,18 +324,18 @@ save(supp2daytab, file = here('tabs/supp2daytab.RData'))
 load(file = here::here('data/sgmods.RData'))
 load(file = here::here('data/sgmodsum.RData'))
 
-suppepccap1 <- paste0('Summary of smoother terms in the Generalized Additive Model used to evaluate seagrass response for the EPC data in relation to temperature (temp) and salinity (sal) stress metrics, with additional smoothers for year (yr) and light attenuation (la).  Separate smoothers were fit for each bay segment.  $s$ = invidual smoother, $ti$ = interaction term. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay. ', modtxt_fun(sgmodsum)$epcmod1, '.')  
+suppepccap1 <- paste0('Summary of smoother terms in the Generalized Additive Model used to evaluate seagrass response for the EPC data in relation to bottom temperature (temp) and salinity (sal) stress metrics, with additional smoothers for year (yr) and light attenuation (la).  Separate smoothers were fit for each bay segment.  $s$ = invidual smoother, $ti$ = interaction term. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay. ', modtxt_fun(sgmodsum)$epcmod1, '.')  
 suppepcmod1tab <- gam_table(sgmods$epcmod1, cap = suppepccap1)
 save(suppepcmod1tab, file = here('tabs/suppepcmod1tab.RData'))
 
-suppepccap2 <- paste0('Summary of smoother terms in the Generalized Additive Model used to evaluate seagrass response for the EPC data in relation to "both" stressors (both temperature above and salinity below thresholds), with additional smoothers for year (yr) and light attenuation (la).  Separate smoothers were fit for each bay segment.  $s$ = invidual smoother, $ti$ = interaction term. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay. ', modtxt_fun(sgmodsum)$epcmod2, '.')    
[…]

### 2024-05-14 manu-draft.qmd
DK and BF revision comments added

@@
 The monitoring and management of seagrasses in coastal environments has received substantial attention on a global scale.
-Seagrasses are fundamental indicators of coastal ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023].
+Seagrasses are fundamental indicators of coastal ecosystem health [@roca2016; @orth2017], while also serving as foundation species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023].
 Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021].
-These losses are comparable to, if not more significant than, other interconnected critical coastal environments such as mangroves and salt marshes [@duarte2008a], which threatens the natural and economic services that these habitats provide.
+These losses are comparable to, if not more significant than, other interconnected critical coastal environments such as mangroves and salt marshes [@duarte2008a].
 Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @hall1999; @orth2006; @burkholder2007].
@@
 In other cases, the reversal of seagrass losses through nutrient reductions alone may not be possible because of system hysteresis, where the path to recovery is not the same as the path to decline [@maxwell2016].
+Top-down effects and associated trophic cascades can also complicate the sustained growth or recovery of seagrass, particularly where herbivores are abundant [e.g., sea turtles, manatees, @heck2007; @fourqurean2010].
 Complementary management actions, in addition to nutrient reductions, are needed in these situations.
@@
 In Chesapeake Bay, the abundance of widgeongrass (*Ruppia maritima*) has responded positively to nutrient reduction and is replacing the formerly dominant but now heat-stressed [*Z.
-marina*, @hensel2023].
+marina*, @hensel2023; also see @bartenfelder2022].
 Changing frequency and severity of precipitation patterns may further alter salinity regimes and with it the distribution and abundance of seagrasses throughout the estuary [@rasheed2011; @webster2021].
@@
 Since 2016, seagrass areal cover has decreased by nearly 1/3 despite relatively stable water quality [@tbep0696; @Beck20a].
-Factors that have influenced this recent decline are unknown and the effects of climate change drivers on physical water quality conditions have been implicated as potential stressors.
+Factors that have influenced this recent decline are unknown and the effects of climate change drivers on physical water quality conditions independent of light environments have been implicated as potential stressors.
 Following global trends, recent work has demonstrated a broad long-term trend of increasing water temperature in Tampa Bay, although at a relatively coarse scale [@nickerson2023].
@@
 This information can fill a critical knowledge gap that can inform regional management activities, while also demonstrating the confounding effects of climate change with ecosystem response to conventional stressors.
+The improvement of bio-optical models that can describe light requirements for seagrasses may also result from this information.
 
@@
 Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia Rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small, channelized tributaries and manmade flood control conveyances [@tbep0623].
-Notably, Old Tampa Bay has restricted circulation from multiple land bridges associated with causeways that traverse the bay, causing relatively longer residence times and accumulation of legacy pollutant loads compared to the other bay segments [@sherwood2015; @luther2022].
+Notably, Old Tampa Bay has restricted circulation from multiple land bridges associated with causeways that traverse the bay, causing longer residence times and accumulation of legacy pollutant loads compared to the other bay segments [@sherwood2015; @luther2022].
 Recurring seasonal harmful algal blooms of the dinoflagellete *Pyrodinium bahamense* have contributed to exceedances of the chlorophyll-a regulatory standard in Old Tampa Bay [@lopez2023].
 By comparison, water quality conditions in Middle Tampa Bay and Lower Tampa Bay are generally better than the upper two bay segments primarily from more frequent water exchanges with the Gulf of Mexico and lower nutrient loading [@tbep0623].
-All bay segments are relatively shallow, with a baywide mean depth of approximately 3 m.
+All bay segments are shallow, with a baywide mean depth of approximately 3 m.
 Light penetration typically reaches bottom habitats under current conditions (Figure [-@fig-seagrasschg]c), although seagrasses were historically limited by high phytoplankton production that affected light environments [@greening2014; @johansson15].
@@
 The long-term recovery of seagrass habitats in Tampa Bay since the 1980s is a nationally recognized success story that demonstrates application of a successful management paradigm through the EPA-administered National Estuary Program [@greening2006; @greening2014; @sherwood2017].
-From 1988 to 2016, seagrasses increased 79% to 41,655 acres (16,857 ha), surpassing the regional management goal of restoring coverage to 95% of a benchmark 1950s coverage.
-Though Tampa Bay was far from pristine in the 1950s, aerial imagery was sufficient to estimate a relatively unimpacted condition for seagrass coverage.
-Since that time, the greatest areal coverage expansions were observed in the upper bay segments (OTB, HB, and MTB; Figure [-@fig-seagrasschg]a).
+From 1988 to 2016, seagrasses increased 79% to 41,655 acres (16,857 ha), surpassing the regional management goal of restoring coverage to 95% of what was present in the 1950s.
+Though Tampa Bay was far from pristine at the time, aerial imagery was sufficient to estimate a relatively unimpacted condition for seagrass coverage.
+Since then, the greatest areal coverage expansions were observed in the upper bay segments (OTB, HB, and MTB; Figure [-@fig-seagrasschg]a).
 Light environments have improved (Figure [-@fig-seagrasschg]c) through reduction of external nitrogen loadings by about 2/3 by the 1980s from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed [@greening2014; @beck2019].
 
-From 2016 to present, dramatic seagrass loss has been observed in Tampa Bay, despite light environments remaining supportive of growth (Figure [-@fig-seagrasschg]c).
+From 2016 to present, dramatic seagrass loss has been observed in Tampa Bay, despite light environments remaining supportive of growth (Figure [-@fig-seagrasschg]c) as defined by historical relationships between nitrogen, chlorophyll-a, and water clarity [@tbep0696; @greening2011].
 Total cover in Tampa Bay has decreased by 28% (11,518 acres/4,661 ha) from the 2016 peak to a total baywide coverage of 30,137 acres (12,196 ha) in 2022.
@@
 The Fisheries Independent Monitoring (FIM) program administered by FWC provides monthly surveys of the entire nekton community in Tampa Bay, including species richness and abundance, using multiple gear types that target different habitats [@schrandt2021].
-A stratified sampling design is used to select sites for 21.3 meter center-bag seines that target shallow habitats (\<1.5 m) where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present, Figure [-@fig-map]c).
-In addition to collecting fish and selected invertebrates, *in situ* physical measurements for water temperature and salinity are collected at the bag, and at the surface and at 1 m intervals to the bottom.
+A stratified sampling design is used to select sites for 21.3-meter center-bag seines that target shallow habitats (\<1.5 m) where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present, Figure [-@fig-map]c).
+In addition to collecting fish and selected invertebrates, *in situ* physical measurements for water temperature and salinity are collected at the bag, and at the surface and at 1-m intervals to the bottom.
 Only measurements from the bottom were used.
@@
 A fundamental challenge describing niche space is identifying the boundaries for optimal conditions.
-In Tampa Bay, three dominant seagrass species occur: *H.
-wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinum* (turtle grass) [@lewis1985; @phillips88].
-Other less common species include *Ruppia maritima* (widgeon grass) and *Halophila engelmanii* (star grass), where the former is often mapped during wet years in the upper bay segments from the aerial surveys.
+In Tampa Bay, three dominant seagrass species occur: *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinum* (turtle grass) [@lewis1985; @phillips88].
+Other less common species include *Ruppia maritima* (widgeon grass) and *Halophila engelmannii* (star grass), where the former is often mapped during wet years in the upper bay segments from the aerial surveys.
 These species co-occur often in mixed beds throughout the bay, although some differences in abundance are observed across salinity ranges.
@@
 Our inclusion of light attenuation in the EPC models was meant to account for how the light environment may be influencing seagrass growth, in addition to climate-related stressors.
-However, light attenuation has improved over the period of record and within the limits estimated to be supportive of seagrass growth in Tampa Bay [@tbep0400], particularly in OTB where the most loss occurred (Figure -[@fig-seagrasschg]c).
+However, light attenuation has improved over the period of record and within the limits estimated to be supportive of seagrass growth in Tampa Bay [@tbep0696; @greening2011], particularly in OTB where the most loss occurred (Figure [-@fig-seagrasschg]c).
 Additional water quality parameters could be included in the models to provide further evidence that light-limitation is not the present driver for seagrass change (e.g., nitrogen loading, chlorophyll-a, color), although nutrient management for the benefit of seagrass growth will likely continue to be a dominant management paradigm for Tampa Bay.
@@
 Ongoing work in OTB using continuous data loggers in shallow areas will provide insights into short-term diurnal changes as potential acute temperature stress on seagrass (see <https://tbep-tech.github.io/otb-temp/tempeval>).
-Morpohological or physiological measurements at the individual level could also provide early indications of heat and osmotic stress.
+Morpohological or physiological measurements at the individual level could also provide early indications of heat and osmotic stress [@congdon2023].
 

### 2024-05-20 revision/review_response.qmd
added revisions doc to revisions folder

---
format: 
  docx:
    reference-doc: response_styles.docx
bibliography: ../refs.bib
csl: ../estuaries-and-coasts.csl

execute: 
  echo: false
  warning: false
  message: false
---

```{r}
#| eval: false
spelling::spell_check_files(here::here('revision/review-response.qmd'))
```

*We sincerely thank the associate editor and all reviewers for providing useful comments on our manuscript. We have made every effort to address these concerns. In particular, we have provided a more thorough assessment of seagrass response to climate-related stressors using more flexible models (GAMs) and by including additional predictors (light attenuation). Please see the specific comments below. Note that the line numbers relate to the original draft and not the revised version.*

### AE comments:

I enjoyed reading this paper and feel it could be an important contribution, especially given its strength in combining many different datasets and analyses to build a comprehensive story and set of conclusions. However, there are some outstanding issues that need to be addressed prior to publication, mainly related to the analyses. My impression is these can be addressed, and so I encourage the authors to consider my points below and those of the reviewers, as they will result in a stronger contribution.

*Response: Thank you for the comments. We have revised the text based on your suggestions below. Please see our specific comments to each response for more details.*

Major comments:

1\. Datasets: I appreciate the use of multiple datasets and the effort taken to conduct multiple analyses that draw on the strengths of these datasets that build towards a comprehensive study conclusion. I did, however, have trouble following which datasets were used for which analysis, and why, and what was included and omitted. I realize all that information is included in the text, but I found it was quite buried and when I tried to refer back from figures and tables to the text it was difficult to find the information. I would suggest inserting a table with this information summarized so that the reader can easily refer to this.

*Response: A new table (Table 1) was added that includes a description of each dataset, the temporal and spatial coverage, and the analyses used for each.*

2\. Justification for period of decline: I struggle with the assertion that 2016 represents the beginning of the dramatic seagrass loss across Tampa Bay, as stated on Line 183. This is especially problematic when observing seagrass coverage in Figure 2a, which represents total species coverage. Here, coverage is relatively stable 1988 to 2012, then increases 2014-16, with levels dropping after this. The extent of these patterns differs per bay segment – in OTB, there is a dramatic increase with some decrease, although only the last year (2022) is below the baseline values 1988 to 2012. For other segments, there is a decline but coverage post 2016 is either similar to the baseline or higher. What are the drivers of those large increases in more recent years, and are declines relevant or is seagrass coverage simply returning to baselines levels that were evident 1988-2012? Please comment in the text. Further, the decline in cover is more evident when observing species specific data, particularly for H. wrightii in OTB and MTB, but not other segments. Should the analyses focus on the dominant species driving these patterns? Please revise text and acknowledge these aspects, perhaps reanalyse the data or interpretation if necessary, to present a more convincing justification.

*Response: The above interpretation is correct in that the patterns of increase/decrease vary by bay segment and dataset (coverage or frequency occurrence). This suggests that multiple factors are influencing these changes and that they potentially vary by location. Our development of more robust models (GAMs) that attempted to better describe these patterns is an explicit acknowledgement of these differences, which were not adequately characterized with the earlier models. Please see our description below of these updated models.*

*Additional text was also added to clarify these nuanced changes and the focus on total species occurrence.*

*Line 191: “As such, trajectories of recovery and decline have varied by bay segment in magnitude and timing of the change, although a consistent decline baywide has been observed since 2016. Change by species has been most notable for Halodule wrightii (shoal grass) in OTB and HB, whereas some losses have been observed for Thalassia testudinum (turtle grass) in LTB, likely related to dominance of each species across the salinity gradient and proximity to hydrologic inputs (Lewis et al. 1985).”*

*Line 364: “Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model and that the majority of seagrass loss in recent years was dominated by H. wrightii in the upper bay segments (Figure 2b).”*

3\. Exclusion of time series and use of pre/post category: Use of pre/post category in GLM analyses with stressors collapses the data across the time series and removes the interannual variability and information contained therein. I don’t understand why this was done instead of explicitly including the time series in the analyses. I would recommend conducting initial GAMs that includes the time series for these data. The GAM approach would be particularly useful to show the relationship smoothers with the stressor values and/or the basic temperature salinity time series (ie smoother relationships). As it stands, it seems that an initial key part of the analyses was omitted and it is not justified to bin into categories. This is my biggest concern with the analyses and I'd like to see this addressed.

*Response: We have revised our modeling approach for describing seagrass change using a GAM approach that uses time as a continuous variable (year) and the inclusion of light attenuation as a predictor when able. For the latter, light attenuation was included only in the EPC model, whereas it was excluded from the FIM and PDEM models due to a large majority of Secchi observations (from which light attenuation was derived) being on the bottom. As such, light attenuation could not be derived for these models, nor would it likely explain much variation in seagrass response since light environments are generally not limiting at most of the locations sampled in the FIM and PDEM datasets. Also note the addition of Figure 2c that shows the improvement in light attenuation over time as shown with the EPC data.*

*Sections 2.7 (links to seagrass) and 3.3 (seagrass response) were revised to describe the updated modeling approach, including the addition of equations 1 to 4 showing the structure for each GAM, Figures 7, 8 , and S7 showing partial effects for each model, and Tables S7 to S10 providing summary statistics for each model. Overall, the general conclusions that climate-related stressors are likely contributing factors to seagrass change remains as before, although the revised approach provides a more complete description of this change using more flexible models that can describe non-linear relationships and additional predictors (time, light attenuation).*

4\. Exclusion of nutrient/light data: A key part of the Tampa Bay story is the success in seagrass recovery through nutrient reduction. I agree with the reviewers that exclusion of these data (or some representative) is perplexing, especially given they are available. I recommend including these in the models, and excluding them only through the model selection process.

*Response: This is an excellent suggestion that we have carefully considered. Several revisions were made to address the concern. First, Figure 2 was updated to include a time series of light attenuation by bay segment over the last fifty years. Regression lines in each plot show that light environments have improved for the period of record covered by the transect monitoring data (1998 to present). This plot provides an overview of the history of bay recovery as well as the perplexing issue that seagrass has changed despite light attenuation being low and continuing to improve. It is also noted that light attenuation in recent years has been below the threshold to support seagrass growth. Second, we have improved our models of seagrass response to climate stressors as noted in the response to the previous comment.*

Minor comments:

Line 205: Define frequency occurrence.

*Response: Text revised as “...cover-abundance, frequency occurrence (number of sample points with seagrass divided by total points on a transect, as in Sherwood et al. 2017), and condition, ...”*

Line 207: My understanding is that both transects and areal maps were done once a year. Clarify here why transect data provided greater temporal resolution.

*Response: As noted on line 195, the areal maps are produced every other year, whereas the transect data are collected every year. The text was revised as follows: “Although the areal maps provide the standard for assessment of restoration goals, the maps are produced every other year. The transect data are collected each year, allowing inter-annual comparison...”*

Line 183: Provide reference or data to show that nutrient loading and light attenuation remained stable during 2016 to present. Better yet, include it in your analyses and exclude through model selection.

*Response: Please see the response to the general comments above.*

Line 368: remind reader why these datasets can’t be used to calculate stressor metrics (point samples?)

*Response: Sentence was revised as follows: “These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location).*

Figure 6: grey points are the actual number of days per year per transect (station), correct? Insert text.

*Response: Edited as “...grey points as actual number of days for each station...”*

### Reviewer #1:

This was a very thorough study, with an impressive amount of data. These types of analyses are greatly needed with all of the long-term datasets that are becoming available in different estuarine systems. The authors were extremely detailed in explaining their methods, and using different tools and models to analyze different datasets. The hang-up for me is the extremely weak, and I think really no convincing links to changes in seagrass. They acknowledge that it is weak and do not try to really oversell it though. However, I am unconvinced as to the omission of any kind of water clarity data into the seagrass models.

*Response: Thank you for the thorough review of our manuscript. Please see our response to your comments below, as well as those in response to the comments from the associate editor. We have re-evaluated seagrass change over time using a more comprehensive modeling approach that includes GAMs and additional predictors (i.e., light attenuation).*

The authors cite Lopez et al. in their site description of Old Tampa Bay. This paper shows that HAB durations have increased from 2011-2020. Since the largest seagrass declines occurred in Old Tampa Bay, could they be more directly related to algal blooms and less directly related to temperature and salinity changes? Based on the 2022 Tampa Bay Water Quality Assessments report that is cited, Old Tampa Bay has not met their chlorophyll criteria for over half the years since seagrass declines started being observed after 2016. In particular, the 3 years in a row from 2019-2021 seem of great interest to this study, and I'm wondering why chlorophyll wasn't included in the models.

*Response: We have included light attenuation in our models when able. Additionally, the FIM and PDEM datasets that target more shallow areas of the bay had a majority of Secchi observations on the bottom, indicating that light environments are generally supportive of seagrass growth in recent years. Also note the addition of Figure 2c that shows the long-term improvement in light environments over the period of record for the EPC dataset.*

Continuing with this same issue, I find the argument to leave out water clarity data fairly weak and at times based on anectodal evidence. For example, Lines 166-167 "Light penetration typically reaches bottom habitats under current conditions" and lines 183-184 "despite nutrient loading and light attenuation remaining relatively stable". What is the justification for these statements? It seems like there is anecdotal evidence used to try to defend the use of not looking at light conditions in the model, but many times citations are not used that show these statements to be true. If light is going to be excluded from the models, I think a stronger justification is needed.

*Response: Please see our response to the previous comment as well as those in response to the general comments from the associate editor.*

Lines 303-305 "we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have been relatively stable in recent years." Here a citation is given, however these mentioned conditions appear to be stable for areas except OTB, which is also the area that saw the greatest declines in seagrass. So again, why leave light availability out of the models?

*Response: Please see our above responses.*

Lines 225-226 "Most samples are collected from mid-morning to early afternoon". That is a major issue with using temperature data that is a snapshot and not continuous. Temperatures can vary greatly between morning to afternoon. Were any attempts made to normalize temp data to time of day? I don't know if it would be possible given the datasets, but something to consider.

*Response: This is certainly an issue worth considering using the temperature data from the monthly discrete samples and was the motivation for including a statement on the need for continuous monitoring data in the original draft (lines 667 - 669). Applying a correction or normalization for time of day would be challenging and we would not be confident in the results to accurately describe maximum or minimum temperature values for a given day. Moreover, we feel the trends in the temperature summaries (e.g., increase in number of days above a threshold over time) would likely be similar. As such, we have not attempted this for the current dataset. However, ongoing work in Old Tampa Bay for a separate project is attempting to describe short-term diurnal variation in temperature with continuous data loggers. We suspect this information will provide further insight into acute temperature stressors that could affect seagrass. We have added a statement and link in the text with a short description of this work, although these loggers are in very shallow areas that likely show different temperature characteristics than the discrete samples at deeper depths.*

*Line 669: “Ongoing work in OTB using continuous data loggers in shallow areas will provide insights into short-term diurnal changes as potential acute temperature stress on seagrass (see https://tbep-tech.github.io/otb-temp/tempeval).”*

Lines 313-314 "Shoal grass is tolerant of a wide range of salinity..." Shoal grass accounted for the largest seagrass declines, according to Figure 2b, so why would decreases in salinity negatively impact this species? Wouldn't a decrease of 0.04-0.06 per year be well within the tolerance range for this species?

*Response: Our previous statement on lines 560-564 suggests this may be a plausible explanation for the weak association of seagrass change with the stressors. The statement was further amended to specifically mention that shoal grass accounted for the largest seagrass decline: “This is especially true for H. wrightii that had the greatest changes over the period of record and is tolerant of a wide range of salinity.”*

Lines 350-352: Did you experiment with different lagged responses?

*Response: Yes, preliminary analyses evaluated lagged associations between the salinity and temperature stressor metrics and seagrass change. Specifically, we evaluated the number of days in 30 day periods from one month to one year when temperature was above or salinity was below a threshold in relation to seagrass change (see code here). No compelling differences were observed for any of the lags compared to the entire year, which was the justification for using the latter in the current analysis. Additional text was added to this effect:*

*Line 354: “Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the transect year summaries. As such, the stressor metrics...”*

Line 384: what was the justification for removing September?

*Response: The assessment of inter annual changes in precipitation during the rainy season showed a weak increase over time, whereas removing the last month (September) showed a more statistically powerful increase over time. The removal of September was meant to show that the trend was driven by an increase in the first three months of the rainy season (Jun - Aug). Text was added to line 386: “...suggesting precipitation increases were driven by the earlier months (June - August).” Also note that our previous estimates of rainfall total were based on an incorrect sum across overlapping areas. We have fixed the issue, although the results and trends remain the same.*

Line 399: "for all" is repeated twice

*Response: Fixed.*

Line 407: Table 1 shows only HB significant for FIM from 2004 to present, not OTB.

*Response: This was corrected.*

### Reviewer #2:

Overall, I found this manuscript to be interesting and useful. It utilizes existing data from various existing, governmental water quality monitoring programs of varying design and coverage to assess if recent declines in seagrass populations in portions of the subtropical Tampa Bay coastal system can be associated with increased water temperatures or lower salinities which might be a result of changing climatic conditions.

Although the results of their conservative, linear modeling analyses determined that the bay waters were only marginally warmer and fresher during the recent period of seagrass declines beginning in 2016, the study provides a good example of how existing monitoring data might be used for similar questions in other areas, as well as what new monitoring or assessment approaches, including evaluation of shorter-term periods of stress might be needed. It reinforces the need for updated coastal water quality management approaches and seagrass restoration criteria. Most existing coastal management restoration paradigms, programs and restoration criteria were developed and first implemented 20-40 years ago or earlier when declines in seagrass populations were first quantified. Consequently, they do not address many of the interactive and potentially negative factors associated with a rapidly changing climate.

The thorough discussion section addresses many of the caveats and limitations of their approach to assessing temporal trends in the Tampa Bay physical environment during the recent period of seagrass decline. Additionally, the authors discuss some of the potentially important unmeasured interactive biotic and abiotic factors which could be important in explaining recent seagrass declines.

Before acceptance I recommend that several points should be addressed to improved the manuscript's clarity and usefulness.

*Response: Thank you for providing useful feedback on our manuscript. Please see our detailed responses below.*

First, although light availability is one of the most principal water quality factors affecting seagrass performance, stability and resilience, data is not presented here to characterize the light environment either spatially or temporally, other than to suggest it has been relatively stable and consistent across the system. However, on L. 120 it is highlighted that "... upper bay segments where marginal light environments exist".

*Response: Line 120 was revised to describe that the upper bay segments are more shallow and receive a majority of hydrologic inflow. The previous was inaccurate. Also, please see the addition of Figure 2c that shows the long-term improvement in the light environment over time. Of particular note is that all bay segments are currently below the threshold supportive for seagrass growth.*

Additionally, on L. 183-184 the authors state, "...dramatic seagrass loss has been observed in Tampa Bay, despite nutrient loading and light attenuation remaining relatively stable (Figure 2a)". However, Figure 2a only shows seagrass abundance changes and Figures S1, S2, S3 only show 2000 to 2020 air and water temperatures, precipitation, and salinity levels and trends. If nutrients and light are to be considered constants and excluded from the analysis (as discussed L. 574-597) the data should be similarly presented in figures so that their exclusion can be better supported. One possibility would be to add to the supplementary materials in a format such as Fig. S1 or S2.

*Response: We have added light attenuation trends in Figure 2 and have included it as a predictor in the seagrass models when able. As noted above, a majority of Secchi observations in the FIM and PDEM datasets have been recorded on the bottom, indicating sufficient light environments for seagrass growth in recent years.*

L. 535-537 states that, "The models did not provide a consistent, nor statistically powerful, explanation that increasing temperature and decreasing salinity were key (or the sole) drivers." Therefore, a weight-of-evidence approach from all the models is used to conclude their effects. This is OK provided that the other potentially important interactive factors can be better supported as being constant. Monitoring data for chlorophyll concentrations nitrogen loadings and other parameters as they relate to light availability to the bottom do exist and should be included in the figures for each of the four regions.

*Response: Please see our above comments regarding the addition of these data in Figure 2 and our use of more robust modeling approaches to describe these potential changes (i.e., GAMs)*

Another important component excluded from analyses or discussion here is identification of the individual seagrass species distributions and their potential effects on seagrass responses or resilience to climate related stresses. Thalassia, which is typically dominant in the lower, higher salinity areas in this system can have greater persistence and resilience to environmental stresses than Halodule and other common seagrass species, however its recovery after loss can also be slow. Since it can grow to greater depths Halodule, its water clarity requirements can be higher. Additionally, at greater depths the effects of temporary air temperature increases on water temperatures can be buffered compared to shoal environments. If the seagrass community consists only of Thalassia then it should be so stated.

*Response: We agree there are likely important distinctions between species that could be considered when evaluating the response to climate-related stressors. As correctly noted by the reviewer, Thalassia is dominant in lower Tampa Bay with coverage decreasing towards the upper bay segments. We did not evaluate lower Tampa Bay in our analysis of seagrass response to climate stressors given that coverage has generally been stable in recent years and salinity has not been changing dramatically. For the remainder of the bay, most of the change in cover has been dominated by Halodule, especially in the upper bay segments where most of the seagrass loss has been observed. An evaluation of individual species response would likely not be more informative than the current evaluation of total frequency occurrence. Also please see our response to a similar comment from the associate editor.*

The definition of "hotter and fresher" is as used here summarized as the average rates of annual increases in water temperature and decreases in salinity. Yet, more importantly, this study presents estimates of the number of days annually when defined thresholds for water temperature above 30C and salinities below 25 occur. While these thresholds which were used for comparisons, were set, in part, on their statistical strength, they do have merit as indexes of episodic stress. These metrics perhaps more important than averages when related to evaluating climate change stressors as referenced here and elsewhere. I suggest these metrics be highlighted and included in the abstract, as they are included and emphasized in the discussion.

*Response: This is a good suggestion and we agree these metrics are useful to highlight in the abstract. We have added this sentence to the abstract, with similar clarification added in section 3.2 and the beginning of the discussion: “Additionally, the number of days when temperature was above 30 °C or salinity was below 25 ppt has increased on average across all bay segments by 48 and 37 days, respectively, since 1975.”*

Additionally, recommendations for future management efforts highlighted in the last sentence of the Abstract (L. 31-34) do not mention recommendations for additional monitoring approaches needed to evaluate future climate stressors as highlighted in the Conclusions. I suggest adding something such as, "...and high-resolution data collection efforts", after "additional, complementary interventions..." to emphasize the need for more precise, spatial and temporal sampling in monitoring programs both here and elsewhere.

*Response: The final sentence of the abstract was revised as follows: “...and warrant additional, complementary interventions and continuous monitoring data to support ecosystem health into the future.”*

### Co-Editor in Chief:

1\) Maybe change the title to "Hot and fresh water:..."

*Response: At the risk of being slightly gimmicky, we have retained the title preface. All co-authors preferred this option. However, we have added “water” later in the title to describe “suboptimal water conditions” to make the focus of the manuscript clear.*

2\) ESCO now requires a “Declarations” section as described in the submission guidelines https://link.springer.com/journal/12237/submission-guidelines. Include declarations on: funding sources; financial or non-financial competing interests, or conflicts of interest; ethics compliance for research involving humans, animals, or plants; data availability and repository DOI; author contributions; pre-print location if applicable; etc.

*Response: The declaration section that was previously after the abstract was added to the end and relevant sections were added (data availability, author contributions, ethics, and funding).*

3\) Lines 384-390 and other places, and Tables 1-4: ESCO has a policy discouraging the use of an arbitrary rejection rate and recommends describing trends and reporting the actual p-values, see Smith (2020) https://doi.org/10.1007/s12237-019-00679-y. This means that you should avoid the phrases “significant,” “p \< 0.05,” or “\*”and replace these phrases with effect size values that represent the trend, the actual p-value, and the sample size. For example: X was 100% higher than Y, p = 0.0001, n=20; or X was 10 units, which is greater than Y at 5 units, p = 0.0001, n=20. Use the same number of significant digits in all your reporting, meaning values \< 0.0001 should be reported as \< 0.0001. Don't forget n. More examples and suggestions are provided in the Smith citation. So, please revise your results section, and any other parts of the manuscript, to address this issue.

*Response: We have removed or revised all instances in the text, tables, and figures to focus on trends, effect sizes, and magnitudes. Any use of “significant” regarding statistical tests was also removed. Sample sizes were reported where missing.*

4\) Figure 2: I think the font on X-axis is too small to read. Consult submission guidelines on figures.

*Response: Font size was increased.*

5\) Fig. 6, 7: There is a charge to print color figures.

*Response: We intend on publishing as open access.*

### 2024-05-21 revision/review_response.qmd
proof revision

@@
 #| eval: false
-spelling::spell_check_files(here::here('revision/review-response.qmd'))
+spelling::spell_check_files(here::here('revision/review_response.qmd'))
 ```
@@
 We have made every effort to address these concerns.
-In particular, we have provided a more thorough assessment of seagrass response to climate-related stressors using more flexible models (GAMs) and by including additional predictors (light attenuation).
-Please see the specific comments below.
-Note that the line numbers relate to the original draft and not the revised version.*
+In particular, we have provided a more thorough assessment of seagrass response to climate-related stressors using more flexible models (GAMs) and including additional predictors (light attenuation).
+Please see the responses below.*
 
@@
 
-*Response: Thank you for the comments. We have revised the text based on your suggestions below. Please see our specific comments to each response for more details.*
+*Response: Thank you for the comments. We have revised the text based on your suggestions. Please see our responses to each comment for more details.*
 
@@
 This suggests that multiple factors are influencing these changes and that they potentially vary by location.
-Our development of more robust models (GAMs) that attempted to better describe these patterns is an explicit acknowledgement of these differences, which were not adequately characterized with the earlier models.
+Our development of more robust models (GAMs) that attempted to better describe these patterns is an explicit acknowledgment of these differences, which were not adequately characterized with the earlier models.
 Please see our description below of these updated models.*
@@
 
-*Response: We have revised our modeling approach for describing seagrass change using a GAM approach that uses time as a continuous variable (year) and the inclusion of light attenuation as a predictor when able.
-For the latter, light attenuation was included only in the EPC model, whereas it was excluded from the FIM and PDEM models due to a large majority of Secchi observations (from which light attenuation was derived) being on the bottom.
+*Response: We have revised our models for describing seagrass change using a GAM approach that uses time as a continuous variable (year) and includes light attenuation as a predictor, when able.
+For the latter, light attenuation was included only in the EPC model, whereas it was excluded from the FIM and PDEM models due to a majority of Secchi observations (from which light attenuation was derived) being on the bottom.
 As such, light attenuation could not be derived for these models, nor would it likely explain much variation in seagrass response since light environments are generally not limiting at most of the locations sampled in the FIM and PDEM datasets.
@@
 
-*Sections 2.7 (links to seagrass) and 3.3 (seagrass response) were revised to describe the updated modeling approach, including the addition of equations 1 to 4 showing the structure for each GAM, Figures 7, 8 , and S7 showing partial effects for each model, and Tables S7 to S10 providing summary statistics for each model.
-Overall, the general conclusions that climate-related stressors are likely contributing factors to seagrass change remains as before, although the revised approach provides a more complete description of this change using more flexible models that can describe non-linear relationships and additional predictors (time, light attenuation).*
+*Sections 2.7 (links to seagrass) and 3.3 (seagrass response) were revised to describe the updated modeling approach, including the addition of equations 1 to 4 showing the structure for each GAM, Figures 7, 8, and S7 showing partial effects for each model, and Tables S7 to S10 providing summary statistics for each model.
+Overall, the general conclusions that climate-related stressors are likely contributing factors to seagrass change remains as before, although the revised approach provides a more complete description using more flexible models that can describe non-linear relationships and additional predictors (time, light attenuation).*
 
@@
 
-*Response: Sentence was revised as follows: “These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location).*
+*Response: Sentence was revised as follows: "These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location)."*
 
@@
 Additionally, the FIM and PDEM datasets that target more shallow areas of the bay had a majority of Secchi observations on the bottom, indicating that light environments are generally supportive of seagrass growth in recent years.
-Also note the addition of Figure 2c that shows the long-term improvement in light environments over the period of record for the EPC dataset.*
+See the revision to section 2.7 for details on the updated methods.
+Also note the addition of Figure 2c that shows the long-term improvement in light environments over the period of record.*
 
@@
 Moreover, we feel the trends in the temperature summaries (e.g., increase in number of days above a threshold over time) would likely be similar.
-As such, we have not attempted this for the current dataset.
+As such, we have not attempted this for the current analysis.
 However, ongoing work in Old Tampa Bay for a separate project is attempting to describe short-term diurnal variation in temperature with continuous data loggers.
@@
 
-*Line 669: “Ongoing work in OTB using continuous data loggers in shallow areas will provide insights into short-term diurnal changes as potential acute temperature stress on seagrass (see https://tbep-tech.github.io/otb-temp/tempeval).”*
+*Line 669: “Ongoing work in OTB using continuous data loggers in shallow areas where seagrass has been gained or lost will provide insights into short-term diurnal changes as potential acute temperature stress (see https://tbep-tech.github.io/otb-temp/tempeval).”*
 
@@
 *Response: Our previous statement on lines 560-564 suggests this may be a plausible explanation for the weak association of seagrass change with the stressors.
-The statement was further amended to specifically mention that shoal grass accounted for the largest seagrass decline: “This is especially true for H.
+The statement was further revised to specifically mention that shoal grass accounted for the largest seagrass decline: “This is especially true for H.
 wrightii that had the greatest changes over the period of record and is tolerant of a wide range of salinity.”*
@@
 *Response: Yes, preliminary analyses evaluated lagged associations between the salinity and temperature stressor metrics and seagrass change.
-Specifically, we evaluated the number of days in 30 day periods from one month to one year when temperature was above or salinity was below a threshold in relation to seagrass change (see code here).
+Specifically, we evaluated the number of days in 30 day periods from one month to one year when temperature was above or salinity was below a threshold in relation to seagrass change (see code here, https://github.com/tbep-tech/temp-manu/blob/9044bb07cd81532a7fc83c02dc035f7d8ff16b2c/R/dat_proc.R#L865, lines 865 to 925).
 No compelling differences were observed for any of the lags compared to the entire year, which was the justification for using the latter in the current analysis.
@@
 
-*Response: The assessment of inter annual changes in precipitation during the rainy season showed a weak increase over time, whereas removing the last month (September) showed a more statistically powerful increase over time.
+*Response: The assessment of inter-annual changes in precipitation during the rainy season showed a weak increase over time, whereas removing the last month (September) showed a more statistically powerful increase over time.
 The removal of September was meant to show that the trend was driven by an increase in the first three months of the rainy season (Jun - Aug).
@@
 *Response: Line 120 was revised to describe that the upper bay segments are more shallow and receive a majority of hydrologic inflow.
-The previous was inaccurate.
+The previous statement was inaccurate.
 Also, please see the addition of Figure 2c that shows the long-term improvement in the light environment over time.
@@
 
-*Response: Please see our above comments regarding the addition of these data in Figure 2 and our use of more robust modeling approaches to describe these potential changes (i.e., GAMs)*
+*Response: Please see our above response regarding the addition of these data in Figure 2 and our use of more robust modeling approaches to describe these potential changes in response to comments from the editor and other reviewer (i.e., GAMs)*
 
@@
 *Response: We agree there are likely important distinctions between species that could be considered when evaluating the response to climate-related stressors.
-As correctly noted by the reviewer, Thalassia is dominant in lower Tampa Bay with coverage decreasing towards the upper bay segments.
+As correctly noted by the reviewer, Thalassia is dominant in Lower Tampa Bay with coverage decreasing towards the upper bay segments.
 We did not evaluate lower Tampa Bay in our analysis of seagrass response to climate stressors given that coverage has generally been stable in recent years and salinity has not been changing dramatically.
@@
 
-*Response: This is a good suggestion and we agree these metrics are useful to highlight in the abstract.
-We have added this sentence to the abstract, with similar clarification added in section 3.2 and the beginning of the discussion: “Additionally, the number of days when temperature was above 30 °C or salinity was below 25 ppt has increased on average across all bay segments by 48 and 37 days, respectively, since 1975.”*
+*Response: We agree these metrics are useful to highlight in the abstract.
+We have added this sentence to the abstract, with similar clarification added in section 3.2 and the beginning of the discussion: “Additionally, the number of days when temperature was above 30 $^\circ$C or salinity was below 25 ppt has increased on average across all bay segments by 48 and 37 days, respectively, since 1975.”*
 
@@
 
-*Response: At the risk of being slightly gimmicky, we have retained the title preface.
+*Response: We have retained the title preface.
 All co-authors preferred this option.

### 2024-05-21 manu-draft.qmd
proof revision

@@
 Species shifts are also expected to occur as changing physical conditions decrease the competitive advantages of historically abundant species.
-In Chesapeake Bay, the abundance of widgeongrass (*Ruppia maritima*) has responded positively to nutrient reduction and is replacing the formerly dominant but now heat-stressed [*Z.
-marina*, @hensel2023; also see @bartenfelder2022].
+In Chesapeake Bay, the abundance of widgeongrass (*Ruppia maritima*) has responded positively to nutrient reduction and is replacing the formerly dominant but now heat-stressed *Z.
+marina* [@hensel2023; also see @bartenfelder2022].
 Changing frequency and severity of precipitation patterns may further alter salinity regimes and with it the distribution and abundance of seagrasses throughout the estuary [@rasheed2011; @webster2021].
@@
 Using this model, mean precipitation for June to August in 1975 was `r prpstr2` mm, whereas current mean precipitation for June to August in 2022 was `r prpend2` mm, showing an overall increase in the period of record of `r prpdif2` mm.
-Notably, rainfall during the dry season (October through May) has decreased slightly over time at `r dryslo` mm per year, although the trend model was weak (`r drysiz`, `r drypvl`, $R^2$ = `r dryrsq`) The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and 2015-2020.
+Notably, rainfall during the dry season (October through May) has decreased slightly over time at `r dryslo` mm per year, although the trend model was weak (`r drysiz`, `r drypvl`, $R^2$ = `r dryrsq`).
+The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and 2015-2020 (third row, [@fig-meteowqraw]).
 
@@
 The EPC dataset was also used to provide detailed information on station-level trends from 1975 to present (@fig-kendall, see Figure S4 for 1998 to 2022).
-All stations had increasing temperature and decreasing salinity from 1975 to present, although several stations in HB did not have weak salinity trends (Figure [-@fig-kendall]a).
+All stations had increasing temperature and decreasing salinity from 1975 to present, although some stations in HB had weak salinity trends (Figure [-@fig-kendall]a).
 Seasonally, most bay segments had more stronger trends in the early fall/winter periods for both temperature and salinity (Figure [-@fig-kendall]b), although some variation was observed throughout the bay.
@@
 Both OTB and HB had models showing an increasing number of days when salinity was below 20 or 25 ppt, whereas MTB only showed an increase in the number of days when salinity was below 25 ppt.
-LTB had no notable salinity models.
-Some salinity models had decreasing trends, although the total number of days at the beginning and end of the period of record were neglible (Table S5).
-Overall, OTB showed the largest increase in the number of days each year when salinity was below a threshold, particularly for 25 ppt, where the change was 86 days per year from 1975 to 2022 (130 days to 216 days per year, @tbl-mixdaytab, Table S5, see Table S6 for 1998 to 2022).
+Some salinity models for LTB had decreasing trends, although the total number of days at the beginning and end of the period of record were neglible (Table S5).
+Overall, OTB showed the largest increase in the number of days each year when salinity was below a threshold, particularly for 25 ppt, where the change was 86 days per year from 1975 to 2022 (130 to 216 days, @tbl-mixdaytab, Table S5, see Table S6 for 1998 to 2022).
 Across all bay segments, the average increase in the number of days salinity was below 25 ppt from 1975 to 2022 was 36 days, although a distinct gradient towards the mouth of the bay was observed.
@@
 The number of days when temperature was above 29 $^\circ$C and salinity was below 25 ppt had the largest slopes of 1.4, 1.1, and 0.7 days per year for OTB, HB, and MTB, respectively.
-Likewise, the average number of days when both temperature was above 29 $^\circ$C and salinity was below 25 ppt from the beginning to the end of the period showed the greatest increase for OTB of 68 days (4 days to 72 days per year from 1975 to 2022).
+Likewise, the average number of days when both temperature was above 29 $^\circ$C and salinity was below 25 ppt from the beginning to the end of the period showed the greatest increase for OTB of 68 days (4 to 72 days per year from 1975 to 2022).
 
 @fig-mixeff provides visual examples of the mixed-effects models for the estimated number of days over time for each bay segment from 1975 to present when temperature was above 30$^\circ$, salinity was below 25 ppt, and when both occurred (see Figure S6 for 1998 to 2022).
-Temperature trends were similar among segments, whereas the number of days when salinity was below the threshold decreased with proximity to the Gulf of Mexico (@tbl-mixdaytab).
+Temperature trends were similar between segments, whereas the number of days when salinity was below the threshold decreased with proximity to the Gulf of Mexico (@tbl-mixdaytab).
 The number of days when both temperature was above and salinity was below the threshold generally followed the trends for the number of days when salinity was below the threshold.
@@
 The reported threshold values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity [@phillips1960; @mcmillan1967; @zieman1975; @lirman2003].
-Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values, conditions in Tampa Bay in recent years are generally within the ecological niche for seagrasses.
+Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values, conditions in Tampa Bay in recent years are likely suboptimal but within the ecological niche for seagrasses.
 This is especially true for *H.
@@
 Our inclusion of light attenuation in the EPC models was meant to account for how the light environment may be influencing seagrass growth, in addition to climate-related stressors.
-However, light attenuation has improved over the period of record and within the limits estimated to be supportive of seagrass growth in Tampa Bay [@tbep0696; @greening2011], particularly in OTB where the most loss occurred (Figure [-@fig-seagrasschg]c).
+However, light attenuation has improved over the period of record and is currently within the limits estimated to be supportive of seagrass growth in Tampa Bay [@tbep0696; @greening2011], particularly in OTB where the most loss occurred (Figure [-@fig-seagrasschg]c).
 Additional water quality parameters could be included in the models to provide further evidence that light-limitation is not the present driver for seagrass change (e.g., nitrogen loading, chlorophyll-a, color), although nutrient management for the benefit of seagrass growth will likely continue to be a dominant management paradigm for Tampa Bay.
@@
 However, elevated summertime temperatures were also implicated in the decline.
-@dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters.
+@dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater inflows.
 In Tampa Bay, the long-term trends of decreasing salinity, especially in the upper bay segments, suggest that the hydrologic loading has had a greater influence on salinity regimes than the effects of sea-level rise.
@@
 These analyses should be supported by additional data collection efforts, particularly high-resolution continuous monitoring data that provide a more precise assessment of diurnal stress across multiple time-scales.
-Ongoing work in OTB using continuous data loggers in shallow areas will provide insights into short-term diurnal changes as potential acute temperature stress on seagrass (see <https://tbep-tech.github.io/otb-temp/tempeval>).
-Morpohological or physiological measurements at the individual level could also provide early indications of heat and osmotic stress [@congdon2023].
+Ongoing work in OTB using continuous data loggers in shallow areas where seagrass has been gained or lost will provide insights into short-term diurnal changes as potential acute temperature stress (see <https://tbep-tech.github.io/otb-temp/tempeval>).
+Morphological or physiological measurements at the individual level could also provide early indications of heat and osmotic stress [@congdon2023].
 
@@
 ![Trends from 1975 to 2022 for bottom water temperature and salinity measurements at long-term monitoring stations in Tampa Bay.
-Results for seasonal Kendall tests by station are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr$^{-1}$).
+Results for seasonal Kendall tests by station are shown in (a) with color, size, and shape corresponding to the estimated annual slope as change per year (yr$^{-1}$).
 Summarized seasonal trends by month are shown as (b) the average magnitude of change (slope) per month for stations in each bay segment for temperature and salinity.
+Average slope across stations as change per year is indicated by color and text scaled by absolute magnitude.
 Bay segment outlines are shown in (a); OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/kendall.png){#fig-kendall}

### 2024-05-21 supplement.qmd
proof revision

@@
 ![Trends from 1998 to 2022 for bottom temperature and salinity measurements at long-term monitoring stations in Tampa Bay.
-Results for seasonal Kendall tests by station and monitoring location (top or bottom of the water column) are shown in (a) with color and shape corresponding to the estimated annual slope as change per year (yr$^{-1}$).
+Results for seasonal Kendall tests by station and monitoring location (top or bottom of the water column) are shown in (a) with color, size, and shape corresponding to the estimated annual slope as change per year (yr$^{-1}$).
 Summarized seasonal trends by month are shown as (b) the average magnitude of change (slope) per month for all stations in each bay segment for temperature and salinity.
+Average slope across stations as change per year is indicated by color and text scaled by absolute magnitude.
 Bay segment outlines are shown in (a); OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppkendall.png){#fig-suppkendall}
@@
 
-![Example of mixed effects models for the estimated number of days per year that bottom temperature (red) or salinity (blue) were above or below thresholds of 30 $^\circ$C or 25 psu, respectively, from 1988 to 2022.
+![Example of mixed effects models for the estimated number of days per year that bottom temperature (red) or salinity (blue) were above or below thresholds of 30 $^\circ$C or 25 psu, respectively, from 1998 to 2022.
 The bottom row (black) shows the number of days when both temperature was above and salinity was below the thresholds.

### 2024-05-21 manu-draft.qmd
more revision proofing

@@
 Compared to the additional datasets described below, the monitoring stations are generally in deeper water beyond where seagrasses occur along the shallow margins of the bay.
-The data were obtained using the tbeptools R package that imports the data directly from a stable web address provided by the EPC [@beck2021].
+The data were obtained using the *tbeptools* R package that imports the data directly from a stable web address provided by the EPC [@beck2021].
 
@@
 For comparison to water quality conditions, daily air temperature (Tampa International Airport [TIA] National Weather Service site) and precipitation (SWFWMD area-weighted watershed summaries) were used to characterize regional conditions for the most consistent period of record covered by the water quality samples (i.e., 1975 to present for the EPC data, @tbl-dattab).
-The rnoaa R package [@chamberlain23] was used to obtain the TIA temperature data.
-Regional precipitation summaries were obtained directly from the SWFWMD (https://www.swfwmd.state.fl.us/resources/data-maps/rainfall-summary-data-region).
+The *rnoaa* R package [@chamberlain23] was used to obtain the TIA temperature data.
+Regional precipitation summaries were obtained directly from the SWFWMD (<https://www.swfwmd.state.fl.us/resources/data-maps/rainfall-summary-data-region>).
 Only rainfall data for the wet season (June to September) were evaluated for trends, whereas the complete record was used to calculate the Standardized Precipitation Index [SPI, @begueria2013] to identify periods of time when rainfall deviated from the long-term average [using the *spei* R package, @begueria23].
@@
 wrightii* in the upper bay segments (Figure [-@fig-seagrasschg]b).
-The GAM equations in mgcv R package notation [@wood17] are below, where $s$ is a smoother function for a single predictor and $ti$ is the tensor product interaction between predictors.
+The GAM equations in *mgcv* R package notation [@wood17] are below, where $s$ is a smoother function for a single predictor and $ti$ is the tensor product interaction between predictors.
 All GAMs were fit using restricted maximum likelihood evaluation [@wood11].
@@
 Results for seasonal Kendall tests by station are shown in (a) with color, size, and shape corresponding to the estimated annual slope as change per year (yr$^{-1}$).
-Summarized seasonal trends by month are shown as (b) the average magnitude of change (slope) per month for stations in each bay segment for temperature and salinity.
-Average slope across stations as change per year is indicated by color and text scaled by absolute magnitude.
+Summarized seasonal trends by month are shown as (b) the average magnitude of change (slope) for stations in each bay segment for temperature and salinity, indicated by color and text scaled by absolute magnitude.
 Bay segment outlines are shown in (a); OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/kendall.png){#fig-kendall}

### 2024-05-21 supplement.qmd
more revision proofing

@@
 Results for seasonal Kendall tests by station and monitoring location (top or bottom of the water column) are shown in (a) with color, size, and shape corresponding to the estimated annual slope as change per year (yr$^{-1}$).
-Summarized seasonal trends by month are shown as (b) the average magnitude of change (slope) per month for all stations in each bay segment for temperature and salinity.
-Average slope across stations as change per year is indicated by color and text scaled by absolute magnitude.
+Summarized seasonal trends by month are shown as (b) the average magnitude of change (slope) for all stations in each bay segment for temperature and salinity, indicated by color and text scaled by absolute magnitude.
 Bay segment outlines are shown in (a); OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppkendall.png){#fig-suppkendall}

### 2024-05-23 revision/review_response.qmd
ready to resubmit

@@
 In particular, we have provided a more thorough assessment of seagrass response to climate-related stressors using more flexible models (GAMs) and including additional predictors (light attenuation).
-Please see the responses below.*
+Please see the responses below (line numbers in responses refer to the revised manuscript).*
 
@@
 
-*Line 191: “As such, trajectories of recovery and decline have varied by bay segment in magnitude and timing of the change, although a consistent decline baywide has been observed since 2016.
+*Line 197: “As such, trajectories of recovery and decline have varied by bay segment in magnitude and timing of the change, although a consistent decline baywide has been observed since 2016.
 Change by species has been most notable for Halodule wrightii (shoal grass) in OTB and HB, whereas some losses have been observed for Thalassia testudinum (turtle grass) in LTB, likely related to dominance of each species across the salinity gradient and proximity to hydrologic inputs (Lewis et al.
@@
 
-*Line 364: “Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model and that the majority of seagrass loss in recent years was dominated by H.
+*Line 372: “Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model and that the majority of seagrass loss in recent years was dominated by H.
 wrightii in the upper bay segments (Figure 2b).”*
@@
 
-*Response: As noted on line 195, the areal maps are produced every other year, whereas the transect data are collected every year.
+*Response: As noted on line 206, the areal maps are produced every other year, whereas the transect data are collected every year.
 The text was revised as follows: “Although the areal maps provide the standard for assessment of restoration goals, the maps are produced every other year.
@@
 
-*Response: This is certainly an issue worth considering using the temperature data from the monthly discrete samples and was the motivation for including a statement on the need for continuous monitoring data in the original draft (lines 667 - 669).
+*Response: This is certainly an issue worth considering using the temperature data from the monthly discrete samples and was the motivation for including a statement on the need for continuous monitoring data in the original draft (lines 679 - 681).
 Applying a correction or normalization for time of day would be challenging and we would not be confident in the results to accurately describe maximum or minimum temperature values for a given day.
@@
 We suspect this information will provide further insight into acute temperature stressors that could affect seagrass.
-We have added a statement and link in the text with a short description of this work, although these loggers are in very shallow areas that likely show different temperature characteristics than the discrete samples at deeper depths.*
+We have added text describing this general need and a statement with a short description of this work, although these loggers are in very shallow areas that likely show different temperature characteristics than the discrete samples at deeper depths.*
 
-*Line 669: “Ongoing work in OTB using continuous data loggers in shallow areas where seagrass has been gained or lost will provide insights into short-term diurnal changes as potential acute temperature stress (see https://tbep-tech.github.io/otb-temp/tempeval).”*
+*Line 582: Without more continuous, diel observations of these metrics over the period of record, we were heavily reliant on these model outputs to determine relevant thresholds for Tampa Bay seagrass.
+This further highlights a long-standing data gap and need for the estuary."*
+
+*Line 681: “Ongoing work in OTB using continuous data loggers in shallow areas where seagrass has been gained or lost will provide insights into short-term diurnal changes as potential acute temperature stress (see https://tbep-tech.github.io/otb-temp/tempeval).”*
 
@@
 
-*Response: Our previous statement on lines 560-564 suggests this may be a plausible explanation for the weak association of seagrass change with the stressors.
+*Response: Our previous statement on lines 571-574 suggests this may be a plausible explanation for the weak association of seagrass change with the stressors.
 The statement was further revised to specifically mention that shoal grass accounted for the largest seagrass decline: “This is especially true for H.
@@
 *Response: Yes, preliminary analyses evaluated lagged associations between the salinity and temperature stressor metrics and seagrass change.
-Specifically, we evaluated the number of days in 30 day periods from one month to one year when temperature was above or salinity was below a threshold in relation to seagrass change (see code here, https://github.com/tbep-tech/temp-manu/blob/9044bb07cd81532a7fc83c02dc035f7d8ff16b2c/R/dat_proc.R#L865, lines 865 to 925).
+Specifically, we evaluated the number of days in 30 day periods from one month to one year when temperature was above or salinity was below a threshold in relation to seagrass change (see code here, https://github.com/tbep-tech/temp-manu/blob/9044bb07cd81532a7fc83c02dc035f7d8ff16b2c/R/dat_proc.R#L865, lines 865 to 925 in the file).
 No compelling differences were observed for any of the lags compared to the entire year, which was the justification for using the latter in the current analysis.
@@
 
-*Line 354: “Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the transect year summaries.
+*Line 360: “Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the transect year summaries.
 As such, the stressor metrics...”*
@@
 The removal of September was meant to show that the trend was driven by an increase in the first three months of the rainy season (Jun - Aug).
-Text was added to line 386: “...suggesting precipitation increases were driven by the earlier months (June - August).” Also note that our previous estimates of rainfall total were based on an incorrect sum across overlapping areas.
+Text was added to line 413: “...suggesting precipitation increases were driven by the earlier months (June - August).” Also note that our previous estimates of rainfall total were based on an incorrect sum across overlapping areas.
 We have fixed the issue, although the results and trends remain the same.*
@@
 
-*Response: Line 120 was revised to describe that the upper bay segments are more shallow and receive a majority of hydrologic inflow.
+*Response: Line 125 was revised to describe that the upper bay segments are more shallow and receive a majority of hydrologic inflow.
 The previous statement was inaccurate.

### 2024-05-23 cover-letter.qmd
ready to resubmit

@@
 
-Enclosed please find our manuscript, titled \`Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary', to be considered as an original article in Estuaries and Coasts.
+Enclosed please find our revised manuscript, titled \`Hot and fresh: evidence of climate-related suboptimal water conditions for seagrass in a large Gulf coast estuary', to be considered as an original article in Estuaries and Coasts.
 
-The long-term recovery of Tampa Bay is a nationally recognized success story that demonstrates how the USEPA National Estuary Program is a valuable model for collaborative management.
-Through large-scale reductions in external nutrient loading, seagrasses have recovered in the bay, reaching an all-time high in 2016.
-Since that time, seagrasses have declined by nearly 1/3 and questions remain regarding the role of climate-related stressors in contributing to this decline.
-We leverage multiple data sources to describe changing water temperature and salinity in the bay and relate these trends to the recent seagrass loss from 2016 to present.
-No such comprehensive assessment has previously been conducted for Tampa Bay and our results provide context for how climate stressors can confound existing management models, particularly related to nutrient management.
+We sincerely appreciate the constructive comments provided by the editors and reviewers and have made every attempt to address these comments in our revision.
+The largest change to the manuscript was a reanalysis of the potential seagrass response to climate stressors, which included more flexible Generalized Additive Models and inclusion of additional predictors (i.e., time as a continuous variable, light attenuation).
+Overall, our conclusions remain the same that climate stressors are potentially driving recent seagrass changes, although our more robust modeling approach provides greater weight of evidence for this conclusion.
+Note that we have also made substantial revisions in response to comments from the co-editor in chief regarding statistical descriptions of results to focus more on effect sizes and magnitudes of change.
 
-This work has not been published elsewhere, nor is it under consideration for publication anywhere else.
-Each author has contributed substantially to this work.
-All authors have contributed to the writing and review of the original draft.
-Analysis was conducted by the first author, with interpretation of results supported by all other co-authors.
-All authors have approved the final draft and the submitted product was prepared according to instructions for authors provided on the journal's website.
-
-We greatly appreciate the opportunity to publish this work in Estuaries and Coasts and are confident readers will consider it a valuable contribution.
+Again, we greatly appreciate the opportunity to publish this work in Estuaries and Coasts and are confident readers will consider it a valuable contribution.
 Please feel free to contact us directly should additional information be needed.

### 2024-05-23 manu-draft.qmd
ready to resubmit

@@
 @lefcheck2017 evaluated interactive effects of water clarity and rising temperatures on seagrasses with over 30 years of data in Chesapeake Bay.
-The environmental stress on seagrasses from acute warming related to climate change was compounded by stress from poor light environments in shallow waters, demonstrating a concerning synergy of stressors most likely to affect Tampa Bay.
+The environmental stress on seagrasses from acute warming related to climate change was compounded by stress from poor light environments in shallow waters, demonstrating a concerning synergy of stressors most likely to also affect Tampa Bay.
 @moreno-marin2018 produced similar results using a multifactorial experiment that considered temperature, light, and nutrient (nitrogen) availability for North Sea eelgrass (*Zostera marina*).
@@
 Tampa Bay is the largest open-water estuary in Florida covering 400 mi$^2$ (1,036 km$^2$) and the second largest in the Gulf of Mexico.
-The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee rivers that contribute to a majority of freshwater inflow to the bay.
-The climate of Tampa Bay is subtropical with warm humid summers and cool less humid winters [@garcia2023].
+The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee rivers contributing the majority of freshwater inflow to the bay.
+The climate of Tampa Bay is subtropical with warm humid summers and cool, less humid winters [@garcia2023].
 Unique to Tampa Bay and the entire central Florida peninsula is that this region is within a transition zone from more temperate weather to the north, similar to the rest of the southeastern United States, and more tropical weather to the south [@morrison2006].
@@
 Since then, the greatest areal coverage expansions were observed in the upper bay segments (OTB, HB, and MTB; Figure [-@fig-seagrasschg]a).
-Light environments have improved (Figure [-@fig-seagrasschg]c) through reduction of external nitrogen loadings by about 2/3 by the 1980s from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed [@greening2014; @beck2019].
+Light environments have improved (Figure [-@fig-seagrasschg]c) through a 2/3 reduction of external nitrogen loadings from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed [@greening2014; @beck2019].
 
@@
 The maps are created by photointerpretation of image signatures coupled with a robust field verification and accuracy assessment.
-The maps provide a spatial estimate of seagrass cover at the landscape scale and is irrespective of species.
+The maps provide a spatial estimate of seagrass cover at the landscape scale, irrespective of species.
 Complementary to the SWFWMD seagrass mapping program is the Tampa Bay Interagency Seagrass Monitoring Program (Figure [-@fig-seagrasschg]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>).
@@
 The transect data are collected each year, allowing inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined.
-As such, the transect data were used below for comparison with temperature and salinity changes for the major bay segments.
+As such, the transect data were used for comparison with temperature and salinity changes for the major bay segments.
 Additional sources of seagrass data are described in the next section.
@@
 
-Separate models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change.
+Two other models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change.
 These models used direct measurements of bottom salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location).
@@
 For the PDEM models, presence/absence was converted to frequency occurrence as the number of sites with seagrass in a year divided by the total number of sites and all data were subset to less than 2 meters to better characterize locations where seagrass occurs.
-The GAM equations for the FIM (@eq-fimmod) and PDEM (@eq-pincomod) datasets, respectively, were as follows.
+The GAM equations for the FIM (@eq-fimmod) and PDEM (@eq-pincomod) datasets, respectively, were as follows:
 
@@
 The number of days when temperature was above 29 $^\circ$C and salinity was below 25 ppt had the largest slopes of 1.4, 1.1, and 0.7 days per year for OTB, HB, and MTB, respectively.
-Likewise, the average number of days when both temperature was above 29 $^\circ$C and salinity was below 25 ppt from the beginning to the end of the period showed the greatest increase for OTB of 68 days (4 to 72 days per year from 1975 to 2022).
+Likewise, the average number of days when both temperature was above 29 $^\circ$C and salinity was below 25 ppt from the beginning to the end of the period showed the greatest increase for OTB of 68 days (Table S5; 4 to 72 days per year from 1975 to 2022).
 
@@
 The GAMs for the FIM and PDEM datasets had notable associations with temperature and salinity stress, where an increase in temperature and a decrease in salinity was often associated with a decline in seagrass depending on bay segment.
-The FIM model (`r sgmodtxt$fimmod`), Figure [-@fig-sgmod2]a, Table S9) showed a slight decrease in seagrass with increasing temperature for HB and MTB, with a similar change as salinity decreased.
+The FIM model (`r sgmodtxt$fimmod`, Figure [-@fig-sgmod2]a, Table S9) showed a slight decrease in seagrass with increasing temperature for HB and MTB, with a similar change as salinity decreased.
 The PDEM model (`r sgmodtxt$pincomod`, Figure [-@fig-sgmod2]b, Table S10) that evaluated only OTB showed a distinct decrease with seagrass as temperature increased, but no association with salinity.
@@
 Our assessment evaluated non-parametric trends (i.e., Kendall tests, less sensitive to outliers) at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to their larger spatial area and model domain.
-Our results showing increases in temperature and decreases in salinity in the fall, winter period provided a finer-scale comparison, where trends were most notable for OTB and northern stations of HB (@fig-kendall), likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.
+Our results showing increases in temperature and decreases in salinity in the fall, winter period provided a finer-scale comparison, where trends were most notable for OTB and northern stations of HB (@fig-kendall) – likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.
 These upper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to more rapid warming with shallower depths.
@@
 Extreme temperature or precipitation events acting individually or in combination are likely captured by the trends in stressor metrics using these lower thresholds, i.e., an increase in a bay segment median number of days also suggests extremes are increasing given the variation around these summary metrics (@fig-mixeff).
-Our thresholds may also be indicative of the potential for chronic sublethal effects of stress on seagrasses, reducing their resilience to other stressors.
-Regardless, our models suggested that temperature and salinity are at least associated with seagrass loss and, if so, long-term trends in both are set to amplify their effect.
+Without more continuous, diel observations of these metrics over the period of record, we were heavily reliant on these model outputs to determine relevant thresholds for Tampa Bay seagrass.
+This further highlights a long-standing data gap and need for the estuary.
+Regardless, our thresholds seem to be indicative of the potential for chronic sublethal effects of stress on seagrasses, reducing their resilience to other stressors.
+Further, our models suggested that temperature and salinity changes are at least associated with seagrass loss and, if so, long-term trends in both are set to amplify their effect in the future.
 
@@
 Natural resource managers should consider how these climate-related stressors may alter the effectiveness of intervention activities aimed at protecting ecological resources in Tampa Bay.
-Management actions that have historically been effective may not be able to maintain ecosystem resilience to climatic change.
-For example, nitrogen load reductions have been effective at restoring seagrass in Tampa Bay [@greening2006; @greening2014].
-As Tampa Bay becomes hotter and fresher, current nutrient loads may no longer be effectively assimilated and algal and seagrass ecology dynamics may shift.
-Strategies that mimic or restore pre-development hydrology or that further reduce allowable load inputs from regulated entities (e.g., additional stormwater controls, hydrological modifications) may be needed to confer additional resilience and adaptive capacity for seagrass in Tampa Bay.
-These considerations are especially critical for upper parts of Tampa Bay where a majority of seagrass loss has occurred and where temperature and salinity trends appear most pronounced.
-Reversal of recent trends may be more likely to occur if aggressive actions and controls are pursued sooner rather than later, given the challenges of restoring these long-lived foundation species once lost, ongoing development in the watershed, and the current climate trajectory.
+Management actions that have historically been effective may become less so, resulting in diminished ecosystem resilience compounded by climate change.
+For example, nitrogen load reductions have historically been the most effective strategy to restore seagrass in Tampa Bay [@greening2006; @greening2014].
+As Tampa Bay becomes hotter and fresher, current nutrient load targets may no longer be effective, resulting in further shifts in algal and seagrass ecology dynamics.
+Strategies that mimic or restore pre-development hydrology or that further reduce allowable load inputs from regulated entities (e.g., additional stormwater controls, hydrological modifications) may be needed to confer additional resilience and adaptive capacity for seagrasses responding to climatic changes in Tampa Bay.
+These considerations are especially critical for upper Tampa Bay where a majority of seagrass loss has occurred and where temperature and salinity trends appear most pronounced.
+Reversal of recent trends may be more likely to occur if aggressive actions and controls are pursued sooner rather than later, especially since the challenges of restoring these long-lived foundation species once lost will be exacerbated by ongoing development in the watershed and the current climate trajectory.
 

### 2024-05-23 R/tabs.R

@@ -24,7 +24,7 @@ epcdat <- epcdata %>%
 totab <- tibble(
   Dataset = c('SWFWMD aerial maps', 'Transect data', 'EPC', 'FIM', 'PDEM', 'Tampa International Airport', 'SWFWMD precipitation'),
   Description = c('Seagrass coverage in acres', 'Seagrass frequency occurrence by species', 'Water quality monitoring samples', 'Nearshore temperature and salinity, seagrass species and cover', 'Water quality and seagrass presence/absence', 'Air temperature', 'Area-weighted precipitation for the wet season (June-September) for the Tampa Bay watershed'),
-  Temporal = c('1988 - 2022, biennial', '1999 - 2022, annual', '1975 - 2022, monthly', '1996 - 2022, monthly', '2003 - 2022, monthly', '1975 - 2022, annual', '1975 - 2022, annual'),
+  Temporal = c('1988 - 2022, ~biennial', '1999 - 2022, annual', '1975 - 2022, monthly', '1996 - 2022, monthly', '2003 - 2022, monthly', '1975 - 2022, annual', '1975 - 2022, annual'),
   Spatial = c('Whole bay', 'Whole bay, 62 transects', 'Whole bay, fixed sites', 'Whole bay shallow, stratified random sites', 'Old Tampa Bay, stratified random sites', '27.979$^\\circ$N, 82.535$^\\circ$W', 'Whole watershed'),
   Analysis = c('Biennial trends by bay segment, visual only', 'Annual trends by bay segment and species, comparison with temperature, salinity, and light attenuation as stressor metrics or observed data at annual scale', 'Trends in annual change and seasonal Kendall tests, estimate of stressor metrics as number of days above/below threshold', 'Trends in annual observed temperature, salinity, comparison to annual seagrass % cover', 'Trends in annual observed temperature, salinity, comparison to annual seagrass frequency occurrence', 'Annual trend', 'Annual trend')
 )

### 2024-06-02 revision2/review_response.qmd
starting second revision

---
format: 
  docx:
    reference-doc: response_styles.docx
bibliography: ../refs.bib
csl: ../estuaries-and-coasts.csl

execute: 
  echo: false
  warning: false
  message: false
---

```{r}
#| eval: false
spelling::spell_check_files(here::here('revision/review_response.qmd'))
```

### AE comments:

Thank you for your revisions on this manuscript, I especially appreciate your well thought out responses to reviewer comments and also the changes made to the analyses (particularly the new GAM models and inclusion of the light attenuation data).  It will be a very nice contribution.  I have only a few minor comments that should be addressed:

*Response: Thank you once again for providing useful comments on our manuscript.  We have addressed the comments below and provide additional interpretation regarding the interaction terms in the GAMs.*

Use of p-values:  I do appreciate what ESCO is trying to address with the move away from our reliance on statistical significance and p-values.  However, I think in this case, it is useful to include both.  For your GAM models, you indicate some smoother relationships are weak or strong, and this is based on the p-value.  Here the p-values indicate the difference of the smoother from a constant straight line and it is an important way to evaluate their significance.  I suggest inserting the smoother p-values into the text in addition to the text that is already present (I found them in the tables but it would be easier in the text).  Further, I feel p-values are important for regression results, because you can have a significant regression (say p=0.01) but a low R2 value, indicating a relationship between the response and predictor is identified but there is lots of spread in the data.  So, I would also suggest inserting these.

*Response:*

For the GAM analysis, you may want to consider removing the interactions with year, to provide more power to define the GAM smoothers.  Most of these interactions are non-significant, and you’ve examined them individually, so you would be justified in removing these. You could mention you did these in preliminary models and no strong patterns were observed.  Or, if there is no difference you could just leave them.  Regardless, it is likely worthwhile to investigate.

*Response:*

The GAM models provided in the methods are really just the R syntax, and this is clear from the tables. I would suggest removing

*Response: Agreed, these equations were removed from the text.*

Line 491: Change “model for the EPC “ to “seagrass cover…”

*Response: Changed*

Line 493: seagrass cover was associated with decreased light attenuation for both HB and MTB segments.  In HB, it was a fairly random pattern with not much biological meaning.  For MT though seagrass cover increased with increased light attenuation, but as you mention, this is still below the threshold.  Some clarification here might be useful.

*Response:*

Line 494: “associations were weak” – add smoother p-values here

*Response:*

Line 498: both metric did not add additional value – however, it is significant for HB.  Not sure if this is relevant or not

*Response:*

Line 555: “For the EPC model none of the stressor metrics were associated with seagrass change”.  This isn’t quite correct: – salinity OTB was, and also the general trends in increasing sg with increasing temperature should be acknowledged explicitly. Although these were not significant they do suggest a relationships.  I appreciate the distinction between the temperature metrics used for the different datasets as you outline, but what would the mechanism be?  Could the increase with temperature in the EPC data indicate increased vegetative growth in anomalous temperature conditions?  Several studies have observed this (DuBois et al. 2020, Wong and Dowd 2023 etc)

*Response:*

Line 611: I think this explanation is really important, can it be highlighted above too.  What is the depth difference and does this ameliorate solar heating effects experienced by the seagrass.

*Response:*

### Co-Editor in Chief:

Now that we are moving towards production, Tables should be moved after the References, Figure Legends should be placed on one page and moved after the Tables, and the Figures should be moved after the Figure Legends.

*Response: Tables have been moved after the references, figure captions after the tables, and figures after the captions.*

### 2024-06-02 manu-draft.qmd
starting second revision

@@
 A single smooth term was used for each predictor using a thin plate regression spline, including a tensor product interaction term with a cubic regression spline that evaluated the potential interacting effects of each predictor with year.
-Two models were evaluated, one with the bottom temperature and salinity metrics together (@eq-epcmod1) and another with the both metric (@eq-epcmod2).
+Two models were evaluated, one with the bottom temperature and salinity metrics together and another with the both metric.
 All models excluded Lower Tampa Bay because of minimal seagrass change over time.
@@
 wrightii* in the upper bay segments (Figure [-@fig-seagrasschg]b).
-The GAM equations in *mgcv* R package notation [@wood17] are below, where $s$ is a smoother function for a single predictor and $ti$ is the tensor product interaction between predictors.
-All GAMs were fit using restricted maximum likelihood evaluation [@wood11].
-
-$$
-\begin{aligned}
-Seagrass \sim s(year, by = bay \ segment) + s(light \ att, by = bay \ segment) + 
-\\
-s(temperature, by = bay \ segment) + s(salinity, by = bay \ segment) + 
-\\
-ti(light \ att, \ year, by = bay \ segment) + ti(temperature, year, by = bay \ segment) +
-\\ 
-ti(salinity, year, by = bay \ segment)
-\end{aligned}
-$$ {#eq-epcmod1}
-
-$$
-\begin{aligned}
-Seagrass \sim s(year, by = bay \ segment) + s(light \ att, by = bay \ segment) + 
-\\
-s(both, by = bay \ segment) +
-\\
-ti(light \ att, \ year, by, bay \ segment) + ti(both, year, by = bay \ segment)
-\end{aligned}
-$$ {#eq-epcmod2}
+All GAMs were fit using restricted maximum likelihood evaluation [@wood11] with the *mgcv* R package \[@wood17.
 
@@
 For the PDEM models, presence/absence was converted to frequency occurrence as the number of sites with seagrass in a year divided by the total number of sites and all data were subset to less than 2 meters to better characterize locations where seagrass occurs.
-The GAM equations for the FIM (@eq-fimmod) and PDEM (@eq-pincomod) datasets, respectively, were as follows:
-
-$$
-\begin{aligned}
-Seagrass \sim s(year, by = bay \ segment) + s(temperature, by = bay \ segment) + 
-\\
-s(salinity, by = bay \ segment) + 
-\\
-ti(temperature, year, by = bay \ segment) + ti(salinity, year, by = bay \ segment)
-\end{aligned}
-$$ {#eq-fimmod}
-
-$$
-\begin{aligned}
-Seagrass \sim s(year) + s(temperature) + s(salinity) + 
-\\
-ti(temperature, year) + ti(salinity, year)
-\end{aligned}
-$$ {#eq-pincomod}
 
@@
 GAMs to assess the potential effects of year, light attenuation, temperature and salinity on seagrass change provided some evidence that seagrass change was influenced by climate stressors, particularly for the FIM and PDEM datasets.
-The model for the EPC data (`r sgmodtxt$epcmod1`) had notable associations with year for the OTB and MTB bay segments and for light attenuation for the HB bay segment (@fig-sgmod1, Table S7).
+The seagrass cover data (`r sgmodtxt$epcmod1`) had notable associations with year for the OTB and MTB bay segments and for light attenuation for the HB bay segment (@fig-sgmod1, Table S7).
 Associations of seagrass change with the temperature and salinity metrics were weak and somewhat contrary to expectation (i.e., an increasing partial effect on seagrass as the stressor metrics increased).

### 2024-06-03 revision2/review_response.qmd
revision2 done

@@
 #| eval: false
-spelling::spell_check_files(here::here('revision/review_response.qmd'))
+spelling::spell_check_files(here::here('revision2/review_response.qmd'))
 ```
@@
 
-*Response: Thank you once again for providing useful comments on our manuscript.
-We have addressed the comments below and provide additional interpretation regarding the interaction terms in the GAMs.*
+*Response: Thank you once again for providing useful comments on our manuscript.  We have addressed them below and provide additional interpretation regarding the interaction terms in the GAMs.*
 
@@
 So, I would also suggest inserting these.
-
-*Response:*
+*Response: We have added p-values to the text when describing the relationship of predictors to the seagrass response variables.  Please see section 3.3 for updates.*
 
@@
 
-*Response:*
+*Response: We have removed all interaction terms from the GAMs and updated the figures, tables, and text.
+The results were very similar and we agree it was not worth including these in the models.
+We have made a note in the text that initial models evaluated interaction terms, but the results were similar.*
+
+*Lines 370-373: Initial models also included a tensor product interaction term with a cubic regression spline that evaluated the potential interacting effects of each predictor with year.
+However, most of the interaction terms were not significant and were excluded from the final models.*
 
@@
 
-*Response:*
+*Response: The following was added on lines 489-493: "Associations with light attenuation were also observed for the HB (p < 0.001) and MTB (p = 0.004) bay segments, although the relationship in HB was flat beyond 0.75 m$^{-1}$ and opposite as expected for MTB (Figure 7, Table S7).
+As noted above, light attenuation in HB and MTB in recent years is well withing the range supportive of seagrass growth (Figure 2c), such that these associations may not be describing meaningful biological relationships."*
 
@@
 
-*Response:*
+*Response: Added.*
 
@@
 
-*Response:*
+*Response: Removing the interaction terms also showed a significant association in OTB, although again this is contrary to expectation as an increase in the metric was associated with an increase in seagrass.
+The response to the following comment provides some explanation for this anomaly.*
 
@@
 
-*Response:*
+*Response: Yes, this was an explanation that was initially considered for the temperature metric, but not stated in original the text.
+It is also difficult to describe the positive association of seagrass change with the salinity metric in OTB.
+Our description later in the paragraph regarding the spatial mismatch between the water quality and seagrass data is a possible explanation.*
+
+*The following text was added to the discussion (lines 557-561): "The positive association of seagrass change with the temperature (or 'both') metric could be a signal of increased vegetative growth with temperature anomalies (e.g., Wong and Dowd 2023), whereas the positive association with the salinity metric in OTB is not easily explained since extreme precipitation events have been linked to seagrass loss in Tampa Bay (Greening and Janicki 2006)."*
 
@@
 
-*Response:*
+*Response: The mean depth of all EPC sampling stations (excluding LTB) is 3.88 meters, whereas the mean depth of points along each transect and sample date is 0.76 meters.
+The following was added to lines 363-365 in section 2.7 of the methods: "...; note that the seagrass transect data were collected at different locations and depths ($\bar{Z}$ = 0.76 m) than the EPC water quality data ($\bar{Z}$ = 3.88 m, excludes LTB not evaluated)."*
 

### 2024-06-03 cover-letter.qmd
revision2 done

@@
 
-Enclosed please find our revised manuscript, titled \`Hot and fresh: evidence of climate-related suboptimal water conditions for seagrass in a large Gulf coast estuary', to be considered as an original article in Estuaries and Coasts.
+Enclosed please find the second revision of our manuscript, titled \`Hot and fresh: evidence of climate-related suboptimal water conditions for seagrass in a large Gulf coast estuary', to be considered as an original article in Estuaries and Coasts.
 
-We sincerely appreciate the constructive comments provided by the editors and reviewers and have made every attempt to address these comments in our revision.
-The largest change to the manuscript was a reanalysis of the potential seagrass response to climate stressors, which included more flexible Generalized Additive Models and inclusion of additional predictors (i.e., time as a continuous variable, light attenuation).
-Overall, our conclusions remain the same that climate stressors are potentially driving recent seagrass changes, although our more robust modeling approach provides greater weight of evidence for this conclusion.
-Note that we have also made substantial revisions in response to comments from the co-editor in chief regarding statistical descriptions of results to focus more on effect sizes and magnitudes of change.
+We have addressed the remaining comments from Dr.
+Wong, as noted in our attached responses.
+In particular, we have removed the interaction terms from the GAMs and included appropriate statistics when describing the resulting relationships.
+Our main conclusions remain the same.
+Additional minor revisions are noted in our responses.
 
@@
 
-Marcus W. Beck, Tampa Bay Estuary Program, St. Petersburg, Florida, USA, [mbeck@tbep.org](mailto:mbeck@tbep.org)
+Marcus W. Beck, Tampa Bay Estuary Program, St. Petersburg, Florida, USA, [mbeck\@tbep.org](mailto:mbeck@tbep.org)
 

### 2024-06-03 manu-draft.qmd
revision2 done

@@
 library(tidyr)
+library(broom)
+library(mgcv)
 
@@
 
+```{r}
+data('transect', package = 'tbeptools')
+data('epcdata', package = 'tbeptools')
+sgdep <- round(-1 * mean(transect$Depth, na.rm = T) / 100, 2)
+epcdep <- epcdata %>% filter(bay_segment != 'LTB') %>% summarise(dep = mean(Total_Depth_m, na.rm = T)) %>% .[[1]] %>% round(2)
+```
+
 For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends above.
@@
 Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the transect year summaries.
-As such, the stressor metrics were compared to frequency occurrence (all species) each year by bay segment.
+As such, the stressor metrics were compared to frequency occurrence (all species) each year by bay segment; note that the seagrass transect data were collected at different locations and depths ($\bar{Z}$ = `r sgdep` m) than the EPC water quality data ($\bar{Z}$ = `r epcdep` m, excludes LTB not evaluated).
 GAMs were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, or both.
@@
 For 1998 to 2022, `r onb$epc` of the Secchi observations were recorded on the bottom.
-A single smooth term was used for each predictor using a thin plate regression spline, including a tensor product interaction term with a cubic regression spline that evaluated the potential interacting effects of each predictor with year.
+A single smooth term was used for each predictor using a thin plate regression spline.
+Initial models also included a tensor product interaction term with a cubic regression spline that evaluated the potential interacting effects of each predictor with year.
+However, most of the interaction terms were not significant and were excluded from the final models.
 Two models were evaluated, one with the bottom temperature and salinity metrics together and another with the both metric.
@@
 GAMs were used for the FIM data to evaluate annual average seagrass percent cover using year, temperature, and salinity as independent variables.
-Separate smoothed terms for each predictor and interactions with year for each predictor for each bay segment were included as above.
+Separate smoothed terms for each predictor and each bay segment were included as above.
 GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data.
@@
 sgmodtxt <- modtxt_fun(sgmodsum)
+epc1pvl <- gam_pvals(sgmods$epcmod1)
+epc2pvl <- gam_pvals(sgmods$epcmod2)
+fimpvl <- gam_pvals(sgmods$fimmod)
+pincopvl <- gam_pvals(sgmods$pincomod)
 ```
 
-GAMs to assess the potential effects of year, light attenuation, temperature and salinity on seagrass change provided some evidence that seagrass change was influenced by climate stressors, particularly for the FIM and PDEM datasets.
-The seagrass cover data (`r sgmodtxt$epcmod1`) had notable associations with year for the OTB and MTB bay segments and for light attenuation for the HB bay segment (@fig-sgmod1, Table S7).
-Associations of seagrass change with the temperature and salinity metrics were weak and somewhat contrary to expectation (i.e., an increasing partial effect on seagrass as the stressor metrics increased).
-The EPC model using the "both" stressor metric was similar (`r sgmodtxt$epcmod2`, Figure S7, Table S8), suggesting no added benefit of evaluating times when both temperature was above and salinity was below the threshold.
+GAMs to assess the potential effects of year, light attenuation, temperature, and salinity on seagrass change provided some evidence that seagrass change was influenced by climate stressors, particularly for the FIM and PDEM datasets.
+The seagrass cover data (`r sgmodtxt$epcmod1`, EPC model) had notable associations with year for all bay segments (OTB: `r epc1pvl[['yr:bay_segmentOTB']]`, HB: `r epc1pvl[['yr:bay_segmentHB']]`, MTB: `r epc1pvl[['yr:bay_segmentMTB']]`) that followed long-term changes in @fig-seagrasschg.
+Associations with light attenuation were also observed for the HB (`r epc1pvl[['la:bay_segmentHB']]`) and MTB (`r epc1pvl[['la:bay_segmentMTB']]`) bay segments, although the relationship in HB was flat beyond 0.75 m$^{-1}$ and opposite as expected for MTB (@fig-sgmod1, Table S7).
+As noted above, light attenuation in HB and MTB in recent years is well within the range supportive of seagrass growth (Figure [-@fig-seagrasschg]c), such that these associations may not be describing meaningful biological relationships.
+Associations of seagrass change with the temperature (OTB: `r epc1pvl[['temp:bay_segmentOTB']]`, HB `r epc1pvl[['temp:bay_segmentHB']]`, MTB `r epc1pvl[['temp:bay_segmentMTB']]`) and salinity (OTB: `r epc1pvl[['sal:bay_segmentOTB']]`, HB `r epc1pvl[['sal:bay_segmentHB']]`, MTB `r epc1pvl[['sal:bay_segmentMTB']]`) metrics were not present or contrary to expectation (e.g., increase in seagrass with an increasing salinity metric in OTB, slight increase in seagrass cover with an increasing temperature metric in all bay segments).
+Similarly, the EPC model using the "both" stressor metric (`r sgmodtxt$epcmod2`, Figure S7, Table S8) had positive associations of the "both" metric with seagrass change in OTB (`r epc2pvl[['both:bay_segmentOTB']]`) and HB (`r epc2pvl[['both:bay_segmentHB']]`).
 
 The GAMs for the FIM and PDEM datasets had notable associations with temperature and salinity stress, where an increase in temperature and a decrease in salinity was often associated with a decline in seagrass depending on bay segment.
-The FIM model (`r sgmodtxt$fimmod`, Figure [-@fig-sgmod2]a, Table S9) showed a slight decrease in seagrass with increasing temperature for HB and MTB, with a similar change as salinity decreased.
-The PDEM model (`r sgmodtxt$pincomod`, Figure [-@fig-sgmod2]b, Table S10) that evaluated only OTB showed a distinct decrease with seagrass as temperature increased, but no association with salinity.
-Some of the interaction terms of the predictors with year suggested the relationships with seagrass may have changed over time, but the results were not consistent between bay segments or models (Tables S7 to S10).
+The FIM model (`r sgmodtxt$fimmod`, Figure [-@fig-sgmod2]a, Table S9) showed a slight decrease in seagrass with increasing temperature in MTB (`r fimpvl[['temp:bay_segmentMTB']]`) and a similar change as salinity decreased in MTB (`r fimpvl[['sal:bay_segmentMTB']]`) and OTB (`r fimpvl[['sal:bay_segmentOTB']]`).
+The PDEM model (`r sgmodtxt$pincomod`, Figure [-@fig-sgmod2]b, Table S10) that evaluated only OTB showed a slight decrease with seagrass as temperature increased (`r pincopvl[['temp']]`), but no association with salinity.
 
@@
 Model results for the FIM and PDEM datasets both suggested that increasing temperature and decreasing salinity were associated with potential seagrass loss in recent years (@fig-sgmod2).
-For the EPC model, none of the stressor metrics were associated with seagrass change.
+For the EPC model, the salinity metric was positively associated with seagrass in OTB and all segments showed a slight increase in seagrass with increases in the temperature metric.
+The positive association of seagrass change with the temperature (or "both") metric could be a signal of increased vegetative growth with temperature anomalies [e.g., @wong2023], whereas the positive association with the salinity metric in OTB is not easily explained since extreme precipitation events have been linked to seagrass loss in Tampa Bay [@greening2006].
 An important distinction between the EPC and other models is that the former evaluated the number of days above/below thresholds each year to quantify annual temperature or salinity stress, whereas the latter evaluated observed temperature and salinity values at the time of seagrass sampling.
-Additionally, the EPC models evaluated water quality changes from data collected at relatively deeper locations than where seagrass typically grows (Figure [-@fig-map]b), suggesting a potential disconnect in relating the two.
+Additionally, the EPC models evaluated water quality changes from data collected at relatively deeper locations than where seagrass typically grows (Figure \[-@fig-map\]b), suggesting a potential disconnect in relating the two.
 As such, both models attempted to describe the role of these stressors on potential seagrass change, but use different independent variables given the different sampling designs of each monitoring program.
@@
 Partial effects describe the modeled association between each predictor and seagrass frequency occurrence after accounting for the effects of the other predictors.
-The model also included smoothed interaction terms with year (not shown).
 See Table S7 for additional model fit statistics.
@@
 Partial effects describe the modeled association between each predictor and seagrass change after accounting for the effects of the other predictors.
-The models also included smoothed interaction terms with year (not shown).
 See Tables S9 and S10 for additional model fit statistics.

### 2024-06-03 R/dat_proc.R

@@ -1500,9 +1500,9 @@ load(file = here('data/pincotemp.RData'))
 tomod <- epccmbdat %>% 
   filter(!bay_segment %in% c('LTB'))
 
-epcmod1 <- gam(total ~ s(yr, by = bay_segment) + s(la, by = bay_segment) + s(temp, by = bay_segment) + s(sal, by = bay_segment) + ti(la, yr, by = bay_segment) + ti(temp, yr, by = bay_segment) + ti(sal, yr, by = bay_segment), data = tomod, method = 'REML')
+epcmod1 <- gam(total ~ s(yr, by = bay_segment) + s(la, by = bay_segment) + s(temp, by = bay_segment) + s(sal, by = bay_segment), data = tomod, method = 'REML')
 
-epcmod2 <- gam(total ~ s(yr, by = bay_segment) + s(la, by = bay_segment) + s(both, by = bay_segment) + ti(la, yr, by = bay_segment) + ti(both, yr, by = bay_segment), data = tomod, method = 'REML')
+epcmod2 <- gam(total ~ s(yr, by = bay_segment) + s(la, by = bay_segment) + s(both, by = bay_segment), data = tomod, method = 'REML')
 
 tomod <- fimsgtempdat %>% 
   filter(!bay_segment %in% c('LTB')) %>%
@@ -1527,8 +1527,7 @@ tomod <- fimsgtempdat %>%
     )
   )
 
-fimmod <- gam(sgcov ~ s(yr, by = bay_segment) + s(temp, by = bay_segment) + s(sal, by = bay_segment) + 
-             ti(sal, yr, by = bay_segment) + ti(temp, yr, by = bay_segment), data = tomod, method = 'REML')
+fimmod <- gam(sgcov ~ s(yr, by = bay_segment) + s(temp, by = bay_segment) + s(sal, by = bay_segment), data = tomod, method = 'REML')
 
 tomod <- pincotemp %>% 
   filter(depth_m < 2) %>% 
@@ -1550,13 +1549,12 @@ tomod <- pincotemp %>%
     la = 1.49 / secchi_m
   )
 
-pincomod <- gam(allsg ~ s(yr) + s(temp) + s(sal) + ti(yr, temp) + ti(yr, sal), 
-           data = tomod, method = 'REML')
+pincomod <- gam(allsg ~ s(yr) + s(temp) + s(sal), 
+                data = tomod, method = 'REML')
 
[…]

### 2024-06-03 R/funcs.R

@@ -47,7 +47,8 @@ gam_table <- function(mod, cap = NULL){
     tidy() %>% 
     rowwise() %>% 
     mutate(
-      p.value = p_txt(p.value, addp = F)
+      p.value = p_txt(p.value, addp = F) ,
+      term = gsub('^s\\(|\\)', '', term)
     ) %>% 
     ungroup() %>% 
     mutate_if(is.numeric, round, 3) %>%
@@ -58,6 +59,24 @@ gam_table <- function(mod, cap = NULL){
       'p' = p.value
     ) %>%
     kable(digits = 3, align = 'lrrrr', caption = cap)
+  
+  return(out)
+  
+}
+
+gam_pvals <- function(mod){
+  
+  out <- mod %>% 
+    broom::tidy() %>% 
+    rowwise() %>% 
+    mutate(
+      p.value = p_txt(p.value, addp = T) ,
+      term = gsub('^s\\(|\\)', '', term)
+    ) %>% 
+    ungroup() %>%
+    select(term, p.value) %>% 
+    split(.$term) %>% 
+    purrr::map(~ .x$p.value)
   
   return(out)
   
@@ -131,8 +150,8 @@ gamplo_fun <- function(sgmods){
         
         yrng <- 0.9 * max(abs(range(data$.estimate, data$.lower_ci, data$.upper_ci)))
         
-        if(bay_segment == 'HB' & xvar == 'sal')
-          yrng <-100
+        # if(bay_segment == 'HB' & xvar == 'sal')
+        #   yrng <-100
         
         ttl <- ifelse(bay_segment == 'OTB' & xvar == 'yr', 
                       '(a) FIM', '')

### 2024-06-03 R/tabs.R

@@ -324,18 +324,18 @@ save(supp2daytab, file = here('tabs/supp2daytab.RData'))
 load(file = here::here('data/sgmods.RData'))
 load(file = here::here('data/sgmodsum.RData'))
 
-suppepccap1 <- paste0('Summary of smoother terms in the Generalized Additive Model used to evaluate seagrass response for the EPC data in relation to bottom temperature (temp) and salinity (sal) stress metrics, with additional smoothers for year (yr) and light attenuation (la).  Separate smoothers were fit for each bay segment.  $s$ = invidual smoother, $ti$ = interaction term. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay. ', modtxt_fun(sgmodsum)$epcmod1, '.')  
+suppepccap1 <- paste0('Summary of smoother terms in the Generalized Additive Model used to evaluate seagrass response for the EPC data in relation to bottom temperature (temp) and salinity (sal) stress metrics, with additional smoothers for year (yr) and light attenuation (la). Separate smoothers were fit for each bay segment. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay. ', modtxt_fun(sgmodsum)$epcmod1, '.')  
 suppepcmod1tab <- gam_table(sgmods$epcmod1, cap = suppepccap1)
 save(suppepcmod1tab, file = here('tabs/suppepcmod1tab.RData'))
 
-suppepccap2 <- paste0('Summary of smoother terms in the Generalized Additive Model used to evaluate seagrass response for the EPC data in relation to "both" stressors (both bottom temperature above and salinity below thresholds), with additional smoothers for year (yr) and light attenuation (la).  Separate smoothers were fit for each bay segment.  $s$ = invidual smoother, $ti$ = interaction term. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay. ', modtxt_fun(sgmodsum)$epcmod2, '.')    
[…]

### 2024-06-03 revision2/review_response.qmd
second revision submitted

@@
 So, I would also suggest inserting these.
+
 *Response: We have added p-values to the text when describing the relationship of predictors to the seagrass response variables.  Please see section 3.3 for updates.*
@@
 
-*Lines 370-373: Initial models also included a tensor product interaction term with a cubic regression spline that evaluated the potential interacting effects of each predictor with year.
-However, most of the interaction terms were not significant and were excluded from the final models.*
+*Lines 370-373: "Initial models also included a tensor product interaction term with a cubic regression spline that evaluated the potential interacting effects of each predictor with year.
+However, most of the interaction terms were not significant and were excluded from the final models."*
 
@@
 
-*Response: Changed*
+*Response: Changed.*
 
@@
 *Response: The following was added on lines 489-493: "Associations with light attenuation were also observed for the HB (p < 0.001) and MTB (p = 0.004) bay segments, although the relationship in HB was flat beyond 0.75 m$^{-1}$ and opposite as expected for MTB (Figure 7, Table S7).
-As noted above, light attenuation in HB and MTB in recent years is well withing the range supportive of seagrass growth (Figure 2c), such that these associations may not be describing meaningful biological relationships."*
+As noted above, light attenuation in HB and MTB in recent years is well within the range supportive of seagrass growth (Figure 2c), such that these associations may not be describing meaningful biological relationships."*
 
@@
 
-*Response: Yes, this was an explanation that was initially considered for the temperature metric, but not stated in original the text.
+*Response: Yes, this was an explanation that was initially considered for the temperature metric, but not stated in the original text.
 It is also difficult to describe the positive association of seagrass change with the salinity metric in OTB.
@@
 
-*Response: The mean depth of all EPC sampling stations (excluding LTB) is 3.88 meters, whereas the mean depth of points along each transect and sample date is 0.76 meters.
+*Response: The mean depth of all EPC sampling stations (excluding LTB) is 3.88 meters, whereas the mean depth of points along each transect is 0.76 meters.
 The following was added to lines 363-365 in section 2.7 of the methods: "...; note that the seagrass transect data were collected at different locations and depths ($\bar{Z}$ = 0.76 m) than the EPC water quality data ($\bar{Z}$ = 3.88 m, excludes LTB not evaluated)."*

### 2024-06-03 manu-draft.qmd
second revision submitted

@@
 wrightii* in the upper bay segments (Figure [-@fig-seagrasschg]b).
-All GAMs were fit using restricted maximum likelihood evaluation [@wood11] with the *mgcv* R package \[@wood17.
+All GAMs were fit using restricted maximum likelihood evaluation [@wood11] with the *mgcv* R package \[@wood17\].
 
@@
 As noted above, light attenuation in HB and MTB in recent years is well within the range supportive of seagrass growth (Figure [-@fig-seagrasschg]c), such that these associations may not be describing meaningful biological relationships.
-Associations of seagrass change with the temperature (OTB: `r epc1pvl[['temp:bay_segmentOTB']]`, HB `r epc1pvl[['temp:bay_segmentHB']]`, MTB `r epc1pvl[['temp:bay_segmentMTB']]`) and salinity (OTB: `r epc1pvl[['sal:bay_segmentOTB']]`, HB `r epc1pvl[['sal:bay_segmentHB']]`, MTB `r epc1pvl[['sal:bay_segmentMTB']]`) metrics were not present or contrary to expectation (e.g., increase in seagrass with an increasing salinity metric in OTB, slight increase in seagrass cover with an increasing temperature metric in all bay segments).
+Associations of seagrass change with the temperature (OTB: `r epc1pvl[['temp:bay_segmentOTB']]`, HB: `r epc1pvl[['temp:bay_segmentHB']]`, MTB: `r epc1pvl[['temp:bay_segmentMTB']]`) and salinity (OTB: `r epc1pvl[['sal:bay_segmentOTB']]`, HB: `r epc1pvl[['sal:bay_segmentHB']]`, MTB: `r epc1pvl[['sal:bay_segmentMTB']]`) metrics were not present or contrary to expectation (e.g., increase in seagrass with an increasing salinity metric in OTB, slight increase in seagrass cover with an increasing temperature metric in all bay segments).
 Similarly, the EPC model using the "both" stressor metric (`r sgmodtxt$epcmod2`, Figure S7, Table S8) had positive associations of the "both" metric with seagrass change in OTB (`r epc2pvl[['both:bay_segmentOTB']]`) and HB (`r epc2pvl[['both:bay_segmentHB']]`).
